### 2024-04-08 README.Rmd
1. Added functions to start simple simulations more easily 2. Added vignette 3. Sara is now package maintainer

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MethEvolSIM

<!-- badges: start -->
<!-- badges: end -->

Software for simulation of DNA methylation dynamics on different genomic structures under the basic model for the evolution of DNA methylation along genealogies. The methylation state of each CpG dinucleotide is represented as 1 for unmethylated, 2 for partially methylated and 3 for methylated.

# Installation

You can use the development version of MethEvolSIM from git. The development is being implemented with the `devtools` package. It can be loaded and checked using the following:

```{r}
library(devtools)
load_all()
```

`check()` runs the unit tests. The unit tests that test the functions used for writing output files:
- *validationResults.txt*: Is generated in case an error relative to the characteristics of the objects happens and contains both the error and the object.

# Manual of Use

## Structural distribution of CpG sites

The initial spatial distribution of CpG sites is given by the user. A data frame with 3 columns named "start", "end" and "globalState" encodes the structural information. Each row provides the information of a different structure. The number of CpG positions is given by $end - start +1$, and each row start value needs to be as the previous row end value + 1. The globalState encodes island and non-island structures with the character value "U" for islands and "M" for non-islands.

```{r}
# Example 3 structures of length 100: non-island, island, non-island
infoStr <- data.frame(start = c(1, 101, 201),
                        end = c(100, 200, 300),
                        globalState = c("M", "U", "M"))
infoStr
```

```{r}
# Example 2 islands: first of length 1, second of lenght 15
infoStr <- data.frame(start = c(1, 2),
                        end = c(1, 16),
                        globalState = c("U", "U"))
infoStr
```

## Initial methylation state distribution

Can be sampled as specified in the section "Model parameters" when infoStr has only start, end and globalState columns,
or given by the user as additional columns of the data frame. Note that when given, each of the structures frequencies needs to sum up to 1.

```{r}
# Example 3 structures of length 100 with customized initial methylation state equilibrium frequencies
infoStr <- data.frame(start = c(1, 101, 201),
                        end = c(100, 200, 300),
                        globalState = c("M", "U", "M"),
                        u_eqFreq = c(0.1, 0.8, 0.1),
                        p_eqFreq = c(0.1, 0.1, 0.1),
                        m_eqFreq = c(0.8, 0.1, 0.8))
infoStr
```

## Parameter values

The default parameter values used by MethEvolSIM can be obtained using the function `get_model_parameters()`, that returns a data frame with the columns named as specified in the section "Model parameters":

```{r}
default_paramValues <- get_parameterValues()
default_paramValues
```

The section "Model parameters" provides further information about each of the parameters, including each parameter range of possible values. To simulate data with a different parameterization, the new value of a chosen parameter can be modified in the data frame.

```{r}
# Example: modification of parameter iota to value 0.2
default_paramValues$iota <- 0.2
default_paramValues
```

## Initial DNA Methylation Data

### Simulation

The initial data is generated in an object of class combiStructureGenerator.

```{r}
# Example with customized initial methylation frequencies and customized parameter values
custom_infoStr <- data.frame(start = c(1, 101, 201),
                        end = c(100, 200, 300),
                        globalState = c("M", "U", "M"),
                        u_eqFreq = c(0.1, 0.8, 0.1),
                        p_eqFreq = c(0.1, 0.1, 0.1),
                        m_eqFreq = c(0.8, 0.1, 0.8))
custom_params <- get_parameterValues()
custom_params$mu <- 0.005
initialD <- simulate_initialData(infoStr = custom_infoStr, params = custom_params)
# Returns customized parameters and simulated data
initialD$params
initialD$data
```

```{r}
# Example with simulated initial methylation frequencies and default parameter values
custom_infoStr <- data.frame(start = c(1, 101, 201),
                        end = c(100, 200, 300),
                        globalState = c("M", "U", "M"))
initialD <- simulate_initialData(infoStr = custom_infoStr)
# Returns default parameters
initialD$params
initialD$data
```

To get the parameter values of a combiStructureGenerator object, it can be given as argument to the function `get_parameterValues()`

```{r}
combiStr_object <- initialD$data
get_parameterValues(combiStr_object)
```

### Exploration of initial data

The methods of the class combiStructureGenerator can be used to access each of the contained structures:

```{r}
# E.g. access fist contained structure
combiStr_object$get_singleStr(1)
```

Each contained structure is an object of singleStructureGenerator and its methods can be used to get the contained information.

```{r}
# E.g. get the equilibrium frequencies of the first singleStructureGenerator
combiStr_object$get_singleStr(1)$get_eqFreqs()
```

```{r}
# E.g. get the sequence of methylation states of the second singleStructureGenerator
combiStr_object$get_singleStr(1)$get_seq()
```

## Simulate DNA Methylation Data Evolution

The evolution data is generated in an object of class treeMultiRegionSimulator.

```{r}
# Example with customized initial methylation frequencies, customized parameter values and default dt (0.01)
tree <- "((a:1, b:1):2, c:2, (d:3.7, (e:4, f:1):3):5);"
custom_infoStr <- data.frame(start = c(1, 101, 201),
                        end = c(100, 200, 300),
                        globalState = c("M", "U", "M"),
                        u_eqFreq = c(0.1, 0.8, 0.1),
                        p_eqFreq = c(0.1, 0.1, 0.1),
                        m_eqFreq = c(0.8, 0.1, 0.8))
custom_params <- get_parameterValues()
custom_params$mu <- 0.005
evolD <- simulate_evolData(infoStr = custom_infoStr, tree = tree, params = custom_params)
# Returns customized parameters, tree used, time step length for SSE process used (dt) and simulated data
evolD$data
evolD$dt
evolD$tree
evolD$params
```

If the evolution of given data at the root is simulated, and it wants to be done with customized parameter values, then the customized parameters are given to `simulate_initialData()` instead to `simulate_evolData()`.

```{r}
# Example with customized initial methylation frequencies, customized parameter values and default dt (0.01)
tree <- "((a:1, b:1):2, c:2, (d:3.7, (e:4, f:1):3):5);"
custom_infoStr <- data.frame(start = c(1, 101, 201),
                        end = c(100, 200, 300),
                        globalState = c("M", "U", "M"),
                        u_eqFreq = c(0.1, 0.8, 0.1),
                        p_eqFreq = c(0.1, 0.1, 0.1),
                        m_eqFreq = c(0.8, 0.1, 0.8))
custom_params <- get_parameterValues()
custom_params$mu <- 0.005
initialD <- simulate_initialData(infoStr = custom_infoStr, params = custom_params)$data
evolD <- simulate_evolData(rootData =initialD, tree = tree)
```
### Exploration of evolved data

The treeMultiStructureGenerator object contains a list `$Branch` with the combiStructureGenerator representing the data at the each node after a tree branch (but for the data at the root).

To know which offspring branches a node has, the combiStructureGenerator object method `$get_offspring_index()` can be used.

```{r}
treeData <- evolD$data
# E.g. offspring of data at root 
treeData$Branch[[1]]$get_offspring_index()
```
If the node is an inner node, its name is set as NULL. It can be accessed with the combiStructureGenerator method `$get_name()`

```{r}
treeData$Branch[[1]]$get_name()
```
But the leafs are named as in the given tree.

```{r}
treeData$Branch[[4]]$get_name()
```
If there were IWE events sampled in a branch, it can be accessed with the combiStructureGenerator method `$get_IWE_events()`

```{r}
treeData$Branch[[4]]$get_IWE_events()
```
In the case we want to know the equilibrium frequencies of a singleStructureGenerator object contained in the combiStructureGenerator we can use the combiStructureGenerator method to access the singleStructureGenerator object of interest `get_singleStr(i)` followed by the singleStructureGenerator method `get_eqFreqs()`

```{r}
# E.g. get the ewuilibrium frequencies of the second structure at the leaf "b"
treeData$Branch[[4]]$get_singleStr(2)$get_eqFreqs()
```

# Model parameters

## Initial methylation state distribution

Structures are categorized as CpG islands or non-islands. In the root of the genealogy we assume that each structure samples an equilibrium probability triple $(\pi_u,\pi_p, \pi_m)$. Current parameter values favor prevalent homogeneity in methylation levels within structures, most CpG islands with a majority predominantly unmethylated and a small proportion predominantly methylated, and most non-islands with a majority predominantly methylated and a small proportion predominantly unmethylated. The sampling of the equilibrium probability tripple is done using two distinct Beta distributions to sample the corresponding methylation frequencies.

**Island parameterization: alpha_pI, beta_pI, alpha_mI, beta_mI**

MethEvolSIM samples $\pi_p$ from the first Beta distribution. Current parameterization uses as initial parameter values a relatively small $\alpha_{I_p}$ and relatively large $\beta_{I_p}$ to encourage minor assignment of partially methylated states.

The second Beta distribution, scaled by $1 - \pi_p$,  corresponds to the proportion of methylated and unmethylated states. Currently both parameter values ($\alpha_{I_m}$ and $\beta_{I_m}$) set to small values while ${\alpha_{I_m} < \beta_{I_m}}$ to favour shifting the stochastic choice toward homogeneous states with a higher proportion of predominantly unmethylated islands, and assigned to $\pi_m$. Finally, we set $\pi_u$ as $1 - \pi_p - \pi_m$.

**Non-island parameterization: alpha_pNI, beta_pNI, alpha_mNI, beta_mNI**

Following the same rationale as for island parameterization, all the regions not belonging to an island sample a common equilibrium probability triple from two Beta distributions with parameters $\alpha_{NI_p}$, $\beta_{NI_p}$, $\alpha_{NI_m}$ and $\beta_{NI_m}$. The parameter values are currently set to favor the characteristic high overall methylation level.

You can simulate DNA methylation along a tree given some data at root, or by simulating data at root given a structural distribution and letting it evolve.

**Parameter ranges** 
The alpha and beta parameters of a Beta distribution must be greater than 0, that is (0, $\infty$).

## DNA methylation evolution processes

DNA methylation state evolution happens by two structure-specific events: CpG single site events (SSEs) and CpG island wide events (IWEs).

### Island-wide events (IWEs)

IWEs change the island methylation probabilities and some of the CpG sites in the CpG island can simultaneously change their state at the same time. Different CpG islands, however, are assumed to evolve independently of each other. IWEs occur per CpG island at a rate denoted by $\mu$. Consequently, over a branch of length $l$, the expected number of IWEs per island is given by $\mu \cdot l$.

**IWE parameterization: mu. Parameter range:**
Should be a small number, but it can take any value from 0, that is [0, $\infty$)

### Single-site events (SSEs)

In addition to IWEs we allow for single site events (SSEs), which change the methylation states of single CpG sites within and between CpG islands. In our model SSEs account for two dynamics: independent of each other (SSEi), and  assuming correlations between adjacent sites (SSEc).

As SSEc transition probabilities are dependent on adjacent sites we need to approximate SSE based evolution by short time steps, allowing for CpG sites to actualize their state. Therefore, we first discretize the branch intervals between IWEs in time intervals of length 0.01. For each CpG position and each time interval the rate matrix for the transitions between the states unmethylated, partially methylated and methylated results from the addition of the rate matrices of the two types of SSE events, so that $\mathbb{E}(R_i + R_c)=1$, $R_i$ and $R_c$ representing the rate factors of independent and correlated processes. The proportion of each of the two SSE types is determined by the parameter $\iota$.

Different positions $j$ have different SSEi rates $R_{i,j}$ coming from a discretized gamma distribution with 3 categories, expected value $\iota$ and shape parameters $\alpha_{R_{i}}$ and ${\beta_{R_{i}}}=\frac{\alpha_{R_{i}}}{\iota}$. The probability to be in each respective rate category is 1/3. For the rate of collaborative SSEs $R_c = 1-\iota$, the probability of considering left of right neighbour is equal.

**SSE process parameterization: iota, alpha_Ri**
**Parameter ranges** 
The alpha parameter of a Gamma distribution must be greater than 0, that is (0, $\infty$). Iota must be between 0 (non-included) and 1, that is (0,1].

### 2024-04-08 R/multiRegion_SIM.R

@@ -3,48 +3,47 @@
 #'
 #' @description
 #' an R6 class representing a single genomic structure
-#' @export
 singleStructureGenerator <-
   R6::R6Class("singleStructureGenerator",
               portable = FALSE,
               private = list(
-                #' @field my_combiStructure Private attribute: Object of class combiStructureGenerator containing it
+                ## @field my_combiStructure Private attribute: Object of class combiStructureGenerator containing it
                 my_combiStructure = NULL,
-                #' @field combiStructure_index Private attribute: Index in Object of class combiStructureGenerator
+                ## @field combiStructure_index Private attribute: Index in Object of class combiStructureGenerator
                 combiStructure_index = NULL,
-                #' @field alpha_pI Private attribute: Model parameter for sampling island equilibrium frequencies
+                ## @field alpha_pI Private attribute: Model parameter for sampling island equilibrium frequencies
                 alpha_pI = 0.1,
-                #' @field beta_pI Private attribute: Model parameter for sampling island equilibrium frequencies
+                ## @field beta_pI Private attribute: Model parameter for sampling island equilibrium frequencies
                 beta_pI = 1,
-                #' @field alpha_mI Private attribute: Model parameter for sampling island equilibrium frequencies
+                ## @field alpha_mI Private attribute: Model parameter for sampling island equilibrium frequencies
                 alpha_mI = 0.1,
-                #' @field beta_mI Private attribute: Model parameter for sampling island equilibrium frequencies
+                ## @field beta_mI Private attribute: Model parameter for sampling island equilibrium frequencies
                 beta_mI = 0.5,
-                #' @field alpha_pNI Private attribute: Model parameter for sampling non-island equilibrium frequencies
+                ## @field alpha_pNI Private attribute: Model parameter for sampling non-island equilibrium frequencies
                 alpha_pNI = 0.1,
-                #' @field beta_pNI Private attribute: Model parameter for sampling non-island equilibrium frequencies
+                ## @field beta_pNI Private attribute: Model parameter for sampling non-island equilibrium frequencies
                 beta_pNI = 1,
-                #' @field alpha_mNI Private attribute: Model parameter for sampling non-island equilibrium frequencies
+                ## @field alpha_mNI Private attribute: Model parameter for sampling non-island equilibrium frequencies
                 alpha_mNI = 0.5,
-                #' @field beta_mNI Private attribute: Model parameter for sampling non-island equilibrium frequencies
+                ## @field beta_mNI Private attribute: Model parameter for sampling non-island equilibrium frequencies
[…]

### 2024-04-08 R/user_functions.R

@@ -1,43 +1,34 @@
-# Island equilibrium frequencies sampling strategia
-singleStructureGenerator$set("public", "get_alpha_pI", function() private$alpha_pI)
-singleStructureGenerator$set("public", "get_beta_pI", function() private$beta_pI)
-singleStructureGenerator$set("public", "get_alpha_mI", function() private$alpha_mI)
-singleStructureGenerator$set("public", "get_beta_mI", function() private$beta_mI)
-# Non-island equilibrium frequencies sampling strategia
-singleStructureGenerator$set("public", "get_alpha_pNI", function() private$alpha_pNI)
-singleStructureGenerator$set("public", "get_beta_pNI", function() private$beta_pNI)
-singleStructureGenerator$set("public", "get_alpha_mNI", function() private$alpha_mNI)
-singleStructureGenerator$set("public", "get_beta_mNI", function() private$beta_mNI)
-# DNA methylation state evolution
-## IWE
-combiStructureGenerator$set("public", "get_mu", function() private$mu)
-## SSE
-singleStructureGenerator$set("public", "get_alpha_Ri", function() private$alpha_Ri)
-singleStructureGenerator$set("public", "get_iota", function() private$iota)
-
-
 #' Get Default Parameter Values
 #'
 #' This function retrieves parameter values for the DNA methylation simulation.
 #'
-#' @param rootData NULL to return default parameter values. For data parameter values, provide rootData$data as the output of simulate_initialData().
+#' @param rootData NULL to return default parameter values. For data parameter values, provide rootData as the output of simulate_initialData()$data.
 #'
 #' @return A data frame containing default parameter values.
 #'
 #' @details The function called without arguments returns default parameter values.
 #' When rootData (as $data output of simulate_initialData()) is given, it returns data parameter values.
 #'
+#' @examples
+#' # Get default parameter values
+#' default_values <- get_parameterValues()
+#'
+#' # Get parameter values of simulate_initialData() output
+#' custom_params <- get_parameterValues()
+#' infoStr <- data.frame(n = c(5, 10), globalState = c("M", "U"))
+#' rootData <- simulate_initialData(infoStr = infoStr, params = custom_params)$data
+#' rootData_paramValues <- get_parameterValues(rootData = rootData)
+#'
 #' @export
 get_parameterValues <- function(rootData = NULL){
   if(is.null(rootData)){
[…]

### 2024-04-08 tests/testthat.R

@@ -7,6 +7,6 @@
 # * https://testthat.r-lib.org/articles/special-files.html
 
 library(testthat)
-library(MethEvolRSIM)
+library(MethEvolSIM)
 
-test_check("MethEvolRSIM")
+test_check("MethEvolSIM")

### 2024-04-08 tests/testthat/test-multiRegion_SIM.R

@@ -9,8 +9,7 @@ test_that("singleStructureGenerator initialization",{
   expect_true(is.null(get_private(single_obj)$my_combiStructure),
               info = "direct singleStructureGenerator initialization assigns not null $my_combiStructure")
 
-  infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+  infoStr <- data.frame(n = c(100, 100, 100),
                         globalState= c("M", "U", "M"))
 
   combi_obj <- combiStructureGenerator$new(infoStr)
@@ -31,17 +30,12 @@ test_that("singleStructureGenerator initialization",{
 })
 
 test_that("combiStructureGenerator initialization", {
-  infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+  infoStr <- data.frame(n = c(100, 100, 100),
                         globalState= c("M", "U", "M"))
 
   combi_obj <- combiStructureGenerator$new(infoStr)
   expect_equal(length(get_private(combi_obj)$singleStr), nrow(infoStr),
                info = "length of $singleStr not equal to row number in input dataframe")
-  expect_equal(length(get_private(combi_obj)$singleStr_start), nrow(infoStr),
-               info = "length of $singleStr_start not equal to row number in input dataframe")
-  expect_equal(length(get_private(combi_obj)$singleStr_end), nrow(infoStr),
-               info = "length of $singleStr_end not equal to row number in input dataframe")
   expect_equal(length(get_private(combi_obj)$singleStr_globalState), nrow(infoStr),
                info = "length of $singleStr_globalState not equal to row number in input dataframe")
   for (i in 1:nrow(infoStr)){
@@ -51,8 +45,7 @@ test_that("combiStructureGenerator initialization", {
 })
 
 test_that("singleStructureGenerator get_seq()", {
-  infoStr <- data.frame(start = c(1, 14, 27),
-                        end = c(13, 26, 39),
+  infoStr <- data.frame(n = c(13, 13, 13),
                         globalState = c("M", "U", "M"))
   combi_obj <- combiStructureGenerator$new(infoStr, testing = TRUE)
   # Define expected $seq under testing mode
@@ -66,8 +59,7 @@ test_that("singleStructureGenerator get_seq()", {
 
[…]

### 2024-04-08 tests/testthat/test-user_functions.R

@@ -5,8 +5,7 @@ get_private <- function(x) {
 
 test_that("get_parameterValues()",{
   # Test error when rootData is not instance of combiStructureGenerator
-  infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+  infoStr <- data.frame(n = c(100, 100, 100),
                         globalState= c("M", "U", "M"))
   output <- get_parameterValues()
   output$alpha_pI <- 0.5
@@ -16,8 +15,7 @@ test_that("get_parameterValues()",{
   output <- get_parameterValues()
   expect_true(is.data.frame(output), info = "fails to output dataframe")
   # Test return rootData parameter values
-  infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+  infoStr <- data.frame(n = c(100, 100, 100),
                         globalState= c("M", "U", "M"))
   output$alpha_pI <- 0.5
   rootData <- simulate_initialData(infoStr = infoStr, params = output)$data
@@ -27,28 +25,27 @@ test_that("get_parameterValues()",{
 
 test_that("simulate_initialData",{
   # Test errors for incorrect input
+
   ## a) Incomplete infoStr argument
-  infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300))
-  expect_error(simulate_initialData(infoStr = infoStr), info ="fails to throw error 'infoStr should be a dataframe with columns: 'start', 'end', 'globalState''")
-  ## b) infoStr with eqFreqs: incorrect values
-  infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+  infoStr <- data.frame(n = c(100, 100, 100))
+  expect_error(simulate_initialData(infoStr = infoStr), info ="fails to throw error 'infoStr should be a dataframe with columns: 'n', 'globalState''")
+
+   ## b) infoStr with eqFreqs: incorrect values
+  infoStr <- data.frame(n = c(100, 100, 100),
                         globalState = c("M", "U", "M"),
                         u_eqFreq = c(0.1, 0.8, 0.1),
                         p_eqFreq = c(NA, 0.1, 0.1),
                         m_eqFreq = c(0.8, 0.1, 0.8))
   expect_error(simulate_initialData(infoStr = infoStr), info = "fails to throw error when given eqFreqs in infoStr have missing values")
-  infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
[…]

### 2024-07-04 README.Rmd

@@
 
-# MethEvolSIM
+```{r setup}
+library(MethEvolSIM)
+```
 
-<!-- badges: start -->
-<!-- badges: end -->
+MethEvolSIM is an R package for simulating DNA methylation dynamics
+on different genomic structures along a given tree.
+It allows simulating methylation and demethylation events at loci
+within two different types of genomic structures: regions with low methylation or with high methylation.
+The data can be interpreted as:
 
-Software for simulation of DNA methylation dynamics on different genomic structures under the basic model for the evolution of DNA methylation along genealogies.
-The methylation state of each CpG dinucleotide is represented as 1 for unmethylated, 2 for partially methylated and 3 for methylated.
+1- CpG sites inside or outside CpG islands or
 
-# Installation
+2- any type of methylation site in regions of higher or lower methylation or
 
-You can use the development version of MethEvolSIM from git. The development is being implemented with the `devtools` package. It can be loaded and checked using the following:
+3- CpG islands inside or outside genomic archipelagos (regions with many CpG islands).
+
+In this document, unless specified otherwise, we use the first interpretation.
+
+# A first example
+As a simple example with default parameter values we show a simulation of
+evolution of a simplified genomic structure on a tree:
 
 ```{r}
-library(devtools)
-load_all()
+# Define the tree:
+tree <- "((a:1, b:1):2, c:2);"
+# Define the genomic structure. Here 1 island ("U") containing 100 CpGs 
+# surrounded by 2 non-island structures ("M") each containing 10 CpGs
+infoStr <- data.frame(n = c(10, 100, 10),
+                      globalState = c("M", "U", "M"))
+# Simulate 1 replicate of data at the tree tips
+output <- simulate_evolData(infoStr = infoStr, tree = tree)
 ```
 
-`check()` runs the unit tests. The unit tests that test the functions used for writing output files:
-- *validationResults.txt*: Is generated in case an error relative to the characteristics of the objects happens and contains both the error and the object.
+In the simulation output 0 stands for unmethylated, 1 for methylated and 0.5 for
+partially methylated or hemimethylated.
 
-# Manual of Use
+```{r}
+# In replicate 1, tip 2: name of the tip of the tree
+output$data[[1]][[2]]$name
+# In replicate 1, tip 2: sequence of methylation states of left non-island structure
+output$data[[1]][[2]]$seq[[1]]
+# In replicate 1, tip 2: sequence of methylation states of island structure
+output$data[[1]][[2]]$seq[[2]]
+# In replicate 1, tip 2: sequence of methylation states of right non-island structure
+output$data[[1]][[2]]$seq[[3]]
+```
+
+# Model parameters
+
+## Initial methylation state distribution
+
+Structures are categorized as CpG islands or non-islands (alternatively regions with low/high methylation or archipelagos/non-archipelagos). In the root of the genealogy we
+assume that each structure samples an equilibrium probability triple $(\pi_u,\pi_p, \pi_m)$ for the methylation states: unmethylated, partially-methylated and methylated, respectively.
+Current parameter values favor prevalent homogeneity in methylation levels within structures,
+most CpG islands predominantly unmethylated and a small proportion predominantly
+methylated, and most non-islands predominantly methylated and a small
+proportion predominantly unmethylated.
+The sampling of the equilibrium probability triple is done using two distinct Beta
+distributions to sample the corresponding methylation frequencies.
+
+**Island parameterization: alpha_pI, beta_pI, alpha_mI, beta_mI**
+
+MethEvolSIM samples $\pi_p$ from the first Beta distribution. Current parameterization
+uses as initial parameter values a relatively small $\alpha_{I_p}$ and relatively large
+$\beta_{I_p}$ to encourage minor assignment of partially methylated states.
+
+The second Beta distribution, scaled by $1 - \pi_p$,  corresponds to the proportion
+of methylated and unmethylated states. Currently both parameter values ($\alpha_{I_m}$
+and $\beta_{I_m}$) set to small values while ${\alpha_{I_m} < \beta_{I_m}}$ to favour
+shifting the stochastic choice toward homogeneous states with a higher proportion of
+predominantly unmethylated islands, and assigned to $\pi_m$.
+Finally, we set $\pi_u$ as $1 - \pi_p - \pi_m$.
+
+**Non-island parameterization: alpha_pNI, beta_pNI, alpha_mNI, beta_mNI**
+
+Following the same rationale as for island parameterization, each non-island structure 
+samples a common equilibrium probability triple from two Beta distributions
+with parameters $\alpha_{NI_p}$, $\beta_{NI_p}$, $\alpha_{NI_m}$ and $\beta_{NI_m}$.
+The parameter values are currently set to favor the characteristic high overall methylation level.
+
+**Parameter ranges** 
+The alpha and beta parameters of a Beta distribution must be greater than 0,
+that is in (0, $\infty$).
+
+## DNA methylation evolution processes
+
+DNA methylation state evolution happens by two structure-specific events: CpG single
+site events (SSEs) and CpG island wide events (IWEs).
+
+### Island-wide events (IWEs)
+
+IWEs change the island methylation probabilities and some of the CpG sites in the
+CpG island can simultaneously change their state at the same time. Different CpG
+islands, however, are assumed to evolve independently of each other.
+IWEs occur per CpG island at a rate denoted by $\mu$. Consequently,
+over a branch of length $l$, the expected number of IWEs per island is given by $\mu \cdot l$.
+
+**IWE parameterization: mu. Parameter range:**
+Should be a small number, but it can take any value from 0, that is in $[0, \infty)$
+
+### Single-site events (SSEs)
+
+In addition to IWEs we allow for single site events (SSEs), which change the
+methylation states of single CpG sites within and between CpG islands.
+In our model SSEs account for two dynamics: independent of each other (SSEi),
+and  assuming correlations between adjacent sites (SSEc).
+
+As SSEc transition probabilities are dependent on adjacent sites we use Gillespie's $\tau$-leap approximation.
+to allow for CpG sites to o update their neighbouring state and rates of change after short time steps.
+Therefore, we first discretize the branch intervals between IWEs in time intervals
+of default length $dt =0.01$.
+For each CpG position and each time interval the rate matrix for
+the transitions between the states unmethylated, partially methylated and methylated
+results from the addition of the rate matrices of the two types of SSE events,
+so that $\mathbb{E}(R_i + R_c)=1$, $R_i$ and $R_c$ representing the rate factors of
+independent and correlated processes.
+The independent process samples the new methylation state from the structure's equilibrium frequencies and the correlated process copies the left or right neighbor state with equal probabilities.
+The proportion of each of the two SSE types
+is determined by the model parameter $\iota$.
+
+Different positions $j$ have different SSEi rates $R_{i,j}$ coming from a
+discretized gamma distribution with 3 categories, expected value $\iota$ and
+shape parameters $\alpha_{R_{i}}$ and ${\beta_{R_{i}}}=\frac{\alpha_{R_{i}}}{\iota}$.
+The probability to be in each respective rate category is 1/3.
+For the rate of collaborative SSEs $R_c = 1-\iota$, the probability of considering
+left of right neighbour is equal.
+
+**SSE process parameterization: iota, alpha_Ri**
+**Parameter ranges** 
+The alpha parameter of a Gamma distribution must be greater than 0, that is (0, $\infty$).
+Iota must be between 0 (non-included) and 1, that is in (0,1].
+
+# Simulation Setup
 
@@
 
-The initial spatial distribution of CpG sites is given by the user.
-A data frame with 3 columns named "start", "end" and "globalState" encodes the structural information.
-Each row provides the information of a different structure.
-The number of CpG positions is given by $end - start +1$, and each row start value needs to be as the previous row end value + 1.
-The globalState encodes island and non-island structures with the character value "U" for islands and "M" for non-islands.
+The initial spatial distribution of CpG sites in islands and non-island structures is given by the user through a data frame with 2 columns named "n" and "globalState"
+, which encode the structural information. 
+Each row provides the information of a different genomic structure.
+
+- **n** number of positions for each structure (e.g. number of CpG sites).
+
+- **globalState** encodes island and non-island structures with the character
+value "U" for unmethylated regions (e.g. CpG islands) and "M" for methylated regions (e.g. non-CpG islands).
 
 ```{r}
-# Example 3 structures of length 100: non-island, island, non-island
-infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+# Example 3 structures of length 100 each:
+# one non-island containing 100 CpG sites,
+# one island containing 100 CpG sites,
+# and one non-island containing 100 CpG sites
+infoStr <- data.frame(n = c(100, 100, 100),
                         globalState = c("M", "U", "M"))
@@
 ```{r}
-# Example 2 islands: first of length 1, second of lenght 15
-infoStr <- data.frame(start = c(1, 2),
-                        end = c(1, 16),
+# Example for 2 islands: first of length 1, second of length 15
+infoStr <- data.frame(n = c(1, 15),
                       globalState = c("U", "U"))
@@
 
-Can be sampled as specified in the section "Model parameters" when infoStr has only start, end and globalState columns,
-or given by the user as additional columns of the data frame. Note that when given, each of the structures frequencies needs to sum up to 1.
+The initial methylation state distribution assigns a triple of methylation equilibrium frequencies() $\pi_u$ for unmethylated, $\pi_p$ for partially-methylated, and $\pi_m$ for methylated) to each of the simulated genomic structures.
+This assignation can be done in two ways:
+
+1 - Sampled as specified in the section "Model parameters: Initial methylation state distribution" when the given `infoStr` only contains `n` and `globalState` columns.
+
+2 - Determined by the user providing in `infoStr` additional columns: `u_eqFreq` for $\pi_u$, `p_eqFreq` for $\pi_p$ and `m_eqFreq` for $\pi_m$.
+Note that when given, the triple of frequency values for each of the structures needs to sum up to 1, so that for each row $i$, $\pi_u(i)+\pi_p(i)+\pi_m(i)=1$.
 
 ```{r}
-# Example 3 structures of length 100 with customized initial methylation state equilibrium frequencies
-infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+# Example 3 structures of length 100 with customized initial methylation state
+# equilibrium frequencies
+infoStr <- data.frame(n = c(100, 100, 100),
                       globalState = c("M", "U", "M"),
@@
 
-The default parameter values used by MethEvolSIM can be obtained using the function `get_model_parameters()`, that returns a data frame with the columns named as specified in the section "Model parameters":
+The default parameter values used by MethEvolSIM can be obtained using the
+function `get_parameterValues()`, which returns a data frame with the columns named
+as specified in the section "Model parameters":
 
@@
 
-The section "Model parameters" provides further information about each of the parameters, including each parameter range of possible values.
-To simulate data with a different parameterization, the new value of a chosen parameter can be modified in the data frame.
+Additionally, it returns the three rates of change for SSEi process,`Ri1`, `Ri2`, and `Ri3` (see section "Model Parameters: Single-site events (SSEs)").
+
+The section "Model parameters" provides further information about each of the parameters,
+including the range of possible values for each parameter.
+To simulate data with a different parameterization, the new value of the chosen parameter(s) can be modified in the data frame, but for the cases of `Ri1`, `Ri2`, and `Ri3`, which are modified through `iota` and `alpha_Ri`.
 
@@
 
-## Initial DNA Methylation Data
+Additional details regarding the `get_parameterValues()` function can be found in its documentation.
 
-### Simulation
+# Simulation of DNA Methylation Data
 
-The initial data is generated in an object of class combiStructureGenerator.
+DNA Methylation Data is simulated using three R6 classes: `singleStructureGenerator`, `combiStructureGenerator`, and `treeMultiRegionSimulator`. 
+Instances of the `singleStructureGenerator` class represent individual genomic structures (e.g.
+islands or non-islands) and are encapsulated within instances of the `combiStructureGenerator` class, which, in turn, represent the simulated data structure (e.g.
+a sequence consisting of a non-island containing 100 CpGs, an island containing 100 CpGs, and a non-island containing 100 CpGs).
+Finally, instances of the `treeMultiRegionSimulator` class contain a `combiStructureGenerator` object at each node of a given tree, including its root and leaves.
+
+Generally, simulations are conducted as explained in subsection "Simulation of Data Evolution", with a documented user-interface function explaining the output structure: `simulate_evolData()`.
+Alternativelly, the user can simulate initial data so that the root of the tree is fixed as explained in subsection "Simulation of Initial Data".
+The latter option generates objects of class `combiStructureGenerator` containing objects of class `singleStructureGenerator`.
+In the subsection "Simulation of Initial Data: Exploration of Initial Data" an example to interact with the classes is given, while further information about the attributes and methods of each class is provided in their documentation.
+
+## Simulation of Initial Data
+
+The initial data can be generated as an object of class `combiStructureGenerator`.
 
@@
 # Example with customized initial methylation frequencies and customized parameter values
-custom_infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+custom_infoStr <- data.frame(n = c(100, 100, 100),
                              globalState = c("M", "U", "M"),
@@
 custom_params$mu <- 0.005
-initialD <- simulate_initialData(infoStr = custom_infoStr, params = custom_params)
+initial_customD <- simulate_initialData(infoStr = custom_infoStr, params = custom_params)
 # Returns customized parameters and simulated data
-initialD$params
-initialD$data
+initial_customD$params
+initial_customD$data
 ```
@@
 ```{r}
-# Example with simulated initial methylation frequencies and default parameter values
-custom_infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+# Example with sampled initial methylation frequencies and default parameter values
+custom_infoStr <- data.frame(n = c(100, 100, 100),
                              globalState = c("M", "U", "M"))
@@
 
-To get the parameter values of a combiStructureGenerator object, it can be given as argument to the function `get_parameterValues()`
+To get the parameter values of a `combiStructureGenerator` object, it can alternativelly be given as argument to the function `get_parameterValues()`.
 
@@
 combiStr_object <- initialD$data
-get_parameterValues(combiStr_object)
+get_parameterValues(rootData = combiStr_object)
 ```
 
-### Exploration of initial data
+### Exploration of Initial Data
 
-The methods of the class combiStructureGenerator can be used to access each of the contained structures:
+The methods of the class `combiStructureGenerator` can be used to access each of the contained structures:
 
-```{r}
-# E.g. access fist contained structure
+```{r, eval=FALSE}
+# E.g. access fist structure (non-island)
 combiStr_object$get_singleStr(1)
@@
 
-Each contained structure is an object of singleStructureGenerator and its methods can be used to get the contained information.
+Each contained structure is an object of `singleStructureGenerator` and its methods can be used to get the contained information.
 
 ```{r}
-# E.g. get the equilibrium frequencies of the first singleStructureGenerator
+# E.g. get the methylation equilibrium frequencies of the first singleStructureGenerator
 combiStr_object$get_singleStr(1)$get_eqFreqs()
@@
 # E.g. get the sequence of methylation states of the second singleStructureGenerator
-combiStr_object$get_singleStr(1)$get_seq()
+# Encoded as: 1 for unmethylated, 2 for partially-methylated, 3 for methylated
+combiStr_object$get_singleStr(2)$get_seq()
 ```
 
-## Simulate DNA Methylation Data Evolution
+## Simulation of Data Evolution
 
-The evolution data is generated in an object of class treeMultiRegionSimulator.
+The evolution data is generated with the function `simulate_evolData()`.
+It returns  the parameters used (`$params`), the length of the time step used for the Gillespie's $\tau$-leap approximation (`$dt`, default 0.01), the tree used (`$tree`), and the simulated data (`$data`).
+The description of the simulated data is available in the function's documentation.
+
+It takes as parameters:
+
+- `tree`: A string representing a tree in Newick format.
+
+- `infoStr`: A dataframe with the information of the number of CpGs corresponding to each island and non-island region (columns named `n` and `globalState`).
+Additionally the initial methylation frequencies can be specified adding the columns `u_eqFreq`, `p_eqFreq` and `m_eqFreq`.
+
+- `dt`: The length of time steps for Gillespie's $\tau$-leap approximation. Numerical, default is 0.01.
+
+- `n_rep`: The number of replicates. Integer, default is 1.
+
+- `only_tip`: Logical indicating whether to extract data only for tips (TRUE, default) or to extract the information for all the tree branches (FALSE).
+
+- `params`: Default NULL for simulation with default parameter values.
+When customized parameter values are used the output dataframe of `get_parameterValues()` can be modified and given to the function.
+E.g:
 
 ```{r}
-# Example with customized initial methylation frequencies, customized parameter values and default dt (0.01)
+# Example with customized initial methylation frequencies, customized parameter values 
+# and default dt (0.01)
 tree <- "((a:1, b:1):2, c:2, (d:3.7, (e:4, f:1):3):5);"
-custom_infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+custom_infoStr <- data.frame(n = c(100, 100, 100),
                              globalState = c("M", "U", "M"),
@@
 custom_params$mu <- 0.005
-evolD <- simulate_evolData(infoStr = custom_infoStr, tree = tree, params = custom_params)
-# Returns customized parameters, tree used, time step length for SSE process used (dt) and simulated data
+evolD <- simulate_evolData(infoStr = custom_infoStr, tree = tree, params = custom_params, n_rep = 3, only_tip = TRUE)
+```
+
+```{r, eval=FALSE}
+# Returns customized parameters, tree used, time step length for SSE process used (dt) 
+# and simulated data
 evolD$data
@@
 
-If the evolution of given data at the root is simulated, and it wants to be done with customized parameter values, then the customized parameters are given to `simulate_initialData()` instead to `simulate_evolData()`.
+```{r}
+# The simulated data is returned as a list. Each element of the list corresponds to a
+# simulation replicate:
+length(evolD$data)
+rep1 <- evolD$data[[1]]
+rep2 <- evolD$data[[2]]
+rep3 <- evolD$data[[3]]
+```
+
+```{r}
+# When only_tip is TRUE, each replicate contains for each tip: its name and the 
+# sequence of methylation states. E.g.
+rep1_tip1 <- rep1[[1]]
+rep1_tip1$name
+# Seq is a list with the methylation states of each simulated structure.
+# In this example, 3 structures: non-island, island and non-island ...
+length(rep1_tip1$seq)
+# Each with 100 CpGs
+length(rep1_tip1$seq[[1]])
+```
+
+When simulations are conducted with `only_tip = FALSE` the `$data` output contains more information.
+
+```{r}
+tree <- "((a:1, b:1):2, c:2, (d:3.7, (e:4, f:1):3):5);"
+custom_infoStr <- data.frame(n = c(100, 100, 100),
+                             globalState = c("M", "U", "M"),
+                             u_eqFreq = c(0.1, 0.8, 0.1),
+                             p_eqFreq = c(0.1, 0.1, 0.1),
+                             m_eqFreq = c(0.8, 0.1, 0.8))
+custom_params <- get_parameterValues()
+custom_params$mu <- 0.005
+evolD <- simulate_evolData(infoStr = custom_infoStr, tree = tree, params = custom_params, n_rep = 3, only_tip = FALSE)
+```
+
+```{r}
+# When only_tip is FALSE. $data output contains the simulated data and the 
+# information on the relationship between tree branches.
+names(evolD$data)
+```
+
+```{r}
+# The information of the tree branches can be accessed using $branchInTree. 
+treeStr <- evolD$data$branchInTree
+# It is a list in which each element index represents the index of the branch
+root <- treeStr[[1]] # First branch (or tree root)
+b2 <- treeStr[[2]] # Second branch 
+# ...
+# Each branch contains information of the offspring and parent indexes.
+root$parent_index # The tree root does not have parent branches
+root$offspring_index # The tree root has 3 daughter branches: 2, 5 and 6. 
+b2$parent_index # Branch 2 therefore has as parent branch the root.
+b2$offspring_index # Branch 2 also has 2 daughter branches: 3 and 4
+```
+
+```{r}
+# The simulated data can be accessed using $sim_data.
+sim_data <- evolD$data$sim_data
+# As before, each list element corresponds to a simulation replicate. 
+# E.g. replicate 3
+rep3 <- sim_data[[3]]
+# In each replicate, each element of the list corresponds to a tree branch
+# the indexes correspond to the information in $branchInTree
+root <- rep3[[1]]
+b2 <- rep3[[2]]
+b3 <- rep3[[3]]
+```
+
+```{r}
+# Each tree branch contains:
+# - the branch name (NULL for tree root and inner nodes and 
+# tip name for the tree tips):
+root$name
+b2$name
+b3$name
+```
+```{r}
+# - the information on IWE events that happened in that branch. 
+root$IWE # The root always has NULL because its branch length is 0
+# The rest of the branches have FALSE when no IWE happened or a list containing
+# $islands corresponds to the island index(es) to which the event(s) happened 
+# $times corresponds to the branch time in which the event happened.
+b2$IWE
+b3$IWE
+```
+
+```{r}
+# - A list with each element representing the sequence of methylation states of 
+# each simulated island and non-island
+# Encoded as 0 for unmethylated, 0.5 for partially methylated and 1 for methylated
+root$seq[[1]] # First sequence (non-island)
+root$seq[[2]] # Second sequence (island)
+```
+
+```{r}
+# - A list with each element representing the methylation frequencies for the states: unmethylated, partially-methylated and methylated
+root$eqFreqs[[1]][1] # First structure of root (non-island) equilibrium frequency for unmethylated state
+root$eqFreqs[[1]][2] # First structure of root (non-island) equilibrium frequency for partially-methylated state
+root$eqFreqs[[1]][3] # First structure of root (non-island) equilibrium frequency for methylated state
+```
+
+- `rootData`: Alternative argument to `infoStr` (only one can be given).
+It enables simulating replicates of evolution along a tree with fixed data at root.
+When used, it takes the output of `simulate_initialData()$data`.
+If customized parameter values are to be used, then the customized parameters are given to `simulate_initialData()` instead to `simulate_evolData()`.
 
@@
 tree <- "((a:1, b:1):2, c:2, (d:3.7, (e:4, f:1):3):5);"
-custom_infoStr <- data.frame(start = c(1, 101, 201),
-                        end = c(100, 200, 300),
+custom_infoStr <- data.frame(n = c(100, 100, 100),
                              globalState = c("M", "U", "M"),
@@
 ```
-### Exploration of evolved data
-
-The treeMultiStructureGenerator object contains a list `$Branch` with the combiStructureGenerator representing the data at the each node after a tree branch (but for the data at the root).
-
-To know which offspring branches a node has, the combiStructureGenerator object method `$get_offspring_index()` can be used.
-
-```{r}
-treeData <- evolD$data
-# E.g. offspring of data at root 
-treeData$Branch[[1]]$get_offspring_index()
-```
-If the node is an inner node, its name is set as NULL. It can be accessed with the combiStructureGenerator method `$get_name()`
-
-```{r}
-treeData$Branch[[1]]$get_name()
-```
-But the leafs are named as in the given tree.
-
-```{r}
-treeData$Branch[[4]]$get_name()
-```
-If there were IWE events sampled in a branch, it can be accessed with the combiStructureGenerator method `$get_IWE_events()`
-
-```{r}
-treeData$Branch[[4]]$get_IWE_events()
-```
-In the case we want to know the equilibrium frequencies of a singleStructureGenerator object contained in the combiStructureGenerator we can use the combiStructureGenerator method to access the singleStructureGenerator object of interest `get_singleStr(i)` followed by the singleStructureGenerator method `get_eqFreqs()`
-
-```{r}
-# E.g. get the ewuilibrium frequencies of the second structure at the leaf "b"
-treeData$Branch[[4]]$get_singleStr(2)$get_eqFreqs()
-```
-
-# Model parameters
-
-## Initial methylation state distribution
-
-Structures are categorized as CpG islands or non-islands.
-In the root of the genealogy we assume that each structure samples an equilibrium probability triple $(\pi_u,\pi_p, \pi_m)$.
-Current parameter values favor prevalent homogeneity in methylation levels within structures, most CpG islands with a majority predominantly unmethylated and a small proportion predominantly methylated, and most non-islands with a majority predominantly methylated and a small proportion predominantly unmethylated.
-The sampling of the equilibrium probability tripple is done using two distinct Beta distributions to sample the corresponding methylation frequencies.
-
-**Island parameterization: alpha_pI, beta_pI, alpha_mI, beta_mI**
-
-MethEvolSIM samples $\pi_p$ from the first Beta distribution.
-Current parameterization uses as initial parameter values a relatively small $\alpha_{I_p}$ and relatively large $\beta_{I_p}$ to encourage minor assignment of partially methylated states.
-
-The second Beta distribution, scaled by $1 - \pi_p$,  corresponds to the proportion of methylated and unmethylated states.
-Currently both parameter values ($\alpha_{I_m}$ and $\beta_{I_m}$) set to small values while ${\alpha_{I_m} < \beta_{I_m}}$ to favour shifting the stochastic choice toward homogeneous states with a higher proportion of predominantly unmethylated islands, and assigned to $\pi_m$.
-Finally, we set $\pi_u$ as $1 - \pi_p - \pi_m$.
-
-**Non-island parameterization: alpha_pNI, beta_pNI, alpha_mNI, beta_mNI**
-
-Following the same rationale as for island parameterization, all the regions not belonging to an island sample a common equilibrium probability triple from two Beta distributions with parameters $\alpha_{NI_p}$, $\beta_{NI_p}$, $\alpha_{NI_m}$ and $\beta_{NI_m}$.
-The parameter values are currently set to favor the characteristic high overall methylation level.
-
-You can simulate DNA methylation along a tree given some data at root, or by simulating data at root given a structural distribution and letting it evolve.
-
-**Parameter ranges** 
-The alpha and beta parameters of a Beta distribution must be greater than 0, that is (0, $\infty$).
-
-## DNA methylation evolution processes
-
-DNA methylation state evolution happens by two structure-specific events: CpG single site events (SSEs) and CpG island wide events (IWEs).
-
-### Island-wide events (IWEs)
-
-IWEs change the island methylation probabilities and some of the CpG sites in the CpG island can simultaneously change their state at the same time.
-Different CpG islands, however, are assumed to evolve independently of each other.
-IWEs occur per CpG island at a rate denoted by $\mu$.
-Consequently, over a branch of length $l$, the expected number of IWEs per island is given by $\mu \cdot l$.
-
-**IWE parameterization: mu. Parameter range:**
-Should be a small number, but it can take any value from 0, that is [0, $\infty$)
-
-### Single-site events (SSEs)
-
-In addition to IWEs we allow for single site events (SSEs), which change the methylation states of single CpG sites within and between CpG islands.
-In our model SSEs account for two dynamics: independent of each other (SSEi), and  assuming correlations between adjacent sites (SSEc).
-
-As SSEc transition probabilities are dependent on adjacent sites we need to approximate SSE based evolution by short time steps, allowing for CpG sites to actualize their state.
-Therefore, we first discretize the branch intervals between IWEs in time intervals of length 0.01.
-For each CpG position and each time interval the rate matrix for the transitions between the states unmethylated, partially methylated and methylated results from the addition of the rate matrices of the two types of SSE events, so that $\mathbb{E}(R_i + R_c)=1$, $R_i$ and $R_c$ representing the rate factors of independent and correlated processes.
-The proportion of each of the two SSE types is determined by the parameter $\iota$.
-
-Different positions $j$ have different SSEi rates $R_{i,j}$ coming from a discretized gamma distribution with 3 categories, expected value $\iota$ and shape parameters $\alpha_{R_{i}}$ and ${\beta_{R_{i}}}=\frac{\alpha_{R_{i}}}{\iota}$.
-The probability to be in each respective rate category is 1/3.
-For the rate of collaborative SSEs $R_c = 1-\iota$, the probability of considering left of right neighbour is equal.
-
-**SSE process parameterization: iota, alpha_Ri**
-**Parameter ranges** 
-The alpha parameter of a Gamma distribution must be greater than 0, that is (0, $\infty$). Iota must be between 0 (non-included) and 1, that is (0,1].

### 2024-07-04 tests/testthat/test-anikolas.R
Adding lines on test-anikolas.R

@@ -12,10 +12,10 @@ test_num <- 1000
     old_freqs <- random_vector1
     new_freqs <- random_vector2
     
-    if(all.equal(old_freqs, new_freqs)!=TRUE&& sum(old_freqs)==1){
-      expect_equal(sum(old_freqs),1)
+    if(all.equal(old_freqs, new_freqs)!=TRUE){
+      #expect_equal(sum(old_freqs),1)
       #creating a new ssg object to test get_SrMatrix
-      ssg <- singleStructureGenerator$new(globalState = "U", n = x,eqFreqs = old_freqs,testing=FALSE)
+      ssg <- singleStructureGenerator$new(globalState = "U", n = x,testing=FALSE)
       out <- ssg$get_SrMatrix(old_eqFreqs = old_freqs, new_eqFreqs = new_freqs, testing=TRUE)
     }
     else{

### 2024-07-04 R/nikolas_extenstion.R
Adding info to nikolas_extension.R

@@ -268,4 +268,8 @@ singleStructureGenerator$set("public","RE_within",function(Y_seq,Y_eqFreqs,testi
 #print(ssg$RE_within(Y_seq=c(3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3),Y_eqFreqs=c(0.7,0.1,0.2)))
 
 
+my_str <- data.frame(n = c(100,100,100), globalState = c("U", "U", "U"))
+combi_1 <- combiStructureGenerator$new(infoStr = my_str)
+combi_2 <- combiStructureGenerator$new(infoStr = my_str)
+
 

### 2024-07-20 R/multiRegion_SIM.R
Cleaned code from error handling and documented last public methods implemented

@@ -229,10 +229,9 @@ singleStructureGenerator <-
                 ## @return A list with the 3 Ri_values
                 ##
                 init_Ri_values = function(){
-                  # Values under alpha_Ri 1e-3 lead to quantile values so close to 0 that they are rounded to 0
+                  # Values under alpha_Ri 1e-2 lead to quantile values so close to 0 that they are rounded to 0
                   # To prevent that:
                   private$alpha_Ri <- max(private$alpha_Ri, 1e-2)
-                  #print(private$alpha_Ri)
                   
                   # Set the values that divide the gamma distribution in 3 equal probability categories
                   qGamma_oneThird <- stats::qgamma(1/3, shape = private$alpha_Ri, scale= private$iota / private$alpha_Ri)
@@ -800,61 +799,16 @@ singleStructureGenerator <-
                                 position <- i
                                 old_St <- private$seq[i]
                             }
-                                        # assign new sequence position state with probability given by the relative rates of changing to each of the 2 other states
-                            # Define a global variable to store error information
-                            error_info_global <- NULL
                             
-                            # Try-catch block with error handling
-                            error_info <- tryCatch({
-                              # Attempt to run sample()
-                              private$seq[i] <<- sample(1:3, size = 1, prob = sapply(Q[[private$siteR[i]]][[private$neighbSt[i]]][private$seq[i], ], max, 0))
-                              NULL # If no error, return NULL
-                            }, error = function(e) {
-                              # Capture error information in a list
-                              singleStrcloned <- self$clone()
-                              list(
-                                message = conditionMessage(e),
-                                ratetree = private$ratetree[[1]][1],
-                                M = M,
-                                probabilities = sapply(Q[[private$siteR[i]]][[private$neighbSt[i]]][private$seq[i], ], max, 0),
-                                Q = Q[[private$siteR[i]]][[private$neighbSt[i]]][private$seq[i], ],
-                                singleStructure_index = private$combiStructure_index,
-                                CpG_position = i,
-                                singleStrcloned = singleStrcloned
-                              )
-                            })
+                            # assign new sequence position state with probability given by the relative rates of changing to each of the 2 other states
+                            private$seq[i] <<- sample(1:3, size=1, prob=sapply(Q[[private$siteR[i]]][[private$neighbSt[i]]][private$seq[i],], max, 0))
                             
-                            # If error_info is not NULL, it means an error occurred
-                            if (!is.null(error_info)) {
-                              print("Error occurred:")
-                              print(error_info$message)
[…]

### 2024-07-20 R/multiRegion_SIM.R
Finished complete implementation of update_ratetree

@@ -955,7 +955,11 @@ singleStructureGenerator <-
                             transPropMC_validationResults(validationStates)
                         }
                         
-                        
+                        # Sample $seq accordint to transition probablities
+                        newseq <- rep(0, length(private$seq))
+                        for(i in 1:length(newseq)) {
+                          newseq[i] <- sample(1:3, size=1, prob=as.vector(Mk[private$seq[i],]))
+                        }
                         
 
                                         # Update $seq and $neighbSt

### 2024-07-30 R/multiRegion_SIM.R
Set minimum value of parameter iota to avoid numerical errors

@@ -232,15 +232,18 @@ singleStructureGenerator <-
                   # Values under alpha_Ri 1e-2 lead to quantile values so close to 0 that they are rounded to 0
                   # To prevent that:
                   private$alpha_Ri <- max(private$alpha_Ri, 1e-2)
+                  # Very small values of iota lead to error in integrate for Ri3 due to a very small probability of the tail
+                  # To prevent that:
+                  private$iota <- max(private$iota, 1e-2)
                   
                   # Set the values that divide the gamma distribution in 3 equal probability categories
                   qGamma_oneThird <- stats::qgamma(1/3, shape = private$alpha_Ri, scale= private$iota / private$alpha_Ri)
                   qGamma_twoThird <- stats::qgamma(2/3, shape = private$alpha_Ri, scale= private$iota / private$alpha_Ri)
 
                   # Calculate the center of gravity of each gamma distribution category
-                  Ri1 <- stats::integrate(function(x) x*stats::dgamma(x, shape =  alpha_Ri, scale =  iota / alpha_Ri) * 3, 0, qGamma_oneThird)$value
-                  Ri2 <- stats::integrate(function(x) x*stats::dgamma(x, shape =  alpha_Ri, scale =  iota / alpha_Ri) * 3, qGamma_oneThird, qGamma_twoThird)$value
-                  Ri3 <- stats::integrate(function(x) x*stats::dgamma(x, shape =  alpha_Ri, scale =  iota / alpha_Ri) * 3, qGamma_twoThird, Inf)$value
+                  Ri1 <- stats::integrate(function(x) x*stats::dgamma(x, shape =  private$alpha_Ri, scale =  private$iota / private$alpha_Ri) * 3, 0, qGamma_oneThird)$value
+                  Ri2 <- stats::integrate(function(x) x*stats::dgamma(x, shape =  private$alpha_Ri, scale =  private$iota / private$alpha_Ri) * 3, qGamma_oneThird, qGamma_twoThird)$value
+                  Ri3 <- stats::integrate(function(x) x*stats::dgamma(x, shape =  private$alpha_Ri, scale =  private$iota / private$alpha_Ri) * 3, qGamma_twoThird, Inf)$value
                   return(c(Ri1, Ri2, Ri3))
                 },
                 ## @field Rc_values Private attribute: Vector containing 2 Rc rate values

### 2024-08-14 R/multiRegion_SIM.R
Fixing asignation of customized alpha_pNI

@@ -654,7 +654,7 @@ singleStructureGenerator <-
                     private$beta_pI <- params$beta_pI
                     private$alpha_mI <- params$alpha_mI
                     private$beta_mI <- params$beta_mI
-                    private$alpha_pNI <- params$beta_pNI
+                    private$alpha_pNI <- params$alpha_pNI
                     private$beta_pNI <- params$beta_pNI
                     private$alpha_mNI <- params$alpha_mNI
                     private$beta_mNI <- params$beta_mNI

### 2024-11-25 extensive_tests/draft-abc-inferences-vignette.Rmd
Adding draft of vignettes and snakemake workflow

---
title: "ABC inferences"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to MethEvolSIM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE}
## TODO: Update 
library(devtools)
load_all()
```

# Worflow for simulation-based inferences

This workflow uses [Snakemake](https://snakemake.readthedocs.io) to manage the steps for simulating DNA methylation patterns, assessing ABC accuracy, and inferring model parameters.

## Requirements

- Snakemake installed (`conda create -c conda-forge -c bioconda -n snakemake snakemake`)

- R packages: optparse (add to dependencies), parallel

- Access to the provided `Snakefile` and associated scripts.

- An .RData file with a dataframe specifying the distribution of methylation sites in the genomic region.

## Steps in the Workflow

1. Save as simulation design the genomic structure, tree(s) -- GIVEN -- and sampled parameter combinations.

2. Run the simulations. TODO: Before submitting load package without devtools

3. Compute the selected summary statistics. TODO: When added sumStats functions to package, stop sourcing script.

4. Assess ABC accuracy using cross-validation.

5. Infer parameters from empirical DNA methylation data.

### 2024-11-25 extensive_tests/draft-sumstat-vignette.Rmd
Adding draft of vignettes and snakemake workflow

---
title: "Summary Statistics"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to MethEvolSIM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE}
## TODO: Update 
library(devtools)
load_all()
```

Interpretation of arguments:

 - tip can be interpreted as sample or replicate

 - structure can be interpreted as island or non-island

## Frequency of methylation states per genomic structure type

### Global Mean

- **Partially-methylated sites**: Let $x_{i,t}$ and $y_{j,t}$ be the frequency of methylation state $p$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$ at tip $t$, respectively. We define the mean global frequency of $x$ and $y$ as:

\[
\overline{x}=\frac1I\sum_{i=1}^{I}\frac1T\sum_{t=1}^{T}x_{i,t} \\
\overline{y}=\frac1J\sum_{j=1}^{J}\frac1T\sum_{t=1}^{T}y_{j,t}
\]

, where $x_{i,t}$ is the frequency of state $p$ in island $i$ and tip $t$ and $y_{j,t}$ is the frequency of state $p$ in non-island $j$ and tip $t$.

- **Methylated sites**: Let $w_{i,t}$ and $z_{j,t}$ be the frequency of methylation state $m$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$ at tip $t$, respectively. We define the mean global frequency of $w$ and $z$ as:

\[
\overline{w}=\frac1I\sum_{i=1}^{I}\frac1T\sum_{t=1}^{T}w_{i,t} \\
\overline{z}=\frac1J\sum_{j=1}^{J}\frac1T\sum_{t=1}^{T}z_{j,t}
\]

, where $w_{i,t}$ is the frequency of state $p$ in island $i$ and tip $t$ and $z_{j,t}$ is the frequency of state $m$ in non-island $j$ and tip $t$.

These values can be computed with the functions `get_islandMeanFreqP`, `get_nonislandMeanFreqP`,`get_islandMeanFreqM` and `get_nonislandMeanFreqM` as in the following examples:

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 0, .5)) # tip 1
get_islandMeanFreqP(index_islands, data, sample_n)
get_nonislandMeanFreqP(index_nonislands, data, sample_n)
get_islandMeanFreqM(index_islands, data, sample_n)
get_nonislandMeanFreqM(index_nonislands, data, sample_n)
```

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 0, .5)), # tip 1
  list(c(0, .5, .5, 1, 1, .5), c(1, .5, 1, .5), c(0, .5, .5), c(1, .5, 1)) # tip 2
  )
get_islandMeanFreqP(index_islands, data, sample_n)
get_nonislandMeanFreqP(index_nonislands, data, sample_n)
get_islandMeanFreqM(index_islands, data, sample_n)
get_nonislandMeanFreqM(index_nonislands, data, sample_n)
```

### Mean SD of observed frequencies per structure type at tree tips

- **Partially-methylated sites**: Let $x_{i,t}$ and $y_{j,t}$ be the frequency of methylation state $p$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$, respectively. We define the standard deviation of the frequency of $x$ and $y$ at tip $t$ as:

\[
\sigma_{t}(x)=\sqrt{\frac{1}{I-1}\sum_{i=1}^{I}(x_{i,t}-\overline{x_{.,t}})^2} \\
\sigma_{t}(y)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{i,t}-\overline{y_{.,t}})^2}
\]

, where $\overline{x_{.,t}}$ is the mean frequency of state $p$ at tip $t$ in islands and $\overline{y_{.,t}}$ is the mean frequency of state $p$ at tip $t$ in non-islands:

\[
\overline{x_{.,t}}=\frac{1}{I}\sum_{i=1}^{I}x_{i,t} \\
\overline{y_{.,t}}=\frac{1}{J}\sum_{j=1}^{J}x_{j,t}
\]

Therefore, the mean standard deviation across tips is:

\[
\hat{\sigma}(x)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(x) \\
\hat{\sigma}(y)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(y) \\
\]

- **Methylated sites**: Let $w_{i,t}$ and $z_{j,t}$ be the frequency of methylation state $m$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$, respectively. We define the standard deviation of the frequency of $w$ and $z$ at tip $t$ as:

\[
\sigma_{t}(w)=\sqrt{\frac{1}{I-1}\sum_{i=1}^{I}(x_{i,t}-\overline{x_{.,t}})^2} \\
\sigma_{t}(z)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{i,t}-\overline{y_{.,t}})^2}
\]

, where $\overline{x_{.,t}}$ is the mean frequency of state $m$ at tip $t$ in islands and $\overline{z_{.,t}}$ is the mean frequency of state $m$ at tip $t$ in non-islands:

\[
\overline{w_{.,t}}=\frac{1}{I}\sum_{i=1}^{I}w_{i,t} \\
\overline{z_{.,t}}=\frac{1}{J}\sum_{j=1}^{J}z_{j,t}
\]

Therefore, the mean standard deviation across tips is:

\[
\hat{\sigma}(w)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(w) \\
\hat{\sigma}(z)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(z) \\
\]

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(c(.5, .5, 0, 1, 1, .5), c(.5, 0, 1, .5), c(.5, .5, 0), c(0, 0, .5)) # tip 1
get_islandSDFreqP(index_islands, data, sample_n)
get_nonislandSDFreqP(index_nonislands, data, sample_n)
get_islandSDFreqM(index_islands, data, sample_n)
get_nonislandSDFreqM(index_nonislands, data, sample_n)
```

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  list(c(.5, .5, 0, 0, 0, 1), c(.5, 0, 0, .5), c(1, .5, 0), c(0, 0, .5)), # tip 1
  list(c(0, .5, .5, 1, 1, .5), c(1, .5, 1, .5), c(0, .5, .5), c(1, .5, 1)) # tip 2
  )
get_islandSDFreqP(index_islands, data, sample_n)
get_nonislandSDFreqP(index_nonislands, data, sample_n)
get_islandSDFreqM(index_islands, data, sample_n)
get_nonislandSDFreqM(index_nonislands, data, sample_n)
```

Note that we average across tips as there can be correlations between the mean sd of tips (e.g. under the case of an event affecting the states at two tips).

# Mean correlation of a segment of methylation states per genomic structure type

Let $s$ be the length of the 2 adjacent segments of central positions to consider.
Let $l$ represent a number of positions to exclude at the start and end of each island/non-island structure.
Define the index positions $p=l+1$ and $p' = s-l$.
Let $x_{i,t,[p,p'-1]}$ and $x_{i,t,[p+1,p']}$ be the methylation states $x \in \{u, p, m\}$ in island $i \in \{1, \cdots ,I\}$ at tip $t$ for the given segments. 
Let $y_{j,t,[p,p'-1]}$ and $y_{j,t,[p+1,p']}$ be the methylation states $y \in \{u, p, m\}$ in non-island $j \in \{1, \cdots ,J\}$ at tip $t$ for the given segments.

We define the mean correlation as:

\[
\overline{Cor(x)_{s,l}}=\frac1I\frac1T\sum_{i=1}^{I}\sum_{t=1}^{T}Cor(x_{i,t,[p:p'-1]}, x_{i,t,[p+1:p']}) \\
\overline{Cor(y)_{s,l}}=\frac1J\frac1T\sum_{j=1}^{J}\sum_{t=1}^{T}Cor(y_{j,t,[p:p'-1]}, y_{j,t,[p+1:p']})
\]

where correlations are computed only for segment pairs where the standard deviation of each segment is non-zero:

\[
\sigma(x_{i,t,[p:p'-1]}) \neq 0 \text{ and } \sigma(x_{i,t,[p+1:p']})\neq 0 \\
\sigma(y_{j,t,[p:p'-1]}) \neq 0 \text{ and } \sigma(y_{j,t,[p+1:p']})\neq 0
\]

In this formulation:

- Only structures with a minimum length of $n = s + 2 \cdot l$ are considered.

- If $l = 0$, all the positions are considered.

*Correlation measures how much the state of each site says about the next. So, when there is no variation (e.g. in middle segment state u of a site is followed by state u of the next), there is no correlation. Therefore, avoiding the shores leads to lower correlation values. Maybe we could test with/out borders.*

### 2024-11-25 extensive_tests/abc_inferences/scripts/compute_sumStats.R

@@ -1,0 +1,43 @@
+library(optparse)
+# TODO: ADD when source is eliminated
+# library(devtools); load_all()
+
+## TODO: Once functions are commented and tested and belong to the package, this should be library(MethEvolSIM) / while developing library(devtools); load_all()
+source("../../functions_summaryStats.R") # Load functions to compute summary statistics
+
+if (FALSE){
+  # Define the command-line options
+  option_list <- list(
+    make_option(c("-d", "--data-dir"), type = "character", default = NULL,
+                help = "Full path to the directory containing the .RData files", metavar = "character"),
+    make_option(c("-f", "--design-file"), type = "character", default = NULL,
+                help = "Full path to the simulation design file", metavar = "character"),
+    make_option(c("-s", "--stats"), type = "character", default = "all",
+                help = "Comma-separated list of summary statistics to compute (default: all). Options: meanFreqP_i, meanFreqP_ni, sdFreqP_i, sdFreqP_ni, meanFreqM_i, meanFreqM_ni, sdFreqM_i, sdFreqM_ni, meanFracMoverMU_i, meanFracMoverMU_ni, sdFracMoverMU_i, sdFracMoverMU_ni, FChangeCherry_i, FChangeCherry_ni, Fitch, Transitions, meanCor", metavar = "character"),
+    make_option(c("-n", "--sample-n"), type = "integer", default = NULL,
+                help = "Number of samples per file", metavar = "integer"),
+    make_option(c("-p", "--pattern"), type = "character", default = NULL,
+                help = ".RData files start name pattern", metavar = "character"),
+    make_option(c("-u", "--update-file"), type = "character", default = NULL,
+                help = "Full path to existing summary statistics file", metavar = "character")
+  )
+}
+
+
+
+# Define command-line options
+option_list <- list(
+  make_option("--n-sim", type = "integer", default = NULL,
+              help = "Number of simulations", metavar = "integer")
+)
+
+# Parse the arguments
+opt_parser <- OptionParser(option_list = option_list)
+opt <- parse_args(opt_parser)
+
+pad_n <- nchar(as.character(opt[["n-sim"]])) + 1
+
+RData_files <- list.files(pattern = paste0("^abc_dataSIM_\\d{", pad_n, "}\\.RData"))
+
+# Print the input files
+print(RData_files)

### 2024-11-25 extensive_tests/abc_inferences/scripts/run_sim.R

@@ -1,0 +1,89 @@
+library(optparse)
+library(parallel)
+library(devtools); load_all()
+
+# Define command-line options
+option_list <- list(
+  make_option("--input", type = "character", default = NULL,
+              help = "Design file, full path", metavar = "character"),
+  make_option("--n-cores", type = "integer", default = NULL,
+              help = "Number of cores", metavar = "integer"),
+  make_option("--n-tips", type = "integer", default = NULL,
+              help = "Number of tree tips", metavar = "integer")
+)
+
+# Parse the arguments
+opt_parser <- OptionParser(option_list = option_list)
+opt <- parse_args(opt_parser)
+
+# Load the simulation design file
+load(opt[["input"]])
+
+# Define the number of runs and cores
+n_sim <- nrow(sampled_params)
+n_cores <- opt[["n-cores"]]
+
+# Generate pad_n based on the number of digits in n_sim
+# To save the files with padded numbers so that are later listed in order with list.files()
+pad_n <- nchar(as.character(n_sim)) + 1
+
+run_simulation_subset <- function(start_idx, end_idx) {
+  
+  # Generate the padded start and end indexes for writing the output
+  padded_start <- formatC(start_idx, width = pad_n, format = "d", flag = "0")
+  padded_end <- formatC(end_idx, width = pad_n, format = "d", flag = "0")
+  
+  # Set the name for the .out file for the entire subset
+  out_file <- paste0("abc_dataSIM_", padded_start, "_to_", padded_end, ".out")
+  
+  # Open the log file once for the entire subset
+  log_connection <- file(out_file, open = "w")
+  
+  # Redirect both regular output and message output to the same file
+  sink(log_connection, type = "output")
+  sink(log_connection, type = "message")
+  
+  # Run the simulations for the current subset
+  for (sim in start_idx:end_idx) {
+    print(paste("Running simulation number:", sim))
+    
+    padded_index <- formatC(sim, width = pad_n, format = "d", flag = "0")
+    
+    # Set a seed
+    set.seed(sim)
+    
+    # Simulate data
+    output <- simulate_evolData(infoStr = spatial_str, tree = tree, params = sampled_params[sim,])
+    
+    # Extract methylation data at the tree tips
+    data <- list()
+    for (tip in 1:opt[["n-tips"]]) {
+      data[[tip]] <- output$data[[1]][[tip]]$seq
+    }
[…]

### 2024-11-25 extensive_tests/abc_inferences/scripts/set_sim_design.R

@@ -1,0 +1,70 @@
+library(optparse)
+
+# Define command-line options with unique letters
+option_list <- list(
+  make_option("--dir", type = "character", default = NULL,
+              help = "Full path to save the design file", metavar = "character"),
+  make_option("--genome-dist", type = "character", default = NULL,
+              help = "Distribution of methylation sites in the genomic region", metavar = "character"),
+  make_option("--tree", type = "character", default = "all",
+              help = "Tree in Newick format", metavar = "character"),
+  make_option("--n-sim", type = "integer", default = NULL,
+              help = "Number of simulations", metavar = "integer"),
+  make_option("--seed", type = "integer", default = NULL,
+              help = "Seed value", metavar = "integer"),
+  make_option("--alpha_pI_shape", type="numeric", default=3, help="Shape parameter for alpha_pI"),
+  make_option("--beta_pI_shape", type="numeric", default=3, help="Shape parameter for beta_pI"),
+  make_option("--alpha_pI_scale", type="numeric", default=as.numeric(1/30), help="Scale parameter for alpha_pI"),
+  make_option("--beta_pI_scale", type="numeric", default=as.numeric(1/3), help="Scale parameter for beta_pI"),
+  make_option("--alpha_pNI_shape", type="numeric", default=3, help="Shape parameter for alpha_pNI"),
+  make_option("--beta_pNI_shape", type="numeric", default=3, help="Shape parameter for beta_pNI"),
+  make_option("--alpha_pNI_scale", type="numeric", default=as.numeric(1/30), help="Scale parameter for alpha_pNI"),
+  make_option("--beta_pNI_scale", type="numeric", default=as.numeric(1/3), help="Scale parameter for beta_pNI"),
+  make_option("--alpha_mI_shape", type="numeric", default=3, help="Shape parameter for alpha_mI"),
+  make_option("--beta_mI_shape", type="numeric", default=3, help="Shape parameter for beta_mI"),
+  make_option("--alpha_mI_scale", type="numeric", default=as.numeric(1/30), help="Scale parameter for alpha_mI"),
[…]

### 2024-11-25 extensive_tests/compute_summaryStats.R

@@ -1,23 +1,6 @@
 # Load necessary libraries
 library(optparse)
-## TODO: Once functions are commented and tested and belong to the package, this should be library(MethEvolSIM) / while developing library(devtools); load_all()
-source("functions_summaryStats.R") # Load functions to compute summary statistics
 
-# Define the command-line options
-option_list <- list(
-  make_option(c("-d", "--data-dir"), type = "character", default = NULL,
-              help = "Full path to the directory containing the .RData files", metavar = "character"),
-  make_option(c("-f", "--design-file"), type = "character", default = NULL,
-              help = "Full path to the simulation design file", metavar = "character"),
-  make_option(c("-s", "--stats"), type = "character", default = "all",
-              help = "Comma-separated list of summary statistics to compute (default: all). Options: meanFreqP_i, meanFreqP_ni, sdFreqP_i, sdFreqP_ni, meanFreqM_i, meanFreqM_ni, sdFreqM_i, sdFreqM_ni, meanFracMoverMU_i, meanFracMoverMU_ni, sdFracMoverMU_i, sdFracMoverMU_ni, FChangeCherry_i, FChangeCherry_ni, Fitch, Transitions, meanCor", metavar = "character"),
-  make_option(c("-n", "--sample-n"), type = "integer", default = NULL,
-              help = "Number of samples per file", metavar = "integer"),
-  make_option(c("-p", "--pattern"), type = "character", default = NULL,
-              help = ".RData files start name pattern", metavar = "character"),
-  make_option(c("-u", "--update-file"), type = "character", default = NULL,
-              help = "Full path to existing summary statistics file", metavar = "character")
-)
 
 
 # Parse the arguments
@@ -64,12 +47,15 @@ if(length(unique(spatial_str$n)) == 1){
 
 # List simulation output files
 ## TODO: the pattern should be given complete, because sometimes it is followed by only evol number, sometimes by reps, sometimes by nothing
-  RData_files <- list.files(opt[["data-dir"]], pattern = paste0("^", opt[["pattern"]], "_\\d{4}\\.RData$"), full.names = TRUE)
+print(paste("data dir", opt[["data-dir"]]))
+print(paste("pattern", opt[["pattern"]]))
+  RData_files <- list.files(opt[["data-dir"]], pattern = paste0("^", opt[["pattern"]], "_\\d{6}\\.RData$"), full.names = TRUE)
 
 
 
 # Set the number of simulations to process
 n_sim <- length(RData_files)
[…]

### 2024-11-25 extensive_tests/functions_summaryStats.R

@@ -1,220 +1,4 @@
-##### SET FUNCTIONS TO EXTRACT SUMMARY STATISTICS ##############################
 
-#### #### #### Frequency of partially methylated sites #### #### ####
-##### Mean frequency of partially methylated sites in islands
-## index_islands: vector with structural indices for islands
-## data: list with methylation states at tree tips for each structure 
-## data[[tip]][[structure]] when the number of tips is >1, 
-## or data[[structure]] when there is only one tip. Methylation states are vectors
-## sample_n: number of given tips/samples
-get_islandMeanFreqP <- function(index_islands, data, sample_n){
-  # Restructure data as nested list
-  if (sample_n == 1){
-    data_list <- list()
-    data_list[[1]] <- data
-    data <- data_list
-  }
-  mean_island <- c()
-  island_counter <- 1
-  for (i in index_islands){
-    mean_tip <- c()
-    for(s in 1:sample_n){
-      mean_tip[s] <- mean(data[[s]][[i]]==0.5)
-    }
-    mean_island[island_counter] <- mean(mean_tip)
-    island_counter <- island_counter + 1
-  }
-  return(mean(mean_island))
-}
-
-##### Mean frequency of partially methylated sites in non-islands
-## index_nonislands: vector with structural indeces for non-islands
-## data: list with methylation states at tree tips for each structure 
-## data[[tip]][[structure]] when the number of tips is >1, 
-## or data[[structure]] when there is only one tip. Methylation states are vectors
-## sample_n: number of given tips/samples
-get_nonislandMeanFreqP <- function(index_nonislands, data, sample_n){
-  # Restructure data as nested list
-  if (sample_n == 1){
-    data_list <- list()
-    data_list[[1]] <- data
-    data <- data_list
-  }
-  mean_nonisland <- c()
-  nonisland_counter <- 1
-  for (i in index_nonislands){
-    mean_tip <- c()
-    for(s in 1:sample_n){
-      mean_tip[s] <- mean(data[[s]][[i]]==0.5)
-    }
-    mean_nonisland[nonisland_counter] <- mean(mean_tip)
-    nonisland_counter <- nonisland_counter + 1
-  }
-  return(mean(mean_nonisland))
-}
-
-##### Mean standard deviation of the frequency of partially methylated sites in islands
-## index_islands: vector with structural indeces for islands
-## data: list with methylation states at tree tips for each structure (island / non-island)
[…]

### 2024-11-26 extensive_tests/draft-sumstat-vignette.Rmd
Updating sumstats.R script with functions to compute mean correlation within structure type and updating draft summary statistics accordingly

@@
 
-Note that we average across tips as there can be correlations between the mean sd of tips (e.g. under the case of an event affecting the states at two tips).
+Note that we average across tips as there can be correlations between the mean standard deviation of tips (e.g. under the case of an event affecting the states at two tips).
 
@@
 
+```{r}
+# 1 tip / sample / replicate
+sample_n <- 1
+index_islands <- c(1, 3)
+index_nonislands <- c(2, 4) 
+data <- list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
+               c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5), # 25 sites
+               c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1), # 40 sites
+               c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, .5, 0, 0, 0, .5)) # 35 sites
+compute_meanCor_i(index_islands, minN_CpG = 10, shore_length = 5, data, sample_n = 1)
+compute_meanCor_ni(index_nonislands, minN_CpG = 10, shore_length = 5, data, sample_n = 1)
+```
+
+```{r}
+# 2 tip / sample / replicate
+sample_n <- 2
+index_islands <- c(1, 3)
+index_nonislands <- c(2, 4)
+data <- list(
+  # tip 1
+    list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
+         c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5), # 25 sites
+         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1), # 40 sites
+         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, .5, 0, 0, 0, .5)), # 35 sites
+  # tip 2
+    list(c(.5, 0, 0, .5, .5, .5, 0, 0, .5, 1, .5, 0, 0, 0, 0, .5, .5, 1, 1, 1, .5, 0, 0, 0, .5, .5, 1, 1, 1, 1), # 30 sites
+         c(.5, .5, 1, 1, .5, .5, 1, 1, 1, .5, .5, 0, 0, 0, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5), # 25 sites
+         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, .5, .5, 0, 0, .5, 1, 1, 1, 1, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1), # 40 sites
+         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, 1, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, .5, .5, 0, .5, .5, .5, .5, 0, .5)) # 35 sites
+  )
+compute_meanCor_i(index_islands, minN_CpG = 10, shore_length = 5, data, sample_n = 2)
+compute_meanCor_ni(index_nonislands, minN_CpG = 10, shore_length = 5, data, sample_n = 2)
+```
+
 *Correlation measures how much the state of each site says about the next.
 So, when there is no variation (e.g.
-in middle segment state u of a site is followed by state u of the next), there is no correlation.
-Therefore, avoiding the shores leads to lower correlation values.
-Maybe we could test with/out borders.*
+in middle segment state u of a site is followed by state u of the next), there is no correlation.*

### 2024-11-26 extensive_tests/abc_inferences/scripts/compute_sumStats.R

@@ -35,9 +35,9 @@ option_list <- list(
 opt_parser <- OptionParser(option_list = option_list)
 opt <- parse_args(opt_parser)
 
+# List the simulated files
 pad_n <- nchar(as.character(opt[["n-sim"]])) + 1
-
 RData_files <- list.files(pattern = paste0("^abc_dataSIM_\\d{", pad_n, "}\\.RData"))
 
-# Print the input files
-print(RData_files)
+# Check that the number of simulated files is equal to the number of simulations
+if (length(RData_files) != opt[["n-sim"]]) stop("Number of simulated files not equal to number of simulations")

### 2024-11-26 extensive_tests/functions_summaryStats.R

@@ -631,90 +631,7 @@ if (FALSE) {
 }
 
 
-#### #### #### Correlations within structure #### #### ####
 
-
-##### Compute the mean correlation in methylation state in islands
-## index_islands: vector with structural indices for islands
-## data: list with methylation states at tree tips for each structure 
-## data[[tip]][[structure]] when the number of tips is >1, 
-## or data[[structure]] when there is only one tip. Methylation states are vectors
-## minN_CpG: minimum number of central CpGs to use
-## shore_length: number of CpGs at each island side to not use
-## sample_n: number of given tips/samples
-## Note: function only uses islands with a minimum length equal to 2*shore_length+minN_CpG
-compute_meanCor_i <- function(index_islands, minN_CpG, shore_length, data, sample_n){
-  # Restructure data as nested list
-  if (sample_n == 1){
-    data_list <- list()
-    data_list[[1]] <- data
-    data <- data_list
-  }
-  str_counter <- 1
-  cor <- c() 
-  for (tip in 1:sample_n){
-    for (i in index_islands){
-      if(length(data[[tip]][[i]] >= 2*shore_length + minN_CpG)){
-        # Define start and end indices to extract the middle segment
-        start1 <- shore_length + 1
-        start2 <- shore_length + 2
-        end1 <- length(data[[tip]][[i]]) - shore_length
-        end2 <- length(data[[tip]][[i]]) - shore_length + 1
-        # Extract the sequence info
-        segment1 <- data[[tip]][[i]][start1:end1]
-        segment2 <- data[[tip]][[i]][start2:end2]
-        # Compute the correlation of those segments with methylation state variation
-        if (sd(segment1) != 0 && sd(segment2) != 0) {
-          cor[str_counter] <- cor(segment1, segment2)
-          str_counter <- str_counter + 1
-        }
-      }
-      
-    }
-  }
-  return(mean(cor))
-}
-
-##### Compute the mean correlation in methylation state in non-islands
-## index_islands: vector with structural indices for islands
-## data: list with methylation states at tree tips for each structure 
-## data[[tip]][[structure]] when the number of tips is >1, 
-## or data[[structure]] when there is only one tip. Methylation states are vectors
-## minN_CpG: minimum number of central CpGs to use
-## shore_length: number of CpGs at each island side to not use
-## sample_n: number of given tips/samples
[…]

### 2024-12-05 R/sumstats.R
Adding to git script sumstats.R

@@ -1,0 +1,468 @@
+#' Calculate the Mean Frequency of Partially Methylated Sites in Islands
+#'
+#' This function computes the mean frequency of partially methylated sites (with methylation state 0.5) 
+#' for a set of structures identified as islands.
+#'
+#' @param index_islands A vector containing the structural indices for islands.
+#' @param data A list containing methylation states at tree tips for each structure (island / non-island) 
+#' For a single tip: \code{data[[structure]]}. For multiple tips: \code{data[[tip]][[structure]]}.
+#' @param sample_n The number of samples (tips) to process.
+#'
+#' @return A numeric value representing the mean frequency of partially methylated sites in the islands.
+#' @examples
+#' # Example usage:
+#' index_islands <- c(1, 2)
+#' data <- list(
+#'   list(c(0.5, 1, 0.5), c(0, 0.5, 1)), # tip 1
+#'   list(c(0.5, 0.5, 0), c(1, 0.5, 0.5)) # tip 2
+#' )
+#' sample_n <- 2
+#' get_islandMeanFreqP(index_islands, data, sample_n)
+#' 
+#' @export
+get_islandMeanFreqP <- function(index_islands, data, sample_n){
+  # Restructure data as nested list
+  if (sample_n == 1){
+    data_list <- list()
+    data_list[[1]] <- data
+    data <- data_list
+  }
+  mean_island <- c()
+  island_counter <- 1
+  for (i in index_islands){
+    mean_tip <- c()
+    for(s in 1:sample_n){
+      mean_tip[s] <- mean(data[[s]][[i]]==0.5)
+    }
+    mean_island[island_counter] <- mean(mean_tip)
+    island_counter <- island_counter + 1
+  }
+  return(mean(mean_island))
+}
+
+#' Calculate the Mean Frequency of Partially Methylated Sites in Non-Islands
+#'
+#' This function computes the mean frequency of partially methylated sites (with methylation state 0.5) 
+#' for a set of structures identified as non-islands.
+#'
+#' @param index_nonislands A vector containing the structural indices for non-islands.
+#' @param data A list containing methylation states at tree tips for each structure (island / non-island).
+#' For a single tip: \code{data[[structure]]}. For multiple tips: \code{data[[tip]][[structure]]}.
+#' @param sample_n The number of samples (tips) to process.
+#'
[…]

### 2024-12-05 tests/testthat/test-sumstats.R
Adding test-sumstats file

@@ -1,0 +1,882 @@
+test_that("get_MeanFreqP structures with equal length islands and non-islands", {
+  # Two tips and two islands
+  index_islands <- c(1, 2)
+  index_nonislands <- c(1, 2)
+  data <- list(
+    list(c(0, 0, 0), c(0, 0, 0)), # tip 1
+    list(c(0, 0, 0), c(1, 1, 1)) # tip 2
+  )
+  sample_n <- 2
+  expect_equal(get_islandMeanFreqP(index_islands, data, sample_n), 0,
+               info = "incorrect mean in two tips and two islands, freq 0")
+  expect_equal(get_nonislandMeanFreqP(index_nonislands, data, sample_n), 0,
+               info = "incorrect mean in two tips and two non-islands, freq 0")
+  
+  # Two tips and two islands
+  index_islands <- c(1, 2)
+  index_nonislands <- c(1, 2)
+  data <- list(
+    list(c(.5, .5, 0), c(0, 0, .5)), # tip 1
+    list(c(0, .5, .5), c(1, .5, 1)) # tip 2
+  )
+  sample_n <- 2
+  expect_equal(get_islandMeanFreqP(index_islands, data, sample_n), .5,
+               info = "incorrect mean in two tips and two islands, freq .5")
+  expect_equal(get_nonislandMeanFreqP(index_nonislands, data, sample_n), .5,
+               info = "incorrect mean in two tips and two non-islands, freq .5")
+  
+  # Two tips and 1 island
+  index_islands <- c(2) # island index 2
+  index_nonislands <- c(2) # non-island index 2
+  data <- list(
+    list(c(.5, .5, 0), c(0, 0, .5)), # tip 1
+    list(c(0, .5, .5), c(1, .5, 1)) # tip 2
+  )
+  sample_n <- 2
+  expect_equal(get_islandMeanFreqP(index_islands, data, sample_n), 1/3,
+               info = "incorrect mean in two tips and one islands, freq 1/3")
+  expect_equal(get_nonislandMeanFreqP(index_nonislands, data, sample_n), 1/3,
+               info = "incorrect mean in two tips and one non-island, freq 1/3")
+  index_islands <- c(1) # island index 1
+  index_nonislands <- c(1) # island index 1
+  expect_equal(get_islandMeanFreqP(index_islands, data, sample_n), 2/3,
[…]

### 2024-12-05 extensive_tests/draft-sumstat-vignette.Rmd
Updating changes in sumstats

@@
 in middle segment state u of a site is followed by state u of the next), there is no correlation.*
+
+## Comparison of tree tips
+
+### Frequency of Methylation State Changes at Single Sites
+
+We compare the methylation state of single sites at the tree cherries.
+A cherry is a pair of leaf nodes (also called tips or terminal nodes) in a phylogenetic tree that share a direct common ancestor.
+In other words, if two leaves are connected to the same internal node and no other leaves are connected to that internal node, they form a cherry.
+
+get_FChange_cherryData (uses find_cherries and cherry_diffs)

### 2024-12-05 R/sumstats.R

@@ -1,3 +1,5 @@
+#### #### #### Frequency of methylation states per structure type #### #### ####
+
 #' Calculate the Mean Frequency of Partially Methylated Sites in Islands
 #'
 #' This function computes the mean frequency of partially methylated sites (with methylation state 0.5) 
@@ -404,6 +406,8 @@ compute_meanCor_i <- function(index_islands, minN_CpG, shore_length, data, sampl
   return(mean(cor))
 }
 
+#### #### #### Mean correlation per structure type #### #### ####
+
 #' Compute the Mean Correlation of Methylation State in Non-islands
 #'
 #' This function calculates the mean correlation of methylation states within 
@@ -465,4 +469,96 @@ compute_meanCor_ni <- function(index_nonislands, minN_CpG, shore_length, data, s
   return(mean(cor))
 }
 
+#### #### #### Tree cherries comparisons #### #### ####
 
+find_cherries <- function(tree) {
+  ## tree must be an ape tree whose tip labels are numbers.
+  ## returns a list of vectors (a, b, x), where  (a, b) are a cherry of tree,
+  ## that is a pair of leaves that are sisters in the tree,
+  ## and x is their distance, that is, sum of the branch lengths between them 
+  outlist <- list()
+  d <- cophenetic.phylo(tree)
+  n <- nrow(d)
+  for(i in 1:n) {
+    j <- which(d[i,]==min(d[i, -i]))
+    if(length(j)==1) {
+      k <- which(d[j,]==min(d[j, -j]))
+      if(length(k)==1) {
+        if(k==i & i<j) {
+          ## outlist[[length(outlist)+1]] <- c(as.integer(rownames(d)[i]), as.integer(colnames(d)[j]), d[i,j])
+          outlist[[length(outlist)+1]] <- c(i, j, d[i,j])
+        }
+      }
+    }
+  }
+  outlist
+}
+
+
+
+## TODO: come back here and check whether the tree needs to
+## have the tips with numeric labels or that can be internally controlled. 
+## For that check previous (old) function 
+## tree needs to be an object of ape's phylo class or a character string in
+## parenthetic format known as the Newick or New Hampshire format
+get_cherryDist <- function(tree, testing = FALSE){
+  if(class(tree) != "phylo") tree <- ape::read.tree(text = tree)
[…]

### 2024-12-05 tests/testthat/test-sumstats.R

@@ -879,4 +879,34 @@ test_that("meanCor", {
                info = "MeanCor_ni returns non-correct value (shore 10 2 tips non-equal sizes)")
 })
 
+test_that("get_cherryDist", {
+  expect_error(get_cherryDist(testing = T),
+               info = "function fails to throw an error when no tree is given")
+  newick_tree <- "((1:1,2:1):1,3:2);"
+  expect_true(class(get_cherryDist(newick_tree, testing = T)$tree) == "phylo",
+              info = "fails to convert newick tree into ape's class phylo")
+  
+  # Test case single cherry
+  expect_true(all(get_cherryDist(newick_tree)[[1]] == c(1,2,2)),
+              info = "incorrect output for newick tree with single cherry")
+  ape_tree <- ape::read.tree(text = newick_tree)
+  expect_true(all(get_cherryDist(ape_tree)[[1]] == c(1,2,2)),
+              info = "incorrect output for ape tree with single cherry")
+  
+  # Test case 2 cherries
+  newick_tree <- "((1:1.5,2:1.5):2,(3:2,4:2):1.5);"
+  ape_tree <- ape::read.tree(text = newick_tree)
+  expect_true(all(get_cherryDist(newick_tree)[[1]] == c(1,2,3)),
+              info = "incorrect output for newick tree with two cherries (1st cherry)")
+  expect_true(all(get_cherryDist(newick_tree)[[2]] == c(3,4,4)),
+              info = "incorrect output for newick tree with two cherries (2nd cherry)")
+  expect_true(all(get_cherryDist(ape_tree)[[1]] == c(1,2,3)),
+              info = "incorrect output for ape tree with two cherries (1st cherry)")
+  expect_true(all(get_cherryDist(ape_tree)[[2]] == c(3,4,4)),
+              info = "incorrect output for ape tree with two cherries (2nd cherry)")
+})
 
+
+if(FALSE){
+  debug(get_cherryDist)
+}

### 2024-12-05 tests/testthat/test-sumstats.R
Updating changes in sumstats

@@ -907,6 +907,12 @@ test_that("get_cherryDist", {
 })
 
 
-if(FALSE){
-  debug(get_cherryDist)
-}
+test_that("get_cherryMethDiff", {
+  newick_tree <- "((1:1,2:1):1,3:2);"
+  data <- list(
+    list(rep(1,10), rep(0,10), rep(1,10)),
+    list(rep(1,10), rep(0.5,10), rep(0,10)),
+    list(rep(1,10), rep(0.5,10), rep(0,10)))
+  cherries <- get_cherryDist(newick_tree)
+  get_cherryMethDiff(cherries, data)
+})

### 2024-12-06 R/sumstats.R
Updating changes

@@ -501,7 +501,8 @@ find_cherries <- function(tree) {
 ## For that check previous (old) function 
 ## tree needs to be an object of ape's phylo class or a character string in
 ## parenthetic format known as the Newick or New Hampshire format
-get_cherryDist <- function(tree, testing = FALSE){
+get_cherryDist <- function(tree, sample_n, testing = FALSE){
+  if(sample_n < 2){stop("Minimum number of tips/samples needed: 2")}
   if(class(tree) != "phylo") tree <- ape::read.tree(text = tree)
   # compute the pairwise distances between the tips from a phylogenetic tree
   dist <- ape::cophenetic.phylo(tree)
@@ -537,28 +538,74 @@ get_cherryDist <- function(tree, testing = FALSE){
 
 
 #undebug(get_cherryDist)
-
-get_cherryMethDiff <- function(cherries, data) {
+## TODO: Assumes that each str has the same number of sites at the different tips
+## Think about how to control for pop cases in which there may be insertions/deletions
+## or whether this should be controlled by the user
+count_cherryMethDiff <- function(cherries, data) {
   ## returns a data frame with one line for each cherry, containing the branch distance between the
   ## cherries and for each structure two numbers: that of the number of sites that are m in one in
   ## u in the other tip and the number of sites that are p in one and m or u in the other tip.
-  n <- length(data)[[1]] # set the number of structures
+
+  str_n <- length(data[[1]]) # set the number of structures
   df <- data.frame(tips=character(), dist=numeric()) # start df
-  for(j in 1:n) {
+  for(str in 1:str_n) {
     # e.g. df 3 structures: tips dist 1_f 1_h 2_f 2_h 3_f 3_h
     df <- cbind(df, integer(), integer())
-    names(df)[c(ncol(df)-1, ncol(df))] <- paste0(j,"_",c("f", "h"))
+    names(df)[c(ncol(df)-1, ncol(df))] <- paste0(str,"_",c("f", "h"))
   }
   for(i in 1:length(cherries)) {
     ch <- cherries[[i]]
     df[i, 1] <- paste0(ch[1],"-",ch[2]) # add tips info
     df[i, 2] <- ch[3] # add dist between tips
-    for(j in 1:n) {
[…]

### 2024-12-06 tests/testthat/test-sumstats.R

@@ -880,39 +880,121 @@ test_that("meanCor", {
 })
 
 test_that("get_cherryDist", {
-  expect_error(get_cherryDist(testing = T),
-               info = "function fails to throw an error when no tree is given")
+  # Set tree
   newick_tree <- "((1:1,2:1):1,3:2);"
-  expect_true(class(get_cherryDist(newick_tree, testing = T)$tree) == "phylo",
+  
+  # Test wrong input
+  expect_error(get_cherryDist(),
+               info = "function fails to throw an error when no tree/sample_n is given")
+  expect_error(get_cherryDist(sample_n = 2),
+               info = "function fails to throw an error when no tree is given")
+  expect_error(get_cherryDist(tree = newick_tree),
+               info = "function fails to throw an error when no sample_n is given")
+  
+  
+  # Test error when tree has only one tip
+  newick_tree <- "(1:1);"
+  expect_error(get_cherryDist(tree = newick_tree, sample_n = 1),
+               info = "fails to throw an error when number of tips is < 2")
+  
+  # Test conversion from newick tree to ape tree
+  newick_tree <- "((1:1,2:1):1,3:2);"
+  expect_true(class(get_cherryDist(newick_tree, sample_n = 3, testing = T)$tree) == "phylo",
               info = "fails to convert newick tree into ape's class phylo")
   
   # Test case single cherry
-  expect_true(all(get_cherryDist(newick_tree)[[1]] == c(1,2,2)),
+  expect_true(all(get_cherryDist(newick_tree, sample_n = 3)[[1]] == c(1,2,2)),
               info = "incorrect output for newick tree with single cherry")
   ape_tree <- ape::read.tree(text = newick_tree)
-  expect_true(all(get_cherryDist(ape_tree)[[1]] == c(1,2,2)),
+  expect_true(all(get_cherryDist(ape_tree, sample_n = 3)[[1]] == c(1,2,2)),
               info = "incorrect output for ape tree with single cherry")
   
   # Test case 2 cherries
   newick_tree <- "((1:1.5,2:1.5):2,(3:2,4:2):1.5);"
   ape_tree <- ape::read.tree(text = newick_tree)
[…]

### 2024-12-09 R/sumstats.R
Updating sumstats and test sumstats

@@ -541,7 +541,7 @@ get_cherryDist <- function(tree, sample_n, testing = FALSE){
 ## TODO: Assumes that each str has the same number of sites at the different tips
 ## Think about how to control for pop cases in which there may be insertions/deletions
 ## or whether this should be controlled by the user
-count_cherryMethDiff <- function(cherries, data) {
+countSites_cherryMethDiff <- function(cherries, data) {
   ## returns a data frame with one line for each cherry, containing the branch distance between the
   ## cherries and for each structure two numbers: that of the number of sites that are m in one in
   ## u in the other tip and the number of sites that are p in one and m or u in the other tip.
@@ -567,45 +567,74 @@ count_cherryMethDiff <- function(cherries, data) {
   df
 }
 
-#undebug(count_cherryMethDiff)
+#undebug(countSites_cherryMethDiff)
+#undebug(get_cherryDist)
 
 ##TODO: n_CpG not given but calculated. Each structure is assumed to have the same number of sites at all tips
-get_FChange_cherryData <- function(tree, data, sample_n, testing = FALSE){
+get_siteFChange_cherry_perMethDiffType <- function(tree, data, sample_n){
   if(sample_n < 2){stop("Minimum number of tips/samples needed: 2")}
-  cherries <- get_cherryDist(tree, sample_n)
-  FChange_cherryData <- count_cherryMethDiff(cherries = cherries, data = data)
+  cherries <- get_cherryDist(tree = tree, sample_n = sample_n)
+  siteFChange_cherry_perMethDiffType <- countSites_cherryMethDiff(cherries = cherries, data = data)
   str_n <- length(data[[1]]) # set the number of structures
   sites_n <- numeric(length = str_n)
   for(str in 1:str_n) sites_n[str] <- length(data[[1]][[str]])
   # Each value is a count number of full methylation changes _f or changes in one strand _h.
   # Each structure has a number of CpGs. Then, if we wanna count the proportion
   # of sites, we divide each value by the number of CpGs. 
-  # Explicitly match sites_n to the columns of FChange_cherryData
[…]

### 2024-12-09 extensive_tests/compute_summaryStats.R

@@ -219,7 +219,6 @@ for(sim in 1:n_sim){
   if ( "all" %in% stats_to_compute || "FChangeCherry_ni" %in% stats_to_compute) {
     meanFChangeCherry_fit_ni <- tryCatch({
       fit_meanFChange_cherry_ni(FChange_cherryData = FChange_cherryData, index_nonislands = index_nonislands)
-    }, error = function(e) {
       error_msg <- paste("Simulation", sim, "- Error in fit_meanFChange_cherry_ni:", e$message)
       error_log[[length(error_log) + 1]] <- error_msg
       list(rate = NA, scale = NA)

### 2024-12-09 tests/testthat/test-sumstats.R

@@ -924,7 +924,7 @@ test_that("get_cherryDist", {
 })
 
 
-test_that("count_cherryMethDiff", {
+test_that("countSites_cherryMethDiff", {
   # Test case one cherry
   newick_tree <- "((1:1,2:1):1,3:2);"
   data <- list(
@@ -932,7 +932,7 @@ test_that("count_cherryMethDiff", {
     list(rep(1,10), rep(0.5,10), rep(0,10)),
     list(rep(1,10), rep(0.5,10), rep(0,10)))
   cherries <- get_cherryDist(newick_tree, sample_n = 3)
-  count_cherryMethDiff(cherries, data)
+  countSites_cherryMethDiff(cherries, data)
   ##TODO: test content output:
   #tips dist 1_f 1_h 2_f 2_h 3_f 3_h
   #1  1-2    2   0   0   0  10  10   0
@@ -945,7 +945,7 @@ test_that("count_cherryMethDiff", {
     list(rep(1,10), rep(0.5,10), rep(0,10)),
     list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1, rep(0.5, 5)), c(0.5, 1, rep(0, 8))))
   cherries <- get_cherryDist(newick_tree, sample_n = 4)
-  count_cherryMethDiff(cherries, data)
+  countSites_cherryMethDiff(cherries, data)
   ##TODO: test content output:
   #  tips dist 1_f 1_h 2_f 2_h 3_f 3_h
   #1  1-2    3   0   0   0  10  10   0
@@ -953,12 +953,12 @@ test_that("count_cherryMethDiff", {
   
 })
 
-test_that("get_FChange_cherryData", {
+test_that("get_siteFChange_cherry_perMethDiffType", {
   
   # Test error when tree has only one tip
   newick_tree <- "(1:1);"
   data <- list(rep(1,10), rep(0,10), rep(1,10))
-  expect_error(get_FChange_cherryData(tree = newick_tree, data = data, sample_n = 1),
+  expect_error(get_siteFChange_cherry_perMethDiffType(tree = newick_tree, data = data, sample_n = 1),
                info = "fails to throw an error when number of tips is < 2")
   
   # Test error when arguments are missing
@@ -967,11 +967,11 @@ test_that("get_FChange_cherryData", {
     list(rep(1,10), rep(0,10), rep(1,10)),
[…]

### 2024-12-09 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
Adding files for test CFTP convergence workflow for reproducibility

--- 2024-11-25 extensive_tests/draft-abc-inferences-vignette.Rmd
@@
 ---
-title: "ABC inferences"
+title: "CFTP Convergence"
 output: 
@@
 vignette: >
-  %\VignetteIndexEntry{Introduction to MethEvolSIM}
+  %\VignetteIndexEntry{Test for CFTP Convergence}
   %\VignetteEngine{knitr::rmarkdown}
@@
 
-- R packages: optparse (add to dependencies), parallel
+- R packages:
 
@@
 
-1. Save as simulation design the genomic structure, tree(s) -- GIVEN -- and sampled parameter combinations.
+1. Bla
 
-2. Run the simulations. TODO: Before submitting load package without devtools
+### Scripts
 
-3. Compute the selected summary statistics. TODO: When added sumStats functions to package, stop sourcing script.
-
-4. Assess ABC accuracy using cross-validation.
-
-5. Infer parameters from empirical DNA methylation data.
+- `run_sim.R`. TODO: Update way in which the package is loaded before submitting.

### 2024-12-09 extensive_tests/abc_inferences/scripts/run_sim.R

@@ -1,5 +1,6 @@
 library(optparse)
 library(parallel)
+## TODO: Modify how the package is loaded before submitting
 library(devtools); load_all()
 
 # Define command-line options
@@ -9,7 +10,9 @@ option_list <- list(
   make_option("--n-cores", type = "integer", default = NULL,
               help = "Number of cores", metavar = "integer"),
   make_option("--n-tips", type = "integer", default = NULL,
-              help = "Number of tree tips", metavar = "integer")
+              help = "Number of tree tips", metavar = "integer"),
+  make_option("--CFTP", type = "logical", default = FALSE, 
+    metavar = "TRUE/FALSE", help = "Use CFTP algorithm at tree root [default: %default]")
 )
 
 # Parse the arguments
@@ -53,7 +56,7 @@ run_simulation_subset <- function(start_idx, end_idx) {
     set.seed(sim)
     
     # Simulate data
-    output <- simulate_evolData(infoStr = spatial_str, tree = scaled_trees[[sim]], params = sampled_params[sim,])
+    output <- simulate_evolData(infoStr = spatial_str, tree = scaled_trees[[sim]], params = sampled_params[sim,], CFTP = opt[["CFTP"]])
     
     # Extract methylation data at the tree tips
     data <- list()

### 2024-12-09 extensive_tests/CFTP_convergence/run_sim.R
Updating changes

@@ -58,6 +58,8 @@ simul_CFTP_branch <- function(custom_params, index_params, b_length, start, end,
   
   if(start == 1){
     
+    ##TODO: Maybe this alternative should be better: simulate_evolData(infoStr = spatial_str, tree = tree, CFTP = TRUE) with tree like "(1:0);"
+    
     print("Generating combiStructureGenerator instance")
     combi <- combiStructureGenerator$new(infoStr = spatial_str, params = custom_params)
     

### 2024-12-11 extensive_tests/CFTP_convergence/scripts/run_sim.R
Adding spatial structure RData file for CFTP convergence test

@@ -18,6 +18,8 @@ option_list <- list(
               help = "number of data replicates to simulate", metavar = "integer")
 )
 
+
+
 # Parse the arguments
 
 opt_parser <- OptionParser(option_list = option_list)
@@ -42,132 +44,114 @@ n_sim <- nrow(sampled_params)
 
 # Generate pad_n based on the number of digits in the number or simulations and the number of replicates
 # To save the files with padded numbers so that are later listed in order with list.files()
-params_pad_n <- nchar(as.character(n_sim)) + 1
-rep_pad_n <- nchar(as.character(opt[["replicate-n"]])) + 1
+params_pad_n <- nchar(as.character(n_sim))
 
 
-print(paste("params_pad_n", params_pad_n))
-print(paste("rep_pad_n", rep_pad_n))
 
-##TODO: Modify paramComb-n, name-pattern: CFTP_testConvergence, test-n
 
-if(FALSE){
-  simul_CFTP_branch <- function(custom_params, index_params, b_length, start, end, out_digit_n, spatial_str, test_n, replicate_n){
+simul_CFTP_branch <- function(custom_params, params_pad_n, index_params, b_length, start, end, step_pad_n, spatial_str, replicate_n){
+  
+  # Set the name for the output file with the padded parameter index
+  padded_index_params <- formatC(index_params, width = params_pad_n, format = "d", flag = "0")
+  rep_pad_n <- nchar(as.character(replicate_n))
+  padded_replicate_n <- formatC(replicate_n, width = rep_pad_n, format = "d", flag = "0")
+  out_file <- paste0("CFTP_testConvergence_paramsID_", padded_index_params, "_rep_", replicate_n, ".out")
+  
+  # Redirect both the stout and stderr to the same file
+  sink(out_file, type = c("output", "message"), append = TRUE)
+  print(paste("Running CFTP_testConvergence. paramsID:", padded_index_params, ". Replicate:", padded_replicate_n))
+  print("Given customized parameter values:")
+  print(custom_params)
+  
+  # Set seed
+  set.seed(index_params)
+  
+  if(start == 1){
     
-    # Set the name for the output file with the padded parameter index
-    padded_index_params <- formatC(index_params, width = params_pad_n, format = "d", flag = "0")
+    print("Generating combiStructureGenerator instance")
+    combi <- combiStructureGenerator$new(infoStr = spatial_str, params = custom_params)
+    
+    ## Save initial instance state and methylation data ##
+    data <- list()
[…]

### 2024-12-12 extensive_tests/draft-abc-inferences-vignette.Rmd
Updating last changes
Adding files and updates on draft abc inferences

@@
 vignette: >
-  %\VignetteIndexEntry{Introduction to MethEvolSIM}
+  %\VignetteIndexEntry{ABC Inferences with MethEvolSIM}
   %\VignetteEngine{knitr::rmarkdown}
@@
 
-- R packages: optparse (add to dependencies), parallel
+- R packages: optparse (add to dependencies), parallel, ape
 
@@
 - An .RData file with a dataframe specifying the distribution of methylation sites in the genomic region.
+
+- An .RData file with a tree in ape format (TODO: modify this and check whether scrm is also producing trees in the same format)
 
@@
 5. Infer parameters from empirical DNA methylation data.
+
+## Selection of the genomic region
+
+1.
+[Download](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000238955.4/) the *M.
+zebra* reference genome (GCF_000238955.4_M_zebra_UMD2a_genomic.fa) from the NCBI Assembly database.
+This genome is part of the M.
+zebra UMD2a assembly, which includes associated data such as gene annotations and other resources.
+
+2. Transform CGI annotation (add citation) in txt format to BED
+
+```{bash, eval = FALSE}
+awk 'NR > 1 {print $1 "\t" $2-1 "\t" $3}' CGI-Mzebra.txt > CGI-Mzebra.bed
+```

### 2024-12-15 extensive_tests/abc_inferences/draft-abc-inferences-vignette.Rmd
Adding script for rule to get CpG counts at island and non island structures for a reference genome and updating draft abc inferences vignette

---
title: "ABC inferences"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{ABC Inferences with MethEvolSIM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE}
## TODO: Update 
library(devtools)
load_all()
```

# Selection of genomic region(s)

-Requires: CGI annotation file (.txt file obtained with makeCGI) **Add reference**

```{r}
# Load data
file_path <- "CGI-Mzebra.txt" # Adjust the path if necessary
cgi_data <- read.table(file_path, header = TRUE, sep = "\t")

# View the first lines
head(cgi_data)
# View the last lines
tail(cgi_data)
```

The prefixes `NC_` and `NW_` in chromosome or genomic sequence identifiers come from the RefSeq database and indicate different types of assembly or annotation status for genomic sequences:

- `NC_` refers to non-redundant complete genomic molecules, which typically represent well-characterized, curated, and assembled sequences.

- `NW_` refers to scaffolds or unplaced sequences that are part of the genome assembly but cannot yet be confidently assigned to a specific chromosome or region.

```{r}
# Count the number of CGIs per chromosome
cgi_count <- table(cgi_data$chr)

# View results grouped by type of chromosome assembly

# Split into groups based on chromosome prefix
groups <- ifelse(grepl("^NC_", names(cgi_count)), "NC", "NW")

# Create a dataframe
cgi_data <- data.frame(
  group = groups,
  count = as.numeric(cgi_count)
)

# Plot boxplots
boxplot(count ~ group, data = cgi_data,
        col = c("skyblue", "salmon"),
        main = "Distribution of Counts for NC_ and NW_ Chromosomes",
        xlab = "Chromosome Type",
        ylab = "Counts")

```

- *Smad4a*: LOC101481185. Fixed but not associated with differential expression.

Genes reported as fixed DMR with significant transcriptional downregulation in benthic livers and associated with fixed benthic-specific hypermethylated levels at their promoters, suggesting retained epigenetic divergence associated with different diets.

- *Dpp4*: Part of the insulin methabolic pathway. Location according to NCBI annotation release 104 (LOC101483164): NC_036801.1 (29502362..29517382)

- *Ces2*: Liver carboxylesterase 2, part of the fatty acid methabolic pathway. Location according to NCBI annotation release 104 (LOC101473569):  	NW_020192811.1 (49164..53741, complement)

```{r}
head(cgi_data[cgi_data$chr == "NC_036801.1",])
```

## How to generate the structural distribution .RData file from the reference genome and the CGI annotation

Requires the following packages (already included in environment.yaml):
```{r, eval = FALSE}
install.packages("BiocManager")
BiocManager::install("Biostrings")
```

```{r}
load("genomic_dist/CpG_count.RData")

```
29502362..29517382

```{r}
# Filter the data for the given chromosome and region
dpp4_chr <- "NC_036801.1"
filtered_data <- CpG_count[CpG_count$chr == dpp4_chr & CpG_count$start > 2.8e7 & CpG_count$end < 3e7, ]

# Set up the plot area
plot(filtered_data$start, filtered_data$CpG_count, 
     type = "n", # No points, just the axis
     xlab = "Position (bp)", ylab = "CpG Count", 
     main = "CpG Count by Island/Non-Island Regions", 
     xlim = c(min(filtered_data$start), max(filtered_data$end)), 
     ylim = c(0, max(filtered_data$CpG_count) * 1.1))

# Add bars with color based on 'str' (island vs non-island)
for (i in 1:nrow(filtered_data)) {
  # Calculate the width of each region (difference between start and end)
  region_width <- filtered_data$end[i] - filtered_data$start[i]

  if (filtered_data$str[i] == 0) {
    # Non-island (blue)
    rect(filtered_data$start[i], 0, filtered_data$end[i], 
         filtered_data$CpG_count[i], col = "blue", border = "blue")
  } else {
    # Island (red)
    rect(filtered_data$start[i], 0, filtered_data$end[i], 
         filtered_data$CpG_count[i], col = "red", border = "red")
  }
}

# Add a legend
legend("topright", legend = c("Non-Island", "Island"), 
       fill = c("blue", "red"), title = "Island Type")

```

```{r}

# Filter the data for the given chromosome and region
dpp4_chr <- "NC_036801.1"
dpp4_start <- 29502362
dpp4_end <- 29517382
filtered_data <- CpG_count[CpG_count$chr == dpp4_chr & CpG_count$start > dpp4_start - 5e5 & CpG_count$end < dpp4_end + 5e5, ]

# Calculate the CpG density for each region (CpG count / region length)
filtered_data$CpG_density <- filtered_data$CpG_count / (filtered_data$end - filtered_data$start)

# Plot CpG density vs. position (using 'start' as x-axis)

# Set up the plot area with no points initially
plot(filtered_data$start, filtered_data$CpG_density, 
     type = "n", # No points, just the axis
     xlab = "Position (bp)", ylab = "CpG Density", 
     main = "CpG Density by Island/Non-Island Regions", 
     xlim = c(min(filtered_data$start), max(filtered_data$end)), 
     ylim = c(0, max(filtered_data$CpG_density) * 1.1))

# Add points colored by 'str' (0 for non-island, 1 for island)
points(filtered_data$start, filtered_data$CpG_density, 
       col = ifelse(filtered_data$str == 1, "red", "blue"), 
       pch = 16)  # pch=16 gives filled circles

# Add vertical lines showing where the gene is
abline(v = dpp4_start)
abline(v = dpp4_end)

# Add a legend
legend("topright", legend = c("Non-Island", "Island"), 
       fill = c("blue", "red"), title = "Type")

```

If using this region the number of CpG sites to simulate would be:

```{r}
sum(filtered_data$CpG_count)
```

with number of non-island (0) and island (1) structures:

```{r}
# Count the number of occurrences of each factor level in the 'str' column
table(filtered_data$str)
```

# Worflow for simulation-based inferences

This workflow uses [Snakemake](https://snakemake.readthedocs.io) to manage the steps for simulating DNA methylation patterns, assessing ABC accuracy, and inferring model parameters.

## Requirements

- Snakemake installed (`conda create -c conda-forge -c bioconda -n snakemake snakemake`)

- R packages: optparse (add to dependencies), parallel, ape

- Access to the provided `Snakefile` and associated scripts.

- An .RData file with a dataframe specifying the distribution of methylation sites in the genomic region.

- An .RData file with a tree in ape format (TODO: modify this and check whether scrm is also producing trees in the same format)

## Steps in the Workflow

1. Save as simulation design the genomic structure, tree(s) -- GIVEN -- and sampled parameter combinations.

2. Run the simulations. TODO: Before submitting load package without devtools

3. Compute the selected summary statistics. TODO: When added sumStats functions to package, stop sourcing script.

4. Assess ABC accuracy using cross-validation.

5. Infer parameters from empirical DNA methylation data.

### 2024-12-15 extensive_tests/abc_inferences/scripts/get_CpG_count.R

@@ -1,0 +1,129 @@
+library(optparse)
+library(Biostrings) # For reading the genome sequence
+
+option_list <- list(
+  make_option(c("--CGI-annotation"), type = "character", default = NULL,
+              help = "Path to the CGI annotation design file (.txt)", metavar = "character"),
+  make_option(c("--reference-genome"), type = "character", default = NULL,
+              help = "Path to the reference genome file (.fa)", metavar = "character"),
+  make_option(c("--output"), type = "character", default = NULL,
+              help = "Output path", metavar = "character")
+)
+
+# Parse the arguments
+
+opt_parser <- OptionParser(option_list = option_list)
+opt <- parse_args(opt_parser)
+
+# Get the names of required options (you can update this based on what's mandatory)
+required_options <- c("CGI-annotation", "reference-genome", "output")
+
+# Check that all required options are not NULL
+missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
+
+
+# Step 1: Load Reference Genome and CGI Annotation
+cgi_data <- read.table(opt[["CGI-annotation"]], header = TRUE, sep = "\t")
+genome <- readDNAStringSet(opt[["reference-genome"]], format = "fasta")
+
+# Step 2: Define a function to compute CpG counts in a region
+count_CpG <- function(sequence) {
+  Biostrings::countPattern("CG", sequence)
+}
+
+# Step 3: Generate a Count Data Frame
+
+# Initialize the dataframe
+CpG_count <- data.frame(
+  chr = character(),
+  str = factor(),
+  start = numeric(),
+  end = numeric(),
+  CpG_count = numeric(),
+  metadata = character(),
+  stringsAsFactors = FALSE
+)
+
+# Iterate over the chromosomes in the genome
+for (chrom in names(genome)) {
+  # Get chromosome sequence
+  chr_seq <- genome[[chrom]]
+  chr_len <- length(chr_seq)
+  
+  # Filter CGI data for this chromosome
+  cleaned_chrom <- strsplit(chrom, " ")[[1]][1]
+  cgi_chr <- cgi_data[cgi_data$chr == cleaned_chrom, ]
+  
+  # Check if there are any CGIs for this chromosome
+  if (nrow(cgi_chr) == 0) {
+    next  # Skip to the next chromosome if no data is available for this one
+  }
+  
+  # Extract metadata
+  metadata <- paste(strsplit(chrom, " ")[[1]][-1], collapse = " ")
+  
[…]

### 2024-12-16 extensive_tests/abc_inferences/draft-abc-inferences-vignette.Rmd
Adding script to get spatial structure

@@
 # Load data
-file_path <- "CGI-Mzebra.txt" # Adjust the path if necessary
+file_path <- "genomic_dist/CGI-Mzebra.txt" # Adjust the path if necessary
 cgi_data <- read.table(file_path, header = TRUE, sep = "\t")
@@
 
+## Fixed DMR found in paper: Associated genes
+
+Add reference
+
 - *Smad4a*: LOC101481185. Fixed but not associated with differential expression.
@@
 
+Lets explore *Dpp4*, as it is associated with differential gene expression and it is located in a better characterized genomic region:
+
 ```{r}
-head(cgi_data[cgi_data$chr == "NC_036801.1",])
+dpp4_chr <- "NC_036801.1"
+dpp4_start <- 29502362
+dpp4_end <- 29517382
 ```
 
-## How to generate the structural distribution .RData file from the reference genome and the CGI annotation
+## Explore the CpG counts in islands and non islands generated from the reference genome and the CGI annotation
 
@@
 load("genomic_dist/CpG_count.RData")
-
 ```
-29502362..29517382
 
@@
 # Filter the data for the given chromosome and region
-dpp4_chr <- "NC_036801.1"
-filtered_data <- CpG_count[CpG_count$chr == dpp4_chr & CpG_count$start > 2.8e7 & CpG_count$end < 3e7, ]
+filtered_data <- CpG_count[CpG_count$chr == dpp4_chr & CpG_count$start > dpp4_start - 5e5 & CpG_count$start < dpp4_end + 5e5, ]
 
@@
 
-  if (filtered_data$str[i] == 0) {
+  if (filtered_data$str[i] == "Non-island") {
     # Non-island (blue)
@@
 
+# Add vertical lines showing where the gene is
+abline(v = dpp4_start)
+abline(v = dpp4_end)
+
 # Add a legend
@@
 # Filter the data for the given chromosome and region
-dpp4_chr <- "NC_036801.1"
-dpp4_start <- 29502362
-dpp4_end <- 29517382
 filtered_data <- CpG_count[CpG_count$chr == dpp4_chr & CpG_count$start > dpp4_start - 5e5 & CpG_count$end < dpp4_end + 5e5, ]

### 2024-12-16 extensive_tests/abc_inferences/scripts/get_spatial_str.R

@@ -1,0 +1,42 @@
+library(optparse)
+
+option_list <- list(
+  make_option(c("--input"), type = "character", default = NULL,
+              help = "Input CpG_count.RData path", metavar = "character"),
+  make_option(c("--name-selected-region"), type = "character", default = NULL,
+              help = "Name of selected genomic region", metavar = "character"),
+  make_option(c("--chr"), type = "character", default = NULL,
+              help = "Chromosome ID of selected region", metavar = "character"),
+  make_option(c("--start"), type = "integer", default = NULL,
+              help = "Start index of selected genomic region", metavar = "integer"),
+  make_option(c("--end"), type = "integer", default = NULL,
+              help = "End index of selected genomic region", metavar = "integer"),
+  make_option(c("--side-bp"), type = "integer", default = NULL,
+              help = "Number of bp to select structures starting after start - side-bp and ending before end + side-bp", metavar = "integer")
+)
+
+# Parse the arguments
+
+opt_parser <- OptionParser(option_list = option_list)
+opt <- parse_args(opt_parser)
+
+# Get the names of required options (you can update this based on what's mandatory)
+required_options <- c("input", "name-selected-region", "chr", "start", "end", "side-bp")
+
+# Check that all required options are not NULL
+missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
+
+if (length(missing_options) > 0) {
+  print_help(opt_parser)
+  stop(paste("The following arguments need to be provided:", paste(missing_options, collapse = ", ")))
+}
+
+# Load CpG counts
+load(opt[["input"]])
+
+# Filter the data for the given chromosome and region
+spatial_str <- CpG_count[CpG_count$chr == opt[["chr"]] & CpG_count$start >= opt[["start"]] - opt[["side-bp"]] & CpG_count$start <= opt[["end"]] + opt[["side-bp"]], ]
+
+# Save filtered data
+output_name <- paste0(paste(opt[["chr"]], opt[["start"]] - opt[["side-bp"]], opt[["end"]] + opt[["side-bp"]], opt[["name-selected-region"]], sep = "_"), ".RData")
+save(spatial_str, file = output_name)

### 2024-12-16 extensive_tests/abc_inferences/draft-abc-inferences-vignette.Rmd
Updating environment.yaml

@@
 
+# Workflow dependencies
+
+Are listed under the file `environment.yaml`. Before running the workflow, a conda environment can be created with:
+
+```{bash, eval = FALSE}
+conda env create -n abc_inferences -f environment.yaml
+```
+
 # Selection of genomic region(s)
 
+Add info on how to download the reference genome from progress report.
+
 -Requires: CGI annotation file (.txt file obtained with makeCGI) **Add reference**
+
+Both under subdirectory genomic_dist
 
@@
 
-Requires the following packages (already included in environment.yaml):
-```{r, eval = FALSE}
-install.packages("BiocManager")
-BiocManager::install("Biostrings")
-```
-
 ```{r}
@@
 # Filter the data for the given chromosome and region
-filtered_data <- CpG_count[CpG_count$chr == dpp4_chr & CpG_count$start > dpp4_start - 5e5 & CpG_count$start < dpp4_end + 5e5, ]
+filtered_data <- CpG_count[CpG_count$chr == dpp4_chr & CpG_count$start >= dpp4_start - 5e5 & CpG_count$start <= dpp4_end + 5e5, ]
 
@@
 ```
+
+X-axis corresponds to the positions in chromosome `r dpp4_chr`
+dpp4_start
+dpp4_end
 

### 2024-12-16 extensive_tests/abc_inferences/draft-abc-inferences-vignette.Rmd
Updating script get_spatial_str of abc_inferences workflow

@@
     toc: true
+bibliography: references.bib
 vignette: >
@@
   collapse = TRUE,
-  comment = "#>"
+  comment = "#>",
+  fig.width = 7,  # Set the figure width to match the text width
+  fig.height = 5, # Set the figure height
+  out.width = "100%"  # Ensure it takes up 100% of the available width in HTML
 )
@@
 
-Add info on how to download the reference genome from progress report.
+-Requires: CGI annotation file (`.txt` file) obtained with makeCGI v1.3.4 [@hiddenMarkov]
 
--Requires: CGI annotation file (.txt file obtained with makeCGI) **Add reference**
+- The genome assembly file: Obtain the FASTA file ([GCF_000238955.4_M_zebra_UMD2a_genomic.fa](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000238955.4/)) from NCBI (file source *RefSeq*, file type *Genome Sequences (FASTA)*)
 
@@
 
-X-axis corresponds to the positions in chromosome `r dpp4_chr`
-dpp4_start
-dpp4_end
+Genomic region close to gene *dpp4*.
+X-axis corresponds to the positions in chromosome `r dpp4_chr`.
+Y-axis corresponds to CpG counts at each genomic region.
+Each genomic region is represented by a bar starting and ending in the corresponding chromosome position.
+Vertical black lines represent start and end of gene *dpp4*.
 
@@
 
-If using this region the number of CpG sites to simulate would be:
+Genomic region close to gene *dpp4*.
+X-axis corresponds to the positions in chromosome `r dpp4_chr`.
+Y-axis corresponds to CpG density at each genomic region.
+Each genomic region is represented by a dot at its corresponding genome start position.
+Vertical black lines represent start and end of gene *dpp4*.
+
+When using this region the number of CpG sites to simulate would be:
 
@@
 
-with number of non-island (0) and island (1) structures:
+with number of non-island and island structures:
 
@@
 5. Infer parameters from empirical DNA methylation data.
+
+# References

### 2024-12-16 extensive_tests/abc_inferences/scripts/get_spatial_str.R

@@ -35,8 +35,14 @@ if (length(missing_options) > 0) {
 load(opt[["input"]])
 
 # Filter the data for the given chromosome and region
-spatial_str <- CpG_count[CpG_count$chr == opt[["chr"]] & CpG_count$start >= opt[["start"]] - opt[["side-bp"]] & CpG_count$start <= opt[["end"]] + opt[["side-bp"]], ]
+filtered_data <- CpG_count[CpG_count$chr == opt[["chr"]] & CpG_count$start >= opt[["start"]] - opt[["side-bp"]] & CpG_count$start <= opt[["end"]] + opt[["side-bp"]], ]
 
-# Save filtered data
+# Create the data frame spatial_str
+spatial_str <- data.frame(n = filtered_data$CpG_count)
+
+# Add the globalState column based on filtered_data$str values
+spatial_str$globalState <- ifelse(filtered_data$str == "Island", "U", "M")
+
+# Save spatial structure
 output_name <- paste0(paste(opt[["chr"]], opt[["start"]] - opt[["side-bp"]], opt[["end"]] + opt[["side-bp"]], opt[["name-selected-region"]], sep = "_"), ".RData")
 save(spatial_str, file = output_name)

### 2024-12-16 extensive_tests/abc_inferences/draft-abc-inferences-vignette.Rmd
Updating rule run_sim

@@
 conda env create -n abc_inferences -f environment.yaml
+```
+
+I got error with:
+
+- `r-parallel`         # parallel for R
+
+But it can be alternativelly installed by:
+
+```{bash, eval = FALSE}
+conda activate abc_inferences
+R
+install.packages("parallel")
+## Another package it is still dependent on is
+install.packages("devtools")
+```
+
+Then, the workflow can be run by:
+
+```{bash, eval = FALSE}
+conda activate abc_inferences
+snakemake --use-conda
 ```

### 2024-12-16 extensive_tests/abc_inferences/scripts/run_sim.R

@@ -21,10 +21,17 @@ opt <- parse_args(opt_parser)
 
 # Load the simulation design file
 load(opt[["input"]])
+print("here")
+print(head(sampled_params))
+print(scaled_trees)
+print(spatial_str)
 
 # Define the number of runs and cores
 n_sim <- nrow(sampled_params)
 n_cores <- opt[["n-cores"]]
+print("here 2")
+print(opt[["n-cores"]])
+print(opt[["CFTP"]])
 
 # Generate pad_n based on the number of digits in n_sim
 # To save the files with padded numbers so that are later listed in order with list.files()
@@ -77,7 +84,7 @@ run_simulation_subset <- function(start_idx, end_idx) {
 }
 
 
-
+print("here3")
 
 # Calculate the number of runs per chore
 runs_per_chore <- ceiling(n_sim / n_cores)

### 2024-12-16 extensive_tests/abc_inferences/scripts/run_sim.R
Updating rule run_sim

@@ -21,17 +21,9 @@ opt <- parse_args(opt_parser)
 
 # Load the simulation design file
 load(opt[["input"]])
-print("here")
-print(head(sampled_params))
-print(scaled_trees)
-print(spatial_str)
 
 # Define the number of runs and cores
 n_sim <- nrow(sampled_params)
-n_cores <- opt[["n-cores"]]
-print("here 2")
-print(opt[["n-cores"]])
-print(opt[["CFTP"]])
 
 # Generate pad_n based on the number of digits in n_sim
 # To save the files with padded numbers so that are later listed in order with list.files()
@@ -84,7 +76,6 @@ run_simulation_subset <- function(start_idx, end_idx) {
 }
 
 
-print("here3")
 
 # Calculate the number of runs per chore
 runs_per_chore <- ceiling(n_sim / n_cores)

### 2024-12-16 extensive_tests/abc_inferences/draft-abc-inferences-vignette.Rmd
Updating snakefile to use conda as global directive

@@
 snakemake --use-conda
+```
+
+I also got the following error when trying to run the rule `run_sim`:
+
+```{bash, eval = FALSE}
+Conda must be version 24.7.1 or later, found version 22.11.1.
+Please update conda to the latest version.
+Note that you can also install conda into the snakemake environment without modifying your main conda installation.
+# Then I installed conda into the snakemake environment with
+conda activate abc_inferences
+conda install -c conda-forge conda 
 ```

### 2024-12-16 extensive_tests/abc_inferences/scripts/get_spatial_str.R
Updating paths until run_sim

@@ -3,6 +3,8 @@ library(optparse)
 option_list <- list(
   make_option(c("--input"), type = "character", default = NULL,
               help = "Input CpG_count.RData path", metavar = "character"),
+  make_option(c("--output"), type = "character", default = NULL,
+              help = "Output full path with file name", metavar = "character"),
   make_option(c("--name-selected-region"), type = "character", default = NULL,
               help = "Name of selected genomic region", metavar = "character"),
   make_option(c("--chr"), type = "character", default = NULL,
@@ -21,7 +23,7 @@ opt_parser <- OptionParser(option_list = option_list)
 opt <- parse_args(opt_parser)
 
 # Get the names of required options (you can update this based on what's mandatory)
-required_options <- c("input", "name-selected-region", "chr", "start", "end", "side-bp")
+required_options <- c("input", "output","name-selected-region", "chr", "start", "end", "side-bp")
 
 # Check that all required options are not NULL
 missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
@@ -44,5 +46,5 @@ spatial_str <- data.frame(n = filtered_data$CpG_count)
 spatial_str$globalState <- ifelse(filtered_data$str == "Island", "U", "M")
 
 # Save spatial structure
-output_name <- paste0(paste(opt[["chr"]], opt[["start"]] - opt[["side-bp"]], opt[["end"]] + opt[["side-bp"]], opt[["name-selected-region"]], sep = "_"), ".RData")
-save(spatial_str, file = output_name)
+#output_name <- paste0(paste(opt[["chr"]], opt[["start"]] - opt[["side-bp"]], opt[["end"]] + opt[["side-bp"]], opt[["name-selected-region"]], sep = "_"), ".RData")
+save(spatial_str, file = opt[["output"]])

### 2024-12-16 extensive_tests/abc_inferences/scripts/run_sim.R

@@ -7,6 +7,8 @@ library(devtools); load_all()
 option_list <- list(
   make_option("--input", type = "character", default = NULL,
               help = "Design file, full path", metavar = "character"),
+  make_option("--output-path", type = "character", default = NULL,
+              help = "Full path for output files", metavar = "character"),
   make_option("--n-cores", type = "integer", default = NULL,
               help = "Number of cores", metavar = "integer"),
   make_option("--n-tips", type = "integer", default = NULL,
@@ -36,7 +38,7 @@ run_simulation_subset <- function(start_idx, end_idx) {
   padded_end <- formatC(end_idx, width = pad_n, format = "d", flag = "0")
   
   # Set the name for the .out file for the entire subset
-  out_file <- paste0("abc_dataSIM_", padded_start, "_to_", padded_end, ".out")
+  out_file <- file.path(opt[["output-path"]], paste0("abc_dataSIM_", padded_start, "_to_", padded_end, ".out"))
   
   # Open the log file once for the entire subset
   log_connection <- file(out_file, open = "w")
@@ -63,7 +65,7 @@ run_simulation_subset <- function(start_idx, end_idx) {
       data[[tip]] <- output$data[[1]][[tip]]$seq
     }
     # Save the simulated data
-    save(data, file = paste0("abc_dataSIM_", padded_index, ".RData"))
+    save(data, file = file.path(opt[["output-path"]], paste0("abc_dataSIM_", padded_index, ".RData")))
     
   }
   

### 2024-12-16 application_examples/abc_inferences/draft-abc-inferences-vignette.Rmd
Updating changes in location of abc_inferences workflow

--- 2024-12-16 extensive_tests/abc_inferences/draft-abc-inferences-vignette.Rmd
@@
 Note that you can also install conda into the snakemake environment without modifying your main conda installation.
-# Then I installed conda into the snakemake environment with
+# Then I tried to install conda into the snakemake environment with
 conda activate abc_inferences
-conda install -c conda-forge conda 
+(abc_inferences) saracv@australia:~/MasterThesis/MethEvolSIM/extensive_tests/abc_inferences$ conda install conda=24.11.0
+Collecting package metadata (current_repodata.json): done
+Solving environment: done
+==> WARNING: A newer version of conda exists. <==                                                                                                                          
+  current version: 22.11.1                                                                                                               
+  latest version: 24.11.0
+Please update conda by running
+    $ conda update -n base -c conda-forge conda
+Or to minimize the number of packages updated during conda update use
+     conda install conda=24.11.0
+
+# All requested packages already installed.
+
+(abc_inferences) saracv@australia:~/MasterThesis/MethEvolSIM/extensive_tests/abc_inferences$ conda --version
+conda 22.11.1
+
+```
+
+**This does not seem to work. I tried it out with**:
+
+```{bash, eval = FALSE}
+conda activate snakemake
+nohup nice snakemake --use-conda run_sim > run_sim.out &
+```
+
+and seems to be stuck in:
+
+```{bash, eval = FALSE}
+cat run_sim.out 
+Injecting conda environment /home/saracv/MasterThesis/MethEvolSIM/extensive_tests/abc_inferences/environment.yaml.
 ```

### 2024-12-17 extensive_tests/draft-sumstat-vignette.Rmd
Updating changes
Updating sumstats and test sumstats
Updating work on siteFChange

@@
 vignette: >
-  %\VignetteIndexEntry{Introduction to MethEvolSIM}
+  %\VignetteIndexEntry{Summary Statistics with MethEvolSIM}
   %\VignetteEngine{knitr::rmarkdown}
@@
 
-get_FChange_cherryData (uses find_cherries and cherry_diffs)
+We calculate the frequency of sites that changes for each structure.
+
+get_FChange_cherryData (uses find_cherries --get_cherryDist-- and cherry_diffs -- now count_cherryMethDiff )
+
+NOTE: Summary statistics related with tree cherries will require a

### 2024-12-18 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
Updating config CFTP convergence test to run simulations in australia
Adding script for rule to get CpG counts at island and non island structures for a reference genome and updating draft abc inferences vignette
Updating snakemake workflow for test CFTP convergence until run_sim

@@
 
-- R packages:
+- Dependencies are on: `environment.yaml`
 
@@
 
-- An .RData file with a dataframe specifying the distribution of methylation sites in the genomic region.
+- An .RData file with a dataframe specifying the distribution of methylation sites in the genomic region (under the same Snakemake directory)
+
+-
 
@@
 
-- `run_sim.R`. TODO: Update way in which the package is loaded before submitting.
+- `run_sim.R`. TODO: Update way in which the package is loaded before submitting. Uses as many cores as the config param n_sim
+
+## Running the workflow with nohup
+
+When using nohup one needs to activate the environment in the command.
+
+```{bash, eval = FALSE}
+# E.g. running the run_sim rule
+nohup conda run -n cftp_testConvergence snakemake run_sim &
+```

### 2024-12-18 extensive_tests/CFTP_convergence/scripts/get_spatial_str.R

@@ -22,5 +22,5 @@ spatial_str <- data.frame(n = rep(opt[["CpG-n"]], 2*opt[["str-n"]]),
                           globalState = rep(c("U", "M"), opt[["str-n"]]))
 
 # Save the spatial structure
-file_path <- file.path(opt[["CpG-n"]], opt[["file-name"]])
+file_path <- file.path(opt[["dir"]], opt[["file-name"]])
 save(spatial_str, file = file_path)

### 2024-12-18 extensive_tests/CFTP_convergence/scripts/run_sim.R

@@ -4,8 +4,11 @@ load_all()
 library(parallel)
 library(optparse)
 
+
 # Define the command-line options
 option_list <- list(
+  make_option("--dir", type = "character", default = NULL,
+              help = "Full path to save the output file", metavar = "character"),
   make_option(c("-f", "--design-file"), type = "character", default = NULL,
               help = "Full path to the simulation design file", metavar = "character"),
   make_option(c("-b", "--branch-length"), type = "numeric", default = NULL,
@@ -26,7 +29,7 @@ opt_parser <- OptionParser(option_list = option_list)
 opt <- parse_args(opt_parser)
 
 # Get the names of required options (you can update this based on what's mandatory)
-required_options <- c("design-file", "branch-length", "start", "end", "replicate-n")
+required_options <- c("dir", "design-file", "branch-length", "start", "end", "replicate-n")
 
 # Check that all required options are not NULL
 missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
@@ -49,16 +52,20 @@ params_pad_n <- nchar(as.character(n_sim))
 
 
 
-simul_CFTP_branch <- function(custom_params, params_pad_n, index_params, b_length, start, end, step_pad_n, spatial_str, replicate_n){
+simul_CFTP_branch <- function(custom_params, params_pad_n, index_params, b_length, start, end, step_pad_n, spatial_str, replicate_n, replicate_pad_n){
   
   # Set the name for the output file with the padded parameter index
   padded_index_params <- formatC(index_params, width = params_pad_n, format = "d", flag = "0")
-  rep_pad_n <- nchar(as.character(replicate_n))
-  padded_replicate_n <- formatC(replicate_n, width = rep_pad_n, format = "d", flag = "0")
-  out_file <- paste0("CFTP_testConvergence_paramsID_", padded_index_params, "_rep_", replicate_n, ".out")
+  padded_replicate_n <- formatC(replicate_n, width = replicate_pad_n, format = "d", flag = "0")
+  out_file <- file.path(opt[["dir"]], paste0("CFTP_testConvergence_paramsID_", padded_index_params, "_rep_", padded_replicate_n, ".out"))
+
+  # Open the log file once for the given parameters and replicate number
+  log_connection <- file(out_file, open = "w")
   
-  # Redirect both the stout and stderr to the same file
-  sink(out_file, type = c("output", "message"), append = TRUE)
[…]

### 2024-12-18 extensive_tests/CFTP_convergence/scripts/set_sim_design.R

@@ -56,6 +56,7 @@ sampled_params <- data.frame(
   iota = runif(opt[["n-sim"]], min = opt[["iota_min"]], max = opt[["iota_max"]])
 )
 
+
 # Due to numerical limitations, the smallest value of iota and alpha_Ri used by MethEvolSIM is 0.01
 # Therefore we correct the sampled values to represent the ones that will be used to simulate data
 sampled_params$alpha_Ri[sampled_params$alpha_Ri<1e-2] <- 1e-2
@@ -65,6 +66,6 @@ sampled_params$iota[sampled_params$iota<1e-2] <- 1e-2
 load(file = opt[["genome-dist"]])
 
 # Save the sampled parameters, the tree and the distribution of methylation sites in the design file
-save(spatial_str, sampled_params, file = opt[["dir"]])
+save(spatial_str, sampled_params, file = opt[["output"]])
 
 

### 2024-12-19 application_examples/abc_inferences/draft-abc-inferences-vignette.Rmd
Updating changes before pulling

@@
 ```
+
+**IMPORTANT: BOTH HERE AND IN CFTP ENVIRONMENT.YAML BEFORE SUBMITTING CHANGE DEVTOOLS BY METHEVOLSIM**
 
@@
 points(filtered_data$start, filtered_data$CpG_density, 
-       col = ifelse(filtered_data$str == 1, "red", "blue"), 
+       col = ifelse(filtered_data$str == "Island", "red", "blue"), 
        pch = 16)  # pch=16 gives filled circles

### 2024-12-24 extensive_tests/CFTP_convergence/scripts/plot.R
Updating workflow step plot_results and adding plots

@@ -82,14 +82,17 @@ for(index_params in 1:n_sim){
   
   # Plot non-island correlations
   plot(branchEvol_sumstats[[1]]$meanCor_ni, main = "Mean Cor. Non-Island", xlab = "Step Number", ylab = "Value", pch = 1, col = transparent_colors[1])
-  abline(h = cftp_sumstats$meanCor_i[1], col = colors[1], lty = 2, lwd = 2)
+  abline(h = cftp_sumstats$meanCor_ni[1], col = colors[1], lty = 2, lwd = 2)
   for (i in 2:n_rep){
-    points(branchEvol_sumstats[[i]]$meanCor_in, pch = i, col = transparent_colors[i])
-    abline(h = cftp_sumstats$meanCor_i[i], col = colors[i], lty = 2, lwd = 2)
+    points(branchEvol_sumstats[[i]]$meanCor_ni, pch = i, col = transparent_colors[i])
+    abline(h = cftp_sumstats$meanCor_ni[i], col = colors[i], lty = 2, lwd = 2)
   }
   
   # Add global title for all plots after the first plot is created
   mtext("Local correlations", outer = TRUE, line = 1, cex = 1.5)
+
+  # Add global subtitle
+  mtext(paste("Relative proportion of correlated single-site changes:", round(1 - sampled_params$iota[index_params], 3)), outer = TRUE, line = 0, cex = 1)
   
   dev.off()
 }

### 2024-12-24 application_examples/abc_inferences/archipelago-definition.Rmd
Adding file archipelago-definition

---
title: "Definition of Archipelagos"
author: "Sara"
date: "`r Sys.Date()`"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# If missing, install required packages

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

if (!requireNamespace("rtracklayer", quietly = TRUE)) {
  BiocManager::install("rtracklayer")
}

library(rtracklayer)
```

# Archipelago definition

We define archipelagos as genomic regions containing clusters of CGIs. We can interpret archipelagos as regions of the genome larger than CGI that together may regulate a gene or set of genes. Therefore, an approach to define archipelagos can be to choose thresholds that lead to clustering CGI in a way so that each archipelago overlaps with a promoter region (or maximizes cases of overlap single-archipelago and promoter).

Our approach then has the following steps:

1. Definition of promoter regions.
2. Filtering of CGI by promoter overlap.

## Definition of promoter regions

As in [@vernaz2022epigenetic], since no functional annotation exists for Lake Malawi cichlid genomes, we define promoter regions in silico as regions ±1 kbp around the TSS.

For that, we use [Annotation Features File (GTF) from the NCBI RefSeq Assembly](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000238955.4/)

```{r}
# Load the GTF file
gtf_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gtf" 
gtf <- import(gtf_file)
head(gtf)
```

The gtf file is structured as a GRanges object containing genomic annotation information with the following key components (columns):

- `seqnames`: Chromosome or scaffold name (e.g., NC_036780.1).

- `ranges`: Start and end positions of the genomic feature.

- `strand`: Strand of the feature (+ for forward strand, - for reverse strand).

- `source`: Annotation source (e.g., Gnomon).

- `type`: Feature type (e.g., gene, transcript, exon).

- `Metadata`: Additional columns include gene_id, transcript_id, gene, product, gene_biotype, exon_number, etc., which describe the feature in detail.

Example Entries:

- Row 1: A gene on the + strand, identified as LOC106676409, classified as lncRNA.

- Row 2: A transcript corresponding to the same gene.

- Rows 3-5: Exons belonging to the transcript.

- Row 6: A gene on the - strand, classified as protein_coding.

We filter for transcripts:

```{r}
# Filter for transcripts
gtf_transcripts <- gtf[gtf$type == "transcript", ]
head(gtf_transcripts)
```

And define the TSS by accounting for the strand orientation so that:

- For Features on the + Strand: The TSS is the smallest genomic coordinate (start).

- For Features on the - Strand: The TSS is the largest genomic coordinate (end), as transcription occurs in the reverse direction.

```{r}
# Extract TSS (assuming + strand)
promoters <- data.frame(
  chr = seqnames(gtf_transcripts),
  tss = ifelse(strand(gtf_transcripts) == "+", start(gtf_transcripts), end(gtf_transcripts)),
  strand = strand(gtf_transcripts)
)
```

### 2025-01-04 application_examples/abc_inferences/archipelago-definition.Rmd
Adding result plots for CFTP convergence test

@@
 
-```{r}
+```{r, warning=FALSE, message=FALSE}
 # If missing, install required packages
@@
 
+if (!requireNamespace("GenomicRanges", quietly = TRUE)) {
+  BiocManager::install("GenomicRanges")
+}
+
+library(GenomicRanges)
 library(rtracklayer)
@@
 
-## Definition of promoter regions
+## Step 1: Definition of promoter regions
 
@@
 ```{r}
-# Extract TSS (assuming + strand)
+# Extract TSS
 promoters <- data.frame(
@@
 )
+head(promoters)
 ```
+
+The promoters data frame contains:
+
+- `chr`: Chromosome names
+
+- `tss`: Transcription start sites
+
+- `strand`: Strand orientation of the transcripts
+
+## Step 2: Filtering of CGI by promoter overlap
+
+CGI annotation file (`.txt` file) obtained with makeCGI v1.3.4 [@hiddenMarkov]
+
+```{r}
+# Load the CGI annotation file
+cgi_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/CGI-Mzebra.txt" 
+cgif <- read.table(cgi_file, header = TRUE, sep = "\t")
+head(cgif)
+```
+
+Convert the promoter and CGI data to `GRanges` objects for genomic interval operations:
+
+```{r}
+# Create GRanges for promoters
+promoter_gr <- GRanges(
+  seqnames = promoters$chr,
+  ranges = IRanges(start = promoters$tss, end = promoters$tss),
+  strand = promoters$strand
+)
+
+# Create GRanges for CGI
+cgi_gr <- GRanges(
+  seqnames = cgif$chr,
+  ranges = IRanges(start = cgif$start, end = cgif$end)
+)
+```
+
+Use `findOverlaps` to find promoters within or near CpG islands:
+
+```{r}
+# Find overlaps
+overlaps <- findOverlaps(promoter_gr, cgi_gr)
+
+# Extract overlapping promoter indices
+promoters_with_cgi <- promoters[queryHits(overlaps), ]
+cgi_overlaps <- cgif[subjectHits(overlaps), ]
+
+# Combine data for output
+promoter_cgi_overlap <- cbind(promoters_with_cgi, cgi_overlaps)
+head(promoter_cgi_overlap)
+```
+
+The `promoter_cgi_overlap` dataset combines information about promoters and CpG islands (CGIs) where they overlap. It contains the following columns:
+
+- `chr`: The chromosome where the promoter is located.
+
+- `tss`: The transcription start site (TSS) position of the promoter.
+
+- `strand`: The strand of the promoter (+ or -).
+
+- `chr`: The chromosome where the CGI is located (same as the promoter in the overlap context).
+
+- `start`: The start position of the overlapping CGI.
+
+- `end`: The end position of the overlapping CGI.
+
+- `length`: The length of the CGI in base pairs.
+
+- `CpGcount`: The number of CpG dinucleotides within the CGI.
+
+- `GCcount`: The number of G and C nucleotides within the CGI.
+
+- `pctGC`: The percentage of G and C nucleotides within the CGI.
+
+- `obsExp`: The observed-to-expected ratio of CpG dinucleotides in the CGI.
+
+Each row corresponds to an overlap between a promoter and a CGI. For example:
+
+- The first row shows that a promoter at TSS 755,445 on the negative strand overlaps with a CGI on chromosome NC_036780.1 spanning positions 755,397 to 756,689.
+This CGI has a length of 1,293 bp, contains 123 CpG dinucleotides, and has a GC content of 63.19% with an observed-to-expected CpG ratio of 0.954.
+
+- The overlap is detailed for both the promoter and the CGI, including their respective properties.
+
+```{r}
+#write.table(promoter_cgi_overlap, file = "promoter_cgi_overlap.txt", sep = "\t", row.names = FALSE)
+promoter_cgi_overlap <- read.table("/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/promoter_cgi_overlap.txt")
+```
+
+### Check TSS overlap density
+
+```{r}
+plot(
+  promoter_cgi_overlap$length, 
+  promoter_cgi_overlap$pctGC, 
+  xlab = "CGI Length (bp)", 
+  ylab = "GC Content (%)", 
+  main = "Relationship Between CGI Length and GC Content",
+  pch = 16,  # Solid circle points
+  col = "blue"  # Point color
+)
+
+```
+
+```{r}
+
+```

### 2025-01-07 extensive_tests/draft-sumstat-vignette.Rmd

@@
 
-We calculate the frequency of sites that changes for each structure.
+We calculate the frequency of sites that changes for each structure (island and non-island structures).
+
+Let $S_k$ be the set of sites in the cherry structure of interest, where $k$ represents the index of the cherry structure.
+For each site in $S_k$, let the methylation state at a given tip $t$ be denoted as $m_{k,s,t}$, where $s \in S_k$ is the index of the site, and $m_{k,s,t} \in {0, 0.5, 1}$ represents the methylation state at tip $t$ (unmethylated, partially-methylated, or methylated, respectively).
+
+For a given cherry structure, we compute in the proportion of sites that have a different methylation state between two tips, say $t_1$ and $t_2$.
+
+Define the indicator variable:
+\[
+\delta_{k,s,t_1,t_2} =
+\begin{cases}
+1, & \text{if } m_{k,s,t_1} \neq m_{k,s,t_2}, \\
+0, & \text{if } m_{k,s,t_1} = m_{k,s,t_2}.
+\end{cases}
+\]
+
+Then, the summary statistic for the proportion of sites with different methylation states between tips $t_1$ and $t_2$ in the cherry structure $k$ is:
+\[
+\text{Proportion}_k(t_1, t_2) = \frac{1}{|S_k|} \sum_{s \in S_k} \delta_{k,s,t_1,t_2}
+\]
+
+where $|S_k|$ is the total number of sites in the cherry structure $k$, and the sum counts the number of sites with different methylation states between tips $t_1$ and $t_2$.
+
+This statistic gives the proportion of sites in the cherry structure $k$ where the methylation state differs between the two tips of interest.
 
@@
 
-NOTE: Summary statistics related with tree cherries will require a
+NOTE: Summary statistics related with tree cherries require a tree with tip labels as:

### 2025-01-07 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
Updating workflow step plot_results and adding plots
Updating draft-test_CFTP_convergence.Rmd after christmas
Updating draft of vignette CFTP convergence

@@
 
-```{r, echo=FALSE}
-## TODO: Update 
-library(devtools)
-load_all()
-```
+# CFTP
 
-# Worflow for simulation-based inferences
+Theoretical intro.
+
+## Implementation
+
+Currently neglects IWEs but allows to sample a sequence of methylation states from the equilibrium given by the equilibrium frequencies of the genomic regions, the correlated and independent SSE.
+
+## Test
 
@@
 
-## Requirements
+**Add summary**
 
-- Snakemake installed (`conda create -c conda-forge -c bioconda -n snakemake snakemake`)
+### Requirements
 
-- Dependencies are on: `environment.yaml`
+-   Dependencies are listed on the file `environment.yaml`. Before running the workflow, a conda environment can be created with:
 
-- Access to the provided `Snakefile` and associated scripts.
+```{bash, eval = FALSE}
+conda env create -n cftp_testConvergence -f environment.yaml
+```
 
-- An .RData file with a dataframe specifying the distribution of methylation sites in the genomic region (under the same Snakemake directory)
+Then, the workflow can be run by:
 
--
+```{bash, eval = FALSE}
+conda activate cftp_testConvergence
+snakemake --use-conda
+```
+
+Or, if using `nohup` by:
+
+```{bash, eval = FALSE}
+conda activate cftp_testConvergence
+nohup nice snakemake --use-conda run_sim > run_sim.out &
+```
+
+-   Provided `Snakefile`, `config.yaml` and associated scripts.
 
@@
 
-1. Bla
+All the given parameters to the workflow steps are taken from the `config.yaml` file.
+
+1.
+Rule `get_spatial_str` run by script `get_spatial_str.R`.
+The test simulates data on a simplified genomic structure alternating island and non-island regions each a given number of times and each (`Str_n`) with a constant number of CpGs  (`CpG_n`).
+The output is a file named as `methSite_genomicDist` under the directory `dir`.
+
+2.
+Rule `set_design` run by script `set_sim_design.R`.
+Takes as input the output of the previous rule.
+For a number of model parameter combinations `n_sim` and setting a seed `seed` for reproducibility, samples the parameter combinations from the prior distributions as defined in `prior_distributions`.
+The output is a file named "design-RData" under the directory `dir` containing both the spatial structure defined in the first rule and the sampled parameters.
+*Note that this script already corrects the values of* $\iota$ *and* $\alpha_{R_i}$ *in the sampled parameters to be a minimum of 1e-2*.
 
@@
 ```
+
+## Results
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_01.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_02.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_03.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_04.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_05.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_06.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_07.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_08.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_09.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_10.pdf")
+```

### 2025-01-07 extensive_tests/CFTP_convergence/scripts/plot.R
Updating script to plot results in test CFTP convergence

@@ -51,7 +51,35 @@ replicate_pad_n = nchar(as.character(n_rep))
 colors <- rainbow(n_rep)
 transparent_colors <- sapply(colors, function(col) adjustcolor(col, alpha.f = 0.2))  # alpha.f is a transparency factor (0=fully transparent, 1=opaque)
 
+# Function to generate the plot
+generate_plot <- function(branchEvol_sumstats, cftp_sumstats, colors, transparent_colors, sampled_params, index_params, n_rep) {
+  # Set up layout for 1 rows, 2 columns, and space for a global title (oma parameter)
+  par(mfrow = c(1, 2), oma = c(0, 0, 3, 0))  # oma reserves space at the top for the title
+  
+  # Plot island correlations
+  plot(branchEvol_sumstats[[1]]$meanCor_i, main = "Mean Cor. Island", xlab = "Step Number", ylab = "Value", pch = 1, col = transparent_colors[1])
+  abline(h = cftp_sumstats$meanCor_i[1], col = colors[1], lty = 2, lwd = 2)
+  for (i in 2:n_rep) {
+    points(branchEvol_sumstats[[i]]$meanCor_i, pch = i, col = transparent_colors[i])
+    abline(h = cftp_sumstats$meanCor_i[i], col = colors[i], lty = 2, lwd = 2)
+  }
+  
+  # Plot non-island correlations
+  plot(branchEvol_sumstats[[1]]$meanCor_ni, main = "Mean Cor. Non-Island", xlab = "Step Number", ylab = "Value", pch = 1, col = transparent_colors[1])
+  abline(h = cftp_sumstats$meanCor_ni[1], col = colors[1], lty = 2, lwd = 2)
+  for (i in 2:n_rep) {
+    points(branchEvol_sumstats[[i]]$meanCor_ni, pch = i, col = transparent_colors[i])
+    abline(h = cftp_sumstats$meanCor_ni[i], col = colors[i], lty = 2, lwd = 2)
+  }
+  
+  # Add global title for all plots
+  mtext("Local correlations", outer = TRUE, line = 1, cex = 1.5)
+  
+  # Add global subtitle
+  mtext(paste("Relative proportion of correlated single-site changes:", round(1 - sampled_params$iota[index_params], 3)), outer = TRUE, line = 0, cex = 1)
+}
 
+# Loop to generate the plots
 for(index_params in 1:n_sim){
[…]

### 2025-01-09 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
Updating draft test CFTP convergence

@@
 
-Theoretical intro.
+The initialization of the DNA methylation data without the CFTP method samples for each genomic structure (island or non-island) the methylation state for each of the CpG sites within from the structure's equilibrium frequencies independently.
+Therefore, the initial distribution of methylation states lacks the effect of the neighbor correlations introduced by the correlated SSE process.
+This correlations would then build during evolution along a branch.
+
+In order to be able to start simulating evolution along a tree with the sequence of states at the tree root accounting for the neighbouring correlations we implemented the CFTP method, which samples a sequence of methylation states from the equilibrium given by the model (accounting for the correlated SSE process while neglecting the IWE process.)
 
@@
 
-This workflow uses [Snakemake](https://snakemake.readthedocs.io) to manage the steps for simulating DNA methylation patterns, assessing ABC accuracy, and inferring model parameters.
+This workflow uses [Snakemake](https://snakemake.readthedocs.io) to manage the steps to test that the CFTP method samples a sequence of methylation states from the equilibrium given by the model accounting for the correlated SSE process (while neglecting the IWE process).
 
-**Add summary**
+For that it samples a number of parameter combinations and for each it simulates initial data for a number of replicates.
+It uses as summary statistics the mean correlation of a central segment of methylation states per genomic structure type.
+We expect that the value of this summary statistics increases with the increase of evolutionary time until it converges towards the equilibrium given by the corresponding parameter combination.
+We track its value by discretizing the evolution along the branch in small steps (length as config.yaml file value for `branch_length`).
+Then, we call the CFTP method on the initial data to sample the sequence of methylation states from the equilibrium and compute its corresponding value for the summary statistics.
+We expect the distribution of the summary statistics obtained by calling the CFTP method to be representative of the equilibrium obtained after the branch evolution has reached the equilibrium.
+
+**Add summary and graphical image**
 
@@
 
-Or, if using `nohup` by:
+When using nohup one needs to activate the environment in the command.
 
 ```{bash, eval = FALSE}
-conda activate cftp_testConvergence
-nohup nice snakemake --use-conda run_sim > run_sim.out &
+# E.g. running the run_sim rule
+nohup conda run -n cftp_testConvergence snakemake run_sim &
 ```
@@
 For a number of model parameter combinations `n_sim` and setting a seed `seed` for reproducibility, samples the parameter combinations from the prior distributions as defined in `prior_distributions`.
-The output is a file named "design-RData" under the directory `dir` containing both the spatial structure defined in the first rule and the sampled parameters.
+The output is a file named "design.RData" under the directory `dir` containing both the spatial structure defined in the first rule and the sampled parameters.
 *Note that this script already corrects the values of* $\iota$ *and* $\alpha_{R_i}$ *in the sampled parameters to be a minimum of 1e-2*.
 
-### Scripts
+3.
+Rule `run_sim` run by script `run_sim.R`.
+Takes as input the output of the previous rule.
+For each parameter combination it simulates a number of replicates (`replicate_n`) and (1) simulates evolution along a branch of length `branch_length` a number of times from `start` to `end` and (2) calls the CFTP method on the initial data.
+Each replicate is simulated with a seed obtained as the index of the corresponding parameter combination * 1000 + the replicate number for reproducibility.
+The output files are written under the directory `dir` and named "CFTP_testConvergence_paramsID_{parameter_combination_number}_rep_{replicate_number}_{step_n}.RData" for the output of evolution along the branch and "CFTP_testConvergence_paramsID_{parameter_combination_number}_rep_{replicate_number}_cftp.RData" for the output of the cftp method.
+**Need to update the way in which the package is loaded before submitting**
 
--   `run_sim.R`. TODO: Update way in which the package is loaded before submitting. Uses as many cores as the config param n_sim
+4.
+Rule `compute_meanCor` run by script `compute_meanCor.R`.
+Takes as input the output of the previous rule.
+For each branch step output and cftp output it computes the mean correlation of a segment of methylation states per genomic structure type (**summary statistics described in the summaryStats vignette**).
+The output files are written under the directory `dir` and are named as "summaryStats_CFTP_testConvergence_paramsID_{parameter_combination_number}_cftp.RData" for the output of the cftp method and ""summaryStats_CFTP_testConvergence_paramsID_{parameter_combination_number}_rep_{replicate_number}.RData" for the output of evolution along the branch.
 
-## Running the workflow with nohup
-
-When using nohup one needs to activate the environment in the command.
-
-```{bash, eval = FALSE}
-# E.g. running the run_sim rule
-nohup conda run -n cftp_testConvergence snakemake run_sim &
-```
+5.
+Rule `plot_results` run by script `plot.R`.
+Takes as input the output of the previous rule.
+The output files are written under a subdirectory `Figures` generated in the same directory of the snakemake workflow, and are named "CFTP_testConvergence_paramsID_{parameter_combination_number}" both in `.pdf` and `.png` format.
 

### 2025-01-10 extensive_tests/Nikolas_simulation/scripts/nikolas_sumstats.R
removed Nikolas' files from develop branch

@@ -1,321 +1,0 @@
-#!/usr/bin/env Rscript
-library(optparse)
-#source("C:\\Users\\ricca\\OneDrive\\Desktop\\Bachelor\\MES_GitClone30_10\\R\\multiRegion_SIM.R")
-#source("C:\\Users\\ricca\\OneDrive\\Desktop\\Bachelor\\MES_GitClone30_10\\R\\nikolas_extenstion.R")
-source("/Users/nikolas/Library/CloudStorage/OneDrive-Persönlich/Desktop/Bachelor/MES_GitClone30_10/R/multiRegion_SIM.R")
-source("/Users/nikolas/Library/CloudStorage/OneDrive-Persönlich/Desktop/Bachelor/MES_GitClone30_10/R/nikolas_extenstion.R")
-
-option_list <- list(
-  make_option("--tipdir", type = "character", default = NULL,
-              help = "Dir of tips list"),
-  make_option("--output", type = "character", default = NULL,
-              help = "Dir of output"),
-  make_option("--nsim", type = "integer", default = NULL,
-              help = "Number of simulations"),
-  make_option("--recrates", type = "character", default = NULL,
-              help = "Recrates location"),
-  make_option("--samples", type = "integer", default = NULL,
-              help = "Sample number")
-  )
-
-convert_global_methylation = function(seq){
-  seq[seq==1] <- 0
-  seq[seq==2] <- 0.5
-  seq[seq==3] <- 1
-  
-  return(seq)
-}
-opt_parser <- OptionParser(option_list = option_list)
-opt <- parse_args(opt_parser)
-tiplist <- list.files(path = opt[["tipdir"]],full.names = TRUE)
-
-rec_rate_df <- read.csv(opt[["recrates"]])
-
-samples <- opt[["samples"]]
-nsim <- opt[["nsim"]]
-pad_n <- nchar(as.character(length(tiplist))) + 1
-
-#results dataframe
-results <- data.frame(sim_num = integer(), rec_rate = numeric(),n_rec = numeric(), tajima_island = numeric(), tajima_non_island = numeric(), wattersons_island = numeric(), wattersons_nonisland = numeric(), watterson_island_pair = numeric(), tajima_i1 = numeric(), tajima_i2 = numeric(), stringsAsFactors = FALSE)
-
-tips_list <- list.files(path = opt[["tipdir"]],full.names = TRUE)
-
-
[…]

### 2025-01-10 application_examples/abc_inferences/archipelago-definition.Rmd
Updating git in order to pull

@@
 # Extract TSS
-promoters <- data.frame(
+tss <- data.frame(
   chr = seqnames(gtf_transcripts),
@@
 )
-head(promoters)
+head(tss)
 ```
 
-The promoters data frame contains:
+The tss data frame contains:
 
@@
 
+# Dirk idea
+
+## Explore the data
+
 ```{r}
+unique(gtf$type)
+```
+
+The gtf contains: gene, transcript, exon, CDS, start_codon and stop_codon
+
+Lets filter the genes:
+
+```{r}
+genes <- gtf[gtf$type == "gene",]
+dim(as.data.frame(genes))
+```
+
+Are the genes duplicated?
+
+```{r}
+length(genes$gene_id)
+length(unique(genes$gene_id))
+```
+
+Lets explore the annotation sources:
+
+```{r}
+unique(genes$source)
+```
+
+Summary table:
+
+| Source        | Primary Focus                       | Reliability          | Features                                         |
+|---------------|-------------------------------------|----------------------|--------------------------------------------------|
+| Gnomon        | Computational gene prediction       | Medium               | Predicted genes, transcripts, and exons         |
+| cmsearch      | RNA homology                        | Medium-High          | Non-coding RNAs (e.g., ribosomal RNAs)          |
+| tRNAscan-SE   | tRNA genes                          | High                 | tRNA annotations                                 |
+| BestRefSeq    | High-confidence curated annotations | High                 | Protein-coding and non-coding genes             |
+| BestRefSeq, Gnomon | Combined curated and predicted annotations | Medium-High | Hybrid of RefSeq and Gnomon predictions         |
+| RefSeq        | Broader RefSeq annotations          | Mixed (High to Low)  | Protein-coding and non-coding genes             |
+
+## Define the 200 kb upstream regions
+
+First, save chromosome lengths:
+
+```{r}
+gbff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gbff" 
+file_lines <- readLines(gbff_file)
+# Extract the LOCUS line
+locus_lines <- file_lines[grep("^LOCUS", file_lines)]
+head(locus_lines)
+```
+
+They dont seem to end with the dot and the number (version) as in the case of the `gtf_genes` data. Lets check if the `gtf_genes` (or the dataframe obtained from it) has several versions:
+
+```{r}
+length(unique(genes_df$seqnames))
+# Remove version suffix from seqnames
+length(unique(sub("\\.\\d+$", "", genes_df$seqnames)))
+```
+
+It does not have several versions.
+
+```{r}
+# Extract chromosome names and lengths
+chrom_info <- gsub("^LOCUS\\s+(\\S+)\\s+(\\d+)\\s+.*$", "\\1,\\2", locus_lines)
+chrom_info <- strsplit(chrom_info, ",")
+chrom_names <- sapply(chrom_info, `[`, 1)
+chrom_lengths <- as.numeric(sapply(chrom_info, `[`, 2))
+# Create a named vector
+chrom_lengths <- setNames(chrom_lengths, chrom_names)
+
+# Remove version numbers from seqnames in genes_df
+genes_df$seqnames <- sub("\\.\\d+$", "", genes_df$seqnames)
 
 ```
+
+Calculate the upstream start and end to prevent overlap with the previous gene.
+
+```{r}
+# Ensure genes_df is sorted by seqnames and start position
+genes_df <- genes_df[order(genes_df$seqnames, genes_df$start), ]
+
+# Initialize the upstream_start column
+genes_df$upstream_start <- NA
+
+# Iterate over genes to adjust upstream_start
+for (i in 1:nrow(genes_df)) {
+  current_gene <- genes_df[i, ]
+  current_chrom <- current_gene$seqnames
+  current_strand <- current_gene$strand
+  current_start <- current_gene$start
+  current_end <- current_gene$end
+
+  # Find the previous gene on the same chromosome
+  previous_gene <- genes_df[i - 1, ]
+  if (i > 1 && previous_gene$seqnames == current_chrom) {
+    previous_end <- previous_gene$end
+    previous_start <- previous_gene$start
+  } else {
+    previous_end <- NA
+    previous_start <- NA
+  }
+
+  # Adjust upstream_start based on strand and previous gene position
+  if (current_strand == "+") {
+    if (!is.na(previous_end)) {
+      genes_df$upstream_start[i] <- max(current_start - 200000, previous_end)
+    } else {
+      genes_df$upstream_start[i] <- max(current_start - 200000, 1)
+    }
+  } else if (current_strand == "-") {
+    if (!is.na(previous_start)) {
+      genes_df$upstream_start[i] <- min(current_end + 200000, previous_start)
+    } else {
+      genes_df$upstream_start[i] <- min(current_end + 200000, chrom_lengths[current_chrom])
+    }
+  }
+}
+
+```
+
+Filter for "curated" chromosomes:
+
+```{r}
+# Filter genes_df to keep only those rows where seqnames starts with "NC"
+genes_df_filtered <- genes_df[grep("^NC", genes_df$seqnames), ]
+
+# Check the result
+head(genes_df_filtered)
+
+```
+
+Still it seems there is nothing but genes. Maybe I should have filtered for something else.
+
+In a GTF (Gene Transfer Format) file, the gene feature type represents the entire gene, including all its exons, introns, and regulatory regions.
+The transcript feature type corresponds to a specific transcript variant of the gene, encompassing all exons and introns of that transcript.
+The exon feature type denotes individual exonic regions within a transcript.
+The CDS (Coding Sequence) feature type indicates the portion of the gene or transcript that codes for protein.
+The start_codon and stop_codon feature types mark the beginning and end of the coding sequence, respectively.
+**Try gff**

### 2025-01-10 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
Updating git in order to pull

--- 2025-01-07 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
@@
 
-```{r, fig.align='center',  out.width='50%'}
+```{r, out.width="0.8\\linewidth", include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
 knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_01.pdf")
 ```
-
-```{r, fig.align='center',  out.width='50%'}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_02.pdf")
-```
-
-```{r, fig.align='center',  out.width='50%'}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_03.pdf")
-```
-
-```{r, fig.align='center',  out.width='50%'}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_04.pdf")
-```
-
-```{r, fig.align='center',  out.width='50%'}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_05.pdf")
-```
-
-```{r, fig.align='center',  out.width='50%'}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_06.pdf")
-```
-
-```{r, fig.align='center',  out.width='50%'}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_07.pdf")
-```
-
-```{r, fig.align='center',  out.width='50%'}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_08.pdf")
-```
-
-```{r, fig.align='center',  out.width='50%'}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_09.pdf")
-```
-
-```{r, fig.align='center',  out.width='50%'}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_10.pdf")
-```

### 2025-01-10 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
Updating git in order to be able to pull from chile

--- 2025-01-10 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
@@
----
+q---
 title: "CFTP Convergence"
@@
 
-```{r, out.width="0.8\\linewidth", include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
+```{r, fig.align='center',  out.width='50%'}
 knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_01.pdf")
 ```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_02.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_03.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_04.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_05.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_06.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_07.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_08.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_09.pdf")
+```
+
+```{r, fig.align='center',  out.width='50%'}
+knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_10.pdf")
+```

### 2025-01-15 application_examples/abc_inferences/archipelago-definition.Rmd
Updating last changes to be able to pull

@@
 **Try gff**
+
+```{r}
+gtf_df[gtf_df$seqnames=="NC_036801.1" & gtf_df$type == "gene",]
+gtf_df[gtf_df$gene_id == "LOC101483164",]
+```
+
+```{r}
+dpp4_chr <- "NC_036801.1"
+dpp4_start <- 29502362
+dpp4_end <- 29517382
+gtf_df[gtf_df$seqnames=="NC_036801.1" & gtf_df$type == "gene" & gtf_df$start > dpp4_start - 5e5 & gtf_df$end > dpp4_end + 5e5  ,]
+```

### 2025-01-23 application_examples/abc_inferences/archipelago-definition.Rmd
Updating changes in archipelago-definition

--- 2025-01-04 application_examples/abc_inferences/archipelago-definition.Rmd
@@
 
-if (!requireNamespace("GenomicRanges", quietly = TRUE)) {
-  BiocManager::install("GenomicRanges")
-}
-
-library(GenomicRanges)
 library(rtracklayer)
@@
 
-## Step 1: Definition of promoter regions
+We will perform this steps for the genomic region of interest:
+
+```{r}
+dpp4_chr <- "NC_036801.1"
+dpp4_start <- 29502362
+dpp4_end <- 29517382
+region_start <- dpp4_start - 5e5
+region_end <- dpp4_end + 5e5
+```
+
+- DPP4 gene location:
+
+  - Chromosome: NC_036801.1
+
+      - Start: 29,502,362
+
+      - End: 29,517,382
+
+- Extended Region: Includes 500,000 base pairs upstream and downstream of the DPP4 gene.
+
+## Step 1: Definition of "promoter/upstream" regions
 
@@
 
-For that, we use [Annotation Features File (GTF) from the NCBI RefSeq Assembly](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000238955.4/)
+### Choice of gene annotiation file format
+
+File formats from [NCBI](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000238955.4/).
+Selected RefSeq only (no Ensembl annotation).
+More info [here](https://www.biobam.com/differences-between-gtf-and-gff-files-in-genomic-data-analysis/), [here](http://www.ensembl.org/info/website/upload/gff.html) and [here](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-6707-9).
+
+- The General Feature Format (GFF): GFF is a standard file format used for storing genomic sequences and annotations, developed by the Sanger Centre (v2) and the Sequence Ontology Project (v3).
+It describes features and annotations of DNA, RNA, and protein sequences in a plain text format with nine fields separated by tabs.
+Focuses on the hierarchical structure of genome annotations and is often less verbose.
+
+- The Gene Transfer Format (GTF): GTF, often seen as a derivative of GFF, is more specific in its application.
+It was generally termed GTF around 2000 and developed as part of collaborative genome annotation projects.
+GTF borrows from GFF but has additional structures for detailed gene annotations, often used to describe transcripts and their coding sequences.
+
+The `.gtf` file contains more information, but for our purpose we can use the `.gff` file.
 
 ```{r}
-# Load the GTF file
-gtf_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gtf" 
-gtf <- import(gtf_file)
-head(gtf)
+# Load the GFF file
+gff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gff" 
+gff <- as.data.frame(import(gff_file))
+
 ```
 
-The gtf file is structured as a GRanges object containing genomic annotation information with the following key components (columns):
+- `seqnames`: Chromosome.
 
-- `seqnames`: Chromosome or scaffold name (e.g., NC_036780.1).
+- `start`, `end`: Coordinates of the gene.
 
-- `ranges`: Start and end positions of the genomic feature.
+- `strand`: Gene orientation.
 
-- `strand`: Strand of the feature (+ for forward strand, - for reverse strand).
+- `type`: Feature type (here filtered for genes).
 
-- `source`: Annotation source (e.g., Gnomon).
+- `Gene`: Gene name or ID.
 
-- `type`: Feature type (e.g., gene, transcript, exon).
-
-- `Metadata`: Additional columns include gene_id, transcript_id, gene, product, gene_biotype, exon_number, etc., which describe the feature in detail.
-
-Example Entries:
-
-- Row 1: A gene on the + strand, identified as LOC106676409, classified as lncRNA.
-
-- Row 2: A transcript corresponding to the same gene.
-
-- Rows 3-5: Exons belonging to the transcript.
-
-- Row 6: A gene on the - strand, classified as protein_coding.
-
-We filter for transcripts:
+- `gene_biotype`: Functionality (e.g., protein_coding)
 
 ```{r}
-# Filter for transcripts
-gtf_transcripts <- gtf[gtf$type == "transcript", ]
-head(gtf_transcripts)
+table(gff$source)
 ```
 
-And define the TSS by accounting for the strand orientation so that:
+- `source`: indicates the source of the annotation for each feature in the file. These sources represent the tools, pipelines, or databases that generated the corresponding annotation.
 
-- For Features on the + Strand: The TSS is the smallest genomic coordinate (start).
+  - `RefSeq`: These annotations come directly from the RefSeq database.
+RefSeq annotations are high-quality, manually curated or computationally derived features that are part of NCBI's Reference Sequence project.
+Examples: Protein-coding genes, RNA genes, pseudogenes.
 
-- For Features on the - Strand: The TSS is the largest genomic coordinate (end), as transcription occurs in the reverse direction.
+   - `Gnomon`: Gnomon is NCBI's gene prediction tool, which generates computationally predicted annotations.
+These annotations may include protein-coding genes, non-coding RNAs, or other features predicted based on sequence analysis.
+Gnomon is used when there is no manually curated annotation available for certain regions or genes.
+
+   - `cmsearch`: This refers to the Infernal package (specifically the cmsearch tool), which identifies RNA features such as non-coding RNAs or structured RNA elements.
+Often used for annotating RNA families based on covariance models.
+
+  - `tRNAscan-SE`: tRNAscan-SE is a specialized tool for identifying and annotating tRNA genes in genome sequences. Annotations with this source indicate regions identified as tRNAs through this tool.
+
+   - `BestRefSeq`: Features with this source are part of the BestRefSeq annotation, which integrates high-quality manual curation with computational predictions to produce a "best" annotation for a gene or transcript.
+BestRefSeq annotations are typically the most reliable.
+
+  - BestRefSeq,Gnomon (or `BestRefSeq%2CGnomon`): This indicates that a feature is annotated by both the BestRefSeq and Gnomon pipelines.
+Such annotations may combine curated data with computational predictions.
+
+### Filter region of interest
 
 ```{r}
-# Extract TSS
-promoters <- data.frame(
-  chr = seqnames(gtf_transcripts),
-  tss = ifelse(strand(gtf_transcripts) == "+", start(gtf_transcripts), end(gtf_transcripts)),
-  strand = strand(gtf_transcripts)
-)
-head(promoters)
+region_gff <- gff[
+  gff$seqnames == "NC_036801.1" & 
+  gff$type == "gene" & 
+  gff$start <= region_end & 
+  gff$end >= region_start, 
+]
+
+head(region_gff)
 ```
 
-The promoters data frame contains:
+```{r}
+table(region_gff$source)
+```
 
-- `chr`: Chromosome names
+```{r}
+region_gff[grep("GeneID:101483164", region_gff$Dbxref), ]
+```
 
-- `tss`: Transcription start sites
+```{r}
+length(unique(region_gff$ID))
+unique(region_gff$ID)
+```
 
-- `strand`: Strand orientation of the transcripts
+```{r}
+# Extract relevant columns
+gene_data <- region_gff[, c("strand","start", "end", "gene")]
+
+# Create a y-axis index for unique gene names
+gene_data$y_index <- match(gene_data$gene, unique(gene_data$gene))
+
+# Set gene colors based on strand information (red for negative strand, blue for positive strand)
+gene_data$gene_color <- ifelse(gene_data$strand == "-", "red", "blue")
+
+```
+
+```{r}
+# Set up larger plotting area and adjust margins
+par(mar=c(6, 10, 4, 2), xpd=TRUE)
+
+# Create the plot with adjusted y-axis
+plot(c(min(gene_data$start), max(gene_data$end)), 
+     c(0, max(gene_data$y_index) + 1), 
+     type="n", 
+     xlab="Genomic Coordinates", 
+     ylab="", 
+     xaxt="n", 
+     yaxt="n")
+
+# Plot each gene as a line, with thicker line for GeneID:101483164
+for (i in 1:nrow(gene_data)) {
+  # Set line width for GeneID:101483164 (2x as thick)
+  lwd_value <- ifelse(gene_data$gene[i] == "LOC101483164", 4, 1)
+
+  # Plot the gene line with the appropriate color and line width
+  segments(gene_data$start[i], gene_data$y_index[i], gene_data$end[i], gene_data$y_index[i], 
+           col=gene_data$gene_color[i], lwd=lwd_value)
+}
+
+# Add gene names to the y-axis
+axis(2, at=gene_data$y_index, labels=gene_data$gene, las=1, cex.axis=0.7)
+# Add x-axis with genomic coordinates
+axis(1)
+
+# Optionally, add a grid for better visualization
+grid()
+
+# Add a legend at the top-left corner
+legend("topleft", 
+       legend=c("Positive strand (blue)", "Negative strand (red)"), 
+       fill=c("blue", "red"), 
+       bty="n", cex=0.8)
+```
+
+```{r}
+# Sort gene_data by strand and then by start position
+gene_data_sorted <- gene_data[order(gene_data$strand, gene_data$start), ]
+
+# Initialize a vector to store overlap results
+gene_data_sorted$overlap <- FALSE
+
+# Loop over genes, comparing each gene with the next one in the same strand
+for (i in 1:(nrow(gene_data_sorted) - 1)) {
+  # Check if genes are on the same strand and overlap
+  if (gene_data_sorted$strand[i] == gene_data_sorted$strand[i + 1] &&
+      gene_data_sorted$end[i] >= gene_data_sorted$start[i + 1]) {
+    gene_data_sorted$overlap[i] <- TRUE
+    gene_data_sorted$overlap[i + 1] <- TRUE
+  }
+}
+
+# View the genes with overlap information
+head(gene_data_sorted)
+
+```
+
+```{r}
+gene_data_sorted[which(gene_data_sorted$overlap==TRUE),"gene"]
+```
+
+```{r}
+region_gff[region_gff$gene == "nop58",]
+region_gff[region_gff$gene == "LOC111501074",]
+```
+
+The second one corresponds to the source `cmsearch`, so it may be the one that should be taken out.
+
+```{r}
+region_gff <- region_gff[region_gff$gene != "LOC111501074",]
+```
+
+### Define the 200 kb upstream regions
+
+Calculate the upstream start and end to prevent overlap with the previous gene.
+*This is done for each strand from the second gene onwards as we dont know where the gene before the first gene is as it is not included in the region dataframe*.
+
+- For genes on the positive strand, the upstream region starts 20kb before the gene's start position, or at the end of the previous gene if there is one.
+
+- For genes on the negative strand, the upstream region starts 20kb after the gene's end position, or at the start of the next gene if there is one.
+
+```{r}
+# Separate genes by strand
+positive_strand_genes <- region_gff[region_gff$strand == "+", ]
+negative_strand_genes <- region_gff[region_gff$strand == "-", ]
+
+# Sort genes by start position within each strand
+positive_strand_genes <- positive_strand_genes[order(positive_strand_genes$start), ]
+negative_strand_genes <- negative_strand_genes[order(negative_strand_genes$end, decreasing = TRUE), ]
+
+# Initialize the new columns for upstream start and end
+positive_strand_genes$upstream_start <- NA
+positive_strand_genes$upstream_end <- NA
+
+negative_strand_genes$upstream_start <- NA
+negative_strand_genes$upstream_end <- NA
+
+# Calculate the upstream region for positive strand genes
+for (i in 2:nrow(positive_strand_genes)) {
+  positive_strand_genes$upstream_start[i] <- max(positive_strand_genes$start[i] - 20000, positive_strand_genes$end[i-1] + 1)
+  positive_strand_genes$upstream_end[i] <- positive_strand_genes$start[i] - 1
+}
+
+# Calculate the upstream region for negative strand genes
+for (i in 2:nrow(negative_strand_genes)) {
+  negative_strand_genes$upstream_start[i] <- min(negative_strand_genes$end[i] + 20000, negative_strand_genes$start[i-1] - 1)
+  negative_strand_genes$upstream_end[i] <- negative_strand_genes$end[i] + 1
+}
+
+# Combine the results back
+region_gff <- rbind(positive_strand_genes, negative_strand_genes)
+
+# View the updated dataframe with upstream regions
+head(region_gff)
+
+```
+
+#### Chromosome lengths
+
+In case the upstream regions were to be defined for all chromosomes/scaffolds, the chromosome lengths would be needed to be the maximum for the start of the upstream region of the genes on the negative strand.
+
+Load chromosome lengths:
+
+```{r, eval = FALSE}
+gbff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gbff" 
+file_lines <- readLines(gbff_file)
+# Extract the LOCUS line
+locus_lines <- file_lines[grep("^LOCUS", file_lines)]
+head(locus_lines)
+```
+
+They don't seem to end with the dot and the number (version) as in the case of the `gff` data. Lets check if the `gff` has several versions:
+
+```{r}
+length(unique(gff$seqnames))
+# Remove version suffix from seqnames
+length(unique(sub("\\.\\d+$", "", gff$seqnames)))
+```
+
+It does not have several versions.
+
+```{r, eval = FALSE}
+# Extract chromosome names and lengths
+chrom_info <- gsub("^LOCUS\\s+(\\S+)\\s+(\\d+)\\s+.*$", "\\1,\\2", locus_lines)
+chrom_info <- strsplit(chrom_info, ",")
+chrom_names <- sapply(chrom_info, `[`, 1)
+chrom_lengths <- as.numeric(sapply(chrom_info, `[`, 2))
+# Create a named vector
+chrom_lengths <- setNames(chrom_lengths, chrom_names)
+
+# Remove version numbers from seqnames in the region-filtered gff
+region_gff$seqnames <- sub("\\.\\d+$", "", region_gff$seqnames)
+
+```
 
@@
 
-Convert the promoter and CGI data to `GRanges` objects for genomic interval operations:
+```{r}
+region_cgi <- cgif[cgif$chr==dpp4_chr,]
+```
 
 ```{r}
-# Create GRanges for promoters
-promoter_gr <- GRanges(
-  seqnames = promoters$chr,
-  ranges = IRanges(start = promoters$tss, end = promoters$tss),
-  strand = promoters$strand
+# Create a matrix of CGI intervals for faster lookup
+cgi_intervals <- cbind(region_cgi$start, region_cgi$end)
+
+# Function to efficiently check if positions are within any CGI
+is_within_cgi_vector <- function(positions, cgi_intervals) {
+  # Apply the check for each position against all intervals
+  sapply(positions, function(pos) {
+    any(pos >= cgi_intervals[,1] & pos <= cgi_intervals[,2])
+  })
+}
+
+# Initialize results list for better memory management
+results_list <- vector("list", nrow(region_gff))
+
+# Process each gene
+for (i in 1:nrow(region_gff)) {
+  upstream_start <- region_gff$upstream_start[i]
+  upstream_end <- region_gff$upstream_end[i]
+
+  # Skip if NA values
+  if (is.na(upstream_start) | is.na(upstream_end)) {
+    next
+  }
+
+  # Create sequence of positions (handling both forward and reverse cases)
+  if (upstream_start > upstream_end) {
+    positions <- seq(from = upstream_start, to = upstream_end, by = -1)
+  } else {
+    positions <- seq(from = upstream_start, to = upstream_end, by = 1)
+  }
+
+  # Get CGI status for all positions at once
+  in_cgi <- is_within_cgi_vector(positions, cgi_intervals)
+
+  # Calculate distances properly based on strand
+  distances <- if(region_gff$strand[i] == "+") {
+    # For positive strand: distance from gene start
+    region_gff$start[i] - positions  # Positive values will be upstream
+  } else {
+    # For negative strand: distance from gene end
+    positions - region_gff$end[i]    # Positive values will be upstream
+  }
+
+  # Create result data frame for this gene
+  results_list[[i]] <- data.frame(
+    gene_id = region_gff$gene[i],
+    position = positions,
+    in_cgi = in_cgi,
+    distance = distances,
+    strand = region_gff$strand[i]
   )
+}
 
-# Create GRanges for CGI
-cgi_gr <- GRanges(
-  seqnames = cgif$chr,
-  ranges = IRanges(start = cgif$start, end = cgif$end)
-)
-```
+# Combine all results efficiently
+upstream_data <- do.call(rbind, results_list[!sapply(results_list, is.null)])
 
-Use `findOverlaps` to find promoters within or near CpG islands:
+# Calculate counts and probabilities
+position_counts <- table(upstream_data$distance)  # Using table() for correct counts
+distance_seq <- as.numeric(names(position_counts))
 
-```{r}
-# Find overlaps
-overlaps <- findOverlaps(promoter_gr, cgi_gr)
-
-# Extract overlapping promoter indices
-promoters_with_cgi <- promoters[queryHits(overlaps), ]
-cgi_overlaps <- cgif[subjectHits(overlaps), ]
-
-# Combine data for output
-promoter_cgi_overlap <- cbind(promoters_with_cgi, cgi_overlaps)
-head(promoter_cgi_overlap)
-```
-
-The `promoter_cgi_overlap` dataset combines information about promoters and CpG islands (CGIs) where they overlap. It contains the following columns:
-
-- `chr`: The chromosome where the promoter is located.
-
-- `tss`: The transcription start site (TSS) position of the promoter.
-
-- `strand`: The strand of the promoter (+ or -).
-
-- `chr`: The chromosome where the CGI is located (same as the promoter in the overlap context).
-
-- `start`: The start position of the overlapping CGI.
-
-- `end`: The end position of the overlapping CGI.
-
-- `length`: The length of the CGI in base pairs.
-
-- `CpGcount`: The number of CpG dinucleotides within the CGI.
-
-- `GCcount`: The number of G and C nucleotides within the CGI.
-
-- `pctGC`: The percentage of G and C nucleotides within the CGI.
-
-- `obsExp`: The observed-to-expected ratio of CpG dinucleotides in the CGI.
-
-Each row corresponds to an overlap between a promoter and a CGI. For example:
-
-- The first row shows that a promoter at TSS 755,445 on the negative strand overlaps with a CGI on chromosome NC_036780.1 spanning positions 755,397 to 756,689.
-This CGI has a length of 1,293 bp, contains 123 CpG dinucleotides, and has a GC content of 63.19% with an observed-to-expected CpG ratio of 0.954.
-
-- The overlap is detailed for both the promoter and the CGI, including their respective properties.
-
-```{r}
-#write.table(promoter_cgi_overlap, file = "promoter_cgi_overlap.txt", sep = "\t", row.names = FALSE)
-promoter_cgi_overlap <- read.table("/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/promoter_cgi_overlap.txt")
-```
-
-### Check TSS overlap density
-
-```{r}
-plot(
-  promoter_cgi_overlap$length, 
-  promoter_cgi_overlap$pctGC, 
-  xlab = "CGI Length (bp)", 
-  ylab = "GC Content (%)", 
-  main = "Relationship Between CGI Length and GC Content",
-  pch = 16,  # Solid circle points
-  col = "blue"  # Point color
-)
+# Calculate probabilities
+cgi_probabilities <- tapply(upstream_data$in_cgi, 
+                          upstream_data$distance, 
+                          mean)   # Probability of being in CGI
 
@@
 
-```{r}
+```{r, fig.height=12, fig.width=10}
+# Set up the plot with better formatting
+par(mfrow = c(2,1))  # Set up 2 plots vertically
+par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
+# Plot 1: Probability plot
+plot(distance_seq, cgi_probabilities, 
+     xlab = "",  # Remove x-label from top plot
+     ylab = "Probability of Position Being in CGI",
+     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
+     cex.lab = 1.2,      # Larger axis labels
+     cex.axis = 1.1,     # Larger axis numbers
+     cex.main = 1.3,     # Larger title
+     las = 1,            # Horizontal axis labels
+     pch = 16,          # Solid circle points
+     ylim = c(0, 1),    # Force probability y-axis to [0,1]
+     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue
+
+# Add grid
+grid(col = "gray90", lty = "dotted")
+
+# Add a horizontal line at y = 0
+abline(h = 0, col = "gray70", lty = 2)
+
+# Plot 2: Count plot
+par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
+plot(distance_seq, position_counts, type = "h",  # Using the correct counts
+     xlab = "Distance to Gene Start (bp)",
+     ylab = "Number of Positions",
+     main = "Number of Positions at Each Distance",
+     col = "darkgreen",
+     lwd = 1,
+     ylim = c(0, max(position_counts) * 1.05))  # Force y-axis to start at 0 with 5% padding
+grid(col = "gray90", lty = "dotted")
+```
+
+```{r, fig.height=4, fig.width=15}
+# Set up plotting area with adjusted margins
+par(mfrow = c(1,1))  
+par(mar=c(5, 8, 4, 4))  # Increased right margin for legend
+
+# Create y-coordinates for each type
+y_coords <- c(1, 2, 3)  # 1 for neg strand, 2 for pos strand, 3 for CGIs
+
+# Get the x-axis range including both genes and CGIs
+x_range <- range(c(region_gff$start, region_gff$end))
+
+# Create the empty plot
+plot(x_range, c(0.5, 3.5), 
+     type="n", 
+     xlab="Genomic Coordinates",
+     ylab="",
+     yaxt="n",  # Remove y-axis ticks
+     main="Genes and CpG Islands")
+
+# Add custom y-axis labels
+axis(2, at=y_coords, 
+     labels=c("- Strand Genes", "+ Strand Genes", "Islands"),
+     las=1)  # Horizontal labels
+
+# Plot genes on negative strand
+neg_genes <- region_gff[region_gff$strand == "-", ]
+for(i in 1:nrow(neg_genes)) {
+    segments(neg_genes$start[i], y_coords[1], 
+            neg_genes$end[i], y_coords[1],
+            col="red", lwd=2)
+}
+
+# Plot genes on positive strand
+pos_genes <- region_gff[region_gff$strand == "+", ]
+for(i in 1:nrow(pos_genes)) {
+    segments(pos_genes$start[i], y_coords[2], 
+            pos_genes$end[i], y_coords[2],
+            col="blue", lwd=2)
+}
+
+# Plot CGIs
+region_cgis <- cgif[cgif$chr == dpp4_chr & 
+                   cgif$start >= region_start & 
+                   cgif$end <= region_end,]
+
+for(i in 1:nrow(region_cgis)) {
+    segments(region_cgis$start[i], y_coords[3], 
+            region_cgis$end[i], y_coords[3],
+            col="darkgreen", lwd=2)
+}
+
+# Add grid
+grid(col="gray90")
 
 ```
+
+# References

### 2025-01-23 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
Updating last changes to be able to pull
Updating changes in archipelago-definition

--- 2025-01-10 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
@@
 
-# CFTP
+```{r setup}
+#library(MethEvolSIM)
+library(devtools)
+load_all()
+```
+
+**Update way in which MethEvolSIM is loaded**
+
+# Introduction
 
@@
 
-## Implementation
+**Add examples of use**
 
-Currently neglects IWEs but allows to sample a sequence of methylation states from the equilibrium given by the equilibrium frequencies of the genomic regions, the correlated and independent SSE.
+CFTP method is defaulted to FALSE but can be set to TRUE when simulating initial data or data evolution along a tree.
 
-## Test
+Sample initial data with the sequence of methylation states according to the equilibrium.
 
-This workflow uses [Snakemake](https://snakemake.readthedocs.io) to manage the steps to test that the CFTP method samples a sequence of methylation states from the equilibrium given by the model accounting for the correlated SSE process (while neglecting the IWE process).
+```{r}
+# Example with sampled initial methylation frequencies and default parameter values
+custom_infoStr <- data.frame(n = c(100, 100, 100),
+                             globalState = c("M", "U", "M"))
+initialD <- simulate_initialData(infoStr = custom_infoStr, CFTP = TRUE)
+```
 
-For that it samples a number of parameter combinations and for each it simulates initial data for a number of replicates.
-It uses as summary statistics the mean correlation of a central segment of methylation states per genomic structure type.
-We expect that the value of this summary statistics increases with the increase of evolutionary time until it converges towards the equilibrium given by the corresponding parameter combination.
-We track its value by discretizing the evolution along the branch in small steps (length as config.yaml file value for `branch_length`).
-Then, we call the CFTP method on the initial data to sample the sequence of methylation states from the equilibrium and compute its corresponding value for the summary statistics.
+Simulate evolution of data along a tree calling the CFTP method to sample a sequence of methylation states at the root of the tree from the equilibrium.
+
+```{r}
+# Example with customized initial methylation frequencies, customized parameter values 
+# and default dt (0.01)
+tree <- "((a:1, b:1):2, c:2, (d:3.7, (e:4, f:1):3):5);"
+custom_infoStr <- data.frame(n = c(10, 10, 10),
+                             globalState = c("M", "U", "M"),
+                             u_eqFreq = c(0.1, 0.8, 0.1),
+                             p_eqFreq = c(0.1, 0.1, 0.1),
+                             m_eqFreq = c(0.8, 0.1, 0.8))
+custom_params <- get_parameterValues()
+custom_params$mu <- 0.005
+
+evolD <- simulate_evolData(infoStr = custom_infoStr, tree = tree, params = custom_params, CFTP = TRUE)
+```
+
+## Visual Examples
+
+To ease understanding of how applying CFTP results in sampling an initial sequence of methylation states from the equilibrium towards which the data would evolve when initial data neglects local correlations, we provide the plots obtained while testing the method.
+The test was performed on 10 parameter combinations as the model equilibrium is dependent on those.
+It uses as summary statistics the mean correlation of a central segment of methylation states per genomic structure type (see vignette for summary statistics **add link or reference or exact name here**).
+When simulating without using CFTP, the initial sequence of methylation states is given by the independent sampling of the methylation state of the sites from the equilibrium frequencies of the genomic structure to which they belong.
+Then, the local correlations between the sites methylation states build up with evolutionary time.
+
+Therefore, we expect that the value of the increases with the increase of evolutionary time until it converges towards the equilibrium given by the corresponding parameter combination.
+We track its value by discretizing the evolution along a tree branch (representing linear time) in small steps (of length 0.01).
+Additionally, we call the CFTP method on the initial data to sample the sequence of methylation states from the equilibrium and compute its corresponding value for the summary statistics.
 We expect the distribution of the summary statistics obtained by calling the CFTP method to be representative of the equilibrium obtained after the branch evolution has reached the equilibrium.
 
-**Add summary and graphical image**
+In the following figures, the y axis represents the value of the mean correlation summary statistics and the x axis represents the number of discretized time steps.
+Each figure corresponds to one of the 10 sampled parameter combinations and for each parameter combination 10 replicates were simulated, represented with the different colors.
+Dots (with a different shape for each of the replicates) represent the value of the mean correlation at the given time step and the dashed line represents its value after calling CFTP on the initial data.
 
-### Requirements
+![Sampled parameter combination 1](Figures/CFTP_testConvergence_paramsID_01.png)
+
+![Sampled parameter combination 2](Figures/CFTP_testConvergence_paramsID_02.png)
+
+![Sampled parameter combination 3](Figures/CFTP_testConvergence_paramsID_03.png)
+
+![Sampled parameter combination 4](Figures/CFTP_testConvergence_paramsID_04.png)
+
+![Sampled parameter combination 5](Figures/CFTP_testConvergence_paramsID_05.png)
+
+![Sampled parameter combination 6](Figures/CFTP_testConvergence_paramsID_06.png)
+
+![Sampled parameter combination 7](Figures/CFTP_testConvergence_paramsID_07.png)
+
+![Sampled parameter combination 8](Figures/CFTP_testConvergence_paramsID_08.png)
+
+![Sampled parameter combination 9](Figures/CFTP_testConvergence_paramsID_09.png)
+
+![Sampled parameter combination 10](Figures/CFTP_testConvergence_paramsID_10.png)
+
+### Reproducibility
+
+The test used to generate the visual example is provided as a workflow, which can be run using [Snakemake](https://snakemake.readthedocs.io) or by running the scripts separatedly.
+Here, we describe the structure and content of the Snakemake workflow.
+However, the scripts that comprise the workflow are available under `extensive_tests/CFTP_convergence/scripts`.
+
+#### Workflow Requirements
 
@@
 
-When using nohup one needs to activate the environment in the command.
+When using nohup one can activate the environment as in the following command.
 
@@
 
-## Steps in the Workflow
+#### Steps in the Workflow
 
@@
 The output files are written under a subdirectory `Figures` generated in the same directory of the snakemake workflow, and are named "CFTP_testConvergence_paramsID_{parameter_combination_number}" both in `.pdf` and `.png` format.
-
-## Results
-
-```{r, out.width="0.8\\linewidth", include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
-knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_01.pdf")
-```

### 2025-01-27 application_examples/abc_inferences/archipelago-definition.Rmd
Updating Rmd file for archipelago definition

@@
 date: "`r Sys.Date()`"
-output: html_document
+output:
+  html_document:
+    toc: true
+    toc_float: true
+    toc_depth: 3
 bibliography: references.bib
@@
 
-```{r, warning=FALSE, message=FALSE}
+```{r load libraries, warning=FALSE, message=FALSE, echo=FALSE}
 # If missing, install required packages
@@
 
-```{r}
+```{r set region values}
 dpp4_chr <- "NC_036801.1"
@@
 
-```{r}
+```{r load gff}
 # Load the GFF file
@@
 gff <- as.data.frame(import(gff_file))
-
+names(gff)
 ```
@@
 
-```{r}
+Lets keep the relevant columns:
+
+```{r gff filter relevant columns}
+gff <- gff[,c("seqnames", "start", "end", "width", "strand", "type", "source", "ID", "Dbxref", "gene")]
+```
+
+```{r gff table of annotation type}
+table(gff$type)
+```
+
+```{r gff table of annotation source}
 table(gff$source)
@@
 
-### Filter region of interest
+### Filter genes of whole genome
+
+First, we filter the GFF file to keep only genes.
+```{r filter gff genome}
+genome_gff <- gff[
+  gff$type == "gene", 
+]
+```
+
+We can check the source of the filtered annotations:
+```{r gff_genome table of annotation source}
+table(genome_gff$source)
+```
+
+And the number of genes:
+
+```{r gff_genome number of genes}
+nrow(genome_gff)
+length(unique(genome_gff$ID))
+```
+
+We order the entries:
+
+```{r order genome_gff}
+genome_gff <- genome_gff[order(
+  genome_gff$seqnames,                     # First, order by chromosome
+  genome_gff$strand,                       # Then, order by strand
+  ifelse(genome_gff$strand == "+",         # If strand is "+", order by start
+         genome_gff$start, 
+         -genome_gff$end)                  # If strand is "-", order by end in descending order
+), ]
+
+# View the first few rows of the ordered dataframe by strand
+head(genome_gff[genome_gff$strand == "+",])
+head(genome_gff[genome_gff$strand == "-",])
+```
+
+Now, we can filter to keep only genes that are not overlapping with each other. We first look for genes that overlap:
 
 ```{r}
-region_gff <- gff[
-  gff$seqnames == "NC_036801.1" & 
-  gff$type == "gene" & 
-  gff$start <= region_end & 
-  gff$end >= region_start, 
-]
+# Initialize a vector to store overlap results
+genome_gff$overlap <- FALSE
 
-head(region_gff)
+# Loop over genes, comparing each gene with the next one in the same strand
+for (i in 1:(nrow(genome_gff) - 1)) {
+
+  # Check if genes are on the same chromosome/scaffold and same strand
+  if (genome_gff$seqnames[i] == genome_gff$seqnames[i + 1] &&
+      genome_gff$strand[i] == genome_gff$strand[i + 1]){
+
+    # Check overlap if strand is positive
+    if (genome_gff$strand[i] == "+" &&
+        genome_gff$end[i] >= genome_gff$start[i+1]) {
+      genome_gff$overlap[i] <- TRUE
+      genome_gff$overlap[i+1] <- TRUE
+    }
+
+     # Check overlap if strand is negative
+    if (genome_gff$strand[i] == "-" &&
+       genome_gff$start[i] <= genome_gff$end[i+1]) {
+      genome_gff$overlap[i] <- TRUE
+      genome_gff$overlap[i+1] <- TRUE
+    }
+  }
+
+}
+
+# Number of overlapping genes
+nrow(genome_gff[genome_gff$overlap == TRUE,])
 ```
 
+We remove the overlapping genes from `genome_gff` and check the number of remaining genes:
+
+```{r gff_genome remove overlap genes}
+# Remove overlapping genes from genome_gff
+genome_gff <- genome_gff[!genome_gff$overlap, ]
+nrow(genome_gff)
+length(unique(genome_gff$ID))
+```
+
+### Chromosome lengths
+
+To define the upstream regions the chromosome lengths are needed to be the maximum for the start of the upstream region of the genes on the negative strand.
+
+Load chromosome lengths:
+
+```{r load gbff}
+gbff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gbff" 
+file_lines <- readLines(gbff_file)
+# Extract the LOCUS line
+locus_lines <- file_lines[grep("^LOCUS", file_lines)]
+head(locus_lines)
+```
+
+They don't seem to end with the dot and the number (version) as in the case of the `gff` data. Lets check if the `gff` has several versions:
+
+```{r explore genome_gff versions}
+length(unique(genome_gff$seqnames))
+# Remove version suffix from seqnames
+length(unique(sub("\\.\\d+$", "", genome_gff$seqnames)))
+```
+
+It does not have several versions, so we can remove the version numbers from `seqnames` in the `genome_gff` and from variable `dpp4_chr`.
+
+```{r remove version from seqnames}
+# Remove version numbers from seqnames in the filtered files
+genome_gff$seqnames <- sub("\\.\\d+$", "", genome_gff$seqnames)
+dpp4_chr <- "NC_036801"
+```
+
+```{r extract chrom names and lengths}
+# Extract chromosome names and lengths
+chrom_info <- gsub("^LOCUS\\s+(\\S+)\\s+(\\d+)\\s+.*$", "\\1,\\2", locus_lines)
+rm(locus_lines)
+chrom_info <- strsplit(chrom_info, ",")
+chrom_names <- sapply(chrom_info, `[`, 1)
+chrom_lengths <- as.numeric(sapply(chrom_info, `[`, 2))
+# Create a named vector
+chrom_lengths <- setNames(chrom_lengths, chrom_names)
+```
+
+```{r check correspondence genome_gff and chrom_lengths}
+# Check how many chromosomes are in chrom_names but not in genome_gff$seqnames
+length(setdiff(chrom_names, unique(genome_gff$seqnames)))
+
+# You can also check the reverse - how many chromosomes in genome_gff$seqnames are not in chrom_names
+length(setdiff(unique(genome_gff$seqnames), chrom_names))
+```
+
+### Define the 200 kb upstream genes of whole genome
+
 ```{r}
+# Initialize the new columns for upstream start and end in genome_gff
+genome_gff$upstream_start <- NA
+genome_gff$upstream_end <- NA
+genome_gff$upstream_length <- NA
+
+# Process each chromosome separately
+for (chr in unique(genome_gff$seqnames)) {
+  chr_length <- chrom_lengths[chr]
+  chr_genes <- genome_gff[genome_gff$seqnames == chr, ]
+
+  ### Process positive strand genes ###
+
+  pos_genes <- chr_genes[chr_genes$strand == "+", ]
+  if (nrow(pos_genes) > 0) {
+
+    # First gene in chromosome (positive strand)
+    if (pos_genes$start[1] > 1){
+      pos_genes$upstream_end[1] <- pos_genes$start[1] - 1
+      pos_genes$upstream_start[1] <- max(1, pos_genes$start[1] - 20000)
+      pos_genes$upstream_length[1] <- pos_genes$upstream_end[1] - pos_genes$upstream_start[1] + 1
+
+      # Only keep upstream region if it's valid (at least 1 base)
+      if (pos_genes$upstream_end[1] - pos_genes$upstream_start[1] < 0){
+        pos_genes$upstream_start[1] <- NA
+        pos_genes$upstream_end[1] <- NA
+        pos_genes$upstream_length[1] <- NA
+      }
+    }
+
+    # Remaining genes
+    if (nrow(pos_genes) > 1) {
+      for (i in 2:nrow(pos_genes)) {
+        pos_genes$upstream_end[i] <- pos_genes$start[i] - 1
+        pos_genes$upstream_start[i] <- max(pos_genes$start[i] - 20000, pos_genes$end[i-1] + 1)
+        pos_genes$upstream_length[i] <- pos_genes$upstream_end[i] - pos_genes$upstream_start[i] + 1
+
+        # Only keep upstream region if it's valid (at least 1 base)
+        if (pos_genes$upstream_end[i] - pos_genes$upstream_start[i] < 0){
+          pos_genes$upstream_start[i] <- NA
+          pos_genes$upstream_end[i] <- NA
+          pos_genes$upstream_length[i] <- NA
+        }
+      }
+    }
+  }
+
+  ### Process negative strand genes ###
+  neg_genes <- chr_genes[chr_genes$strand == "-", ]
+  if (nrow(neg_genes) > 0) {
+
+    # First gene in chromosome (negative strand)
+    if (neg_genes$end[1] < chr_length) {
+      neg_genes$upstream_start[1] <- min(chr_length, neg_genes$end[1] + 20000)
+      neg_genes$upstream_end[1] <- neg_genes$end[1] + 1
+      neg_genes$upstream_length[1] <- neg_genes$upstream_start[1] - neg_genes$upstream_end[1] + 1
+
+      # Only keep upstream region if it's valid (at least 1 base)
+      if (neg_genes$upstream_start[1] - neg_genes$upstream_end[1] < 0) {
+        neg_genes$upstream_start[1] <- NA
+        neg_genes$upstream_end[1] <- NA
+        neg_genes$upstream_length[1] <- NA
+      }
+    }
+
+    # Remaining genes
+    if (nrow(neg_genes) > 1) {
+      for (i in 2:nrow(neg_genes)) {
+        neg_genes$upstream_start[i] <- min(neg_genes$end[i] + 20000, neg_genes$start[i-1] - 1)
+        neg_genes$upstream_end[i] <- neg_genes$end[i] + 1
+        neg_genes$upstream_length[i] <- neg_genes$upstream_start[i] - neg_genes$upstream_end[i] + 1
+
+        # Only keep upstream region if it's valid (at least 1 base)
+        if (neg_genes$upstream_start[i] - neg_genes$upstream_end[i] < 0) {
+          neg_genes$upstream_start[i] <- NA
+          neg_genes$upstream_end[i] <- NA
+          neg_genes$upstream_length[i] <- NA
+        }
+      }
+    }
+  }
+
+  # Update the main genome_gff dataframe
+  if(nrow(pos_genes)>0){
+    genome_gff[genome_gff$seqnames == chr & genome_gff$strand == "+", c("upstream_start", "upstream_end", "upstream_length")] <- 
+    pos_genes[, c("upstream_start", "upstream_end", "upstream_length")]
+  }
+
+  if(nrow(neg_genes)>0){
+    genome_gff[genome_gff$seqnames == chr & genome_gff$strand == "-", c("upstream_start", "upstream_end", "upstream_length")] <- 
+    neg_genes[, c("upstream_start", "upstream_end", "upstream_length")]
+  }
+
+}
+
+# Check the results
+summary(genome_gff$upstream_length)
+```
+
+#### Subset chromosome of interest
+
+```{r filter gff chromosome alternative}
+chromosome_gff <- genome_gff[
+  genome_gff$seqnames == dpp4_chr,
+]
+```
+
+We can check the source of the filtered annotations:
+```{r gff_chromosome table of annotation source}
+table(chromosome_gff$source)
+```
+
+Check the number of remaining genes:
+
+```{r gff_chromosome remaining genes}
+nrow(chromosome_gff)
+length(unique(chromosome_gff$ID))
+```
+
+Check the summary of the upstream lengths:
+
+```{r}
+summary(chromosome_gff$upstream_length)
+```
+
+#### Subset region of interest
+
+```{r filter gff region alternative}
+region_gff <- chromosome_gff[
+  chromosome_gff$start <= region_end & 
+  chromosome_gff$end >= region_start, 
+]
+```
+
+We can check the source of the filtered annotations:
+```{r gff_region table of annotation source}
 table(region_gff$source)
@@
 
-```{r}
-region_gff[grep("GeneID:101483164", region_gff$Dbxref), ]
+Check the number of remaining genes:
+
+```{r gff_region remaining genes}
+nrow(region_gff)
+length(unique(region_gff$ID))
 ```
 
-```{r}
-length(unique(region_gff$ID))
-unique(region_gff$ID)
-```
+We can visualize the genes on the region of interest:
 
-```{r}
+```{r gff_region plot}
 # Extract relevant columns
@@
 
-```
-
-```{r}
 # Set up larger plotting area and adjust margins
@@
 # Add a legend at the top-left corner
-legend("topleft", 
+legend("topright", 
        legend=c("Positive strand (blue)", "Negative strand (red)"), 
@@
 
-```{r}
-# Sort gene_data by strand and then by start position
-gene_data_sorted <- gene_data[order(gene_data$strand, gene_data$start), ]
+Check the summary of the upstream lengths:
 
-# Initialize a vector to store overlap results
-gene_data_sorted$overlap <- FALSE
-
-# Loop over genes, comparing each gene with the next one in the same strand
-for (i in 1:(nrow(gene_data_sorted) - 1)) {
-  # Check if genes are on the same strand and overlap
-  if (gene_data_sorted$strand[i] == gene_data_sorted$strand[i + 1] &&
-      gene_data_sorted$end[i] >= gene_data_sorted$start[i + 1]) {
-    gene_data_sorted$overlap[i] <- TRUE
-    gene_data_sorted$overlap[i + 1] <- TRUE
-  }
-}
-
-# View the genes with overlap information
-head(gene_data_sorted)
-
-```
-
-```{r}
-gene_data_sorted[which(gene_data_sorted$overlap==TRUE),"gene"]
-```
-
-```{r}
-region_gff[region_gff$gene == "nop58",]
-region_gff[region_gff$gene == "LOC111501074",]
-```
-
-The second one corresponds to the source `cmsearch`, so it may be the one that should be taken out.
-
-```{r}
-region_gff <- region_gff[region_gff$gene != "LOC111501074",]
-```
-
-### Define the 200 kb upstream regions
-
-Calculate the upstream start and end to prevent overlap with the previous gene.
-*This is done for each strand from the second gene onwards as we dont know where the gene before the first gene is as it is not included in the region dataframe*.
-
-- For genes on the positive strand, the upstream region starts 20kb before the gene's start position, or at the end of the previous gene if there is one.
-
-- For genes on the negative strand, the upstream region starts 20kb after the gene's end position, or at the start of the next gene if there is one.
-
-```{r}
-# Separate genes by strand
-positive_strand_genes <- region_gff[region_gff$strand == "+", ]
-negative_strand_genes <- region_gff[region_gff$strand == "-", ]
-
-# Sort genes by start position within each strand
-positive_strand_genes <- positive_strand_genes[order(positive_strand_genes$start), ]
-negative_strand_genes <- negative_strand_genes[order(negative_strand_genes$end, decreasing = TRUE), ]
-
-# Initialize the new columns for upstream start and end
-positive_strand_genes$upstream_start <- NA
-positive_strand_genes$upstream_end <- NA
-
-negative_strand_genes$upstream_start <- NA
-negative_strand_genes$upstream_end <- NA
-
-# Calculate the upstream region for positive strand genes
-for (i in 2:nrow(positive_strand_genes)) {
-  positive_strand_genes$upstream_start[i] <- max(positive_strand_genes$start[i] - 20000, positive_strand_genes$end[i-1] + 1)
-  positive_strand_genes$upstream_end[i] <- positive_strand_genes$start[i] - 1
-}
-
-# Calculate the upstream region for negative strand genes
-for (i in 2:nrow(negative_strand_genes)) {
-  negative_strand_genes$upstream_start[i] <- min(negative_strand_genes$end[i] + 20000, negative_strand_genes$start[i-1] - 1)
-  negative_strand_genes$upstream_end[i] <- negative_strand_genes$end[i] + 1
-}
-
-# Combine the results back
-region_gff <- rbind(positive_strand_genes, negative_strand_genes)
-
-# View the updated dataframe with upstream regions
-head(region_gff)
-
-```
-
-#### Chromosome lengths
-
-In case the upstream regions were to be defined for all chromosomes/scaffolds, the chromosome lengths would be needed to be the maximum for the start of the upstream region of the genes on the negative strand.
-
-Load chromosome lengths:
-
-```{r, eval = FALSE}
-gbff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gbff" 
-file_lines <- readLines(gbff_file)
-# Extract the LOCUS line
-locus_lines <- file_lines[grep("^LOCUS", file_lines)]
-head(locus_lines)
-```
-
-They don't seem to end with the dot and the number (version) as in the case of the `gff` data. Lets check if the `gff` has several versions:
-
-```{r}
-length(unique(gff$seqnames))
-# Remove version suffix from seqnames
-length(unique(sub("\\.\\d+$", "", gff$seqnames)))
-```
-
-It does not have several versions.
-
-```{r, eval = FALSE}
-# Extract chromosome names and lengths
-chrom_info <- gsub("^LOCUS\\s+(\\S+)\\s+(\\d+)\\s+.*$", "\\1,\\2", locus_lines)
-chrom_info <- strsplit(chrom_info, ",")
-chrom_names <- sapply(chrom_info, `[`, 1)
-chrom_lengths <- as.numeric(sapply(chrom_info, `[`, 2))
-# Create a named vector
-chrom_lengths <- setNames(chrom_lengths, chrom_names)
-
-# Remove version numbers from seqnames in the region-filtered gff
-region_gff$seqnames <- sub("\\.\\d+$", "", region_gff$seqnames)
-
+```{r summary of region upstream length}
+summary(region_gff$upstream_length)
 ```
@@
 
-```{r}
+```{r load cgi annotation file}
 # Load the CGI annotation file
@@
 ```
+Lets remove the chromosome version number from the CGI annotation:
 
 ```{r}
-region_cgi <- cgif[cgif$chr==dpp4_chr,]
+cgif$chr <- sub("\\.\\d+$", "", cgif$chr)
 ```
 
-```{r}
-# Create a matrix of CGI intervals for faster lookup
-cgi_intervals <- cbind(region_cgi$start, region_cgi$end)
+Check the summary of the island lengths:
 
-# Function to efficiently check if positions are within any CGI
-is_within_cgi_vector <- function(positions, cgi_intervals) {
-  # Apply the check for each position against all intervals
-  sapply(positions, function(pos) {
-    any(pos >= cgi_intervals[,1] & pos <= cgi_intervals[,2])
-  })
-}
-
-# Initialize results list for better memory management
-results_list <- vector("list", nrow(region_gff))
-
-# Process each gene
-for (i in 1:nrow(region_gff)) {
-  upstream_start <- region_gff$upstream_start[i]
-  upstream_end <- region_gff$upstream_end[i]
-
-  # Skip if NA values
-  if (is.na(upstream_start) | is.na(upstream_end)) {
-    next
-  }
-
-  # Create sequence of positions (handling both forward and reverse cases)
-  if (upstream_start > upstream_end) {
-    positions <- seq(from = upstream_start, to = upstream_end, by = -1)
-  } else {
-    positions <- seq(from = upstream_start, to = upstream_end, by = 1)
-  }
-
-  # Get CGI status for all positions at once
-  in_cgi <- is_within_cgi_vector(positions, cgi_intervals)
-
-  # Calculate distances properly based on strand
-  distances <- if(region_gff$strand[i] == "+") {
-    # For positive strand: distance from gene start
-    region_gff$start[i] - positions  # Positive values will be upstream
-  } else {
-    # For negative strand: distance from gene end
-    positions - region_gff$end[i]    # Positive values will be upstream
-  }
-
-  # Create result data frame for this gene
-  results_list[[i]] <- data.frame(
-    gene_id = region_gff$gene[i],
-    position = positions,
-    in_cgi = in_cgi,
-    distance = distances,
-    strand = region_gff$strand[i]
-  )
-}
-
-# Combine all results efficiently
-upstream_data <- do.call(rbind, results_list[!sapply(results_list, is.null)])
-
-# Calculate counts and probabilities
-position_counts <- table(upstream_data$distance)  # Using table() for correct counts
-distance_seq <- as.numeric(names(position_counts))
-
-# Calculate probabilities
-cgi_probabilities <- tapply(upstream_data$in_cgi, 
-                          upstream_data$distance, 
-                          mean)   # Probability of being in CGI
-
+```{r summary of cgi lengths}
+summary(cgif$length)
 ```
 
-```{r, fig.height=12, fig.width=10}
-# Set up the plot with better formatting
-par(mfrow = c(2,1))  # Set up 2 plots vertically
-par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
-# Plot 1: Probability plot
-plot(distance_seq, cgi_probabilities, 
-     xlab = "",  # Remove x-label from top plot
-     ylab = "Probability of Position Being in CGI",
-     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
-     cex.lab = 1.2,      # Larger axis labels
-     cex.axis = 1.1,     # Larger axis numbers
-     cex.main = 1.3,     # Larger title
-     las = 1,            # Horizontal axis labels
-     pch = 16,          # Solid circle points
-     ylim = c(0, 1),    # Force probability y-axis to [0,1]
-     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue
+Not all chromosomes in `genome_gff` are in the CGI annotations:
 
-# Add grid
-grid(col = "gray90", lty = "dotted")
+```{r}
+# Identify seqnames in genome_gff that are not in cgif
+missing_seqnames <- setdiff(unique(genome_gff$seqnames), unique(cgif$chr))
 
-# Add a horizontal line at y = 0
-abline(h = 0, col = "gray70", lty = 2)
+# Print the missing seqnames
+print(missing_seqnames)
+```
 
-# Plot 2: Count plot
-par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
-plot(distance_seq, position_counts, type = "h",  # Using the correct counts
-     xlab = "Distance to Gene Start (bp)",
-     ylab = "Number of Positions",
-     main = "Number of Positions at Each Distance",
-     col = "darkgreen",
-     lwd = 1,
-     ylim = c(0, max(position_counts) * 1.05))  # Force y-axis to start at 0 with 5% padding
-grid(col = "gray90", lty = "dotted")
+To check when extracting chromosome info.
+```{r}
+nrow(cgif[cgif$chr==missing_seqnames[1],])
 ```
+
+Lets visualize the islands in the genomic region of interest:
 
@@
 
+For each position within each upstream region we check whether it is in a CGI and the distance to the gene.
+
+```{r, eval = FALSE}
+# Function to efficiently check if positions are within any CGI
+is_within_cgi_vector <- function(positions, cgi_intervals) {
+  # Apply the check for each position against all intervals
+  sapply(positions, function(pos) {
+    any(pos >= cgi_intervals[,1] & pos <= cgi_intervals[,2])
+  })
+}
+
+# Initialize results list
+#results_list <- vector("list", nrow(region_gff))
+## TODO: Update comment
+cgiVsDistance <- matrix(data = NA, nrow = nrow(genome_gff), ncol = 20000)
+
+# Get the genome_gff indices for genes with a defined upstream region
+nonNA_upstream <- which(!is.na(genome_gff$upstream_length))
+
+# Set start variable for previous chromosome name
+prev_chrom_name <- 0
+
+for (i in nonNA_upstream){
+  # Get current chromosome name
+  chrom_name <- genome_gff$seqnames[i]
+  # Get the cgi annotations for the current chromosome
+  if (prev_chrom_name != chrom_name){
+    chrom_cgi <- cgif[cgif$chr==chrom_name,]
+    # Create a matrix of CGI intervals for faster lookup
+    chrom_cgi_intervals <- cbind(chrom_cgi$start, chrom_cgi$end)
+  }
+
+  # Get the sequence of upstream positions
+  upstream_start <- genome_gff$upstream_start[i]
+  upstream_end <- genome_gff$upstream_end[i]
+  if (upstream_start > upstream_end) { # Upstream in the negative strand
+    positions <- seq(from = upstream_end, to = upstream_start, by = 1)
+  } else { # Upstream in the positive strand
+    positions <- seq(from = upstream_start, to = upstream_end, by = 1)
+  }
+
+  # Get CGI status for all positions at once
+  in_cgi <- is_within_cgi_vector(positions, chrom_cgi_intervals)
+
+  # Get the sequence of upstream distances to the gene start
+  if(genome_gff$strand[i] == "+") {
+    # For positive strand: distance from gene start
+    distances <- genome_gff$start[i] - positions
+  } else {
+    # For negative strand: distance from gene end
+    distances <- positions - genome_gff$end[i]
+  }
+
+  # Save the CGI status in the distance matrix
+  cgiVsDistance[i,distances] <- in_cgi
+
+  # Update the variable for the previous chromosome name
+  prev_chrom_name <- chrom_name
+}
+```
+
+```{r, eval=FALSE}
+save(cgiVsDistance, file = "cgiVsDistance.RData")
+```
+
+```{r}
+load("cgiVsDistance.RData")
+```
+
+Now we can calculate the probability of each promoter site (described by the distance to the gene) to be in a CGI:
+
+```{r}
+cgiProbByDistance <- colMeans(cgiVsDistance, na.rm = TRUE)
+distCounts <- colSums(!is.na(cgiVsDistance))
+```
+
+```{r, fig.height=12, fig.width=10}
+# Set up the plot with better formatting
+par(mfrow = c(2,1))  # Set up 2 plots vertically
+par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
+# Plot 1: Probability plot
+plot(1:length(cgiProbByDistance), cgiProbByDistance, 
+     xlab = "Distance of Upstream Positions to Gene",  
+     ylab = "Probability of Position Being in CGI",
+     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
+     cex.lab = 1.2,      # Larger axis labels
+     cex.axis = 1.1,     # Larger axis numbers
+     cex.main = 1.3,     # Larger title
+     las = 1,            # Horizontal axis labels
+     pch = 16,          # Solid circle points
+     ylim = c(0, 1),    # Force probability y-axis to [0,1]
+     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue
+
+# Add grid
+grid(col = "gray90", lty = "dotted")
+
+# Add a horizontal line at y = 0
+abline(h = 0, col = "gray70", lty = 2)
+
+# Plot 2: Count plot
+par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
+plot(1:length(distCounts), distCounts, type = "h",  # Using the correct counts
+     xlab = "Distance to Gene Start (bp)",
+     ylab = "Number of Positions",
+     main = "Number of Positions at Each Distance",
+     col = "darkgreen",
+     lwd = 1,
+     ylim = c(0, max(distCounts) * 1.05))  # Force y-axis to start at 0 with 5% padding
+grid(col = "gray90", lty = "dotted")
+```
+
+#### Subset chromosome of interest
+
+```{r}
+cgiProbByDistance <- colMeans(cgiVsDistance[which(genome_gff$seqnames == dpp4_chr),], na.rm = TRUE)
+distCounts <- colSums(!is.na(cgiVsDistance[which(genome_gff$seqnames == dpp4_chr),]))
+```
+
+```{r, fig.height=12, fig.width=10}
+# Set up the plot with better formatting
+par(mfrow = c(2,1))  # Set up 2 plots vertically
+par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
+# Plot 1: Probability plot
+plot(1:length(cgiProbByDistance), cgiProbByDistance, 
+     xlab = "Distance of Upstream Positions to Gene",  
+     ylab = "Probability of Position Being in CGI",
+     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
+     cex.lab = 1.2,      # Larger axis labels
+     cex.axis = 1.1,     # Larger axis numbers
+     cex.main = 1.3,     # Larger title
+     las = 1,            # Horizontal axis labels
+     pch = 16,          # Solid circle points
+     ylim = c(0, 1),    # Force probability y-axis to [0,1]
+     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue
+
+# Add grid
+grid(col = "gray90", lty = "dotted")
+
+# Add a horizontal line at y = 0
+abline(h = 0, col = "gray70", lty = 2)
+
+# Plot 2: Count plot
+par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
+plot(1:length(distCounts), distCounts, type = "h",  # Using the correct counts
+     xlab = "Distance to Gene Start (bp)",
+     ylab = "Number of Positions",
+     main = "Number of Positions at Each Distance",
+     col = "darkgreen",
+     lwd = 1,
+     ylim = c(0, max(distCounts) * 1.05))  # Force y-axis to start at 0 with 5% padding
+grid(col = "gray90", lty = "dotted")
+```
+
+#### Subset region of interest
+
+```{r}
+region_indeces <- genome_gff$seqnames == dpp4_chr & genome_gff$start <= region_end & genome_gff$end >= region_start
+cgiProbByDistance <- colMeans(cgiVsDistance[region_indeces,], na.rm = TRUE)
+distCounts <- colSums(!is.na(cgiVsDistance[region_indeces,]))
+```
+
+```{r, fig.height=12, fig.width=10}
+# Set up the plot with better formatting
+par(mfrow = c(2,1))  # Set up 2 plots vertically
+par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
+# Plot 1: Probability plot
+plot(1:length(cgiProbByDistance), cgiProbByDistance, 
+     xlab = "Distance of Upstream Positions to Gene",  
+     ylab = "Probability of Position Being in CGI",
+     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
+     cex.lab = 1.2,      # Larger axis labels
+     cex.axis = 1.1,     # Larger axis numbers
+     cex.main = 1.3,     # Larger title
+     las = 1,            # Horizontal axis labels
+     pch = 16,          # Solid circle points
+     ylim = c(0, 1),    # Force probability y-axis to [0,1]
+     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue
+
+# Add grid
+grid(col = "gray90", lty = "dotted")
+
+# Add a horizontal line at y = 0
+abline(h = 0, col = "gray70", lty = 2)
+
+# Plot 2: Count plot
+par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
+plot(1:length(distCounts), distCounts, type = "h",  # Using the correct counts
+     xlab = "Distance to Gene Start (bp)",
+     ylab = "Number of Positions",
+     main = "Number of Positions at Each Distance",
+     col = "darkgreen",
+     lwd = 1,
+     ylim = c(0, max(distCounts) * 1.05))  # Force y-axis to start at 0 with 5% padding
+grid(col = "gray90", lty = "dotted")
+```
+
 # References

### 2025-01-31 application_examples/abc_inferences/archipelago-definition.Rmd
Updating state of branch develop to checkout to cftp

@@
 
-# Initialize results list
-#results_list <- vector("list", nrow(region_gff))
-## TODO: Update comment
+# Initialize result dataframe
 cgiVsDistance <- matrix(data = NA, nrow = nrow(genome_gff), ncol = 20000)
@@
 
+## Step 3: Island assignation to closer gene
+
+Neglecting those islands that intersect gene bodies, we get the overlaps of CGI with upstream regions:
+
+```{r}
+library(GenomicRanges)
+
+### Step 1: Filter CGI not within gene bodies ###
+
+# Create GRanges for two sets with different chromosomes
+gr_cgi <- GRanges(seqnames = cgif$chr,
+                  ranges = IRanges(start = cgif$start,
+                                   end = cgif$end))
+
+gr_genes <- GRanges(seqnames = genome_gff$seqnames,
+                    ranges = IRanges(start = genome_gff$start,
+                                     end = genome_gff$end))
+
+# Subset common chromosomes (seqnames)
+common_seqnames <- intersect(seqlevels(gr_cgi), seqlevels(gr_genes))
+gr_cgi <- keepSeqlevels(gr_cgi, common_seqnames, pruning.mode = "coarse")
+gr_genes <- keepSeqlevels(gr_genes, common_seqnames, pruning.mode = "coarse")
+
+# Find overlaps
+overlap_result <- overlapsAny(gr_cgi, gr_genes)
+
+# Subset non-overlapping CG islands
+non_overlapping_cgi <- gr_cgi[!overlap_result]
+
+### Step 2: Find overlaps of CGI with upstream regions ###
+
+# Filter out rows with NA in upstream_start or upstream_end 
+genome_gff_clean <- genome_gff[!is.na(genome_gff$upstream_start) & !is.na(genome_gff$upstream_end), ]
+
+# Rename upstream_start and upstream_end from the "-" strand as in GRanges objects
+# each range must have an end that is greater or equal to its start minus one
+inverted_start <- genome_gff_clean$upstream_end[genome_gff_clean$strand == "-"]
+inverted_end <- genome_gff_clean$upstream_start[genome_gff_clean$strand == "-"]
+genome_gff_clean$upstream_start[genome_gff_clean$strand == "-"] <- inverted_start
+genome_gff_clean$upstream_end[genome_gff_clean$strand == "-"] <- inverted_end
+
+# Create a GRanges object for upstream regions
+upstream_regions <- GRanges(
+  seqnames = genome_gff_clean$seqnames,
+  ranges = IRanges(start = genome_gff_clean$upstream_start, end = genome_gff_clean$upstream_end),
+  gene_id = genome_gff_clean$ID,
+  strand = genome_gff_clean$strand
+)
+
+# Subset common chromosomes (seqnames)
+common_seqnames <- intersect(seqlevels(non_overlapping_cgi), seqlevels(upstream_regions))
+non_overlapping_cgi <- keepSeqlevels(non_overlapping_cgi, common_seqnames, pruning.mode = "coarse")
+upstream_regions <- keepSeqlevels(upstream_regions, common_seqnames, pruning.mode = "coarse")
+
+# Find overlaps between non-overlapping CG islands and upstream regions
+upstream_overlap <- findOverlaps(non_overlapping_cgi, upstream_regions)
+
+### Step 3: Extract details of overlaps ###
+
+# Query (CGI) and subject (upstream regions) indices
+cgi_indices <- queryHits(upstream_overlap)
+upstream_indices <- subjectHits(upstream_overlap)
+
+# Calculate the number of overlaps for each CGI
+cgi_overlap_counts <- table(cgi_indices) # Creates a named vector with CGI indices as names and counts as values
+
+# Create a vector that maps the overlap count to each CGI in overlap_info
+overlap_number <- cgi_overlap_counts[as.character(cgi_indices)] # Map counts back to the corresponding CGI index
+
+# Map upstream regions back to the original genome_gff_clean to get gene start and end
+gene_start <- genome_gff_clean$start[match(mcols(upstream_regions)$gene_id[upstream_indices], genome_gff_clean$ID)]
+gene_end <- genome_gff_clean$end[match(mcols(upstream_regions)$gene_id[upstream_indices], genome_gff_clean$ID)]
+
+# Combine relevant information
+overlap_info <- data.frame(
+  CGI_Chr = seqnames(non_overlapping_cgi)[cgi_indices],
+  CGI_Start = start(non_overlapping_cgi)[cgi_indices],
+  CGI_End = end(non_overlapping_cgi)[cgi_indices],
+  Gene_ID = mcols(upstream_regions)$gene_id[upstream_indices],
+  Gene_Start = gene_start,
+  Gene_End = gene_end,
+  Upstream_Start = start(upstream_regions)[upstream_indices],
+  Upstream_End = end(upstream_regions)[upstream_indices],
+  Upstream_Length = width(upstream_regions)[upstream_indices],
+  Strand = as.character(strand(upstream_regions)[upstream_indices]),
+  Overlap_Number = as.integer(overlap_number)
+)
+
+# Calculate the distance to the gene based on the strand
+distance_to_gene <- ifelse(
+  overlap_info$Strand == "+",
+  overlap_info$Gene_Start - overlap_info$CGI_End, # Positive strand: distance from CGI_End to Gene_Start
+  overlap_info$CGI_Start - overlap_info$Gene_End # Negative strand: distance from CGI_Start to Gene_End
+)
+
+# Add the distance_to_gene column to overlap_info
+overlap_info$Distance_to_Gene <- distance_to_gene
+
+# Inspect the result
+head(overlap_info)
+
+```
+
+Number of upstream regions each island intersects with:
+
+```{r}
+table(overlap_info$Overlap_Number)
+```
+
+Summary of distances from the islands to the associated gene:
+
+```{r}
+summary(overlap_info$Distance_to_Gene)
+```
+
+```{r}
+# Count the number of CGIs assigned to each gene
+gene_cgi_counts <- table(overlap_info$Gene_ID)
+
+# Create a frequency table of these counts
+count_distribution <- table(gene_cgi_counts)
+
+# Plot
+barplot(
+  height = as.numeric(count_distribution), # y-axis: number of genes
+  names.arg = names(count_distribution),   # x-axis: CGI counts per gene
+  xlab = "Number of CGIs assigned to a gene", 
+  ylab = "Number of genes",
+  main = "Distribution of CGI Assignments to Genes",
+  col = "skyblue",
+  border = "black"
+)
+```
+
+Now from those islands that were associated to 2 upstream regions (positive and negative strand), we keep only the entry corresponding to the closer gene:
+
+```{r}
+# Step 1: Split overlap_info by unique CGI coordinates
+cgi_groups <- split(overlap_info, paste(overlap_info$CGI_Chr, overlap_info$CGI_Start, overlap_info$CGI_End))
+
+# Step 2: For each group, keep the row with the smallest Distance_to_Gene
+filtered_overlap_info <- do.call(rbind, lapply(cgi_groups, function(group) {
+  group[which.min(group$Distance_to_Gene), ] # Select the row with the minimum Distance_to_Gene
+}))
+
+# Step 3: Reset row names
+rownames(filtered_overlap_info) <- NULL
+
+# Step 4: Inspect the filtered result
+head(filtered_overlap_info)
+
+```
+
+```{r}
+# Check that the number of entries filtered out corresponds to half of the original number of entries corresponding to islands assigned to 2 genes
+nrow(overlap_info) - nrow(filtered_overlap_info) == table(overlap_info$Overlap_Number)[2]/2
+```
+
+```{r}
+# Count the number of CGIs assigned to each gene
+gene_cgi_counts <- table(filtered_overlap_info$Gene_ID)
+
+# Create a frequency table of these counts
+count_distribution <- table(gene_cgi_counts)
+
+# Plot
+barplot(
+  height = as.numeric(count_distribution), # y-axis: number of genes
+  names.arg = names(count_distribution),   # x-axis: CGI counts per gene
+  xlab = "Number of CGIs assigned to a gene", 
+  ylab = "Number of genes",
+  main = "Distribution of CGI Assignments to Genes",
+  col = "skyblue",
+  border = "black"
+)
+```
+
+**Maybe I should check the order of the filtering, as all those genes overlapping with each other were removed from genome_gff before, so maybe some of the islands that are kept here because they dont overlap gene bodies actually would overlap with those genes that were removed**
+
+*Not now at least, but keep it in mind when looking at the results*
+
 # References

### 2025-01-31 R/multiRegion_SIM.R
Updating state of cftp branch

@@ -1975,12 +1975,6 @@ combiStructureGenerator <-
                   equal <- FALSE
                   counter <- 0
                   while(!equal) {
-                    ###################
-                    if (counter > 5){
-                      save(combi_u, combi_m, file = paste0("/scratch/saracv/CFTP_debug/counter_", counter, ".RData"))
-                    }
-                    ##TODO: check size of vectors with info of the cftp steps
-                    ###################
                     # Sample the CFTP steps 
                     self$cftp_event_generator(steps)
                     # Copy (deep clone) the combiStructure instance to generate 2 instances
@@ -2207,6 +2201,11 @@ treeMultiRegionSimulator <- R6Class("treeMultiRegionSimulator",
                                      self$Branch[[1]] <- self$testing_output$cftp_output$combi_u
                                    } else {
                                      self$Branch[[1]] <- self$Branch[[1]]$cftp()
+                                     ################################################
+                                     ##TODO: delete before merging to develop
+                                     print(paste("Size of combi after CFTP", object.size(self$Branch[[1]])))
+                                     print(paste("Number of CFTP steps:", length(self$Branch[[1]]$CFTP_event)))
+                                     ################################################
                                    }
                                  }
                                  

### 2025-02-06 extensive_tests/draft-sumstat-vignette.Rmd
Updating git in order to pull
Updating Rmd file for archipelago definition
Updating summary statistics state for mean sd of frequencies per structure type and mean correlation of a central segment of methylation states per structure type

@@
 
-\[
+$$
 \sigma_{t}(x)=\sqrt{\frac{1}{I-1}\sum_{i=1}^{I}(x_{i,t}-\overline{x_{.,t}})^2} \\
 \sigma_{t}(y)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{i,t}-\overline{y_{.,t}})^2}
-\]
+$$
 
@@
 
-\[
+$$
 \overline{x_{.,t}}=\frac{1}{I}\sum_{i=1}^{I}x_{i,t} \\
 \overline{y_{.,t}}=\frac{1}{J}\sum_{j=1}^{J}x_{j,t}
-\]
+$$
 
@@
 
-\[
+$$
 \hat{\sigma}(x)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(x) \\
 \hat{\sigma}(y)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(y) \\
-\]
+$$
 
@@
 
-\[
+$$
 \sigma_{t}(w)=\sqrt{\frac{1}{I-1}\sum_{i=1}^{I}(x_{i,t}-\overline{x_{.,t}})^2} \\
 \sigma_{t}(z)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{i,t}-\overline{y_{.,t}})^2}
-\]
+$$
 
@@
 
-\[
+$$
 \overline{w_{.,t}}=\frac{1}{I}\sum_{i=1}^{I}w_{i,t} \\
 \overline{z_{.,t}}=\frac{1}{J}\sum_{j=1}^{J}z_{j,t}
-\]
+$$
 
@@
 
-\[
+$$
 \hat{\sigma}(w)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(w) \\
 \hat{\sigma}(z)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(z) \\
-\]
+$$
 
@@
 
-# Mean correlation of a segment of methylation states per genomic structure type
+# Mean correlation of a central segment of methylation states per genomic structure type
 
@@
 
-NOTE: Summary statistics related with tree cherries require a tree with tip labels as:
+NOTE: Summary statistics related with tree cherries require a tree with tip labels givenas numeric values

### 2025-02-06 R/summaryStatistics.R

@@ -171,3 +171,308 @@ get_nonislandMeanFreqM <- function(index_nonislands, data, sample_n){
   }
   return(mean(mean_nonisland))
 }
+
+
+
+#' Calculate the Mean Standard Deviation of Partially Methylated Sites in Islands
+#'
+#' This function computes the mean standard deviation of partially methylated sites 
+#' (with methylation state 0.5) for a set of genomic structures identified as islands.
+#'
+#' @param index_islands A vector containing the structural indices for islands.
+#' @param data A list containing methylation states at tree tips for each genomic structure (island / non-island).
+#' For a single tip: \code{data[[structure]]}. For multiple tips: \code{data[[tip]][[structure]]}.
+#' @param sample_n The number of samples (tips) to process.
+#'
+#' @return A numeric value representing the mean standard deviation of partially methylated sites in the islands.
+#' @examples
+#' # Example usage:
+#' index_islands <- c(1, 3)
+#' data <- list(
+#'   list(c(0.5, 1, 0.5), c(0, 0.5, 1), c(1, 0, 0.5)), # tip 1
+#'   list(c(0.5, 0.5, 0), c(1, 0.5, 0.5), c(0.5, 0.5, 1)) # tip 2
+#' )
+#' sample_n <- 2
+#' get_islandSDFreqP(index_islands, data, sample_n)
+#' 
+#' @export
+get_islandSDFreqP <- function(index_islands, data, sample_n){
+  # Restructure data as nested list
+  if (sample_n == 1){
+    data_list <- list()
+    data_list[[1]] <- data
+    data <- data_list
+  }
+  sd_tip <- c() 
+  for (s in 1:sample_n){
+    freq_island <- c()
+    island_counter <- 1
+    for (i in index_islands){
+      # Compute proportion of partially methylated sites at each island
+      freq_island[island_counter] <- mean(data[[s]][[i]] == 0.5)
+      island_counter <- island_counter + 1
+    }
+    sd_tip[s] <- stats::sd(freq_island)
+  }
+  return(mean(sd_tip))
+}
+
+#' Calculate the Mean Standard Deviation of Partially Methylated Sites in Non-Islands
+#'
+#' This function computes the mean standard deviation of partially methylated sites 
+#' (with methylation state 0.5) for a set of genomic structures identified as non-islands.
+#'
+#' @param index_nonislands A vector containing the structural indices for non-islands.
[…]

### 2025-02-06 extensive_tests/toClean-sumstats.R

@@ -1,306 +1,5 @@
 
 
-#' Calculate the Mean Standard Deviation of Partially Methylated Sites in Islands
-#'
-#' This function computes the mean standard deviation of partially methylated sites 
-#' (with methylation state 0.5) for a set of structures identified as islands.
-#'
-#' @param index_islands A vector containing the structural indices for islands.
-#' @param data A list containing methylation states at tree tips for each structure (island / non-island).
-#' For a single tip: \code{data[[structure]]}. For multiple tips: \code{data[[tip]][[structure]]}.
-#' @param sample_n The number of samples (tips) to process.
-#'
-#' @return A numeric value representing the mean standard deviation of partially methylated sites in the islands.
-#' @examples
-#' # Example usage:
-#' index_islands <- c(1, 2)
-#' data <- list(
-#'   list(c(0.5, 1, 0.5), c(0, 0.5, 1)), # tip 1
-#'   list(c(0.5, 0.5, 0), c(1, 0.5, 0.5)) # tip 2
-#' )
-#' sample_n <- 2
-#' get_islandSDFreqP(index_islands, data, sample_n)
-#' 
-#' @export
-get_islandSDFreqP <- function(index_islands, data, sample_n){
-  # Restructure data as nested list
-  if (sample_n == 1){
-    data_list <- list()
-    data_list[[1]] <- data
-    data <- data_list
-  }
-  sd_tip <- c() 
-  for (s in 1:sample_n){
-    freq_island <- c()
-    island_counter <- 1
-    for (i in index_islands){
-      # Compute proportion of partially methylated sites at each island
-      freq_island[island_counter] <- mean(data[[s]][[i]] == 0.5)
-      island_counter <- island_counter + 1
-    }
-    sd_tip[s] <- stats::sd(freq_island)
-  }
-  return(mean(sd_tip))
-}
-
-#' Calculate the Mean Standard Deviation of Partially Methylated Sites in Non-Islands
-#'
-#' This function computes the mean standard deviation of partially methylated sites 
-#' (with methylation state 0.5) for a set of structures identified as non-islands.
-#'
-#' @param index_nonislands A vector containing the structural indices for non-islands.
-#' @param data A list containing methylation states at tree tips for each structure (island / non-island).
[…]

### 2025-02-06 extensive_tests/toClean-test-sumstats.R

@@ -1,678 +1,4 @@
 
-test_that("get_SDFreqP structures with equal length islands and non-islands", {
-  # One tip and four islands / non-islands
-  index_islands <- c(1, 2, 3, 4)
-  index_nonislands <- c(1, 2, 3, 4)
-  # 1 tip mean freq p 0.5
-  data <- list(c(0, 0, 0), # freq P 0
-               c(.5, .5, .5), # freq P 1
-               c(0, 0, 0), # freq P 0
-               c(.5, .5, .5)) # freq P 1
-  exp_SD <- sqrt(4*(0.5)^2/3) 
-  sample_n <- 1
-  expect_equal(get_islandSDFreqP(index_islands, data, sample_n), exp_SD,
-               info = "incorrect SD in one tip and 4 islands")
-  expect_equal(get_nonislandSDFreqP(index_nonislands, data, sample_n), exp_SD,
-               info = "incorrect SD in one tip and 4 non-islands")
-  
-  # Two tips and four islands / non-islands
-  index_islands <- c(1, 2, 3, 4)
-  index_nonislands <- c(1, 2, 3, 4)
-  # Tip 1 mean freq p 0.5, tip 2 mean freq p 0.5
-  data <- list(
-    list(c(0, 0, 0), # Tip 1: freq P 0
-         c(.5, .5, .5), # Tip 1: freq P 1
-         c(0, 0, 0), # Tip 1: freq P 0
-         c(.5, .5, .5)), # Tip 1: freq P 1
-    list(c(0, 0, 0), # Tip 2: freq P 0
-         c(1, .5, 1), # Tip 2: freq P 1/3
-         c(.5, 0, .5), # Tip 2: freq P 2/3
-         c(.5, .5, .5)) # Tip 2: freq P 1
-  )
-  exp_SD1 <- sqrt(4*(0.5)^2/3)
-  exp_SD2 <- sqrt((2*(0.5)^2+2*((1/3)-0.5)^2)/3)
-  mean_SD <- mean(c(exp_SD1, exp_SD2))
-  sample_n <- 2
-  expect_equal(get_islandSDFreqP(index_islands, data, sample_n), mean_SD,
-               info = "incorrect SD in two tip and 4 islands")
-  expect_equal(get_nonislandSDFreqP(index_nonislands, data, sample_n), mean_SD,
-               info = "incorrect SD in two tip and 4 non-islands")
-  
[…]

### 2025-02-06 tests/testthat/test-summaryStatistics.R

@@ -204,3 +204,681 @@ test_that("get_MeanFreqM structures with different length (islands and non-islan
                info = "incorrect mean in two tips and two non-islands, freq .5")
 })
 
+
+test_that("get_SDFreqP structures with equal length islands and non-islands", {
+  # One tip and four islands / non-islands
+  index_islands <- c(1, 2, 3, 4)
+  index_nonislands <- c(1, 2, 3, 4)
+  # 1 tip mean freq p 0.5
+  data <- list(c(0, 0, 0), # freq P 0
+               c(.5, .5, .5), # freq P 1
+               c(0, 0, 0), # freq P 0
+               c(.5, .5, .5)) # freq P 1
+  exp_SD <- sqrt(4*(0.5)^2/3) 
+  sample_n <- 1
+  expect_equal(get_islandSDFreqP(index_islands, data, sample_n), exp_SD,
+               info = "incorrect SD in one tip and 4 islands")
+  expect_equal(get_nonislandSDFreqP(index_nonislands, data, sample_n), exp_SD,
+               info = "incorrect SD in one tip and 4 non-islands")
+  
+  # Two tips and four islands / non-islands
+  index_islands <- c(1, 2, 3, 4)
+  index_nonislands <- c(1, 2, 3, 4)
+  # Tip 1 mean freq p 0.5, tip 2 mean freq p 0.5
+  data <- list(
+    list(c(0, 0, 0), # Tip 1: freq P 0
+         c(.5, .5, .5), # Tip 1: freq P 1
+         c(0, 0, 0), # Tip 1: freq P 0
+         c(.5, .5, .5)), # Tip 1: freq P 1
+    list(c(0, 0, 0), # Tip 2: freq P 0
+         c(1, .5, 1), # Tip 2: freq P 1/3
+         c(.5, 0, .5), # Tip 2: freq P 2/3
+         c(.5, .5, .5)) # Tip 2: freq P 1
+  )
+  exp_SD1 <- sqrt(4*(0.5)^2/3)
+  exp_SD2 <- sqrt((2*(0.5)^2+2*((1/3)-0.5)^2)/3)
+  mean_SD <- mean(c(exp_SD1, exp_SD2))
+  sample_n <- 2
+  expect_equal(get_islandSDFreqP(index_islands, data, sample_n), mean_SD,
+               info = "incorrect SD in two tip and 4 islands")
[…]

### 2025-02-10 application_examples/abc_inferences/draft-abc-inferences-vignette.Rmd
Addinf filter_CpG_count.R

@@
 
-1. Save as simulation design the genomic structure, tree(s) -- GIVEN -- and sampled parameter combinations.
+1. Set the output directory.
+
+2. Get the CpG count in islands and non-island regions.
+
+3. Filter CpG count to keep only regions with at least one CpG site and get the CpG position indeces for each region.
+
+1. Save as simulation design the genomic structure for the regions within the selected genomic region, tree(s) -- GIVEN -- and sampled parameter combinations.
 

### 2025-02-10 extensive_tests/draft-sumstat-vignette.Rmd
Addinf filter_CpG_count.R

@@
 $$
+These values can be computed with the functions `get_islandSDFreqP`, `get_nonislandSDFreqP`,`get_islandSDFreqM` and `get_nonislandSDFreqM` as in the following examples:
 
@@
 - If $l = 0$, all the positions are considered.
+
+These values can be computed with the functions `compute_meanCor_i` for island structures and `compute_meanCor_ni` for non-island structures as in the following examples:
 
@@
 
-get_FChange_cherryData (uses find_cherries --get_cherryDist-- and cherry_diffs -- now count_cherryMethDiff )
+Functions:
 
-NOTE: Summary statistics related with tree cherries require a tree with tip labels givenas numeric values
+- `get_cherryDist` to get the distances between the cherry tips.

### 2025-02-10 R/summaryStatistics.R

@@ -476,3 +476,100 @@ compute_meanCor_ni <- function(index_nonislands, minN_CpG, shore_length, data, s
   }
   return(mean(cor))
 }
+
+#### #### #### Tree cherries comparisons #### #### ####
+
+#' Get Cherry Pair Distances from a Phylogenetic Tree
+#'
+#' This function computes the pairwise distances between the tips of a phylogenetic tree 
+#' that are part of cherries. A cherry is a pair of leaves (tips) that are sisters in the tree.
+#' The distance is calculated as the sum of the branch lengths between the two sister tips.
+#'
+#' @param tree A tree in Newick format (as a character string) or an object of class \code{phylo} from the \code{ape} package.
+#' If the input is a character string, it must follow the Newick or New Hampshire format (e.g. \code{"((tip_1:1,tip_2:1):5,tip_3:6);"}).
+#' If an object of class \code{phylo} is provided, it should represent a valid phylogenetic tree.
+#'
+#' @return A data frame with three columns:
+#'   \item{first_tip}{A character string representing the first tip in the cherry (sister pair).}
+#'   \item{second_tip}{A character string representing the second tip in the cherry.}
+#'   \item{dist}{A numeric value representing the sum of the branch lengths between the two tips (i.e., the distance between the cherries).}
+#'   
+#' @details The function first checks if the input is either a character string in the Newick format or an object of class \code{phylo}.
+#'   It then computes the pairwise distances between the tips in the tree and identifies the sister pairs (cherries).
+#'   The distance between each cherry is the sum of the branch lengths leading to the sister tips.
+#'   
+#'   If the tree is provided in Newick format, it will be parsed using the \code{ape::read.tree} function.
+#'   
+#' @importFrom ape read.tree cophenetic.phylo
+#'   
+#' @examples
+#' # Example of a tree in Newick format
+#' newick_tree <- "((a:1,b:2):5,c:6);"
+#' get_cherryDist(newick_tree)
+#' 
+#' # Example of using a phylo object from ape
+#' library(ape)
+#' tree_phylo <- read.tree(text = "((a:1,b:1):5,c:6);")
+#' get_cherryDist(tree_phylo)
+#' 
+#' @export
+get_cherryDist <- function(tree){
+  
+  # Check that the input is a character string or phylo object
+  if (!is.character(tree) && !inherits(tree, "phylo")) {
[…]

### 2025-02-10 application_examples/abc_inferences/scripts/filter_CpG_count.R

@@ -39,16 +39,5 @@ load(opt[["input"]])
 # Filter the data for the given chromosome and region
 CpG_count_region <- CpG_count[CpG_count$chr == opt[["chr"]] & CpG_count$start >= opt[["start"]] - opt[["side-bp"]] & CpG_count$start <= opt[["end"]] + opt[["side-bp"]], ]
 
-# Create the data frame spatial_str
-spatial_str <- data.frame(n = CpG_count_region$CpG_count)
-
-# Add the globalState column based on CpG_count_region$str values
-spatial_str$globalState <- ifelse(CpG_count_region$str == "Island", "U", "M")
-
-# Save spatial structure
-#output_name <- paste0(paste(opt[["chr"]], opt[["start"]] - opt[["side-bp"]], opt[["end"]] + opt[["side-bp"]], opt[["name-selected-region"]], sep = "_"), ".RData")
-save(spatial_str, file = opt[["output"]])
-
-##TODO: save CpG_count_region as region_CGI-Mzebra.RData. 
-# Maybe rename it as region_cgi. Then update name it in get_CpG_positions
-# or if in next script region_CpG_positions.
+# Save the CpG count filtered for the region of interest
+save(CpG_count_region, file = opt[["output"]])

### 2025-02-10 extensive_tests/toClean-sumstats.R

@@ -1,71 +1,16 @@
 
-
-#### #### #### Tree cherries comparisons #### #### ####
-
-find_cherries <- function(tree) {
-  ## tree must be an ape tree whose tip labels are numbers.
-  ## returns a list of vectors (a, b, x), where  (a, b) are a cherry of tree,
-  ## that is a pair of leaves that are sisters in the tree,
-  ## and x is their distance, that is, sum of the branch lengths between them 
-  outlist <- list()
-  d <- cophenetic.phylo(tree)
-  n <- nrow(d)
-  for(i in 1:n) {
-    j <- which(d[i,]==min(d[i, -i]))
-    if(length(j)==1) {
-      k <- which(d[j,]==min(d[j, -j]))
-      if(length(k)==1) {
-        if(k==i & i<j) {
-          ## outlist[[length(outlist)+1]] <- c(as.integer(rownames(d)[i]), as.integer(colnames(d)[j]), d[i,j])
-          outlist[[length(outlist)+1]] <- c(i, j, d[i,j])
-        }
-      }
-    }
-  }
-  outlist
-}
+index_islands <- c(1,3)
+index_nonislands <- c(2,4)
+# Simulated data has all positions
+# But empirical data may have differences in coverage
+# E.g. if in island one positions 1,3,5,7 are CpGs
+# Empirical data may have in first_tip 1,5,7
+# And in second tip 1,3,5,7
+# So that only sites 1,5,7 can be compared
 
 
 
-## TODO: The tree needs to
-## have the tips with numeric labels. ordered from left to as ((1:1,2:1):5,3:6); 
-## For that check previous (old) function 
-## tree needs to be an object of ape's phylo class or a character string in
-## parenthetic format known as the Newick or New Hampshire format
-get_cherryDist <- function(tree, sample_n, testing = FALSE){
-  if(sample_n < 2){stop("Minimum number of tips/samples needed: 2")}
-  if(class(tree) != "phylo") tree <- ape::read.tree(text = tree)
-  # compute the pairwise distances between the tips from a phylogenetic tree
-  dist <- ape::cophenetic.phylo(tree)
-  # get the number of tips
-  n_tips <- nrow(dist)
-  # set list to store the cherry info
-  cherry_list <- list()
-  for (tip in 1:n_tips){
-    # find the closest tip
-    tip_a <- which(dist[tip,] == min(dist[tip, -tip]))
-    # if there is one single closest tip
[…]

### 2025-02-10 extensive_tests/toClean-test-sumstats.R

@@ -1,49 +1,5 @@
 
 
-test_that("get_cherryDist", {
-  # Set tree
-  newick_tree <- "((1:1,2:1):1,3:2);"
-  
-  # Test wrong input
-  expect_error(get_cherryDist(),
-               info = "function fails to throw an error when no tree/sample_n is given")
-  expect_error(get_cherryDist(sample_n = 2),
-               info = "function fails to throw an error when no tree is given")
-  expect_error(get_cherryDist(tree = newick_tree),
-               info = "function fails to throw an error when no sample_n is given")
-  
-  
-  # Test error when tree has only one tip
-  newick_tree <- "(1:1);"
-  expect_error(get_cherryDist(tree = newick_tree, sample_n = 1),
-               info = "fails to throw an error when number of tips is < 2")
-  
-  # Test conversion from newick tree to ape tree
-  newick_tree <- "((1:1,2:1):1,3:2);"
-  expect_true(class(get_cherryDist(newick_tree, sample_n = 3, testing = T)$tree) == "phylo",
-              info = "fails to convert newick tree into ape's class phylo")
-  
-  # Test case single cherry
-  expect_true(all(get_cherryDist(newick_tree, sample_n = 3)[[1]] == c(1,2,2)),
-              info = "incorrect output for newick tree with single cherry")
-  ape_tree <- ape::read.tree(text = newick_tree)
-  expect_true(all(get_cherryDist(ape_tree, sample_n = 3)[[1]] == c(1,2,2)),
-              info = "incorrect output for ape tree with single cherry")
-  
-  # Test case 2 cherries
-  newick_tree <- "((1:1.5,2:1.5):2,(3:2,4:2):1.5);"
-  ape_tree <- ape::read.tree(text = newick_tree)
-  expect_true(all(get_cherryDist(newick_tree, sample_n = 4)[[1]] == c(1,2,3)),
-              info = "incorrect output for newick tree with two cherries (1st cherry)")
-  expect_true(all(get_cherryDist(newick_tree, sample_n = 4)[[2]] == c(3,4,4)),
-              info = "incorrect output for newick tree with two cherries (2nd cherry)")
-  expect_true(all(get_cherryDist(ape_tree, sample_n = 4)[[1]] == c(1,2,3)),
-              info = "incorrect output for ape tree with two cherries (1st cherry)")
-  expect_true(all(get_cherryDist(ape_tree, sample_n = 4)[[2]] == c(3,4,4)),
[…]

### 2025-02-10 tests/testthat/test-summaryStatistics.R

@@ -881,4 +881,96 @@ test_that("meanCor", {
                info = "MeanCor_ni returns non-correct value (shore 10 2 tips non-equal sizes)")
 })
 
+test_that("get_cherryDist input control errors", {
+  
+  # No input
+  expect_error(get_cherryDist(),
+               info = "function fails to throw an error when no tree is given")
+  # Tree not character string or phylo class
+  tree <- 5
+  expect_error(get_cherryDist(tree),
+               info = "function fails to throw error when tree is not character string or phylo object")
+  
+  # Incorrect newick format
+  tree <- "a:1, b:2"
+  expect_error(get_cherryDist(tree),
+               info = "function fails to throw error when given tree has incorrect format")
+  tree <- "a:1, b:2;"
+  expect_error(get_cherryDist(tree),
+               info = "function fails to throw error when given tree has incorrect format")
+  # Test error when tree has only one tip
+  tree <- "(1:1);"
+  expect_error(get_cherryDist(tree),
+               info = "function fails to throw error when given tree has only one tip")
+})
+
+test_that("get_cherryDist processing of different input types", {
+  
+  # One cherry, numeric tip labels ordered
+  type <- "One cherry, numeric tip labels ordered"
+  tree <- "((1:0.1,2:0.1):2,3:3);"
+  output <- get_cherryDist(tree)
+  expect_equal(output$first_tip, "1",
+               info = paste("Fails to output correct $first tip with input type:", type))
+  expect_equal(output$second_tip, "2",
+               info = paste("Fails to output correct $second_tip tip with input type:", type))
+  expect_equal(output$dist, 0.2,
+               info = paste("Fails to output correct $dist tip with input type:", type))
+  
+  # One cherry, numeric tip labels unordered
+  type <- "One cherry, numeric tip labels unordered"
+  tree <- "((3:0.15,1:0.2):2,2:3);"
+  output <- get_cherryDist(tree)
+  expect_equal(output$first_tip, "3",
+               info = paste("Fails to output correct $first tip with input type:", type))
+  expect_equal(output$second_tip, "1",
+               info = paste("Fails to output correct $second_tip tip with input type:", type))
+  expect_equal(output$dist, 0.35,
+               info = paste("Fails to output correct $dist tip with input type:", type))
+  
+  # One cherry, named tips
+  type <- "One cherry, named tips"
+  tree <- "((a:0.8,b:0.2):2,c:3);"
[…]

### 2025-02-10 application_examples/abc_inferences/scripts/get_spatial_str.R
Added rule to filter GpG counts relative to the genomic region of interest

@@ -4,17 +4,7 @@ option_list <- list(
   make_option(c("--input"), type = "character", default = NULL,
               help = "Input CpG_count.RData path", metavar = "character"),
   make_option(c("--output"), type = "character", default = NULL,
-              help = "Output full path with file name", metavar = "character"),
-  make_option(c("--name-selected-region"), type = "character", default = NULL,
-              help = "Name of selected genomic region", metavar = "character"),
-  make_option(c("--chr"), type = "character", default = NULL,
-              help = "Chromosome ID of selected region", metavar = "character"),
-  make_option(c("--start"), type = "integer", default = NULL,
-              help = "Start index of selected genomic region", metavar = "integer"),
-  make_option(c("--end"), type = "integer", default = NULL,
-              help = "End index of selected genomic region", metavar = "integer"),
-  make_option(c("--side-bp"), type = "integer", default = NULL,
-              help = "Number of bp to select structures starting after start - side-bp and ending before end + side-bp", metavar = "integer")
+              help = "Output full path with file name", metavar = "character")
 )
 
 # Parse the arguments
@@ -23,7 +13,7 @@ opt_parser <- OptionParser(option_list = option_list)
 opt <- parse_args(opt_parser)
 
 # Get the names of required options (you can update this based on what's mandatory)
-required_options <- c("input", "output","name-selected-region", "chr", "start", "end", "side-bp")
+required_options <- c("input", "output")
 
 # Check that all required options are not NULL
 missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
@@ -36,9 +26,6 @@ if (length(missing_options) > 0) {
 # Load CpG counts
 load(opt[["input"]])
 
-# Filter the data for the given chromosome and region
-CpG_count_region <- CpG_count[CpG_count$chr == opt[["chr"]] & CpG_count$start >= opt[["start"]] - opt[["side-bp"]] & CpG_count$start <= opt[["end"]] + opt[["side-bp"]], ]
-
 # Create the data frame spatial_str
 spatial_str <- data.frame(n = CpG_count_region$CpG_count)
 
@@ -46,9 +33,4 @@ spatial_str <- data.frame(n = CpG_count_region$CpG_count)
 spatial_str$globalState <- ifelse(CpG_count_region$str == "Island", "U", "M")
 
 # Save spatial structure
[…]

### 2025-02-10 application_examples/abc_inferences/draft-abc-inferences-vignette.Rmd
Adding script get_CpG_positions.R

@@
 
-2. Get the CpG count in islands and non-island regions.
+2. Get the CpG count in islands and non-island regions relative to the reference genome.
 
-3. Filter CpG count to keep only regions with at least one CpG site and get the CpG position indeces for each region.
+3. Filter the CpG count in islands and non-island regions for the genomic region of interest.
 
-1. Save as simulation design the genomic structure for the regions within the selected genomic region, tree(s) -- GIVEN -- and sampled parameter combinations.
+4. Get the spatial structure dataframe from the filtered GpG count to use as input for the simulations.
 
-2. Run the simulations. TODO: Before submitting load package without devtools
+5. Save as simulation design the genomic structure for the regions within the selected genomic region, tree(s) -- GIVEN -- and sampled parameter combinations.
+
+6. Run the simulations. TODO: Before submitting load package without devtools
+
+7. Get the CpG position indices relative to the reference genome (cytosine index relative to positive strand).
 

### 2025-02-10 application_examples/abc_inferences/scripts/get_CpG_positions.R

@@ -1,0 +1,89 @@
+library(optparse)
+library(Biostrings)
+
+option_list <- list(
+  make_option(c("--input"), type = "character", default = NULL,
+              help = "Input CpG_count.RData path", metavar = "character"),
+  make_option(c("--reference-genome"), type = "character", default = NULL,
+              help = "Relative path to the reference genome file (.fa)", metavar = "character"),
+  make_option(c("--output"), type = "character", default = NULL,
+              help = "Output relative path with file name", metavar = "character")
+)
+
+# Parse the arguments
+
+opt_parser <- OptionParser(option_list = option_list)
+opt <- parse_args(opt_parser)
+
+# Get the names of required options (you can update this based on what's mandatory)
+required_options <- c("input", "output")
+
+# Check that all required options are not NULL
+missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
+
+if (length(missing_options) > 0) {
+  print_help(opt_parser)
+  stop(paste("The following arguments need to be provided:", paste(missing_options, collapse = ", ")))
+}
+
+
+# Load Reference Genome and CpG_count
+genome <- readDNAStringSet(opt[["reference-genome"]], format = "fasta")
+load(opt[["input"]])
+
+# Define a function to get CpG positions in a given sequence
+get_CpG_positions <- function(seq, start_pos) {
+  cpg_sites <- start(Biostrings::matchPattern("CG", seq))  # Find "CG" occurrences
+  return(cpg_sites + start_pos - 1)  # Convert to genome-wide positions
+}
+
+# Extract only the chromosome part from reference genome names
+genome_chr_names <- sapply(names(genome), function(x) strsplit(x, " ")[[1]][1])
+
+# Filter only the regions with at least one CpG site
+CpG_count <- CpG_count[CpG_count$CpG_count > 0,]
+
+# Initialize a list to store results
+cpg_positions_list <- list()
+
+# Initialize a variable to keep track of current chromosome
+current_chr <- "no_chromosome"
+
+# Loop over each region in CpG_count
+for (i in 1:nrow(CpG_count)) {
+  region <- CpG_count[i, ]
+  chr_name <- region$chr
+  
+  # Extract chromosome sequence
+  if (chr_name != current_chr){
+    
+    current_chr <- chr_name
+    print(paste("Processing chromosome", current_chr))
+    
[…]

### 2025-02-10 application_examples/abc_inferences/scripts/get_CpG_positions.R
Updating Snakefile with rule to get the CpG positions

@@ -74,6 +74,7 @@ for (i in 1:nrow(CpG_count)) {
   seq_region <- subseq(seq_chr, start = region$start, end = region$end)  # Extract region
   
   # Get CpG positions
+  cpg_positions <- get_CpG_positions(seq_region, region$start)
   
   # Store results
   cpg_positions_list[[i]] <- data.frame(

### 2025-02-10 application_examples/abc_inferences/scripts/filter_CpG_positions.R
Adding scritp filter_CpG_positions.R

@@ -1,0 +1,71 @@
+library(optparse)
+
+option_list <- list(
+  make_option(c("--filtered-CpGcount-region"), type = "character", default = NULL,
+              help = "CpG_count_region.RData relative path", metavar = "character"),
+  make_option(c("--CpG-positions"), type = "character", default = NULL,
+              help = "CpG_positions.RData relative path", metavar = "character"),
+  make_option(c("--output"), type = "character", default = NULL,
+              help = "Output relative path with file name", metavar = "character")
+)
+
+# Parse the arguments
+
+opt_parser <- OptionParser(option_list = option_list)
+opt <- parse_args(opt_parser)
+
+# Get the names of required options (you can update this based on what's mandatory)
+required_options <- c("filtered-CpGcount-region", "CpG-positions", "output")
+
+# Check that all required options are not NULL
+missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
+
+if (length(missing_options) > 0) {
+  print_help(opt_parser)
+  stop(paste("The following arguments need to be provided:", paste(missing_options, collapse = ", ")))
+}
+
+
+# Load the CpG count for the genomic region of interest and the CpG positions 
+# relative to the reference genome (cytosine, relative to the positive strand)
+load(opt[["filtered-CpGcount-region"]])
+load(opt[["CpG-positions"]])
+
+# Initialize a list to store results
+filtered_positions <- list(length = nrow(CpG_count_region))
+
+# Loop over each island and non-island region in the genomic region of interest
+for (i in 1:nrow(CpG_count_region)) {
+  
+  # Extract the relevant information for the current island / non-island region
+  region <- CpG_count_region[i,]
+  chr <- region$chr
+  start <- region$start
+  end <- region$end
+  
+  # Extract the CpG positions within the current region
+  filtered_positions[[i]] <- cpg_positions_df[cpg_positions_df$chr == chr & cpg_positions_df$CpG_position >= start & cpg_positions_df$CpG_position <= end,]
+  
+  # Add the positions for region start and end
+  filtered_positions[[i]]$region_start <- region$start
+  filtered_positions[[i]]$region_end <- region$end
+  
+  # Check the number of filtered positions corresponds to the CpG count for that region
+  if(nrow(filtered_positions[[i]]) != region$CpG_count){
+    print("Numper of filtered positions differs from region CpG_count in region:")
+    print(region)
+  }
+  
[…]

### 2025-02-11 extensive_tests/draft-sumstat-vignette.Rmd
Updating changes after adding function to count the number of methylation changes per structure, testing and documenting it

@@
 
-- `get_cherryDist` to get the distances between the cherry tips.
+- `get_cherryDist(tree)` to get the distances between the tips of each cherry.
+It identifies each cherry by the tip names and the tip indices.
+The tip indices correspond to (a) the index from left to right on the newick string, (b) the order of the tip label in the `phylo_object$tip.label`, and (c) the index in the methylation data list `data[[tip]][[structure]]` as obtained with the function `simulate_evolData()` when the given tree has several tips.
+
+- `countSites_cherryMethDiff(cherryDist, data)` to get for each cherry the counts of sites with half-methylation and full-methylation changes per genomic structure (island or non-island).

### 2025-02-11 R/summaryStatistics.R

@@ -482,21 +482,30 @@ compute_meanCor_ni <- function(index_nonislands, minN_CpG, shore_length, data, s
 #' Get Cherry Pair Distances from a Phylogenetic Tree
 #'
 #' This function computes the pairwise distances between the tips of a phylogenetic tree 
-#' that are part of cherries. A cherry is a pair of leaves (tips) that are sisters in the tree.
-#' The distance is calculated as the sum of the branch lengths between the two sister tips.
+#' that are part of cherries. A cherry is a pair of leaf nodes (also called tips or terminal nodes) 
+#' in a phylogenetic tree that share a direct common ancestor. 
+#' In other words, if two leaves are connected to the same internal node and no other leaves 
+#' are connected to that internal node, they form a cherry.
+#' The distance is calculated as the sum of the branch lengths between the two cherry tips.
 #'
 #' @param tree A tree in Newick format (as a character string) or an object of class \code{phylo} from the \code{ape} package.
 #' If the input is a character string, it must follow the Newick or New Hampshire format (e.g. \code{"((tip_1:1,tip_2:1):5,tip_3:6);"}).
 #' If an object of class \code{phylo} is provided, it should represent a valid phylogenetic tree.
 #'
-#' @return A data frame with three columns:
-#'   \item{first_tip}{A character string representing the first tip in the cherry (sister pair).}
-#'   \item{second_tip}{A character string representing the second tip in the cherry.}
+#' @return A data frame with five columns:
+#'   \item{first_tip_name}{A character string representing the name of the first tip in the cherry.}
+#'   \item{second_tip_name}{A character string representing the name of the second tip in the cherry.}
+#'   \item{first_tip_index}{An integer representing the index of the first tip in the cherry.}
+#'   \item{second_tip_index}{An integer representing the index of the second tip in the cherry.}
 #'   \item{dist}{A numeric value representing the sum of the branch lengths between the two tips (i.e., the distance between the cherries).}
 #'   
 #' @details The function first checks if the input is either a character string in the Newick format or an object of class \code{phylo}.
 #'   It then computes the pairwise distances between the tips in the tree and identifies the sister pairs (cherries).
 #'   The distance between each cherry is the sum of the branch lengths leading to the sister tips.
[…]

### 2025-02-11 application_examples/abc_inferences/scripts/get_CpG_positions.R

@@ -84,7 +84,7 @@ for (i in 1:nrow(CpG_count)) {
   )
 }
 
-# Combine results into a single data frame
+# Combine results into a single data frame and save into output file
 cpg_positions <- do.call(rbind, cpg_positions_list)
 save(cpg_positions, file = opt[["output"]])
 

### 2025-02-11 extensive_tests/toClean-sumstats.R

@@ -7,41 +7,13 @@ index_nonislands <- c(2,4)
 # Empirical data may have in first_tip 1,5,7
 # And in second tip 1,3,5,7
 # So that only sites 1,5,7 can be compared
+# Previous step of filtering data using a list containing for each island and non-island region a vector of the
+# relative indices of the CpGs that are to be compared.
 
 
 
 
 
-
-#undebug(get_cherryDist)
-## TODO: Assumes that each str has the same number of sites at the different tips
-## Think about how to control for pop cases in which there may be insertions/deletions
-## or whether this should be controlled by the user
-countSites_cherryMethDiff <- function(cherries, data) {
-  ## returns a data frame with one line for each cherry, containing the branch distance between the
-  ## cherries and for each structure two numbers: that of the number of sites that are m in one in
-  ## u in the other tip and the number of sites that are p in one and m or u in the other tip.
-
-  str_n <- length(data[[1]]) # set the number of structures
-  df <- data.frame(tips=character(), dist=numeric()) # start df
-  for(str in 1:str_n) {
-    # e.g. df 3 structures: tips dist 1_f 1_h 2_f 2_h 3_f 3_h
-    df <- cbind(df, integer(), integer())
-    names(df)[c(ncol(df)-1, ncol(df))] <- paste0(str,"_",c("f", "h"))
-  }
-  for(i in 1:length(cherries)) {
-    ch <- cherries[[i]]
-    df[i, 1] <- paste0(ch[1],"-",ch[2]) # add tips info
-    df[i, 2] <- ch[3] # add dist between tips
-    for(str in 1:str_n) {
-      # count number of sites with full methylation change (from u to m or m to u)
-      df[i, (str+1)*2-1] <- sum(abs(data[[ch[1]]][[str]]-data[[ch[2]]][[str]])==1)
-      # count number of sites with half methylation change (p in one tip and m or u in the other)
-      df[i, (str+1)*2] <- sum(abs(data[[ch[1]]][[str]]-data[[ch[2]]][[str]])==0.5)
-    }
-  }
-  df
-}
 
 #undebug(countSites_cherryMethDiff)
 #undebug(get_cherryDist)

### 2025-02-11 extensive_tests/toClean-test-sumstats.R

@@ -1,36 +1,5 @@
 
 
-
-test_that("countSites_cherryMethDiff", {
-  # Test case one cherry
-  newick_tree <- "((1:1,2:1):1,3:2);"
-  data <- list(
-    list(rep(1,10), rep(0,10), rep(1,10)),
-    list(rep(1,10), rep(0.5,10), rep(0,10)),
-    list(rep(1,10), rep(0.5,10), rep(0,10)))
-  cherries <- get_cherryDist(newick_tree, sample_n = 3)
-  o <- countSites_cherryMethDiff(cherries, data)
-  expect_equal(nrow(o), 1, info = "case one cherry does not have one row")
-  expect_true(all(as.numeric(o[1,-c(1,2)]) == c(0, 0, 0, 10, 10, 0)),
-              info = "case one cherry incorrect count output")
-  
-  # Test case two cherries
-  newick_tree <- "((1:1.5,2:1.5):2,(3:2,4:2):1.5);"
-  data <- list(
-    list(rep(1,10), rep(0,10), rep(1,10)),
-    list(rep(1,10), rep(0.5,10), rep(0,10)),
-    list(rep(1,10), rep(0.5,10), rep(0,10)),
-    list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1, rep(0.5, 5)), c(0.5, 1, rep(0, 8))))
-  cherries <- get_cherryDist(newick_tree, sample_n = 4)
-  o <- countSites_cherryMethDiff(cherries, data)
-  expect_equal(nrow(o), 2, info = "case two cherries does not have one row")
-  expect_true(all(as.numeric(o[1,-c(1,2)]) == c(0, 0, 0, 10, 10, 0)),
-              info = "case two cherries incorrect count output for first cherry")
-  expect_true(all(as.numeric(o[2,-c(1,2)]) == c(5, 5, 0, 5, 1, 1)),
-              info = "case two cherries incorrect count output for second cherry")
-  
-})
-
 test_that("get_siteFChange_cherry_perMethDiffType", {
   
   # Test error when tree has only one tip

### 2025-02-11 tests/testthat/test-summaryStatistics.R

@@ -910,10 +910,14 @@ test_that("get_cherryDist processing of different input types", {
   type <- "One cherry, numeric tip labels ordered"
   tree <- "((1:0.1,2:0.1):2,3:3);"
   output <- get_cherryDist(tree)
-  expect_equal(output$first_tip, "1",
-               info = paste("Fails to output correct $first tip with input type:", type))
-  expect_equal(output$second_tip, "2",
-               info = paste("Fails to output correct $second_tip tip with input type:", type))
+  expect_equal(output$first_tip_name, "1",
+               info = paste("Fails to output correct $first_tip_name with input type:", type))
+  expect_equal(output$second_tip_name, "2",
+               info = paste("Fails to output correct $second_tip_name with input type:", type))
+  expect_equal(output$first_tip_index, 1,
+               info = paste("Fails to output correct $first_tip_index with input type:", type))
+  expect_equal(output$second_tip_index, 2,
+               info = paste("Fails to output correct $second_tip_index with input type:", type))
   expect_equal(output$dist, 0.2,
                info = paste("Fails to output correct $dist tip with input type:", type))
   
@@ -921,56 +925,247 @@ test_that("get_cherryDist processing of different input types", {
   type <- "One cherry, numeric tip labels unordered"
   tree <- "((3:0.15,1:0.2):2,2:3);"
   output <- get_cherryDist(tree)
-  expect_equal(output$first_tip, "3",
-               info = paste("Fails to output correct $first tip with input type:", type))
-  expect_equal(output$second_tip, "1",
-               info = paste("Fails to output correct $second_tip tip with input type:", type))
+  expect_equal(output$first_tip_name, "3",
+               info = paste("Fails to output correct $first_tip_name with input type:", type))
+  expect_equal(output$second_tip_name, "1",
+               info = paste("Fails to output correct $second_tip_name with input type:", type))
+  expect_equal(output$first_tip_index, 1,
+               info = paste("Fails to output correct $first_tip_index with input type:", type))
+  expect_equal(output$second_tip_index, 2,
+               info = paste("Fails to output correct $second_tip_index with input type:", type))
   expect_equal(output$dist, 0.35,
[…]

### 2025-02-11 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
began breaking lines
Updating changes in draft CFTP convergence

@@
-q---
+---
 title: "CFTP Convergence"
@@
 The output files are written under a subdirectory `Figures` generated in the same directory of the snakemake workflow, and are named "CFTP_testConvergence_paramsID_{parameter_combination_number}" both in `.pdf` and `.png` format.
+
+# TO ADD
+
+- Add references
+
+- Simplify sentences
+
+- Step in the Workflow and then mention Snakemake and conda as an additional option.
+
+- Instead of whan using nohup, if using nohup

### 2025-02-26 extensive_tests/draft-sumstat-vignette.Rmd
Updating functions to compute freq of sites changing at each methylation structure

@@
 - `countSites_cherryMethDiff(cherryDist, data)` to get for each cherry the counts of sites with half-methylation and full-methylation changes per genomic structure (island or non-island).
+
+- `freqSites_cherryMethDiff(tree,data)`.
+The function first validates the tree structure and extracts pairwise distances between cherry tips with `get_cherryDist`.
+It then validates the input data and counts half and full methylation differences for each cherry at each structure using `countSites_cherryMethDiff`.
+Finally, it normalizes these (half and full) counts by the number of sites per structure to compute frequencies.
+.
+
+- `get_siteFChange_cherry(tree,data)`.
+Uses `freqSites_cherryMethDiff(tree,data)` and then computes the frequency of sites with any change of methylation state (full or half) for each cherry at each genomic structure.
+
+```{r}
+# Set example tree and methylation data
+tree <- "((a:1.5,b:1.5):2,(c:2,d:2):1.5);"
+data <- list(
+    list(rep(1,10), rep(0,5), rep(1,8)), # tip a
+    list(rep(1,10), rep(0.5,5), rep(0,8)), # tip b
+    list(rep(1,10), rep(0.5,5), rep(0,8)), # tip c
+    list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1), c(0.5, 1, rep(0, 6)))) # tip d
+
+get_siteFChange_cherry(tree = tree, data = data)
+```

### 2025-02-26 R/summaryStatistics.R

@@ -477,7 +477,63 @@ compute_meanCor_ni <- function(index_nonislands, minN_CpG, shore_length, data, s
   return(mean(cor))
 }
 
+
+
 #### #### #### Tree cherries comparisons #### #### ####
+
+#' Validate and Parse a Phylogenetic Tree
+#'
+#' This function checks whether the input is a valid phylogenetic tree, either as a character string in Newick format
+#' or as an object of class \code{phylo} from the \code{ape} package. If the input is a Newick string, it is parsed into 
+#' a \code{phylo} object. The function also ensures that the tree contains at least two tips.
+#'
+#' @param tree A phylogenetic tree in Newick format (as a character string) or an object of class \code{phylo} from the \code{ape} package.
+#'   - If the input is a character string, it must follow the Newick format (e.g., \code{"((tip_1:1,tip_2:1):5,tip_3:6);"}).
+#'   - If an object of class \code{phylo} is provided, it should be a valid phylogenetic tree.
+#'
+#' @return A \code{phylo} object representing the validated and parsed tree.
+#'
+#' @details 
+#' - The function first verifies that the input is either a valid \code{phylo} object or a character string.
+#' - If the input is a Newick string, it attempts to parse it into a \code{phylo} object using \code{ape::read.tree()}.
+#' - If parsing fails, an informative error message is returned.
+#' - The function also checks that the tree contains at least two tips, as a valid phylogenetic tree should have at least one split.
+#'
+#' @importFrom ape read.tree
+validate_tree <- function(tree){
+  
+  # Check that the input is a character string or phylo object
+  if (!is.character(tree) && !inherits(tree, "phylo")) {
+    stop(paste("Argument 'tree' must be a character string in newick format (e.g. '((tip_1:1,tip_2:1):5,tip_3:6);').\n",
+               "or an object of class 'phylo' from the 'ape' package. For more details, see ?phylo or ?ape."))
+  }
+  
+  # Transform the newick tree in a phylo object managing issues with given tree format
+  if(!inherits(tree, "phylo")) {
+    tryCatch({
+      # Attempt to read the tree
+      tree <- ape::read.tree(text = tree)
+      # Check if the result is NULL, which indicates an issue with the tree format
+      if (is.null(tree)) {
+        stop("Error in ape::read.tree(tree): Tree could not be parsed. Invalid format.")
+      }
+    }, warning = function(w) {
[…]

### 2025-02-26 extensive_tests/toClean-sumstats.R

@@ -13,28 +13,6 @@ index_nonislands <- c(2,4)
 
 
 
-
-
-#undebug(countSites_cherryMethDiff)
-#undebug(get_cherryDist)
-
-##TODO: n_CpG not given but calculated. Each structure is assumed to have the same number of sites at all tips
-get_siteFChange_cherry_perMethDiffType <- function(tree, data, sample_n){
-  if(sample_n < 2){stop("Minimum number of tips/samples needed: 2")}
-  cherries <- get_cherryDist(tree = tree, sample_n = sample_n)
-  siteFChange_cherry_perMethDiffType <- countSites_cherryMethDiff(cherries = cherries, data = data)
-  str_n <- length(data[[1]]) # set the number of structures
-  sites_n <- numeric(length = str_n)
-  for(str in 1:str_n) sites_n[str] <- length(data[[1]][[str]])
-  # Each value is a count number of full methylation changes _f or changes in one strand _h.
-  # Each structure has a number of CpGs. Then, if we wanna count the proportion
-  # of sites, we divide each value by the number of CpGs. 
-  # Explicitly match sites_n to the columns of siteFChange_cherry_perMethDiffType
-  sites_n_repeated <- rep(sites_n, each = 2)
-  siteFChange_cherry_perMethDiffType[,3:ncol(siteFChange_cherry_perMethDiffType)] <- sweep(siteFChange_cherry_perMethDiffType[,3:ncol(siteFChange_cherry_perMethDiffType)], 2, sites_n_repeated, "/")
-  siteFChange_cherry_perMethDiffType
-}
-
 get_siteFChange_cherry <- function(tree, data, sample_n){
   if(sample_n < 2){stop("Minimum number of tips/samples needed: 2")}
   siteFChange_cherry_perMethDiffType <- get_siteFChange_cherry_perMethDiffType(tree = tree, data = data, sample_n = sample_n)

### 2025-02-26 extensive_tests/toClean-test-sumstats.R

@@ -1,53 +1,5 @@
 
 
-test_that("get_siteFChange_cherry_perMethDiffType", {
-  
-  # Test error when tree has only one tip
-  newick_tree <- "(1:1);"
-  data <- list(rep(1,10), rep(0,10), rep(1,10))
-  expect_error(get_siteFChange_cherry_perMethDiffType(tree = newick_tree, data = data, sample_n = 1),
-               info = "fails to throw an error when number of tips is < 2")
-  
-  # Test error when arguments are missing
-  newick_tree <- "((1:1,2:1):1,3:2);"
-  data <- list(
-    list(rep(1,10), rep(0,10), rep(1,10)),
-    list(rep(1,10), rep(0.5,10), rep(0,10)),
-    list(rep(1,10), rep(0.5,10), rep(0,10)))
-  expect_error(get_siteFChange_cherry_perMethDiffType(data = data, sample_n = 3),
-               info = "fails to throw an error when argument 'tree' is not given")
-  expect_error(get_siteFChange_cherry_perMethDiffType(tree = newick_tree, sample_n = 3),
-               info = "fails to throw an error when argument 'data' is not given")
-  expect_error(get_siteFChange_cherry_perMethDiffType(tree = newick_tree, data = data),
-               info = "fails to throw an error when argument 'sample_n' is not given")
-  
-  # Test counts are transformed into frequencies correctly
-  tree <- "((1:1.5,2:1.5):2,(3:2,4:2):1.5);"
-  data <- list(
-    list(rep(1,10), rep(0,5), rep(1,8)),
-    list(rep(1,10), rep(0.5,5), rep(0,8)),
-    list(rep(1,10), rep(0.5,5), rep(0,8)),
-    list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1), c(0.5, 1, rep(0, 6))))
-  o <- get_siteFChange_cherry_perMethDiffType(tree = tree, data = data, sample_n = 4)
-  expect_equal(o$tips[1], "1-2",
-               info = "incorrect tips id in cherry 1")
-  expect_equal(o$tips[2], "3-4",
-               info = "incorrect tips id in cherry 2")
-  expect_equal(o$dist[1], 3,
-               info = "incorrect tips distance in cherry 1")
-  expect_equal(o$dist[2], 4,
[…]

### 2025-02-26 tests/testthat/test-summaryStatistics.R

@@ -881,27 +881,43 @@ test_that("meanCor", {
                info = "MeanCor_ni returns non-correct value (shore 10 2 tips non-equal sizes)")
 })
 
+test_that("validate_tree input errors", {
+  # No input
+  expect_error(validate_tree(),
+               info = "function fails to throw an error when no tree is given")
+  
+  # Tree not character string or phylo class
+  tree <- 5
+  expect_error(validate_tree(tree),
+               info = "function fails to throw error when tree is not character string or phylo object")
+  
+  # Incorrect newick format
+  tree <- "a:1, b:2"
+  expect_error(validate_tree(tree),
+               info = "function fails to throw error when given tree has incorrect format")
+  tree <- "a:1, b:2;"
+  expect_error(validate_tree(tree),
+               info = "function fails to throw error when given tree has incorrect format")
+  # Test error when tree has only one tip
+  tree <- "(1:1);"
+  expect_error(validate_tree(tree),
+               info = "function fails to throw error when given tree has only one tip")
+})
+
+
+test_that("validate_tree returns phylo tree",{
+  tree <- "(a:1,b:2);"
+  o <- validate_tree(tree)
+  expect_equal(class(o), "phylo")
+})
+
+
 test_that("get_cherryDist input control errors", {
   
   # No input
   expect_error(get_cherryDist(),
                info = "function fails to throw an error when no tree is given")
-  # Tree not character string or phylo class
-  tree <- 5
-  expect_error(get_cherryDist(tree),
-               info = "function fails to throw error when tree is not character string or phylo object")
   
-  # Incorrect newick format
-  tree <- "a:1, b:2"
-  expect_error(get_cherryDist(tree),
-               info = "function fails to throw error when given tree has incorrect format")
-  tree <- "a:1, b:2;"
-  expect_error(get_cherryDist(tree),
-               info = "function fails to throw error when given tree has incorrect format")
-  # Test error when tree has only one tip
-  tree <- "(1:1);"
-  expect_error(get_cherryDist(tree),
[…]

### 2025-02-27 extensive_tests/draft-sumstat-vignette.Rmd
Updating summary statistics with finished meanSiteFChange_cherry

@@
 
+- `MeanSiteFChange_cherry(tree, data, index_islands, index_nonislands)`.
+The function computes the per-cherry frequency of sites with different methylation states at each structure (islands and non-islands) using \code{get_siteFChange_cherry}.
+Then it calculates the weighted mean site frequency of methylation changes for each cherry separately for islands and non-islands.
+
 ```{r}
@@
     list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1), c(0.5, 1, rep(0, 6)))) # tip d 
+# Set the index for islands and non-island structures
+index_islands <- c(2)
+index_nonislands <- c(1, 3)
 
-get_siteFChange_cherry(tree = tree, data = data)
+MeanSiteFChange_cherry(data = data, 
+                       tree = tree, 
+                       index_islands = index_islands, 
+                       index_nonislands = index_nonislands)
 ```

### 2025-02-27 R/summaryStatistics.R

@@ -1,3 +1,56 @@
+#' Validate Structure Indices for Island and Non-Island Data
+#'
+#' This function checks whether the provided indices for islands and non-islands 
+#' are within the valid range of structures in the dataset. It also warns if 
+#' any indices are present in both `index_islands` and `index_nonislands`.
+#'
+#' @param data A nested list \code{data[[tip]][[structure]]}. 
+#'   Assumes that the number of structures is consistent across tips and that
+#'   within each structure, all tips have the same number of sites. 
+#'   The number of structures is inferred from `length(data[[1]])`.
+#' @param index_islands An integer vector specifying indices that correspond to island structures.
+#' @param index_nonislands An integer vector specifying indices that correspond to non-island structures.
+#'
+#' @details The funct@exportion performs the following checks:
+#' - Ensures that all indices in `index_islands` and `index_nonislands` are within 
+#'   the range of available structures.
+#' - Throws an error if any index is out of bounds.
+#' - Issues a warning if the same index appears in both `index_islands` and `index_nonislands`.
+#'
+#' @return No return value. The function stops execution if invalid indices are detected.
+validate_structureIndices <- function(data, index_islands, index_nonislands) {
+  
+  # Get the number of structures
+  str_n <- length(data[[1]])
+  
+  # Define the valid range
+  valid_indices <- 1:str_n
+  
+  # Check whether there are invalid indices
+  invalid_island_indices <- index_islands[!(index_islands %in% valid_indices)]
+  invalid_nonisland_indices <- index_nonislands[!(index_nonislands %in% valid_indices)]
+  
+  # If there are any invalid indices, throw an error
+  if (length(invalid_island_indices) > 0) {
+    stop(sprintf("Invalid island indices detected: %s. Number of structures in given data: %d", 
+                 paste(invalid_island_indices, collapse = ", "), str_n))
+  }
+  
+  if (length(invalid_nonisland_indices) > 0) {
+    stop(sprintf("Invalid non-island indices detected: %s. Number of structures in given data: %d", 
+                 paste(invalid_nonisland_indices, collapse = ", "), str_n))
+  }
+  
+  # Check for overlapping indices
[…]

### 2025-02-27 extensive_tests/toClean-sumstats.R

@@ -1,42 +1,4 @@
 
-index_islands <- c(1,3)
-index_nonislands <- c(2,4)
-# Simulated data has all positions
-# But empirical data may have differences in coverage
-# E.g. if in island one positions 1,3,5,7 are CpGs
-# Empirical data may have in first_tip 1,5,7
-# And in second tip 1,3,5,7
-# So that only sites 1,5,7 can be compared
-# Previous step of filtering data using a list containing for each island and non-island region a vector of the
-# relative indices of the CpGs that are to be compared.
-
-
-
-
-get_siteFChange_cherry <- function(tree, data, sample_n){
-  if(sample_n < 2){stop("Minimum number of tips/samples needed: 2")}
-  siteFChange_cherry_perMethDiffType <- get_siteFChange_cherry_perMethDiffType(tree = tree, data = data, sample_n = sample_n)
-  # Identify the unique structures in the column names
-  structures <- unique(sub("_[fh]$", "", names(siteFChange_cherry_perMethDiffType)[-c(1, 2)]))
-  # Create the new dataframe
-  siteFChange_cherry <- data.frame(
-    tips = siteFChange_cherry_perMethDiffType$tips,
-    dist = siteFChange_cherry_perMethDiffType$dist
-  )
-  # Add the summed columns
-  for (str in structures){
-    siteFChange_cherry[[str]] <- rowSums(siteFChange_cherry_perMethDiffType[, grep(paste0("^", str, "_"), names(siteFChange_cherry_perMethDiffType))])
-  }
-  siteFChange_cherry
-}
-
-## For something like before:
-# There are 2 columns per structure in siteFChange_cherry_perMethDiffType
-# _f represents the frequency of full changes (both DNA strands)
-# _h represents the frequency of half changes (one DNA strand)
-#siteFChange_cherry_perMethDiffType[,grep("h", colnames(siteFChange_cherry_perMethDiffType))] <- siteFChange_cherry_perMethDiffType[,grep("h", colnames(siteFChange_cherry_perMethDiffType))]/2
-
-#index islands: vector
 
 SDandMean_SiteFChange_cherry <- function(siteFChange_cherry, index_islands, index_nonislands){
   siteFChange <- siteFChange_cherry[-c(1,2)]

### 2025-02-27 extensive_tests/toClean-test-sumstats.R

@@ -1,105 +1,5 @@
 
 
-test_that("get_siteFChange_cherry", {
-  tree <- "((1:1.5,2:1.5):2,(3:2,4:2):1.5);"
-  data <- list(
-    list(rep(1,10), rep(0,5), rep(1,8)),
-    list(rep(1,10), rep(0.5,5), rep(0,8)),
-    list(rep(1,10), rep(0.5,5), rep(0,8)),
-    list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1), c(0.5, 1, rep(0, 6))))
-  o <- get_siteFChange_cherry(tree = tree, data = data, sample_n = 4)
-  exp_siteFChange_cherry1 <- c(0, 1, 1)
-  exp_siteFChange_cherry2 <- c(1, 1, 2/8)
-  expect_equal(as.numeric(o[1,3:ncol(o)]), exp_siteFChange_cherry1,
-               info = "incorrect frequency of site FChange cherry 1")
-  expect_equal(as.numeric(o[2,3:ncol(o)]), exp_siteFChange_cherry2,
-               info = "incorrect frequency of site FChange cherry 2")
-})
 
-test_that("SDandMean_SiteFChange_cherry", {
-  tree <- "((1:1.5,2:1.5):2,(3:2,4:2):1.5);"
-  data <- list(
-    list(rep(1,10), rep(0,5), rep(1,8)),
-    list(rep(1,10), rep(0.5,5), rep(0,8)),
-    list(rep(1,10), rep(0.5,5), rep(0,8)),
-    list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1), c(0.5, 1, rep(0, 6))))
-  siteFChange_cherry <- get_siteFChange_cherry(tree = tree, data = data, sample_n = 4)
-  
-  # Test case all structures are islands
-  index_islands <- c(1,2,3)
-  index_nonislands <- c()
-  o <- SDandMean_SiteFChange_cherry(siteFChange_cherry = siteFChange_cherry, index_islands = index_islands, index_nonislands = index_nonislands)
-  exp_mean_I_ch1 <- mean(c(0,1,1))
-  exp_mean_I_ch2 <- mean(c(1,1,0.25))
-  exp_sd_I_ch1 <- sd(c(0,1,1))
-  exp_sd_I_ch2 <- sd(c(1,1,0.25))
-  expect_equal(o$mean_I, c(exp_mean_I_ch1, exp_mean_I_ch2),
[…]

### 2025-02-27 tests/testthat/test-summaryStatistics.R

@@ -1,3 +1,21 @@
+test_that("get_MeanFreqP non-valid structureIndices", {
+  
+  data <- list(
+    list(c(0, 0, 0), c(0, 0, 0)), # tip 1
+    list(c(0, 0, 0), c(1, 1, 1)) # tip 2
+  )
+  
+  index_islands <- c(1,3)
+  index_nonislands <- c(1,3)
+  
+  expect_error(get_islandMeanFreqP(data = data, index_islands = index_islands, sample_n =2),
+               "Invalid island indices detected: 3. Number of structures in given data: 2",
+               info = "fails to throw error with incorrect island index")
+  expect_error(get_nonislandMeanFreqP(data = data, index_nonislands = index_nonislands, sample_n =2),
+               "Invalid non-island indices detected: 3. Number of structures in given data: 2",
+               info = "fails to throw error with incorrect nonisland index")
+})
+
 test_that("get_MeanFreqP structures with equal length islands and non-islands", {
   # Two tips and two islands
   index_islands <- c(1, 2)
@@ -104,6 +122,24 @@ test_that("get_MeanFreqP structures with different length (islands and non-islan
  
 })
 
+test_that("get_MeanFreqM non-valid structureIndices", {
+  
+  data <- list(
+    list(c(0, 0, 0), c(0, 0, 0)), # tip 1
+    list(c(0, 0, 0), c(1, 1, 1)) # tip 2
+  )
+  
+  index_islands <- c(1,3)
+  index_nonislands <- c(1,3)
+  
+  expect_error(get_islandMeanFreqM(data = data, index_islands = index_islands, sample_n =2),
+               "Invalid island indices detected: 3. Number of structures in given data: 2",
+               info = "fails to throw error with incorrect island index")
+  expect_error(get_nonislandMeanFreqM(data = data, index_nonislands = index_nonislands, sample_n =2),
+               "Invalid non-island indices detected: 3. Number of structures in given data: 2",
+               info = "fails to throw error with incorrect nonisland index")
+})
+
 
 test_that("get_MeanFreqM structures with equal length islands and non-islands", {
   # Two tips and two islands
@@ -204,6 +240,23 @@ test_that("get_MeanFreqM structures with different length (islands and non-islan
[…]

### 2025-03-01 extensive_tests/draft-sumstat-vignette.Rmd
Updating functions for fitch estimation per island and corresponding tests

@@
 
-### Frequency of Methylation State Changes at Single Sites
+### Mean Frequency of Methylation State Changes at Single Sites per Structure Type
 
-We compare the methylation state of single sites at the tree cherries.
-A cherry is a pair of leaf nodes (also called tips or terminal nodes) in a phylogenetic tree that share a direct common ancestor.
-In other words, if two leaves are connected to the same internal node and no other leaves are connected to that internal node, they form a cherry.
+Let \(S_k\) be the set of sites in the genomic structure indexed by \(k\).
+For each site \(s \in S_k\), let the methylation state at a given tip \(t\) be denoted as \(m_{k,s,t}\), where \(m_{k,s,t} \in \{0, 0.5, 1\}\) represents the unmethylated, partially-methylated, or methylated state, respectively.
 
-We calculate the frequency of sites that changes for each structure (island and non-island structures).
-
-Let $S_k$ be the set of sites in the cherry structure of interest, where $k$ represents the index of the cherry structure.
-For each site in $S_k$, let the methylation state at a given tip $t$ be denoted as $m_{k,s,t}$, where $s \in S_k$ is the index of the site, and $m_{k,s,t} \in {0, 0.5, 1}$ represents the methylation state at tip $t$ (unmethylated, partially-methylated, or methylated, respectively).
-
-For a given cherry structure, we compute in the proportion of sites that have a different methylation state between two tips, say $t_1$ and $t_2$.
+For a given cherry \((t_1, t_2)\), we compute the proportion of sites with differing methylation states between the two tips.
 
 Define the indicator variable:
-\[
+$$
 \delta_{k,s,t_1,t_2} =
@@
 \end{cases}
-\]
+$$
 
-Then, the summary statistic for the proportion of sites with different methylation states between tips $t_1$ and $t_2$ in the cherry structure $k$ is:
-\[
-\text{Proportion}_k(t_1, t_2) = \frac{1}{|S_k|} \sum_{s \in S_k} \delta_{k,s,t_1,t_2}
-\]
+Then, the proportion of sites with different methylation states in structure \(k\) for the cherry \((t_1, t_2)\) is given by:
 
-where $|S_k|$ is the total number of sites in the cherry structure $k$, and the sum counts the number of sites with different methylation states between tips $t_1$ and $t_2$.
+$$
+F_k(t_1, t_2) = \frac{1}{|S_k|} \sum_{s \in S_k} \delta_{k,s,t_1,t_2}
+$$
 
-This statistic gives the proportion of sites in the cherry structure $k$ where the methylation state differs between the two tips of interest.
+where \(|S_k|\) is the total number of sites in structure \(k\), and the sum counts the number of sites where the methylation state differs between tips \(t_1\) and \(t_2\).
+
+This statistic quantifies the proportion of differing methylation states for a given genomic structure in a cherry.
+
+To compute the weighted mean frequency of methylation changes across all structures in a cherry, we separate island and non-island structures.
+
+For a given cherry \((t_1, t_2)\), let \( \mathcal{I} \) be the set of indices for island structures and \( \mathcal{N} \) be the set of indices for non-island structures.
+Each structure \( k \) has a total number of sites \( |S_k| \), which we use as weights.
+
+The weighted mean frequency of methylation changes for island structures is:
+
+$$
+\bar{F}_{\text{island}}(t_1, t_2) = \frac{\sum_{k \in \mathcal{I}} |S_k| F_k(t_1, t_2)}{\sum_{k \in \mathcal{I}} |S_k|}
+$$
+
+Similarly, the weighted mean frequency for non-island structures is:
+
+$$
+\bar{F}_{\text{non-island}}(t_1, t_2) = \frac{\sum_{k \in \mathcal{N}} |S_k| F_k(t_1, t_2)}{\sum_{k \in \mathcal{N}} |S_k|}
+$$
+
+These expressions compute the mean frequency of methylation state differences across all island and non-island structures, weighted by the number of sites in each structure.
 
@@
 ```
+
+# Fitch
+
+- `get_meanMeth_islands` computes the mean methylation state of all the islands. **Maybe this should allow excluding island borders**

### 2025-03-01 R/summaryStatistics.R

@@ -85,6 +85,7 @@ get_islandMeanFreqP <- function(index_islands, data, sample_n){
   
   # Validate index islands
   tryCatch({
+    if(length(index_islands) == 0) stop("'index_islands' has no indices")
     validate_structureIndices(data, index_islands, index_nonislands = c())
   }, warning = function(w) {
     stop(conditionMessage(w))
@@ -139,6 +140,7 @@ get_nonislandMeanFreqP <- function(index_nonislands, data, sample_n){
   
   # Validate index non-islands
   tryCatch({
+    if(length(index_nonislands) == 0) stop("'index_nonislands' has no indices")
     validate_structureIndices(data, index_islands = c(), index_nonislands)
   }, warning = function(w) {
     stop(conditionMessage(w))
@@ -194,6 +196,7 @@ get_islandMeanFreqM <- function(index_islands, data, sample_n){
   
   # Validate index islands
   tryCatch({
+    if(length(index_islands) == 0) stop("'index_islands' has no indices")
     validate_structureIndices(data, index_islands, index_nonislands = c())
   }, warning = function(w) {
     stop(conditionMessage(w))
@@ -246,6 +249,7 @@ get_nonislandMeanFreqM <- function(index_nonislands, data, sample_n){
   
   # Validate index non-islands
   tryCatch({
+    if(length(index_nonislands) == 0) stop("'index_nonislands' has no indices")
     validate_structureIndices(data, index_islands = c(), index_nonislands)
   }, warning = function(w) {
     stop(conditionMessage(w))
@@ -685,6 +689,11 @@ validate_tree <- function(tree){
   
   # Check given tree has minium two tips
   if (length(tree$tip.label)<2) stop("The input 'tree' must have a minimum of 2 tips.")
+  
+  # Check given tree has unique tip labels
+  if (any(duplicated(tree$tip.label))) {
+    stop(paste("The input tree has duplicated tree labels:", paste(unique(tree$tip.label[duplicated(tree$tip.label)]), collapse = ", ")))
+  }
   
   # Return tree as phylo object
   tree
@@ -1200,6 +1209,184 @@ MeanSiteFChange_cherry <- function(data, tree, index_islands, index_nonislands){
   
   MeanSiteFChange_cherry
 }
+
+#### #### #### Fitch estimation of minimum number of global methylation changes per island #### #### ####
+##TODO: Check again name
+
+##TODO: document
+# data: A list containing methylation states at tree tips for each genomic structure 
[…]

### 2025-03-01 extensive_tests/toClean-sumstats.R

@@ -1,28 +1,6 @@
 
 
-SDandMean_SiteFChange_cherry <- function(siteFChange_cherry, index_islands, index_nonislands){
-  siteFChange <- siteFChange_cherry[-c(1,2)]
-  if(length(index_nonislands)>0){
-    mean_NI <- rowMeans(as.data.frame(siteFChange[,index_nonislands]))
-    sd_NI <- apply(as.data.frame(siteFChange[,index_nonislands]), 1, sd)
-  } else {
-    mean_NI = sd_NI <- NA
-  }
-  if(length(index_islands)>0){
-    mean_I <- rowMeans(as.data.frame(siteFChange[,index_islands]))
-    sd_I <- apply(as.data.frame(siteFChange[,index_islands]), 1, sd)
-  } else {
-    mean_I = sd_I <- NA
-  }
-  SDandMean_SiteFChange_cherry  <- data.frame(
-    tips = siteFChange_cherry$tips,
-    dist = siteFChange_cherry$dist,
-    mean_I = mean_I,
-    sd_I = sd_I,
-    mean_NI = mean_NI,
-    sd_NI = sd_NI
-  )
-  SDandMean_SiteFChange_cherry
-}
-#undebug(SDandMean_SiteFChange_cherry)
-##TODO: Add index_nonislands
+
+
+
+

### 2025-03-01 tests/testthat/test-summaryStatistics.R

@@ -14,6 +14,15 @@ test_that("get_MeanFreqP non-valid structureIndices", {
   expect_error(get_nonislandMeanFreqP(data = data, index_nonislands = index_nonislands, sample_n =2),
                "Invalid non-island indices detected: 3. Number of structures in given data: 2",
                info = "fails to throw error with incorrect nonisland index")
+  
+  index_islands <- c()
+  index_nonislands <- c()
+  
+  expect_error(get_islandMeanFreqP(data = data, index_islands = index_islands, sample_n =2),
+               "'index_islands' has no indices")
+  expect_error(get_nonislandMeanFreqP(data = data, index_nonislands = index_nonislands, sample_n =2),
+               "'index_nonislands' has no indices")
+  
 })
 
 test_that("get_MeanFreqP structures with equal length islands and non-islands", {
@@ -138,6 +147,14 @@ test_that("get_MeanFreqM non-valid structureIndices", {
   expect_error(get_nonislandMeanFreqM(data = data, index_nonislands = index_nonislands, sample_n =2),
                "Invalid non-island indices detected: 3. Number of structures in given data: 2",
                info = "fails to throw error with incorrect nonisland index")
+  
+  index_islands <- c()
+  index_nonislands <- c()
+  
+  expect_error(get_islandMeanFreqM(data = data, index_islands = index_islands, sample_n =2),
+               "'index_islands' has no indices")
+  expect_error(get_nonislandMeanFreqM(data = data, index_nonislands = index_nonislands, sample_n =2),
+               "'index_nonislands' has no indices")
 })
 
 
@@ -989,6 +1006,22 @@ test_that("validate_tree input errors", {
   tree <- "(1:1);"
   expect_error(validate_tree(tree),
                info = "function fails to throw error when given tree has only one tip")
+  
+  # Test error when tree has one duplicated tip label
+  tree <- "((a:1,b:1):2,(d:1.25,d:1.25):0.5);"
+  expect_error(validate_tree(tree),
+               "The input tree has duplicated tree labels: d",
[…]

### 2025-03-03 extensive_tests/draft-sumstat-vignette.Rmd
Updating draft vignette summarystatistics

@@
 ```
+# Introduction
 
-Interpretation of arguments:
+This document presents summary statistics related to the evolution of methylation 
+states across different genomic structures. The statistics include mean 
+frequencies, standard deviations, neighbor correlations of methylation states in 
+island and non-island regions. Additionally, for analysis comparing different 
+samples (comparison of tree tips), the included summary statistics compute
+the mean per-site frequency of methylation changes and the Fitch estimate for
+the minimum number of changes per islands leading to different global methylation
+states.
 
- - tip can be interpreted as sample or replicate
+The provided set of functions allows the computation of summary statistics for 
+methylation data in a genomic region with structural categorization. In this 
+vignette, we will refer to these structures as **islands** and 
+**non-island structures**.
 
- - structure can be interpreted as island or non-island
+The `data` argument must be in one of the following formats, depending on the 
+analysis:
 
-## Frequency of methylation states per genomic structure type
+1. **For analyses without tip comparison:** A list where each element represents 
+the methylation states in a given structure (`data[[structure]]`):
+
+```{r}
+# Example: a single sample with 3 genomic structures
+# (1) island with 10 partially-methylated sites
+# (2) non-island with 5 methylated sites
+# (3) island with 15 unmethylated sites
+
+data <- list(rep(0.5, 10),  # Partially methylated
+             rep(1,5),      # Methylated
+             rep(0,15))     # Unmethylated
+data
+```
+
+2. **For analyses comparing multiple tips in a phylogenetic tree:** A nested 
+list where the first level represents the tips, and the second level represents 
+the structures within each tip (`data[[tip]][[structure]]`).
+
+```{r}
+# Example: data from 3 tips of a tree,
+# each with 3 genomic structures
+data <- list(
+    # Tip 1
+    list(c(rep(0.5,5), rep(0,5)),  # 5 partially methylated, 5 unmethylated
+         c(rep(1,4), 0.5),         # 4 methylated, 1 unmethylated
+         c(rep(0,7), rep(0.5,8))), # 7 unmethylated, 8 partially methylated
+    # Tip 2
+    list(c(rep(0.5,9), 1), # 9 partially methylated, 1 methylated
+         c(rep(0.5,4), 1), # 4 partially methylated, 1 methylated
+         c(rep(0,8), rep(0.5,7))), # 8 unmethylated, 7 partially-methylated
+    # Tip 3
+    list(c(1, rep(0,8), 1), # first and last methylated, rest unmethylated
+         c(rep(0.5,3), rep(1,2)), # 3 methylated, 1 unmethylated
+         c(rep(0.5,15)))) # all partially methylated
+```
+
+In this case, `data` must be pre-filtered to include only sites present in all 
+tips, ensuring a valid comparison between species or samples.
+
+Regardless of the format, methylation values should be represented as follows:
+
+- `0` for unmethylated sites,
+
+- `0.5` for partially methylated sites,
+
+- `1` for methylated sites.
+
+Intermediate values (e.g., obtained from pooled empirical data) should be 
+categorized before analysis.
+
+Finally, when phylogenetic data is provided, the order of tips in `data` must 
+match the tip order in the Newick tree format (left to right) or the order of 
+`tree$tip.label()` when the tree is provided as a `phylo` object from the `ape` 
+package.
+
+```{r}
+# Example tree in Newick format for the above data
+newick_tree <- "((tip1:1, tip2:1):1, tip3:2);"
+
+# Example tree as a phylo object from the ape package
+library(ape)
+phylo_tree <- read.tree(text = newick_tree)
+phylo_tree$tip.label
+```
+
+# Summary Statistics
+
+## Frequency of Methylation States per Genomic Structure Type
 
@@
 
-- **Partially-methylated sites**: Let $x_{i,t}$ and $y_{j,t}$ be the frequency of methylation state $p$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$ at tip $t$, respectively.
-We define the mean global frequency of $x$ and $y$ as:
+- **Partially-methylated sites**: Let $x_{i,t}$ and $y_{j,t}$ be the frequency 
+of methylation state $p$ in island $i \in\{1, \cdots, I\}$ and non-island 
+$j \in\{1, \cdots, J\}$ at tip $t$, respectively. We define the mean global 
+frequency of $x$ and $y$ as:
 
-\[
+$$
 \overline{x}=\frac1I\sum_{i=1}^{I}\frac1T\sum_{t=1}^{T}x_{i,t} \\
 \overline{y}=\frac1J\sum_{j=1}^{J}\frac1T\sum_{t=1}^{T}y_{j,t}
-\]
+$$
 
-, where $x_{i,t}$ is the frequency of state $p$ in island $i$ and tip $t$ and $y_{j,t}$ is the frequency of state $p$ in non-island $j$ and tip $t$.
+where $x_{i,t}$ and $y_{j,t}$ represent the respective frequencies at each tip.
 
-- **Methylated sites**: Let $w_{i,t}$ and $z_{j,t}$ be the frequency of methylation state $m$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$ at tip $t$, respectively.
-We define the mean global frequency of $w$ and $z$ as:
+- **Methylated sites**: Similarly, let $w_{i,t}$ and $z_{j,t}$ be the frequency 
+of methylation state $m$ in islands and non-islands. The mean global frequencies 
+are:
 
-\[
+$$
 \overline{w}=\frac1I\sum_{i=1}^{I}\frac1T\sum_{t=1}^{T}w_{i,t} \\
 \overline{z}=\frac1J\sum_{j=1}^{J}\frac1T\sum_{t=1}^{T}z_{j,t}
-\]
+$$
 
-, where $w_{i,t}$ is the frequency of state $p$ in island $i$ and tip $t$ and $z_{j,t}$ is the frequency of state $m$ in non-island $j$ and tip $t$.
-
-These values can be computed with the functions `get_islandMeanFreqP`, `get_nonislandMeanFreqP`,`get_islandMeanFreqM` and `get_nonislandMeanFreqM` as in the following examples:
+These values can be computed using the following functions:
 
@@
 
-### Mean SD of observed frequencies per structure type at tree tips
+### Mean SD of Observed Frequencies per Structure Type at Tree Tips
 
-- **Partially-methylated sites**: Let $x_{i,t}$ and $y_{j,t}$ be the frequency of methylation state $p$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$, respectively.
-We define the standard deviation of the frequency of $x$ and $y$ at tip $t$ as:
+- **Partially-methylated sites**: We define the standard deviation of $x$ and 
+$y$ at tip $t$ as:
 
@@
 \sigma_{t}(x)=\sqrt{\frac{1}{I-1}\sum_{i=1}^{I}(x_{i,t}-\overline{x_{.,t}})^2} \\
-\sigma_{t}(y)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{i,t}-\overline{y_{.,t}})^2}
+\sigma_{t}(y)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{j,t}-\overline{y_{.,t}})^2}
 $$
 
-, where $\overline{x_{.,t}}$ is the mean frequency of state $p$ at tip $t$ in islands and $\overline{y_{.,t}}$ is the mean frequency of state $p$ at tip $t$ in non-islands:
-
-$$
-\overline{x_{.,t}}=\frac{1}{I}\sum_{i=1}^{I}x_{i,t} \\
-\overline{y_{.,t}}=\frac{1}{J}\sum_{j=1}^{J}x_{j,t}
-$$
-
-Therefore, the mean standard deviation across tips is:
+where $\overline{x_{.,t}}$ and $\overline{y_{.,t}}$ represent mean frequencies 
+at tip $t$. The mean standard deviation across tips is:
 
@@
 \hat{\sigma}(x)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(x) \\
-\hat{\sigma}(y)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(y) \\
+\hat{\sigma}(y)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(y)
 $$
 
-- **Methylated sites**: Let $w_{i,t}$ and $z_{j,t}$ be the frequency of methylation state $m$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$, respectively.
-We define the standard deviation of the frequency of $w$ and $z$ at tip $t$ as:
-
-$$
-\sigma_{t}(w)=\sqrt{\frac{1}{I-1}\sum_{i=1}^{I}(x_{i,t}-\overline{x_{.,t}})^2} \\
-\sigma_{t}(z)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{i,t}-\overline{y_{.,t}})^2}
-$$
-
-, where $\overline{x_{.,t}}$ is the mean frequency of state $m$ at tip $t$ in islands and $\overline{z_{.,t}}$ is the mean frequency of state $m$ at tip $t$ in non-islands:
-
-$$
-\overline{w_{.,t}}=\frac{1}{I}\sum_{i=1}^{I}w_{i,t} \\
-\overline{z_{.,t}}=\frac{1}{J}\sum_{j=1}^{J}z_{j,t}
-$$
-
-Therefore, the mean standard deviation across tips is:
+- **Methylated sites**: Similarly, for $w$ and $z$:
 
@@
 \hat{\sigma}(w)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(w) \\
-\hat{\sigma}(z)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(z) \\
+\hat{\sigma}(z)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(z)
 $$
-These values can be computed with the functions `get_islandSDFreqP`, `get_nonislandSDFreqP`,`get_islandSDFreqM` and `get_nonislandSDFreqM` as in the following examples:
+
+These values can be computed using:
 
@@
 index_nonislands <- c(2, 4)
-data <- list(c(.5, .5, 0, 1, 1, .5), c(.5, 0, 1, .5), c(.5, .5, 0), c(0, 0, .5)) # tip 1
+data <- list(c(.5, .5, 0, 1, 1, .5), c(.5, 0, 1, .5), c(.5, .5, 0), c(0, 0, .5))
 get_islandSDFreqP(index_islands, data, sample_n)
@@
 
-Note that we average across tips as there can be correlations between the mean standard deviation of tips (e.g. under the case of an event affecting the states at two tips).
+Note that we average across tips as there can be correlations between the mean 
+standard deviation of tips (e.g. under the case of an event affecting the states 
+at two tips).
 
-# Mean correlation of a central segment of methylation states per genomic structure type
+## Mean Correlation of a Central Segment of Methylation States per Genomic Structure Type
 
-Let $s$ be the length of the 2 adjacent segments of central positions to consider.
-Let $l$ represent a number of positions to exclude at the start and end of each island/non-island structure.
-Define the index positions $p=l+1$ and $p' = s-l$.
-Let $x_{i,t,[p,p'-1]}$ and $x_{i,t,[p+1,p']}$ be the methylation states $x \in \{u, p, m\}$ in island $i \in \{1, \cdots ,I\}$ at tip $t$ for the given segments. 
-Let $y_{j,t,[p,p'-1]}$ and $y_{j,t,[p+1,p']}$ be the methylation states $y \in \{u, p, m\}$ in non-island $j \in \{1, \cdots ,J\}$ at tip $t$ for the given segments.
+Let $s$ be the length of two adjacent segments of central positions to consider. 
+Let $l$ be the number of positions excluded at the start and end of each 
+structure. Define:
 
-We define the mean correlation as:
-
-\[
+$$
 \overline{Cor(x)_{s,l}}=\frac1I\frac1T\sum_{i=1}^{I}\sum_{t=1}^{T}Cor(x_{i,t,[p:p'-1]}, x_{i,t,[p+1:p']}) \\
 \overline{Cor(y)_{s,l}}=\frac1J\frac1T\sum_{j=1}^{J}\sum_{t=1}^{T}Cor(y_{j,t,[p:p'-1]}, y_{j,t,[p+1:p']})
-\]
+$$
 
-where correlations are computed only for segment pairs where the standard deviation of each segment is non-zero:
+Correlations are computed only for segment pairs where standard deviation is 
+non-zero:
 
-\[
+$$
 \sigma(x_{i,t,[p:p'-1]}) \neq 0 \text{ and } \sigma(x_{i,t,[p+1:p']})\neq 0 \\
 \sigma(y_{j,t,[p:p'-1]}) \neq 0 \text{ and } \sigma(y_{j,t,[p+1:p']})\neq 0
-\]
+$$
 
@@
 index_nonislands <- c(2, 4) 
-data <- list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
-               c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5), # 25 sites
-               c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1), # 40 sites
-               c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, .5, 0, 0, 0, .5)) # 35 sites
-compute_meanCor_i(index_islands, minN_CpG = 10, shore_length = 5, data, sample_n = 1)
-compute_meanCor_ni(index_nonislands, minN_CpG = 10, shore_length = 5, data, sample_n = 1)
+data <- list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, 
+               .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
+               c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, 
+                 .5, .5, 1, 1, 1, .5), # 25 sites
+               c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, 
+                 .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 
+                 0, 0, .5, 1), # 40 sites
+               c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, 
+                 .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 
+                 .5, 0, 0, 0, .5)) # 35 sites
+compute_meanCor_i(index_islands, minN_CpG = 10, 
+                  shore_length = 5, data, sample_n = 1)
+compute_meanCor_ni(index_nonislands, minN_CpG = 10, 
+                   shore_length = 5, data, sample_n = 1)
 ```
@@
   # tip 1
-    list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
-         c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5), # 25 sites
-         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1), # 40 sites
-         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, .5, 0, 0, 0, .5)), # 35 sites
+    list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 
+           1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
+         c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, 
+           .5, 1, 1, 1, .5), # 25 sites
+         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, 
+           .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, 
+           .5, 1), # 40 sites
+         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 
+           1, 1, 1, .5, .5, .5, 0, 0, 0, .5, .5, 0, 0, 0, .5)), # 35 sites
   # tip 2
-    list(c(.5, 0, 0, .5, .5, .5, 0, 0, .5, 1, .5, 0, 0, 0, 0, .5, .5, 1, 1, 1, .5, 0, 0, 0, .5, .5, 1, 1, 1, 1), # 30 sites
-         c(.5, .5, 1, 1, .5, .5, 1, 1, 1, .5, .5, 0, 0, 0, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5), # 25 sites
-         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, .5, .5, 0, 0, .5, 1, 1, 1, 1, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1), # 40 sites
-         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, 1, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, .5, .5, 0, .5, .5, .5, .5, 0, .5)) # 35 sites
+    list(c(.5, 0, 0, .5, .5, .5, 0, 0, .5, 1, .5, 0, 0, 0, 0, .5, .5, 1, 1, 1, 
+           .5, 0, 0, 0, .5, .5, 1, 1, 1, 1), # 30 sites
+         c(.5, .5, 1, 1, .5, .5, 1, 1, 1, .5, .5, 0, 0, 0, .5, .5, 1, 1, 1, .5, 
+           .5, 1, 1, 1, .5), # 25 sites
+         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, .5, .5, 0, 0, .5, 1, 
+           1, 1, 1, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, 
+           .5, 1), # 40 sites
+         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, 1, .5, .5, 0, 0, 0, .5, 1, 
+           1, 1, .5, .5, .5, .5, .5, 0, .5, .5, .5, .5, 0, .5)) # 35 sites
   )
-compute_meanCor_i(index_islands, minN_CpG = 10, shore_length = 5, data, sample_n = 2)
-compute_meanCor_ni(index_nonislands, minN_CpG = 10, shore_length = 5, data, sample_n = 2)
+compute_meanCor_i(index_islands, minN_CpG = 10, 
+                  shore_length = 5, data, sample_n = 2)
+compute_meanCor_ni(index_nonislands, minN_CpG = 10, 
+                   shore_length = 5, data, sample_n = 2)
 ```
 
-*Correlation measures how much the state of each site says about the next.
-So, when there is no variation (e.g.
-in middle segment state u of a site is followed by state u of the next), there is no correlation.*
+Note that correlation measures how much the state of each site says about the 
+next. So, when there is no variation (e.g. in middle segment state u of a site 
+is followed by state u of the next), there is no correlation.
 
@@
 
-Let \(S_k\) be the set of sites in the genomic structure indexed by \(k\).
-For each site \(s \in S_k\), let the methylation state at a given tip \(t\) be denoted as \(m_{k,s,t}\), where \(m_{k,s,t} \in \{0, 0.5, 1\}\) represents the unmethylated, partially-methylated, or methylated state, respectively.
+Let $S_k$ be the set of sites in the genomic structure indexed by $k$. 
+For each site $s \in S_k$, let the methylation state at a given tip $t$ be 
+denoted as $m_{k,s,t}$, where $m_{k,s,t} \in \{0, 0.5, 1\}$ represents the 
+unmethylated, partially-methylated, or methylated state, respectively.
 
-For a given cherry \((t_1, t_2)\), we compute the proportion of sites with differing methylation states between the two tips.
+For a given cherry $(t_1, t_2)$, we compute the proportion of sites with 
+differing methylation states between the two tips.
 
@@
 
-Then, the proportion of sites with different methylation states in structure \(k\) for the cherry \((t_1, t_2)\) is given by:
+Then, the proportion of sites with different methylation states in structure $k$ 
+for the cherry $(t_1, t_2)$ is given by:
 
@@
 
-where \(|S_k|\) is the total number of sites in structure \(k\), and the sum counts the number of sites where the methylation state differs between tips \(t_1\) and \(t_2\).
+where $|S_k|$ is the total number of sites in structure \(k\), and the sum 
+counts the number of sites where the methylation state differs between tips 
+$t_1$ and $t_2$.
 
-This statistic quantifies the proportion of differing methylation states for a given genomic structure in a cherry.
+This statistic quantifies the proportion of differing methylation states for a 
+given genomic structure in a cherry.
 
-To compute the weighted mean frequency of methylation changes across all structures in a cherry, we separate island and non-island structures.
+To compute the weighted mean frequency of methylation changes across all 
+structures in a cherry, we separate island and non-island structures.
 
-For a given cherry \((t_1, t_2)\), let \( \mathcal{I} \) be the set of indices for island structures and \( \mathcal{N} \) be the set of indices for non-island structures.
-Each structure \( k \) has a total number of sites \( |S_k| \), which we use as weights.
+For a given cherry $(t_1, t_2)$, let $\mathcal{I}$ be the set of indices 
+for island structures and $\mathcal{N}$ be the set of indices for non-island 
+structures. Each structure $k$ has a total number of sites $|S_k|$,
+which we use as weights.
 
@@
 
-These expressions compute the mean frequency of methylation state differences across all island and non-island structures, weighted by the number of sites in each structure.
+These expressions compute the mean frequency of methylation state differences 
+across all island and non-island structures, weighted by the number of sites in 
+each structure.
 
@@
 It identifies each cherry by the tip names and the tip indices. 
-The tip indices correspond to (a) the index from left to right on the newick string, (b) the order of the tip label in the `phylo_object$tip.label`, and (c) the index in the methylation data list `data[[tip]][[structure]]` as obtained with the function `simulate_evolData()` when the given tree has several tips.
+The tip indices correspond to (a) the index from left to right on the newick 
+string, (b) the order of the tip label in the `phylo_object$tip.label`, and (c) 
+the index in the methylation data list `data[[tip]][[structure]]` as obtained 
+with the function `simulate_evolData()` when the given tree has several tips.
 
-- `countSites_cherryMethDiff(cherryDist, data)` to get for each cherry the counts of sites with half-methylation and full-methylation changes per genomic structure (island or non-island).
+- `countSites_cherryMethDiff(cherryDist, data)` to get for each cherry the 
+counts of sites with half-methylation and full-methylation changes per genomic 
+structure (island or non-island).
 
-- `freqSites_cherryMethDiff(tree,data)`.
-The function first validates the tree structure and extracts pairwise distances between cherry tips with `get_cherryDist`.
-It then validates the input data and counts half and full methylation differences for each cherry at each structure using `countSites_cherryMethDiff`.
-Finally, it normalizes these (half and full) counts by the number of sites per structure to compute frequencies.
-.
+- `freqSites_cherryMethDiff(tree,data)`. The function first validates the tree 
+structure and extracts pairwise distances between cherry tips 
+with `get_cherryDist`. It then validates the input data and counts half and full 
+methylation differences for each cherry at each structure using 
+`countSites_cherryMethDiff`. Finally, it normalizes these (half and full) counts 
+by the number of sites per structure to compute frequencies. .
 
-- `get_siteFChange_cherry(tree,data)`.
-Uses `freqSites_cherryMethDiff(tree,data)` and then computes the frequency of sites with any change of methylation state (full or half) for each cherry at each genomic structure.
+- `get_siteFChange_cherry(tree,data)`. Uses `freqSites_cherryMethDiff(tree,data)` 
+and then computes the frequency of sites with any change of methylation state 
+(full or half) for each cherry at each genomic structure.
 
-- `MeanSiteFChange_cherry(tree, data, index_islands, index_nonislands)`.
-The function computes the per-cherry frequency of sites with different methylation states at each structure (islands and non-islands) using \code{get_siteFChange_cherry}.
-Then it calculates the weighted mean site frequency of methylation changes for each cherry separately for islands and non-islands.
+- `MeanSiteFChange_cherry(tree, data, index_islands, index_nonislands)`. The 
+function computes the per-cherry frequency of sites with different methylation 
+states at each structure (islands and non-islands) using `get_siteFChange_cherry`. 
+Then it calculates the weighted mean site frequency of methylation changes for 
+each cherry separately for islands and non-islands.
 
@@
     list(rep(1,10), rep(0.5,5), rep(0,8)), # tip c
-    list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1), c(0.5, 1, rep(0, 6)))) # tip d 
+    list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1), c(0.5, 1, rep(0, 6)))) # d
+
 # Set the index for islands and non-island structures
@@
 
-# Fitch
+### Fitch Parsimony Estimate for Minimum Global Methylation State Changes at CpG Islands
 
-- `get_meanMeth_islands` computes the mean methylation state of all the islands. **Maybe this should allow excluding island borders**
+Let $I$ be the set of CpG islands in a given genomic region For each island 
+$i \in I$, the mean methylation level across sites is computed with 
+`get_meanMeth_islands(index_islands, data)` and categorized into one of three 
+global methylation states—unmethylated ($u$), partially methylated ($p$), or 
+methylated ($m$)—based on user-defined thresholds $u_{\text{thresh}}$ and 
+$m_{\text{thresh}}$. This categorization is given by the function:
+
+$$
+\operatorname{categorize\_islandGlbSt}(\mu_i, u_{\text{thresh}}, m_{\text{thresh}}) =
+\begin{cases}
+  u, & \text{if } \mu_i \leq u_{\text{thresh}}, \\
+  m, & \text{if } \mu_i \geq m_{\text{thresh}}, \\
+  p, & \text{otherwise}.
+\end{cases}
+$$
+
+Given a phylogenetic tree $T$ with tips corresponding to sampled species or 
+individuals, we aim to estimate the minimum number of state changes required to 
+explain the observed distribution of island methylation states across the tips. 
+This is achieved using Fitch parsimony, a dynamic programming approach 
+implemented in `compute_fitch(meth, tree)` with the following steps:
+
+- Each tip $t$ is assigned the observed methylation state for each island $i$.
+- Internal nodes recursively inherit possible state sets by computing the 
+intersection of the state sets from their child nodes.
+- If the intersection is empty (i.e., no common state between child nodes), the 
+union of both state sets is assigned, and the number of changes is incremented.
+
+The function `computeFitch_islandGlbSt(T, M)`, where $M$ is the matrix of 
+categorized methylation states across tips, returns a vector $C$ where each entry 
+$C_i$ represents the minimum number of changes required for island $i$ under the 
+Fitch parsimony criterion.
+
+```{r}
+# Example with data from a single island structure 
+# and three tips
+
+tree <- "((bla:1,bah:1):2,booh:2);"
+
+data <- list(
+  #Tip 1
+  list(c(rep(1,9), rep(0,1))), # m
+  #Tip 2
+  list(c(rep(0.5,10))), # p
+  #Tip 3
+  list(c(rep(0.5,9), rep(0.5,1)))) # p
+
+index_islands <- c(1)
+
+computeFitch_islandGlbSt(index_islands, data, tree, 
+                         u_threshold = 0.1, m_threshold = 0.9)
+```
+```{r}
+# Example: data from a genomic region consisting on 3 structures with 10 sites each
+# one island, one non-island, one island
+# and a tree with 8 tips
+tree <- "(((a:1,b:1):1,(c:1,d:1):1):1,((e:1,f:1):1,(g:1,h:1):1):1);"
+data <- list(
+  #Tip 1
+  list(c(rep(1,5), rep(0,5)), # p
+       c(rep(0,9), 1), 
+        c(rep(1,8), rep(0.5,2))), # m
+  #Tip 2
+  list(c(rep(0.5,9), rep(0.5,1)), # p
+        c(rep(0.5,9), 1), 
+       c(rep(0,8), rep(0.5,2))), # u
+  #Tip 3
+  list(c(rep(1,9), rep(0.5,1)), # m
+       c(rep(0.5,9), 1), 
+       c(rep(0.5,8), rep(0.5,2))), # p
+  #Tip 4
+  list(c(rep(1,9), rep(0.5,1)), # m
+       c(rep(1,9), 0), 
+       c(rep(0.5,8), rep(0.5,2))), # p
+  #Tip 5
+  list(c(rep(0,5), rep(0,5)), # u
+       c(rep(0,9), 1), 
+       c(rep(0.5,8), rep(0.5,2))), # p
+  #Tip 6
+  list(c(rep(0,9), rep(0.5,1)), # u
+       c(rep(0.5,9), 1), 
+       c(rep(1,8), rep(0.5,2))), # m
+  #Tip 7
+  list(c(rep(0,9), rep(0.5,1)), # u
+       c(rep(0.5,9), 1), 
+       c(rep(0,8), rep(0.5,2))), # u
+  #Tip 8
+  list(c(rep(0,9), rep(0.5,1)), # u
+       c(rep(1,9), 0), 
+       c(rep(0,9), rep(0.5,1)))) # u
+
+index_islands <- c(1,3)
+
+computeFitch_islandGlbSt(index_islands, data, tree, 
+                         u_threshold = 0.1, m_threshold = 0.9)
+```
+
+For genomic regions containing multiple CpG islands, we can summarize the 
+per-island estimates using the mean to obtain an overall measure of the minimum 
+state changes.
+
+```{r}
+mean(computeFitch_islandGlbSt(index_islands, data, tree, 
+                              u_threshold = 0.1, m_threshold = 0.9))
+```

### 2025-03-04 extensive_tests/draft-sumstat-vignette.Rmd
suggest a few changes in vignettes on summary stats

@@
     toc: true
+bibliography: ../vignettes/references.bib
 vignette: >
@@
 
-This document presents summary statistics related to the evolution of methylation 
-states across different genomic structures. The statistics include mean 
+Here we present summary statistics that can be used,
+e.g. in an Approximate Bayesian Computation method [@csillery2012abc],
+to estimate parameters of the methylation dynamics model of MethEvolSIM.
+The statistics include mean 
 frequencies, standard deviations, neighbor correlations of methylation states in 
-island and non-island regions. Additionally, for analysis comparing different 
-samples (comparison of tree tips), the included summary statistics compute
-the mean per-site frequency of methylation changes and the Fitch estimate for
+island and non-island regions.
+Additionally, for analyses comparing different 
+samples (comparison of tree tips), some of the summary statistics are based
+on the mean per-site frequency of methylation changes and
 the minimum number of changes per islands leading to different global methylation
-states.
+states, calculated with the Fitch algorithm [@fitch1971toward].
 
 The provided set of functions allows the computation of summary statistics for 
-methylation data in a genomic region with structural categorization. In this 
-vignette, we will refer to these structures as **islands** and 
+methylation data in a genomic region with structural categorization
+into two types of structures.
+In this  vignette, we will refer to these structures as **islands** and 
 **non-island structures**.
@@
 ```
+Note that in the calculation of the means the relative frequencies $x_{i, t}$, 
+$y_{j, t}$, $w_{i, t}$ and $z_{j, t}$ are **not** weighted by the lengths of 
+island $i$ or non-island $j$.
 
@@
 
-For a given cherry $(t_1, t_2)$, we compute the proportion of sites with 
+For a given cherry $(t_1, t_2)$, that is two tips $t_1$ and $t_2$ that are direct
+offspring of the same internal node, we compute the proportion of sites with 
 differing methylation states between the two tips.
@@
 `countSites_cherryMethDiff`. Finally, it normalizes these (half and full) counts 
-by the number of sites per structure to compute frequencies. .
+by the number of sites per structure to compute frequencies.
 
-- `get_siteFChange_cherry(tree,data)`. Uses `freqSites_cherryMethDiff(tree,data)` 
+- `get_siteFChange_cherry(tree,data)` uses `freqSites_cherryMethDiff(tree,data)` 
 and then computes the frequency of sites with any change of methylation state 
@@
 explain the observed distribution of island methylation states across the tips. 
-This is achieved using Fitch parsimony, a dynamic programming approach 
-implemented in `compute_fitch(meth, tree)` with the following steps:
-
-- Each tip $t$ is assigned the observed methylation state for each island $i$.
-- Internal nodes recursively inherit possible state sets by computing the 
-intersection of the state sets from their child nodes.
-- If the intersection is empty (i.e., no common state between child nodes), the 
-union of both state sets is assigned, and the number of changes is incremented.
+This is achieved using the Fitch algorithm [@fitch1971toward] as
+implemented in `compute_fitch(meth, tree)`.
 
@@
 ```
+
+# References

### 2025-03-05 application_examples/abc_inferences/ABC_cichlids.Rmd
Adding ABC_cichlids.Rmd

---
title: "ABC_cichlids"
author: "Sara"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true  # Add a table of contents
    toc_float: false
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before debug

```{r load params and sumstats}
load("design.RData")
names(sampled_params)[12] <- "TMRCA_smallestCherry"
load("merged_sumStats.RData")
```

```{r set functions to compute expected value and sd of P and M/M+U}
# Set functions to compute expected value and sd of P and M/M+U
    get_expectedValueP <- function(alpha, beta){
      alpha/(alpha+beta)
    }

    get_expectedValueMoverMU <- function(alpha, beta){
      alpha/(alpha+beta)
    }

    get_sdP <- function(alpha, beta){
      var <- (alpha*beta)/((alpha+beta)^2*(alpha+beta+1))
      sqrt(var)
    }

    get_sdMoverMU <- function(alpha, beta){
      var <- (alpha*beta)/((alpha+beta)^2*(alpha+beta+1))
      sqrt(var)
    }
```

```{r filter params corresponding to output files}
# Filter sampled params corresponding to output files
params <- sampled_params[as.integer(rownames(merged_sumStats)),]
print(paste("Number of parameter combinations:", dim(params)[1]))
```

```{r}
# Substitute params alpha and beta for expected value and sd
params_expSD <- params[,c("mu", "iota", "alpha_Ri", "TMRCA_smallestCherry")]
params_expSD$expP_I <- get_expectedValueP(params[,"alpha_pI"], 
                                          params[,"beta_pI"])
params_expSD$expP_NI <- get_expectedValueP(params[,"alpha_pNI"],
                                           params[,"beta_pNI"])   
params_expSD$expMoverMU_I <- get_expectedValueMoverMU(params[,"alpha_mI"],
                                                      params[,"beta_mI"])
params_expSD$expMoverMU_NI <- get_expectedValueMoverMU(params[,"alpha_mNI"],
                                                       params[,"beta_mNI"])
params_expSD$sdP_I <- get_sdP(params[,"alpha_pI"],
                              params[,"beta_pI"])
params_expSD$sdP_NI <- get_sdP(params[,"alpha_pNI"],
                               params[,"beta_pNI"])
params_expSD$sdMoverMU_I <- get_sdMoverMU(params[,"alpha_mI"],
                                          params[,"beta_mI"])
params_expSD$sdMoverMU_NI <- get_sdMoverMU(params[,"alpha_mNI"],
                                           params[,"beta_mNI"])
```

## Params as alpha and beta

```{r, message=FALSE, warning=FALSE}
library(abc)
set.seed(1)
loclinear_cv <- cv4abc(params, merged_sumStats,
                       nval = 100,
                       method = "loclinear",
                       transf = "none",
                       tols = 0.01)
set.seed(1)
ridge_cv <- cv4abc(params, merged_sumStats,
                   nval = 100, 
                   method = "ridge",
                   transf = "none",
                   tols = 0.01)
```

```{r}
loclinear_error <- summary(loclinear_cv)[1,]
ridge_error <- summary(ridge_cv)[1,]
```

```{r}
# Combine into a matrix for grouped bars
error_matrix <- rbind(loclinear_error, ridge_error)

par(mar = c(5, 10, 4, 2))  # Increase left margin for labels
# Create the horizontal bar plot
barplot(error_matrix, 
        horiz = TRUE, 
        beside = TRUE,     # Plot bars side by side
        col = c("red", "steelblue"),  # Red for loclinear, Blue for ridge (cv)
        border = "black",  # Bar border color
        main = "Comparison of Errors",
        xlab = "Error Value",
        names.arg = names(loclinear_error),
        las = 1, 
        cex.names = 0.8,   # Adjust text size
        legend.text = c("Loclinear", "Ridge"),  # Legend
        args.legend = list(x = "topright", bty = "n"))  # Position legend
```

## Params as Expected Value and SD

```{r, message=FALSE, warning=FALSE}
set.seed(1)
loclinear_cv <- cv4abc(params_expSD, merged_sumStats,
                       nval = 100,
                       method = "loclinear",
                       transf = "none",
                       tols = 0.01)
set.seed(1)
ridge_cv <- cv4abc(params_expSD, merged_sumStats,
                   nval = 100, 
                   method = "ridge",
                   transf = "none",
                   tols = 0.01)
```

```{r}
loclinear_error <- summary(loclinear_cv)[1,]
ridge_error <- summary(ridge_cv)[1,]
```

```{r}
# Combine into a matrix for grouped bars
error_matrix <- rbind(loclinear_error, ridge_error)

par(mar = c(5, 10, 4, 2))  # Increase left margin for labels
# Create the horizontal bar plot
barplot(error_matrix, 
        horiz = TRUE, 
        beside = TRUE,     # Plot bars side by side
        col = c("red", "steelblue"),  # Red for loclinear, Blue for ridge (cv)
        border = "black",  # Bar border color
        main = "Comparison of Errors",
        xlab = "Error Value",
        names.arg = names(loclinear_error),
        las = 1, 
        cex.names = 0.8,   # Adjust text size
        legend.text = c("Loclinear", "Ridge"),  # Legend
        args.legend = list(x = "topright", bty = "n"))  # Position legend
```

### Parameter exploration

```{r}
hist(params$mu,
     main = "Frequency of mu in params")
```

```{r}
total_branchLength_unscaled <- 1+1+9+10
total_branchLength_scaled <- params$TMRCA_smallestCherry * total_branchLength_unscaled
expIWE_perIsland <- total_branchLength_scaled * params$mu
hist(expIWE_perIsland,
     main = "Expected Number IWE per Island")
```

```{r}
plot(expIWE_perIsland, merged_sumStats$Fitch_islandGlbSt)
```

### 2025-03-09 application_examples/abc_inferences/draft-abc-inferences-vignette.Rmd
Updating to test new sumstats with ABC

@@
 
-3. Compute the selected summary statistics. TODO: When added sumStats functions to package, stop sourcing script.
+8. Filter the CpG position indices within the genomic region of interest.
 
-4. Assess ABC accuracy using cross-validation.
+9. Compute the selected summary statistics. TODO: When added sumStats functions to package, stop sourcing script.
 
-5. Infer parameters from empirical DNA methylation data.
+10. Assess ABC accuracy using cross-validation.
+
+11. Infer parameters from empirical DNA methylation data.
+
+# Choice of parameters according to region of interest
+
+## Shore length and min number of CpGs
+
+```{r}
+# load the spatial distribution of sites in islands and non-islands for the
+# region of interest
+load("NC_036801.1_29002362_30002362_dpp4.RData")
+```
+
+```{r}
+# Add a column to keep the original structure index after subsetting
+spatial_str$str_index <- 1:dim(spatial_str)[1]
+# Subset islands and non islands
+islands <- spatial_str[spatial_str$globalState == "U",]
+nonislands <- spatial_str[spatial_str$globalState == "M",]
+```
+
+Number of islands and non-islands in the region of interest:
+
+```{r}
+island_n <- dim(islands)[1]
+nonisland_n <- dim(nonislands)[1]
+island_n
+nonisland_n
+```
+
+Summary of the number of sites at each island and non-island structure
+
+```{r}
+summary(islands$n)
+summary(nonislands$n)
+```
+
+```{r}
+# Combine data into a list
+boxplot_data <- list(Islands = islands$n, NonIslands = nonislands$n)
+
+# Create the boxplot
+boxplot(boxplot_data, 
+        main = "Distribution of n in Islands and Non-Islands",
+        ylab = "n",
+        col = c("lightblue", "lightcoral"))
+
+```
+
+```{r}
+load("design.RData")
+```
+
+```{r}
+iota <- sampled_params[1, "iota"]
+iota
+```
+
+```{r}
+load("abc_dataSIM_000001.RData")
+```
+
+```{r}
+index_islands <- which(spatial_str$globalState == "U")
+index_nonislands <- which(spatial_str$globalState == "M")
+shore_length <- 10
+minN_CpG <- 5
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+```
+
+```{r}
+shore_length <- 5
+minN_CpG <- 5
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+```
+```{r}
+shore_length <- 5
+minN_CpG <- 10
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("iota:", iota))
+```
+
+```{r}
+shore_length <- 5
+minN_CpG <- 15
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("iota:", iota))
+```
+
+```{r}
+shore_length <- 10
+minN_CpG <- 10
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("iota:", iota))
+```
+
+```{r}
+shore_length <- 10
+minN_CpG <- 20
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("iota:", iota))
+```
+
+```{r}
+iota <- sampled_params[6, "iota"]
+iota
+```
+
+```{r}
+load("abc_dataSIM_000006.RData")
+```
+
+```{r}
+index_islands <- which(spatial_str$globalState == "U")
+index_nonislands <- which(spatial_str$globalState == "M")
+shore_length <- 10
+minN_CpG <- 5
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+```
+
+```{r}
+shore_length <- 5
+minN_CpG <- 5
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+```
+```{r}
+shore_length <- 5
+minN_CpG <- 10
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("iota:", iota))
+```
+
+```{r}
+shore_length <- 5
+minN_CpG <- 15
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("iota:", iota))
+```
+
+```{r}
+shore_length <- 10
+minN_CpG <- 10
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("iota:", iota))
+```
+
+```{r}
+shore_length <- 10
+minN_CpG <- 15
+print(paste("Number of islands considered out of:", island_n))
+dim(islands[islands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor islands")
+compute_meanCor_i(index_islands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("Number of non-islands considered out of:", nonisland_n))
+dim(nonislands[nonislands$n >= shore_length + minN_CpG,])[1]
+print("MeanCor non-islands")
+compute_meanCor_ni(index_nonislands, minN_CpG, shore_length, data, sample_n = 3)
+print(paste("iota:", iota))
+```
+
+Seems that the combination leading to higher MeanCor values while still including some islands could be shore_length 5, minN_CpG 15 (keeps 9 islands). Lets see the borders in the data
+
+```{r}
+shore_length <- 5
+minN_CpG <- 15
+kept_islands <- islands[which(islands$n >= shore_length + minN_CpG), "str_index"]
+for(tip in 1:3){
+  for(i in 1:length(kept_islands)){
+    print(data[[tip]][[kept_islands[i]]])
+  }
+}
+
+```
+
+## u and m thresholds
+
+```{r}
+load("mean_Meth_islands.RData")
+```
+
+```{r}
+# Extract all methylation values into a single vector
+all_values <- unlist(mean_Meth_islands)
+
+# Plot density
+plot(density(all_values, na.rm = TRUE), 
+     main = "Density of Mean Methylation Values at Islands", 
+     xlab = "Mean Methylation", 
+     col = "blue", lwd = 2)
+
+```
+
+```{r}
+summary(all_values)
+```
+
+```{r}
+# Initialize an empty list to store results
+changes_list <- list()
+
+# Loop through each simulation
+for (i in seq_along(mean_Meth_islands)) {
+
+  # Extract values for the three tips
+  tip1 <- mean_Meth_islands[[i]][[1]]
+  tip2 <- mean_Meth_islands[[i]][[2]]
+  tip3 <- mean_Meth_islands[[i]][[3]]
+
+  # Ensure all vectors are the same length
+  n <- length(tip1)  # Assuming tip1, tip2, tip3 are of equal length
+
+  # Create a data frame for changes between tips
+  changes_list[[i]] <- data.frame(
+    index = rep(index[i], 3 * n), # Repeat index for all comparisons
+    tip_from = rep(c(1, 1, 2), each = n), # Tip of origin
+    tip_to = rep(c(2, 3, 3), each = n), # Tip of destination
+    meanMeth_from = c(tip1, tip1, tip2),
+    meanMeth_to = c(tip2, tip3, tip3),
+    meanMeth_diff = c(tip2 - tip1, tip3 - tip1, tip3 - tip2)
+  )
+}
+
+# Combine all changes into a single data frame
+changes_df <- do.call(rbind, changes_list)
+
+# Print the first few rows
+head(changes_df[changes_df$meanMeth_diff != 0,])
+```
+
+```{r}
+# Filter cases where meanMeth_diff is not zero
+changes_nonzero <- changes_df[changes_df$meanMeth_diff != 0, ]
+
+# Plot histogram
+plot(density(abs(changes_nonzero$meanMeth_diff)), 
+     breaks = 30, 
+     col = "lightblue", 
+     main = "Distribution of meanMeth_diff (Non-Zero Changes)", 
+     xlab = "Methylation Difference",
+     border = "black")
+
+```
+
+```{r}
+changes_nonzero$meanMeth_diff <- abs(changes_nonzero$meanMeth_diff)
+summary(changes_nonzero$meanMeth_diff)
+```
+
+```{r}
+head(changes_nonzero[order(changes_nonzero$meanMeth_diff), "meanMeth_diff"])
+```
+
+```{r}
+changes_overFirstQuantile <- changes_nonzero[changes_nonzero$meanMeth_diff > 0.08333,]
+head(changes_overFirstQuantile[order(changes_overFirstQuantile$meanMeth_diff),])
+```
+
+Lets see how the current summary statistics behave, taking into account that it will not capture most of the global methylation changes as they the state at the different tips will fall under the same category.
+E.g.
+for index 12 tip 1 shows 0.58 and tip 2 and 3 have the same value 0.66.
+Both tips will be categorized as partially methylated but this looks like an IWE.
+
+# Sumstats before debug exploration
+
+```{r}
+load("merged_sumStats.RData")
+load("design.RData")
+```
+
+```{r}
+# Subset parameter combinations in output files
+subset_params <- sampled_params[as.integer(rownames(merged_sumStats)),]
+```
+
+```{r}
+par(mfrow = c(1,2))
+plot(subset_params$iota, merged_sumStats$meanCor_i,
+     xlab = "iota", ylab = "meanCor",
+     main = "Island")
+plot(subset_params$iota, merged_sumStats$meanCor_ni,
+     xlab = "iota", ylab = "meanCor",
+     main = "Non-island")
+```
+
+For this I used as thresholds a minimum number of CpGs of 15 and shore length 5.
+It was then only considering 9 islands.
+Still in non islands it shows a decreasing trend.
+However in the test for CFTP convergence CFTP with higher iota was leading to larger mean correlation values (there all islands and non-islands were considered, as the number of sites per structure was fixed to 100 (and 20 islands and 20 non-islands) and 10 sites was the shore lengthä)
+
+```{r}
+merged_sumStats$meanCor_i[which(subset_params$iota == min(subset_params$iota))]
+```
+
+```{r}
+plot(subset_params$mu, merged_sumStats$Fitch_islandGlbSt,
+     xlab = "mu", ylab = "Fitch",
+     main = "Island")
+```
+
+```{r}
+names(subset_params)[12] <- "TMRCA_smallestCherry"
+par(mfrow = c(1,2))
+plot(subset_params$TMRCA_smallestCherry, merged_sumStats$MeanSiteFChange_i,
+     xlab = "TMRCA_smallestCherry", ylab = "MeanSiteFChange",
+     main = "Island")
+plot(subset_params$TMRCA_smallestCherry, merged_sumStats$MeanSiteFChange_ni,
+     xlab = "TMRCA_smallestCherry", ylab = "MeanSiteFChange",
+     main = "Non-island")
+```
 

### 2025-03-09 R/summaryStatistics.R

@@ -60,8 +60,19 @@ validate_structureIndices <- function(data, index_islands, index_nonislands) {
 #'
 #' @param index_islands A vector containing the structural indices for islands.
 #' @param data A list containing methylation states at tree tips for each genomic structure (island / non-island) 
-#' For a single tip: \code{data[[structure]]}. For multiple tips: \code{data[[tip]][[structure]]}.
+#'   For a single tip: \code{data[[structure]]}. 
+#'   For multiple tips: \code{data[[tip]][[structure]]}.
+#'   Each element contains the methylation states at the sites in a given tip and structure
+#'   represented as 0, 0.5 or 1 (for unmethylated, partially-methylated and methylated). 
+#'   If methylation states are not represented as 0, 0.5, 1 they are categorized
+#'   as 0 when value equal or under 0.2
+#'   0.5 when value between 0.2 and 0.8
+#'   and 1 when value over 0.8.
+#'   For customized categorization thresholds use \code{categorize_siteMethSt}
+#' 
 #' @param sample_n The number of samples (tips) to process.
+#' @param categorized_data Logical defaulted to FALSE. 
+#'   TRUE to skip redundant categorization when methylation states are represented as 0, 0.5, and 1.
 #'
 #' @return A numeric value representing the mean frequency of partially methylated sites in the islands.
 #' @examples
@@ -75,7 +86,10 @@ validate_structureIndices <- function(data, index_islands, index_nonislands) {
 #' get_islandMeanFreqP(index_islands, data, sample_n)
 #' 
 #' @export
-get_islandMeanFreqP <- function(index_islands, data, sample_n){
+get_islandMeanFreqP <- function(index_islands, data, sample_n, categorized_data = FALSE){
+  
+  if(!is.list(data)) stop("Input 'data' is not list.")
+  
   # Restructure data as nested list
   if (sample_n == 1){
     data_list <- list()
@@ -83,10 +97,16 @@ get_islandMeanFreqP <- function(index_islands, data, sample_n){
     data <- data_list
   }
   
-  # Validate index islands
+  
   tryCatch({
+    
+    # Validate index islands
     if(length(index_islands) == 0) stop("'index_islands' has no indices")
     validate_structureIndices(data, index_islands, index_nonislands = c())
+    
+    # Categorize methylation states
+    if(!categorized_data) data <- categorize_siteMethSt(data)
+    
   }, warning = function(w) {
     stop(conditionMessage(w))
   }, error = function(e) {
[…]

### 2025-03-09 extensive_tests/functions_summaryStats.R

@@ -682,33 +682,43 @@ library(ape)
 
 ##### Get a dataframe with distances between each combination of pairs of tips
 ## tree: phylogenetic tree object (as generated with package scrm)
-get_pairwiseDistance <- function(tree){
+get_tipPairwiseDist <- function(tree, input_control = TRUE){
+  
+  # Check input tree format and minium two tips, get tree in phylo format (ape package)
+  if (input_control) tree <- validate_tree(tree)
   
   # Get the number of tree tips
   n_tips <- length(tree$tip.label)
   
   # Generate the vectors for the tip pairwise comparisons
-  tipA <- numeric()
-  tipB <- numeric()
+  first_tip_index <- numeric()
+  second_tip_index <- numeric()
+  first_tip_name <- character()
+  second_tip_name <- character()
   
   for (tA in 1:(n_tips - 1)){
     for (tB in (tA+1):n_tips){
-      tipA <- c(tipA, tA)
-      tipB <- c(tipB, tB)
+      first_tip_index <- c(first_tip_index, tA)
+      second_tip_index <- c(second_tip_index, tB)
+      first_tip_name <- c(first_tip_name, tree$tip.label[tA])
+      second_tip_name <- c(second_tip_name, tree$tip.label[tB])
     }
   }
   
   # Get the matrix with the distances between the tree tips
-  distance <- signif(ape::cophenetic.phylo(tree), 4)
+  #distance <- signif(ape::cophenetic.phylo(tree), 4)
+  dist <- ape::cophenetic.phylo(tree)
   
   # Set a dataframe for the pairwise distances
-  pairwiseD <- data.frame(tipA = tipA,
-                          tipB = tipB,
-                          distance = rep(NA, length(tipA)))
+  pairwiseD <- data.frame(first_tip_name = first_tip_name,
+                          second_tip_name = second_tip_name,
+                          first_tip_index = first_tip_index,
+                          second_tip_index = second_tip_index,
+                          dist = rep(NA, length(tipA)))
   
   # Save the distances
-  for (tip in 1:length(tipA)){
-    pairwiseD$distance[tip] <- distance[tipA[tip],tipB[tip]]
+  for (tip in 1:length(first_tip_index)){
+    pairwiseD$dist[tip] <- dist[first_tip_index[tip],second_tip_index[tip]]
   }
   pairwiseD
 }

### 2025-03-09 tests/testthat/test-summaryStatistics.R

@@ -25,6 +25,27 @@ test_that("get_MeanFreqP non-valid structureIndices", {
   
 })
 
+test_that("get_MeanFreqP non-categorized input of methylation states", {
+  
+  data <- list(
+    list(c(0.1, 0.2, 0.02), c(0.05, 0.25, 0.15)), # tip 1
+    list(c(0.01, 0.7, 0.85), c(0.3, 0.1, 0.98)) # tip 2
+  )
+  
+  index_islands <- c(1,2)
+  index_nonislands <- c(1,2)
+  
+  mean_freqP <- mean(c(0, 0, 0, 0, 0.5, 0, 0, 0.5, 1, 0.5, 0, 1)==0.5)
+  
+  expect_equal(get_islandMeanFreqP(data = data, index_islands = index_islands, sample_n =2),
+               mean_freqP,
+               info = "incorrect value with uncategorized input data")
+  expect_equal(get_nonislandMeanFreqP(data = data, index_nonislands = index_nonislands, sample_n =2),
+               mean_freqP,
+               info = "incorrect value with uncategorized input data")
+  
+})
+
 test_that("get_MeanFreqP structures with equal length islands and non-islands", {
   # Two tips and two islands
   index_islands <- c(1, 2)
@@ -157,6 +178,27 @@ test_that("get_MeanFreqM non-valid structureIndices", {
                "'index_nonislands' has no indices")
 })
 
+test_that("get_MeanFreqM non-categorized input of methylation states", {
+  
+  data <- list(
+    list(c(0.1, 0.2, 0.02), c(0.05, 0.25, 0.15)), # tip 1
+    list(c(0.01, 0.7, 0.85), c(0.3, 0.1, 0.98)) # tip 2
+  )
+  
+  index_islands <- c(1,2)
+  index_nonislands <- c(1,2)
+  
+  mean_freqM <- mean(c(0, 0, 0, 0, 0.5, 0, 0, 0.5, 1, 0.5, 0, 1)==1)
+  
+  expect_equal(get_islandMeanFreqM(data = data, index_islands = index_islands, sample_n =2),
+               mean_freqM,
+               info = "incorrect value with uncategorized input data")
+  expect_equal(get_nonislandMeanFreqM(data = data, index_nonislands = index_nonislands, sample_n =2),
+               mean_freqM,
+               info = "incorrect value with uncategorized input data")
+  
+})
+
 
[…]

### 2025-03-10 application_examples/abc_inferences/ABC_cichlids.Rmd
Updating to test new sumstats with ABC
Implemented modification to approximate cftp after a given step limit, pending testing

@@
 names(sampled_params)[12] <- "TMRCA_smallestCherry"
-load("merged_sumStats.RData")
+load("merged_sumStats_old.RData")
 ```
@@
 
+#### mu
+
 ```{r}
@@
 ```
+
+#### TMRCA
+
+```{r}
+hist(params$TMRCA_smallestCherry)
+```
+
+```{r}
+summary(params$TMRCA_smallestCherry)
+```
+
+```{r}
+plot(params$TMRCA_smallestCherry,
+     merged_sumStats$MeanSiteFChange_i)
+```
+
+```{r}
+plot(params$TMRCA_smallestCherry,
+     merged_sumStats$MeanSiteFChange_ni)
+```
+
+```{r}
+estim <- loclinear_cv$estim$tol0.01[,"TMRCA_smallestCherry"]
+param <- loclinear_cv$true[,"TMRCA_smallestCherry"]
+(1/100)*sum(abs(estim-param)/param)
+```
+```{r}
+estim <- loclinear_cv$estim$tol0.01[,"mu"]
+param <- loclinear_cv$true[,"mu"]
+(1/100)*sum(abs(estim-param)/param)
+```

### 2025-03-10 application_examples/abc_inferences/ABC_cichlids_new.Rmd
Implemented modification to approximate cftp after a given step limit, pending testing

--- 2025-03-10 application_examples/abc_inferences/ABC_cichlids.Rmd
@@
 names(sampled_params)[12] <- "TMRCA_smallestCherry"
-load("merged_sumStats_old.RData")
+load("merged_sumStats.RData")
 ```
@@
 
+## Params as Exp Value and SD. Log scale
+
+```{r}
+transf <- c("log", "none", "log", rep("none", 9))
+names(transf) <- names(params_expSD)
+transf
+```
+
+```{r}
+set.seed(1)
+loclinear_cv <- cv4abc(params_expSD, merged_sumStats,
+                       nval = 100,
+                       method = "loclinear",
+                       transf = transf,
+                       tols = 0.01)
+set.seed(1)
+ridge_cv <- cv4abc(params_expSD, merged_sumStats,
+                   nval = 100, 
+                   method = "ridge",
+                   transf = transf,
+                   tols = 0.01)
+```
+
+```{r}
+# Combine into a matrix for grouped bars
+error_matrix <- rbind(loclinear_error, ridge_error)
+
+par(mar = c(5, 10, 4, 2))  # Increase left margin for labels
+# Create the horizontal bar plot
+barplot(error_matrix, 
+        horiz = TRUE, 
+        beside = TRUE,     # Plot bars side by side
+        col = c("red", "steelblue"),  # Red for loclinear, Blue for ridge (cv)
+        border = "black",  # Bar border color
+        main = "Comparison of Errors",
+        xlab = "Error Value",
+        names.arg = names(loclinear_error),
+        las = 1, 
+        cex.names = 0.8,   # Adjust text size
+        legend.text = c("Loclinear", "Ridge"),  # Legend
+        args.legend = list(x = "topright", bty = "n"))  # Position legend
+```
+
 ### Parameter exploration
@@
 ```{r}
-plot(expIWE_perIsland, merged_sumStats$Fitch_islandGlbSt)
+plot(expIWE_perIsland, merged_sumStats$cherryFreqsChange_i,
+     main = "cherryFreqsChange")
+```
+
+```{r}
+plot(expIWE_perIsland, merged_sumStats$treeFreqsChange_i,
+     main = "cherryFreqsChange")
 ```
@@
 ```
+
+## New prior to try for alpha_Ri
+```{r}
+load("design.RData")
+hist(sampled_params$alpha_Ri) # old
+
+```
+```{r}
+hist(rexp(10000, 0.1), 50) #new
+```
+
+```{r}
+
+```

### 2025-03-10 R/multiRegion_SIM.R

@@ -1328,6 +1328,16 @@ singleStructureGenerator <-
                     list(seq = private$seq,
                          neighbSt = private$neighbSt)
                   }
+                },
+                
+                #' @description
+                #' Public Method. Resets the sequence states by resampling according to the instance's equilibrium frequencies.
+                #' 
+                #' @return NULL. The sequence is updated in place.
+                reset_seq = function(){
+                  n_sites <- length(private$seq)
+                  # Resample seq states according to the instance's equilibrium frequencies
+                  private$seq <- sample(1L:3L, size = n_sites, prob = private$eqFreqs, replace = TRUE)
                 }
               )
               )
@@ -2046,6 +2056,19 @@ combiStructureGenerator <-
                     
                     # Double the number of steps
                     steps <- 2*steps
+                    
+                    ####TODO: draft
+                    if (steps > limit){
+                      # initialize each singleStr with a method for that
+                      # AFTER initializing ALL
+                      ## Initialice the neighbSt encoding for each singleStr
+                      for (str in 1:length(private$singleStr)){
+                        private$singleStr[[str]]$init_neighbSt()
+                      }
+                      # self$cftp_apply_events() 
+                      equal <- TRUE
+                      message("Steps limit reached. CFTP applying approximation.")
+                    }
                     
                     # Increase counter value
                     counter <- counter + 1

### 2025-03-10 tests/testthat/test-multiRegion_SIM.R

@@ -4938,6 +4938,12 @@ test_that("treeMultiRegionSimulator errors in input", {
 
 })
 
+test_that("singleStructureGenerator reset_seq",{
+  # Test my_combiStructure initialization
+  single_obj <- singleStructureGenerator$new("U",100)
+  
+})
+
 
 
 

### 2025-03-13 extensive_tests/draft-sumstat-vignette.Rmd
Updating vignette with sumstat for mean number of significat freq changes per island and cherry

@@
 
+### Mean Number of Significant Methylation Frequency Changes per Island and Cherry
+
+A cherry $c$ is defined as a pair of direct offspring (tips \( t_1 \) and \( t_2 \)) 
+of the same internal node in the phylogenetic tree. 
+For a given cherry $c$, let $\mathcal{I}$ be the set of indices for island structures. 
+For each island structure $i \in \mathcal{I}$ at each cherry tip $t_c \in {t_{c,1}, t_{c,2}$ in each phylogenetic tree cherries, 
+the number of sites in each 
+methylation state (unmethylated \( 0 \), partially-methylated \( 0.5 \), 
+methylated \( 1 \)) is counted. This count is represented as:
+
+$$
+\text{count}_{\text{UPM}}(c,t_c, i) = \left( n_0, n_{0.5}, n_1 \right)
+$$
+
+where \( n_0\), \( n_{0.5}\), and \( n_1\) are the counts of unmethylated, 
+partially-methylated, and methylated sites at tip \( t_c\), respectively.
+
+The distribution of methylation states at each island between the two tips \( t_1 \) 
+and \( t_2 \) of a given cherry is compared using a chi-squared test. 
+The null hypothesis \( H_0 \) is that the distributions of methylation states at 
+the two tips are the same. The test statistic is calculated using the contingency 
+table \( T \) of the methylation state counts for each tip:
+
+$$
+H_0: \text{distribution}(t_{c,1}) = \text{distribution}(t_{c,2})
+$$
+
+The contingency table \( T_{c,i} \) for tips \( t_{c,1} \) and \( t_{c,2} \) is:
+
+$$
+T_{c,i} = \begin{pmatrix}
+n_0^{t_{c,1}} & n_{0.5}^{t_{c,1}} & n_1^{t_{c,1}} \\
+n_0^{t_{c,2}} & n_{0.5}^{t_{c,2}} & n_1^{t_{c,2}}
+\end{pmatrix}
+$$
+
+The p-value for the chi-squared test is calculated via Monte Carlo simulation to 
+improve reliability, even when the expected frequencies do not meet the assumptions 
+of the chi-squared test (i.e., expected counts of at least 5 in each category).
+
+For each cherry $c$ and each island $i$, the methylation data at each cherry tip 
+for the specified genomic structures (islands) are compared using the chi-squared test.
+
+$$
+p_{\text{value}}(c,i) = \chi^2\text{-test}(T_{c,i})
+$$
+
+These p-values are stored in a dataframe for each cherry and for each island.
+
+The significance of the methylation frequency changes is determined based on a 
+user-defined threshold \( p_{\text{threshold}} \). If the p-value is smaller than 
+the threshold, the change is considered significant. This is done for each cherry 
+and each island \( i \in \mathcal{I} \) by comparing each p-value to the threshold:
+
+$$
+\text{Significant Change}_{c,i} = 
+\begin{cases} 
+1, & \text{if } p_{\text{value}}(c,i) < p_{\text{threshold}} \\
+0, & \text{if } p_{\text{value}}(c,i) \ge p_{\text{threshold}}
+\end{cases}
+$$
+
+For each cherry , the mean number of significant changes across all islands 
+\( i \in \mathcal{I} \) is computed as the average of the significant changes for each island:
+
+$$
+\text{Mean Number of Significant Changes per Island}_{c} = \frac{1}{|\mathcal{I}|} \sum_{i \in \mathcal{I}} \text{Significant Change}_{c,i}
+$$
+
+where \( |\mathcal{I}| \) is the total number of islands.
+
 # References

### 2025-03-13 R/summaryStatistics.R

@@ -1875,7 +1875,7 @@ pValue_CherryFreqsChange_i <- function(data, categorized_data = FALSE, index_isl
   cherryDist
 }
 
-#' Count Significant Methylation Frequency Changes in Cherries
+#' Mean Number of Significant Methylation Frequency Changes per Island in Cherries
 #'
 #' Computes the mean number of significant changes per island in phylogenetic tree cherries,
 #' based on a specified p-value threshold.
@@ -1923,11 +1923,11 @@ pValue_CherryFreqsChange_i <- function(data, categorized_data = FALSE, index_isl
 #'        c(rep(0,9), rep(0.5,1)))) 
 #' 
 #' index_islands <- c(1,3)
-#' count_CherryFreqsChange_i(data, categorized_data = TRUE,
+#' mean_CherryFreqsChange_i(data, categorized_data = TRUE,
 #'                           index_islands, tree, pValue_threshold = 0.05)
 #'
 #' @export
-count_CherryFreqsChange_i <- function(data, categorized_data = FALSE, index_islands, tree, pValue_threshold){
+mean_CherryFreqsChange_i <- function(data, categorized_data = FALSE, index_islands, tree, pValue_threshold){
   
   # Validate given threshold for pValue
   if (!(pValue_threshold > 0 & pValue_threshold < 1)) stop("pValue_threshold needs to be between 0 and 1.")
@@ -2097,7 +2097,7 @@ categorize_siteMethSt <- function(data, u_threshold = 0.2, m_threshold = 0.8) {
 
 
 
-#' Count Significant Frequency Changes Across Islands in Tree Tips
+#' Mean Number of Significant Frequency Changes per Island Across all Tree Tips
 #'
 #' This function analyzes the frequency changes of methylation states (unmethylated, partially methylated, methylated)
 #' across tree tips for a given set of islands. It performs a chi-squared test for each island to check for significant 
@@ -2124,6 +2124,7 @@ categorize_siteMethSt <- function(data, u_threshold = 0.2, m_threshold = 0.8) {
 #' @param index_islands A vector of indices of genomic structures corresponding to islands in data.
 #' @param pValue_threshold A numeric value between 0 and 1 that serves as the threshold for statistical significance in 
 #'   the chi-squared test.
[…]

### 2025-03-13 tests/testthat/test-summaryStatistics.R

@@ -2840,13 +2840,13 @@ test_that("pValue_CherryFreqsChange_i", {
                info = "fails to return a dataframe with 2 rows when tree has 2 cherries")
 })
 
-test_that("count_CherryFreqsChange_i input control",{
+test_that("mean_CherryFreqsChange_i input control",{
   wrong_tree <- "(a:1);"
   data <- list(c(rep(0,5), rep(0.5,5)),
                rep(1,10),
                c(rep(0,10)))
   index_islands <- c(1,3)
-  expect_error(count_CherryFreqsChange_i(data, categorized_data = T, index_islands, tree = wrong_tree, pValue_threshold = 0.05),
+  expect_error(mean_CherryFreqsChange_i(data, categorized_data = T, index_islands, tree = wrong_tree, pValue_threshold = 0.05),
                "The input 'tree' must have a minimum of 2 tips.")
   
   tree <- "(a:1,b:1);"
@@ -2859,11 +2859,11 @@ test_that("count_CherryFreqsChange_i input control",{
          c(rep(0,5), rep(0.5,5))))
   
   wrong_index_islands <- c(1,4)
-  expect_error(count_CherryFreqsChange_i(data, categorized_data = T, wrong_index_islands, tree, pValue_threshold = 0.05),
+  expect_error(mean_CherryFreqsChange_i(data, categorized_data = T, wrong_index_islands, tree, pValue_threshold = 0.05),
                "Invalid island indices detected: 4. Number of structures in given data: 3")
 })
 
-test_that("count_CherryFreqsChange_i non-categorized input of methylation states", {
+test_that("mean_CherryFreqsChange_i non-categorized input of methylation states", {
   
   tree <- "(a:1,b:1);"
   data <- list(
@@ -2872,13 +2872,13 @@ test_that("count_CherryFreqsChange_i non-categorized input of methylation states
   )
   index_islands <- c(1,2)
   
-  o <- count_CherryFreqsChange_i(data, categorized_data = F, index_islands, tree, pValue_threshold = 0.05)
+  o <- mean_CherryFreqsChange_i(data, categorized_data = F, index_islands, tree, pValue_threshold = 0.05)
   
   expect_equal(o$FreqsChange, 0.5,
                info = "not correct mean number of changes per island with non-categorized input data")
 })
 
-test_that("count_CherryFreqsChange_i correct output", {
[…]

### 2025-03-13 extensive_tests/draft-sumstat-vignette.Rmd
added line breaks to make comparisons easier
Updating vignette with sumstat for mean number of significat freq changes per island and cherry

@@
 Intermediate values (e.g., obtained from pooled empirical data) should be 
-categorized before analysis.
+categorized before analysis with `categorize_siteMethSt`.
+
+```{r}
+non_categorized_data <- list(
+  # Tip 1
+    list(c(0.1, 0.7, 0.9), # First structure
+         c(0.3, 0.5, 0.9)), # Second structure
+  # Tip 2
+    list(c(0.2, 0.8, 0.6), # First structure
+         c(0.9, 0.4, 0.7)) # Second structure
+  )
+
+# Transform the data with custom thresholds
+categorized_data <- categorize_siteMethSt(data, u_threshold = 0.15, m_threshold = 0.85)
+
+categorized_data
+```
 
@@
 index_nonislands <- c(2, 4)
-data <- list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 0, .5)) # tip 1
-get_islandMeanFreqP(index_islands, data, sample_n)
-get_nonislandMeanFreqP(index_nonislands, data, sample_n)
-get_islandMeanFreqM(index_islands, data, sample_n)
-get_nonislandMeanFreqM(index_nonislands, data, sample_n)
+data <- list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 1, .5)) # tip 1
+get_islandMeanFreqP(index_islands, data, categorized_data = T, sample_n)
+get_nonislandMeanFreqP(index_nonislands, data, categorized_data = T, sample_n)
+get_islandMeanFreqM(index_islands, data, categorized_data = T, sample_n)
+get_nonislandMeanFreqM(index_nonislands, data, categorized_data = T, sample_n)
 ```
@@
   )
-get_islandMeanFreqP(index_islands, data, sample_n)
-get_nonislandMeanFreqP(index_nonislands, data, sample_n)
-get_islandMeanFreqM(index_islands, data, sample_n)
-get_nonislandMeanFreqM(index_nonislands, data, sample_n)
+get_islandMeanFreqP(index_islands, data, categorized_data = T, sample_n)
+get_nonislandMeanFreqP(index_nonislands, data, categorized_data = T, sample_n)
+get_islandMeanFreqM(index_islands, data, categorized_data = T, sample_n)
+get_nonislandMeanFreqM(index_nonislands, data, categorized_data = T, sample_n)
 ```
@@
 data <- list(c(.5, .5, 0, 1, 1, .5), c(.5, 0, 1, .5), c(.5, .5, 0), c(0, 0, .5))
-get_islandSDFreqP(index_islands, data, sample_n)
-get_nonislandSDFreqP(index_nonislands, data, sample_n)
-get_islandSDFreqM(index_islands, data, sample_n)
-get_nonislandSDFreqM(index_nonislands, data, sample_n)
+get_islandSDFreqP(index_islands, data, categorized_data = T, sample_n)
+get_nonislandSDFreqP(index_nonislands, data, categorized_data = T, sample_n)
+get_islandSDFreqM(index_islands, data, categorized_data = T, sample_n)
+get_nonislandSDFreqM(index_nonislands, data, categorized_data = T, sample_n)
 ```
@@
   )
-get_islandSDFreqP(index_islands, data, sample_n)
-get_nonislandSDFreqP(index_nonislands, data, sample_n)
-get_islandSDFreqM(index_islands, data, sample_n)
-get_nonislandSDFreqM(index_nonislands, data, sample_n)
+get_islandSDFreqP(index_islands, data, categorized_data = T, sample_n)
+get_nonislandSDFreqP(index_nonislands, data, categorized_data = T, sample_n)
+get_islandSDFreqM(index_islands, data, categorized_data = T, sample_n)
+get_nonislandSDFreqM(index_nonislands, data, categorized_data = T, sample_n)
 ```
@@
 compute_meanCor_i(index_islands, minN_CpG = 10, 
-                  shore_length = 5, data, sample_n = 1)
+                  shore_length = 5, data, sample_n = 1, categorized_data = T)
 compute_meanCor_ni(index_nonislands, minN_CpG = 10, 
-                   shore_length = 5, data, sample_n = 1)
+                   shore_length = 5, data, sample_n = 1, categorized_data = T)
 ```
@@
 compute_meanCor_i(index_islands, minN_CpG = 10, 
-                  shore_length = 5, data, sample_n = 2)
+                  shore_length = 5, data, sample_n = 2, categorized_data = T)
 compute_meanCor_ni(index_nonislands, minN_CpG = 10, 
-                   shore_length = 5, data, sample_n = 2)
+                   shore_length = 5, data, sample_n = 2, categorized_data = T)
 ```
@@
 MeanSiteFChange_cherry(data = data, 
+                       categorized_data = T, 
                        tree = tree, 
@@
 
+Functions:
+
+- `compare_CherryFreqs(tip1, tip2)` to perform a chi-squared test to compare the
+distribution of methylation states between two cherry tips.
+
+- `pValue_CherryFreqsChange_i(data, index_islands, tree)` uses `compare_CherryFreqs(tip1, tip2)` to get for each cherry 
+and each island the chi-square p-value of the comparison of the distribution of
+methylation states between two cherry tips.
+
+- `mean_CherryFreqsChange_i(data, categorized_data = FALSE, index_islands, tree, pValue_threshold)`. The function
+uses `pValue_CherryFreqsChange_i(data, index_islands, tree)` and counts the mean
+number of significant changes per island at each cherry.
+
+```{r}
+# Set example tree and methylation data
+  tree <- "((a:1,b:1):2,c:2);"
+  data <- list(
+    #Tip a
+    list(c(rep(1,9), rep(0,1)), # Structure 1: island
+         c(rep(0,9), 1), # Structure 2: non-island
+         c(rep(0,9), rep(0.5,1))),  # Structure 3: island
+    #Tip b
+    list(c(rep(0,9), rep(0.5,1)),  # Structure 1: island
+         c(rep(0.5,9), 1), # Structure 2: non-island
+         c(rep(0,9), rep(0,1))), # Structure 3: island
+    #Tip c
+    list(c(rep(1,9), rep(0.5,1)),  # Structure 1: island
+         c(rep(0.5,9), 1), # Structure 2: non-island
+         c(rep(0,9), rep(0.5,1)))) # Structure 3: island
+
+  index_islands <- c(1,3)
+
+  mean_CherryFreqsChange_i(data, categorized_data = T,
+                           index_islands, tree,
+                           pValue_threshold = 0.05)
+```
+
 # References

### 2025-03-13 R/summaryStatistics.R

@@ -2155,7 +2155,6 @@ categorize_siteMethSt <- function(data, u_threshold = 0.2, m_threshold = 0.8) {
 #'        
 #' index_islands <- c(1,3)
 #' 
-#' pValue_threshold <- 0.05
 #' 
 #' mean_TreeFreqsChange_i(tree, data, categorized_data = TRUE,
 #'                             index_islands, 

### 2025-03-13 extensive_tests/draft-sumstat-vignette.Rmd
a few little changes regarding chi-square test null hypothesis

--- 2025-03-13 extensive_tests/draft-sumstat-vignette.Rmd
@@
 For a given cherry $c$, let $\mathcal{I}$ be the set of indices for island structures. 
-For each island structure $i \in \mathcal{I}$ at each cherry tip $t_c \in {t_{c,1}, t_{c,2}$ in each phylogenetic tree cherries, 
+For each island structure $i \in \mathcal{I}$ at each cherry tip
+$t_c \in \{t_{c,1}, t_{c,2}\}$ in each phylogenetic tree, 
 the number of sites in each 
@@
 and \( t_2 \) of a given cherry is compared using a chi-squared test. 
-The null hypothesis \( H_0 \) is that the distributions of methylation states at 
-the two tips are the same. The test statistic is calculated using the contingency 
-table \( T \) of the methylation state counts for each tip:
-
-$$
-H_0: \text{distribution}(t_{c,1}) = \text{distribution}(t_{c,2})
-$$
-
-The contingency table \( T_{c,i} \) for tips \( t_{c,1} \) and \( t_{c,2} \) is:
+The null hypothesis of the test is that the frequency distributions of states
+in the two islands follow the same multinomial distribution.
+Note that not only IWEs can lead to deviations from this null hypothesis but
+also neighbor-dependent SSEs can lead to slight deviations from the assumption
+of multinomial distributions.
+The test statistic is calculated using the contingency 
+table \( T_{c,i} \) for tips \( t_{c,1} \) and \( t_{c,2} \):
 
@@
 improve reliability, even when the expected frequencies do not meet the assumptions 
-of the chi-squared test (i.e., expected counts of at least 5 in each category).
+of the chi-squared approximation (i.e., expected counts of at least 5 in each category).
 
@@
 for the specified genomic structures (islands) are compared using the chi-squared test.
-
-$$
-p_{\text{value}}(c,i) = \chi^2\text{-test}(T_{c,i})
-$$
-
-These p-values are stored in a dataframe for each cherry and for each island.
+The obtained p-values are stored in a dataframe for each cherry and for each island.
 
 The significance of the methylation frequency changes is determined based on a 
-user-defined threshold \( p_{\text{threshold}} \). If the p-value is smaller than 
-the threshold, the change is considered significant. This is done for each cherry 
-and each island \( i \in \mathcal{I} \) by comparing each p-value to the threshold:
+user-defined threshold \( p_{\text{threshold}} \).
+If the p-value is smaller than the threshold, the change is considered significant.
+This is done for each cherry and each island \( i \in \mathcal{I} \) by
+comparing each p-value to the threshold:
 
@@
 For each cherry , the mean number of significant changes across all islands 
-\( i \in \mathcal{I} \) is computed as the average of the significant changes for each island:
+\( i \in \mathcal{I} \) is computed as the average of the significant changes
+for each island:
 
 $$
-\text{Mean Number of Significant Changes per Island}_{c} = \frac{1}{|\mathcal{I}|} \sum_{i \in \mathcal{I}} \text{Significant Change}_{c,i}
+\text{Mean Number of Significant Changes per Island}_{c}
+= \frac{1}{|\mathcal{I}|} \sum_{i \in \mathcal{I}} \text{Significant Change}_{c,i}
 $$

### 2025-03-14 R/summaryStatistics.R
Cleaned subdirectory extensive tests

@@ -2156,9 +2156,10 @@ categorize_siteMethSt <- function(data, u_threshold = 0.2, m_threshold = 0.8) {
 #' index_islands <- c(1,3)
 #' 
 #' 
-#' mean_TreeFreqsChange_i(tree, data, categorized_data = TRUE,
-#'                             index_islands, 
-#'                             pValue_threshold)
+#' mean_TreeFreqsChange_i(tree, 
+#'                        data, categorized_data = TRUE,
+#'                        index_islands, 
+#'                        pValue_threshold = 0.05)
 #' 
 #' @export
 mean_TreeFreqsChange_i <- function(tree, data, categorized_data = FALSE, index_islands, pValue_threshold, testing=FALSE) {

### 2025-03-14 extensive_tests/CFTP_testConvergence.R

@@ -1,161 +1,0 @@
-## TODO: modify how the package is loaded before submitting
-library(devtools)
-load_all()
-library(parallel)
-library(optparse)
-
-# Define the command-line options
-option_list <- list(
-  make_option(c("-o", "--output-dir"), type = "character", default = NULL,
-              help = "Full path to the output directory", metavar = "character"),
-  make_option(c("-f", "--design-file"), type = "character", default = NULL,
-              help = "Full path to the simulation design file", metavar = "character"),
-  make_option(c("-b", "--branch-length"), type = "numeric", default = NULL,
-              help = "length of branch for simulating each evol step", metavar = "numeric"),
-  make_option(c("-s", "--start"), type = "integer", default = NULL,
-              help = "start step number for simulating evol", metavar = "integer"),
-  make_option(c("-e", "--end"), type = "integer", default = NULL,
-              help = "end step number for simulating evol", metavar = "integer"),
-  make_option(c("-p", "--paramComb-n"), type = "integer", default = NULL,
-              help = "Number of parameter combinations", metavar = "integer"),
-  make_option(c("-t", "--test-n"), type = "integer", default = NULL,
-              help = "test number ", metavar = "integer"),
-  make_option(c("-n", "--name-pattern"), type = "character", default = NULL,
-              help = "output .RData files start name pattern", metavar = "character"),
-  make_option(c("-r", "--replicate-n"), type = "integer", default = NULL,
-              help = "number of data replicates to simulate", metavar = "integer")
-)
-
-# Parse the arguments
-
-opt_parser <- OptionParser(option_list = option_list)
-opt <- parse_args(opt_parser)
-
-# Get the names of required options (you can update this based on what's mandatory)
-required_options <- c("output-dir", "design-file", "branch-length", "start", "end", "paramComb-n", "test-n", "name-pattern", "replicate-n")
-
-# Check that all required options are not NULL
-missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
-
-if (length(missing_options) > 0) {
-  print_help(opt_parser)
-  stop(paste("The following arguments need to be provided:", paste(missing_options, collapse = ", ")))
-}
-
-
[…]

### 2025-03-14 extensive_tests/compute_summaryStats_cftpOutput.R

@@ -1,355 +1,0 @@
-# Load necessary libraries
-library(optparse)
-source("functions_summaryStats.R") # Load functions to compute summary statistics
-
-# Define the command-line options
-option_list <- list(
-  make_option(c("-d", "--data-dir"), type = "character", default = NULL,
-              help = "Full path to the directory containing the .RData files", metavar = "character"),
-  make_option(c("-f", "--design-file"), type = "character", default = NULL,
-              help = "Full path to the simulation design file", metavar = "character"),
-  make_option(c("-s", "--stats"), type = "character", default = "all",
-              help = "Comma-separated list of summary statistics to compute (default: all). Options: meanFreqP_i, meanFreqP_ni, sdFreqP_i, sdFreqP_ni, meanFreqM_i, meanFreqM_ni, sdFreqM_i, sdFreqM_ni, meanFracMoverMU_i, meanFracMoverMU_ni, sdFracMoverMU_i, sdFracMoverMU_ni, FChangeCherry_i, FChangeCherry_ni, Fitch, Transitions, meanCor", metavar = "character"),
-  make_option(c("-n", "--sample-n"), type = "integer", default = NULL,
-              help = "Number of samples per file", metavar = "integer"),
-  make_option(c("-p", "--pattern"), type = "character", default = NULL,
-              help = ".RData files start name pattern", metavar = "character"),
-  make_option(c("-u", "--update-file"), type = "character", default = NULL,
-              help = "Full path to existing summary statistics file", metavar = "character")
-)
-
-
-# Parse the arguments
-opt_parser <- OptionParser(option_list = option_list)
-opt <- parse_args(opt_parser)
-
-# Check if the required arguments are provided
-if (is.null(opt[["data-dir"]]) || is.null(opt[["design-file"]]) || is.null(opt[["sample-n"]]) || is.null(opt[["pattern"]])) {
-  print_help(opt_parser)
-  stop("All following arguments need to be provided: data-dir, design-file, sample-n, pattern")
-}
-
-# Convert the stats argument to a list
[…]

### 2025-03-14 extensive_tests/nest_replicates.R

@@ -1,64 +1,0 @@
-
-
-# Define the number of cases, reps, and steps
-#n_cases <- 10    # Number of cases (paramsID_01, paramsID_02, ...)
-id_cases <- 1:10
-n_reps <- 10    # Number of replicates (e.g., rep1, rep2, ..., rep10)
-n_steps <- 1001  # Number of steps (e.g. 251: from 0000 to 0250, inclusive)
-test_n <- 2
-
-# Loop over each case
-#for (case in 1:n_cases) {
-for (case in id_cases) {
-  print(paste("Processing $branch_evol() data. Parameter combination case:", case))
-  
-  # Loop over each time step
-  for (step in 0:(n_steps - 1)) {
-    print(paste("Time step:", step))
-    
-    # Format the time step with leading zeros
-    step_formatted <- sprintf("%04d", step)
-    
-    # Initialize an empty list to store the data for each case and time step
-    rep_data <- list()
-    
-    # Loop over each replicate
-    for (rep in 1:n_reps) {
-      
-      # Format the file name with the case, rep, and step
-      file_name <- sprintf("/scratch/saracv/CFTP_test/test%d/CFTP_testConvergence%d_paramsID_%02d_rep_%02d_%s.RData",test_n ,test_n ,case, rep, step_formatted)
-      
-      # Load the .RData file
-      load(file_name)
-      
-      # Append the data to rep_data list
-      rep_data[[rep]] <- data  
-    }
-    out_file <- sprintf("/scratch/saracv/CFTP_test/test%d/nestedReps_CFTP_testConvergence%d_paramsID_%02d_%s.RData",test_n ,test_n , case, step_formatted)
-    data <- rep_data
-    save(data, file = out_file)
-  }
-}
-
-#for(case in 1:n_cases){
-for (case in id_cases) {
-  print(paste("Processing $cftp() data. Parameter combination case:", case))
-  # Initialize an empty list to store the data for each case 
-  rep_data <- list()
-  
-  # Loop over $cftp() output files
-  for(rep in 1:n_reps){
-    # Format the file name with the case, rep, and step
-    file_name <- sprintf("/scratch/saracv/CFTP_test/test%d/CFTP_testConvergence%d_paramsID_%02d_rep_%02d_cftp.RData",test_n ,test_n , case, rep)
-    
-    # Load the .RData file
-    load(file_name)
-    
-    # Append the data to rep_data list
[…]

### 2025-03-14 extensive_tests/run_CFTP_testConvergence_australia.sh

@@ -1,24 +1,0 @@
-#!/bin/bash
-#SBATCH --job-name=CFTP_testConvergence_local
-
-# First test run: b_length set to 0.1, start to 1, end to 500.  
-
-# Define the test number variable
-test_n=1
-
-# Define the base directory path
-base_dir="/scratch/saracv/CFTP_test/test${test_n}"
-
-  Rscript CFTP_testConvergence.R -o $base_dir\
-                                 -f /scratch/saracv/abc_designSIM.RData \
-                                 -b 0.1 \
-                                 -s 1 \
-                                 -e 1000 \
-                                 -p 10 \
-                                 -t $test_n \
-                                 -n CFTP_testConvergence \
-                                 -r 10
-                                 
-                                 
-
-

### 2025-03-14 extensive_tests/run_CFTP_testConvergence_local.sh

@@ -1,19 +1,0 @@
-#!/bin/bash
-#SBATCH --job-name=CFTP_testConvergence_local
-
-# First test run: b_length set to 10, start to 1, end to 50
-# Second test run: b_length set to 1, start to 1, end to 500
-
-  Rscript CFTP_testConvergence.R -o /home/sara/methylation/MethEvolSIM/extensive_tests/test2 \
-                                 -f /home/sara/methylation/phd_project_saracv/simulation_studies/abc_designSIM.RData \
-                                 -b 1 \
-                                 -s 1 \
-                                 -e 3 \
-                                 -p 10 \
-                                 -t 2 \
-                                 -n CFTP_testConvergence \
-                                 -r 3
-                                 
-                                 
-
-

### 2025-03-14 extensive_tests/run_compute_summaryStats_CFTP_australia.sh

@@ -1,39 +1,0 @@
-#!/bin/bash
-#SBATCH --job-name=CFTP_compute_summaryStats
-
-# Define the test number variable
-test_n=2
-
-# Define the base directory path
-base_dir="/scratch/saracv/CFTP_test/test${test_n}"
-
-#for i in $(seq 1 10)
-for i in 1 2 3 4 5 6 7 8 9 10
-do
-  if [ $i -lt 10 ]
-  then
-    param_id="nestedReps_CFTP_testConvergence${test_n}_paramsID_0$i"
-  else
-    param_id="nestedReps_CFTP_testConvergence${test_n}_paramsID_$i"
-  fi
-    
-    # Print param_id to check its value
-    echo "param_id is set to: $param_id"
-  
-    # Call the R script for $branch_evol output data
-    Rscript compute_summaryStats.R -d $base_dir \
-                                   -p $param_id \
-                                   -f /scratch/saracv/abc_designSIM.RData \
-                                   -n 10 \
-                                   -s meanFreqP_i,meanFreqP_ni,sdFreqP_i,sdFreqP_ni,meanFreqM_i,meanFreqM_ni,sdFreqM_i,sdFreqM_ni,meanCor,Steepness
-                                   
-    # Call the R script for $cftp output data
-    Rscript compute_summaryStats_cftpOutput.R -d $base_dir \
-    					      -p $param_id \
-    					      -f /scratch/saracv/abc_designSIM.RData \
-    					      -n 10 \
-    					      -s meanFreqP_i,meanFreqP_ni,sdFreqP_i,sdFreqP_ni,meanFreqM_i,meanFreqM_ni,sdFreqM_i,sdFreqM_ni,meanCor,Steepness	
-done
-
-
-

### 2025-03-14 extensive_tests/run_compute_summaryStats_abc_local.sh

@@ -1,13 +1,0 @@
-#!/bin/bash
-#SBATCH --job-name=newSumStatsFreqPM_compute_summaryStats
-
-# Compute again the subset of summary statistics for which the functions have been updated
-
-Rscript /home/sara/methylation/MethEvolSIM/extensive_tests/compute_summaryStats.R -d /home/sara/methylation/phd_project_saracv/simulation_studies/parallel_abc \
-										   -p abc_dataSIM_0 \
-                                                                                   -f /home/sara/methylation/phd_project_saracv/simulation_studies/abc_designSIM.RData \
-                                                                                   -n 100 \
-                                                                                   -s meanFreqM_i,meanFreqM_ni \
-                                                                                   -u /home/sara/methylation/MethEvolSIM/extensive_tests/summaryStats_abc_dataSIM_0.RData
-
-

### 2025-03-16 application_examples/abc_inferences/after_debug/run_sim_server.R
Updating changes on compute_sumStats.R

@@ -107,6 +107,7 @@ mclapply(indices, function(i) {
   })
 }, mc.cores = opt[["n-cores"]])
 
+
 if(length(failed_indices) > 0) {
   print("Done. Failed indices:")
   print(failed_indices)

### 2025-03-16 application_examples/abc_inferences/scripts/compute_sumStats.R

@@ -26,7 +26,9 @@ option_list <- list(
   make_option("--shore-length", type = "integer", default = NULL,
               help = "Number of CpGs at each side of an island to exclude when computing mean correlations", metavar = "integer"),
   make_option("--n-cores", type = "integer", default = NULL,
-              help = "Number of cores", metavar = "integer")
+              help = "Number of cores", metavar = "integer"),
+  make_option("--categorized-data", type = "logical", default = FALSE,
+              help = "Methylation states are categorized", metavar = "logical")
   )
 
 
@@ -36,7 +38,7 @@ opt_parser <- OptionParser(option_list = option_list)
 opt <- parse_args(opt_parser)
 
 # Get the names of required options (you can update this based on what's mandatory)
-required_options <- c("data-dir", "design-file", "spatialStr-file", "sample-n", "n-sim", "u-threshold", "m-threshold", "cherry-index", "minN-CpG", "shore-length")
+required_options <- c("data-dir", "design-file", "spatialStr-file", "sample-n", "n-sim", "u-threshold", "m-threshold", "cherry-index", "minN-CpG", "shore-length", "categorized-data")
 
 # Check that all required options are not NULL
 missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
@@ -54,6 +56,7 @@ m_threshold <- opt[["m-threshold"]]
 cherry <- opt[["cherry-index"]]
 minN_CpG <- opt[["minN-CpG"]]
 shore_length <- opt[["shore-length"]]
+categorized_data <- opt[["categorized-data"]]
 pad_n <- nchar(as.character(opt[["n-sim"]])) + 1
 
 load(file.path(dir, opt[["design-file"]]))
@@ -75,20 +78,22 @@ compute_sumStats <- function(input_file, dir) {
   # Compute example summary statistics
   MeanSiteFChange <- MeanSiteFChange_cherry(data, tree = scaled_trees[[index]], index_islands, index_nonislands)
   # Initialize the data frame to store the summary statistics
-  sumStats <- data.frame(islandMeanFreqP = get_islandMeanFreqP(index_islands, data, sample_n),
-                         islandSDFreqP = get_islandSDFreqP(index_islands, data, sample_n),
-                         nonislandMeanFreqP = get_nonislandMeanFreqP(index_nonislands, data, sample_n),
-                         nonislandSDFreqP = get_nonislandSDFreqP(index_nonislands, data, sample_n),
[…]

### 2025-03-16 application_examples/abc_inferences/scripts/compute_sumStats.R
Updating merge_sumstats.R script

@@ -76,23 +76,22 @@ compute_sumStats <- function(input_file, dir) {
   index <- as.integer(sub("abc_dataSIM_(\\d+)\\.RData", "\\1", input_file))
   
   # Compute example summary statistics
-  MeanSiteFChange <- MeanSiteFChange_cherry(data, tree = scaled_trees[[index]], index_islands, index_nonislands)
+  MeanSiteFChange <- MeanSiteFChange_cherry(data = data, categorized_data = categorized_data, tree = scaled_trees[[index]], index_islands = index_islands, index_nonislands = index_nonislands)
   # Initialize the data frame to store the summary statistics
-  sumStats <- data.frame(islandMeanFreqP = get_islandMeanFreqP(index_islands, data, sample_n, categorized_data),
-                         islandSDFreqP = get_islandSDFreqP(index_islands, data, sample_n, categorized_data),
-                         nonislandMeanFreqP = get_nonislandMeanFreqP(index_nonislands, data, sample_n, categorized_data),
-                         nonislandSDFreqP = get_nonislandSDFreqP(index_nonislands, data, sample_n, categorized_data),
-                         islandMeanFreqM = get_islandMeanFreqM(index_islands, data, sample_n, categorized_data),
-                         islandSDFreqM = get_islandSDFreqM(index_islands, data, sample_n, categorized_data),
-                         nonislandMeanFreqM = get_nonislandMeanFreqM(index_nonislands, data, sample_n, categorized_data),
-                         nonislandSDFreqM = get_nonislandSDFreqM(index_nonislands, data, sample_n, categorized_data),
-                         meanCor_i = compute_meanCor_i(index_islands, minN_CpG = 10, shore_length = 5, data, sample_n, categorized_data),
-                         meanCor_ni = compute_meanCor_ni(index_nonislands, minN_CpG = 10, shore_length = 5, data, sample_n, categorized_data),
+  sumStats <- data.frame(islandMeanFreqP = get_islandMeanFreqP(index_islands = index_islands, data = data, categorized_data = categorized_data, sample_n = sample_n),
+                         islandSDFreqP = get_islandSDFreqP(index_islands = index_islands, data = data, categorized_data = categorized_data, sample_n = sample_n),
+                         nonislandMeanFreqP = get_nonislandMeanFreqP(index_nonislands = index_nonislands, data = data, categorized_data = categorized_data, sample_n = sample_n),
[…]

### 2025-03-16 application_examples/abc_inferences/after_debug/run_sim_server.R
Updated version with CFTP method and functions to compute summary statistics

@@ -1,122 +1,0 @@
-# Load required libraries
-library(parallel)
-library(optparse)
-library(devtools); load_all()
-
-# Define command-line options
-option_list <- list(
-  make_option("--input", type = "character", default = NULL,
-              help = "Design file, full path", metavar = "character"),
-  make_option("--output-path", type = "character", default = NULL,
-              help = "Full path for output files", metavar = "character"),
-  make_option("--n-cores", type = "integer", default = NULL,
-              help = "Number of cores", metavar = "integer"),
-  make_option("--n-tips", type = "integer", default = NULL,
-              help = "Number of tree tips", metavar = "integer"),
-  make_option("--CFTP", type = "logical", default = FALSE, 
-              metavar = "TRUE/FALSE", help = "Use CFTP algorithm at tree root [default: %default]"),
-  make_option(c("--server-name"), type = "character", default = NULL, 
-              help = "Server name (fuego, chile, australia, mauritius)")
-)
-
-# Parse command-line arguments
-opt <- parse_args(OptionParser(option_list = option_list))
-
-# Get the names of required options (you can update this based on what's mandatory)
-required_options <- c("input", "output-path", "n-cores", "n-tips", "CFTP", "server-name")
-
-# Check that all required options are not NULL
-missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
-
-if (length(missing_options) > 0) {
-  print_help(opt_parser)
-  stop(paste("The following arguments need to be provided:", paste(missing_options, collapse = ", ")))
-}
-
-# Check if server name is valid
-server_names <- c("fuego", "chile", "australia", "mauritius")
-if (!(opt[["server-name"]] %in% server_names)) {
-  stop("Invalid server name. Choose from: fuego, chile, australia, mauritius")
-}
-
-# Load the simulation design file
-load(opt[["input"]])
-print(head(sampled_params))
-
-# Define index subsets
-all_indices <- 1:nrow(sampled_params)
-subsets <- split(all_indices, cut(all_indices, breaks = length(server_names), labels = server_names))
-
-# Get the corresponding subset
-indices <- subsets[[opt[["server-name"]]]]
-
-# Generate pad_n based on the number of digits in n_sim
-# To save the files with padded numbers so that are later listed in order with list.files()
[…]

### 2025-03-16 application_examples/abc_inferences/run_snakemake.sh

@@ -1,1 +1,0 @@
-snakemake --use-conda

### 2025-03-16 application_examples/abc_inferences/scripts/compute_sumStats.R

@@ -1,117 +1,0 @@
-library(optparse)
-library(parallel)
-# TODO: ADD when source is eliminated
-library(devtools); load_all()
-
-# Define the command-line options
-option_list <- list(
-  make_option("--data-dir", type = "character", default = NULL,
-              help = "Full path to the directory containing the .RData files", metavar = "character"),
-  make_option("--design-file", type = "character", default = NULL,
-              help = "Name of the simulation design file", metavar = "character"),
-  make_option("--spatialStr-file", type = "character", default = NULL,
-              help = "Name of the file with the spatial structure of the genomic region", metavar = "character"),
-  make_option("--sample-n", type = "integer", default = NULL,
-              help = "Number of samples per file", metavar = "integer"),
-  make_option("--n-sim", type = "integer", default = NULL,
-              help = "Total number of simulations", metavar = "integer"),
-  make_option("--u-threshold", type = "numeric", default = NULL,
-              help = "Threshold for category unmethylated", metavar = "numeric"),
-  make_option("--m-threshold", type = "numeric", default = NULL,
-              help = "Threshold for category methylated", metavar = "numeric"),
-  make_option("--cherry-index", type = "integer", default = NULL,
-              help = "Index for cherry to use", metavar = "integer"),
-  make_option("--minN-CpG", type = "integer", default = NULL,
-              help = "Minimum number of CpGs to compute mean correlations", metavar = "integer"),
-  make_option("--shore-length", type = "integer", default = NULL,
-              help = "Number of CpGs at each side of an island to exclude when computing mean correlations", metavar = "integer"),
-  make_option("--n-cores", type = "integer", default = NULL,
-              help = "Number of cores", metavar = "integer"),
-  make_option("--categorized-data", type = "logical", default = FALSE,
-              help = "Methylation states are categorized", metavar = "logical")
-  )
-
-
-
-# Parse command line arguments
-opt_parser <- OptionParser(option_list = option_list)
-opt <- parse_args(opt_parser)
-
-# Get the names of required options (you can update this based on what's mandatory)
[…]

### 2025-03-16 application_examples/abc_inferences/scripts/filter_CpG_positions.R

@@ -1,71 +1,0 @@
-library(optparse)
-
-option_list <- list(
-  make_option(c("--filtered-CpGcount-region"), type = "character", default = NULL,
-              help = "CpG_count_region.RData relative path", metavar = "character"),
-  make_option(c("--CpG-positions"), type = "character", default = NULL,
-              help = "CpG_positions.RData relative path", metavar = "character"),
-  make_option(c("--output"), type = "character", default = NULL,
-              help = "Output relative path with file name", metavar = "character")
-)
-
-# Parse the arguments
-
-opt_parser <- OptionParser(option_list = option_list)
-opt <- parse_args(opt_parser)
-
-# Get the names of required options (you can update this based on what's mandatory)
-required_options <- c("filtered-CpGcount-region", "CpG-positions", "output")
-
-# Check that all required options are not NULL
-missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
-
-if (length(missing_options) > 0) {
-  print_help(opt_parser)
-  stop(paste("The following arguments need to be provided:", paste(missing_options, collapse = ", ")))
-}
-
-
-# Load the CpG count for the genomic region of interest and the CpG positions 
-# relative to the reference genome (cytosine, relative to the positive strand)
-load(opt[["filtered-CpGcount-region"]])
-load(opt[["CpG-positions"]])
-
-# Initialize a list to store results
-filtered_positions <- list(length = nrow(CpG_count_region))
-
-# Loop over each island and non-island region in the genomic region of interest
-for (i in 1:nrow(CpG_count_region)) {
-  
-  # Extract the relevant information for the current island / non-island region
-  region <- CpG_count_region[i,]
-  chr <- region$chr
-  start <- region$start
-  end <- region$end
-  
-  # Extract the CpG positions within the current region
-  filtered_positions[[i]] <- cpg_positions[cpg_positions$chr == chr & cpg_positions$CpG_position >= start & cpg_positions$CpG_position <= end,]
-  
-  # Add the positions for region start and end
-  filtered_positions[[i]]$region_start <- region$start
-  filtered_positions[[i]]$region_end <- region$end
-  
-  # Check the number of filtered positions corresponds to the CpG count for that region
-  if(nrow(filtered_positions[[i]]) != region$CpG_count){
-    print("Numper of filtered positions differs from region CpG_count in region:")
-    print(region)
-  }
-  
[…]

### 2025-03-16 application_examples/abc_inferences/scripts/get_CpG_positions.R

@@ -1,90 +1,0 @@
-library(optparse)
-library(Biostrings)
-
-option_list <- list(
-  make_option(c("--input"), type = "character", default = NULL,
-              help = "Input CpG_count.RData path", metavar = "character"),
-  make_option(c("--reference-genome"), type = "character", default = NULL,
-              help = "Relative path to the reference genome file (.fa)", metavar = "character"),
-  make_option(c("--output"), type = "character", default = NULL,
-              help = "Output relative path with file name", metavar = "character")
-)
-
-# Parse the arguments
-
-opt_parser <- OptionParser(option_list = option_list)
-opt <- parse_args(opt_parser)
-
-# Get the names of required options (you can update this based on what's mandatory)
-required_options <- c("input", "output")
-
-# Check that all required options are not NULL
-missing_options <- required_options[sapply(required_options, function(x) is.null(opt[[x]]))]
-
-if (length(missing_options) > 0) {
-  print_help(opt_parser)
-  stop(paste("The following arguments need to be provided:", paste(missing_options, collapse = ", ")))
-}
-
-
-# Load Reference Genome and CpG_count
-genome <- readDNAStringSet(opt[["reference-genome"]], format = "fasta")
-load(opt[["input"]])
-
-# Define a function to get CpG positions in a given sequence
-get_CpG_positions <- function(seq, start_pos) {
-  cpg_sites <- start(Biostrings::matchPattern("CG", seq))  # Find "CG" occurrences
-  return(cpg_sites + start_pos - 1)  # Convert to genome-wide positions
-}
-
-# Extract only the chromosome part from reference genome names
-genome_chr_names <- sapply(names(genome), function(x) strsplit(x, " ")[[1]][1])
-
-# Filter only the regions with at least one CpG site
-CpG_count <- CpG_count[CpG_count$CpG_count > 0,]
-
-# Initialize a list to store results
-cpg_positions_list <- list()
-
-# Initialize a variable to keep track of current chromosome
-current_chr <- "no_chromosome"
-
-# Loop over each region in CpG_count
-for (i in 1:nrow(CpG_count)) {
-  region <- CpG_count[i, ]
-  chr_name <- region$chr
-  
-  # Extract chromosome sequence
-  if (chr_name != current_chr){
-    
-    current_chr <- chr_name
-    print(paste("Processing chromosome", current_chr))
-    
[…]

### 2025-03-16 application_examples/abc_inferences/scripts/run_sim.R

@@ -1,92 +1,0 @@
-library(optparse)
-library(parallel)
-## TODO: Modify how the package is loaded before submitting
-library(devtools); load_all()
-
-# Define command-line options
-option_list <- list(
-  make_option("--input", type = "character", default = NULL,
-              help = "Design file, full path", metavar = "character"),
-  make_option("--output-path", type = "character", default = NULL,
-              help = "Full path for output files", metavar = "character"),
-  make_option("--n-cores", type = "integer", default = NULL,
-              help = "Number of cores", metavar = "integer"),
-  make_option("--n-tips", type = "integer", default = NULL,
-              help = "Number of tree tips", metavar = "integer"),
-  make_option("--CFTP", type = "logical", default = FALSE, 
-    metavar = "TRUE/FALSE", help = "Use CFTP algorithm at tree root [default: %default]")
-)
-
-# Parse the arguments
-opt_parser <- OptionParser(option_list = option_list)
-opt <- parse_args(opt_parser)
-
-# Load the simulation design file
-load(opt[["input"]])
-
-# Define the number of runs and cores
-n_sim <- nrow(sampled_params)
-
-# Generate pad_n based on the number of digits in n_sim
-# To save the files with padded numbers so that are later listed in order with list.files()
-pad_n <- nchar(as.character(n_sim)) + 1
-
-run_simulation_subset <- function(start_idx, end_idx) {
-  
-  # Generate the padded start and end indexes for writing the output
-  padded_start <- formatC(start_idx, width = pad_n, format = "d", flag = "0")
-  padded_end <- formatC(end_idx, width = pad_n, format = "d", flag = "0")
-  
-  # Set the name for the .out file for the entire subset
-  out_file <- file.path(opt[["output-path"]], paste0("abc_dataSIM_", padded_start, "_to_", padded_end, ".out"))
-  
-  # Open the log file once for the entire subset
-  log_connection <- file(out_file, open = "w")
-  
-  # Redirect both regular output and message output to the same file
-  sink(log_connection, type = "output")
-  sink(log_connection, type = "message")
-  
-  # Run the simulations for the current subset
-  for (sim in start_idx:end_idx) {
-    print(paste("Running simulation number:", sim))
-    
-    padded_index <- formatC(sim, width = pad_n, format = "d", flag = "0")
-    
[…]

### 2025-03-16 application_examples/abc_inferences/scripts/set_sim_design.R

@@ -1,87 +1,0 @@
-library(optparse)
-library(ape)
-
-# Define command-line options with unique letters
-option_list <- list(
-  make_option("--dir", type = "character", default = NULL,
-              help = "Full path to save the design file", metavar = "character"),
-  make_option("--genome-dist", type = "character", default = NULL,
-              help = "Distribution of methylation sites in the selected genomic region", metavar = "character"),
-  make_option("--tree", type = "character", default = "all",
-              help = "Tree in Newick format", metavar = "character"),
-  make_option("--n-sim", type = "integer", default = NULL,
-              help = "Number of simulations", metavar = "integer"),
-  make_option("--seed", type = "integer", default = NULL,
-              help = "Seed value", metavar = "integer"),
-  make_option("--alpha_pI_shape", type="numeric", default=3, help="Shape parameter for alpha_pI"),
-  make_option("--beta_pI_shape", type="numeric", default=3, help="Shape parameter for beta_pI"),
-  make_option("--alpha_pI_scale", type="numeric", default=as.numeric(1/30), help="Scale parameter for alpha_pI"),
-  make_option("--beta_pI_scale", type="numeric", default=as.numeric(1/3), help="Scale parameter for beta_pI"),
-  make_option("--alpha_pNI_shape", type="numeric", default=3, help="Shape parameter for alpha_pNI"),
-  make_option("--beta_pNI_shape", type="numeric", default=3, help="Shape parameter for beta_pNI"),
-  make_option("--alpha_pNI_scale", type="numeric", default=as.numeric(1/30), help="Scale parameter for alpha_pNI"),
-  make_option("--beta_pNI_scale", type="numeric", default=as.numeric(1/3), help="Scale parameter for beta_pNI"),
-  make_option("--alpha_mI_shape", type="numeric", default=3, help="Shape parameter for alpha_mI"),
-  make_option("--beta_mI_shape", type="numeric", default=3, help="Shape parameter for beta_mI"),
-  make_option("--alpha_mI_scale", type="numeric", default=as.numeric(1/30), help="Scale parameter for alpha_mI"),
[…]
